---
title: Analysis of Abau and Ftul infected mice - lungs
author: Eduard Ansaldo
date: '`r Sys.Date()`'
verbose: false
engine: knitr
knitr:
  opts_chunk: 
    echo: false
    warning: false
    message: false
    error: false
    cache: false
    results: markdown
    code-fold: true
    cache-lazy: false
    
processing: true

execute: 
  cache: true


format:
  html:
    embed-resources: true
    fig-cap-location: top
    toc: true
    toc-depth: 3
    toc-expand: 1
    toc-location: left
    toc-title: Table of Contents
    toc-collapse: true
    number-sections: true
    page-layout: full
    fig-width: 6
    fig-height: 6
    fig-align: 'center'
---

::: {.content-visible when-meta="verbose"}

# Initialization

:::

```{r}
# General R and plotting 
library(conflicted)
library(tidyverse)
library(ggplot2)
library(scales)
library(patchwork)
library(cowplot)
library(gridExtra)
library(ggrepel)
library(stringr)
library(VennDiagram)
library(pheatmap)
library(viridis)
library(here)
library(readxl)

# Single Cell Analysis Packages
library(Seurat)
library(scRepertoire)
library(circlize)
library(scCustomize)
library(SingleR)
library(celldex)
library(UCell)
library(presto)
library(scDblFinder)

# DEG, pathway enrichment and visualization packages
library(DESeq2)
library(gprofiler2)
library(clusterProfiler)
library(DOSE)
library(pathview)
library(org.Mm.eg.db)
library(scRepertoire)
library(enrichplot)
library(msigdbr)
library(rlang)

# Python Version 3.9.6!!! 

knitr::dep_prev()
conflict_scout()
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")

```

```{r}
#| results: hide 
# Setting working directory and seed
# renv::update()
# renv::snapshot()
# renv::init()
# renv::status()
set.seed(3514)
i_am('scripts/Analysis.qmd')
here()
results_path <- here('results/')
dir.create(results_path)
results_path <- here('results/')
dir.create(results_path)
figures_path <- here('results/figures/')
dir.create(figures_path)
# data_path <- here('data/')
# dir.create(data_path)
tables_path <- here('results/tables/')
dir.create(tables_path)

# Loading custom functions
source(here('scripts/function_template.r'))
source(here('scripts/gProfiler2_functions.r'))

# Setting up color palettes
diverging_palette <- hcl.colors(n = 20,'Purple-Green',rev = T)
sequential_palette_dotplot <- hcl.colors(n = 20,'YlGn',rev = T)
sequential_palette <- hcl.colors(n = 20,'Mako',rev = T)
```

:::{.content-visible when-meta='processing'}

# Preprocessing Abau MAIT dataset

:::

:::{.content-visible when-meta='verbose'}

## Read in data

:::

```{r}
#| results: hide
#| #Load the dataset from the cellranger outs
#| results: hide
#Load the dataset from the cellranger outs
scdata <- Read10X(data.dir = here("data/cluster_processing_2/Abau_MAIT/outs/per_sample_outs/Abau_MAIT/count/sample_filtered_feature_bc_matrix/"))

#Initialize the seurat object with the raw (non-normalized data)
seurat <- CreateSeuratObject(counts=scdata$'Gene Expression', min.cells = 3)

```

:::{.content-visible when-meta='processing'}
#### Confirming used HTOs - MAIT
:::

```{r}
print('Antibody Capture total reads:')
scdata$`Antibody Capture` |> as.matrix() |>
        rowSums()  |> as.data.frame() 
```

Hashtags Used: 1, 2, 4, 5

```{r}
# Extracting HTO and biotin data
scdata$'Antibody Capture' <-  scdata$'Antibody Capture'[-c(3), ]
# colnames(scdata$'HTO') <- colnames(scdata$'Antibody Capture')
# rownames(scdata$'HTO') <- rownames(scdata$'Antibody Capture')[-c(2, 3, 4, 5, 6, 7, 11, 17, 18, 19)]

# #Add HTO data as a new assay independent from RNA
HTO <- CreateAssayObject(counts = scdata$'Antibody Capture')
seurat[["HTO"]] <- HTO
```

:::{.content-visible when-meta='processing'}

## QC

:::

```{r}
#| layout-ncol: 2
#Quantifying percentage mitochondria
Idents(seurat) <- 'Tissue'
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, assay="RNA",pattern = "mt-")

VlnPlot(seurat, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol=3, pt.size = 0.01)

# Setting thresholds:
nCount_RNA_threshold <- 1500
nFeature_RNA_threshold <- 500
percent_mt_threshold <- 6


p1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nCount_RNA_threshold)

p2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nFeature_RNA_threshold)

p3 <- seurat@meta.data %>% 
  	ggplot(aes(x=percent.mt)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = percent_mt_threshold)

grid.arrange(p1, p2, p3)    

# # Visualize feature relationships
# # plot1 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "percent.mt")

# # plot2 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
# # plot1+plot2

cell_number <- list(nrow(seurat[[]]))
```

Filtering Low Quality Cells

```{r}
# Filter data

cell_number <- list(nrow(seurat[[]]))
seurat <- subset(seurat, subset = nFeature_RNA  > nFeature_RNA_threshold & percent.mt < percent_mt_threshold & nCount_RNA > nCount_RNA_threshold)
cell_number <- append(cell_number, nrow(seurat[[]]))

print("Cell number before and after filterings")
print(cell_number)
```

:::{.content-visible when-meta='verbose'}
## Data Normalization 
:::

```{r}
# Normalize RNA data with SCTransform
# seurat <- SCTransform(seurat, verbose = T)

# Normalize HTO data with CLR
seurat <- NormalizeData(seurat, assay = "HTO",normalization.method = "CLR")
```

:::{.content-visible when-meta='processing'}
## Hashtag Demultiplexing - MAIT
:::

```{r}
#| fig-width: 14
#| fig-height: 10
plot1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot1+plot2

seurat <- HTODemux(seurat, assay = "HTO", positive.quantile = 0.99)

table(seurat$hash.ID)
table(seurat$HTO_maxID)
table(seurat$HTO_classification.global)

#tSNE-visualization
seurat_subset <- subset(seurat, idents = "Negative", invert =TRUE)
DefaultAssay(seurat_subset) <- "HTO"
seurat_subset <- ScaleData(seurat_subset, assay = "HTO",  features = rownames(seurat_subset), verbose = FALSE )
seurat_subset <- RunPCA(seurat_subset, assay = "HTO", rownames(seurat_subset), approx = FALSE)
seurat_subset <- RunTSNE(seurat_subset, assay = "HTO", dims = 1:24,  perplexity = 100, check_duplicates=FALSE)
DimPlot(seurat_subset)
```

Hashtag staining and demultiplexing was very successful, with extremely clear separation between the hashtags.

```{r}
print("Number of Singlets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Singlet')))

print("Number of Doublets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Doublet')))

# Filtering singlets only
Idents(seurat) <- "HTO_classification.global"
seurat <-subset(seurat, idents = "Singlet")
```

:::{.content-visible when-meta='verbose'}
## Create Groups
:::

```{r}

seurat@meta.data <- seurat@meta.data |>
    mutate(Groups = 'Abau') |>
    mutate(Samples = str_c(Groups, hash.ID, sep = '_')) |> 
    mutate(Samples = case_when(
            hash.ID == 'Mouse-4' ~ 'Abau_Mouse-3',
            hash.ID == 'Mouse-5' ~ 'Abau_Mouse-4',
            TRUE ~ Samples
    )) |>
    mutate(Samples = factor(Samples, levels = c(
            'Abau_Mouse-1',
            'Abau_Mouse-2',
            'Abau_Mouse-3',
            'Abau_Mouse-4')))

```

:::{.content-visible when-meta='verbose'}
## Import TCR data
:::

```{r}
contigs <- read.csv(here('data/cluster_processing_2/Abau_MAIT/outs/per_sample_outs/Abau_MAIT/vdj_t/filtered_contig_annotations.csv'))

#Demultiplexing VDJ libraries
contig_list <- createHTOContigList(contigs, seurat, group.by = "Samples")
contig_list <- contig_list[c(levels(seurat$Samples))]

# Combining the Contigs
combined <- combineTCR(contig_list, removeNA = TRUE, samples=c(levels(seurat$Samples)), filterMulti = TRUE)

# # #Adding Samples as variables
combined <- addVariable(combined, variable.name='Samples', variables=c(levels(seurat$Samples)))                                    

## Integrating data with seurat object
cell_names <- Cells(seurat)
groups_merge <- pull(seurat@meta.data, Samples)

#Changing barcodes on the seurat object for the merge with VDJ data
new_cell_names <- paste(groups_merge, cell_names, sep='_')
seurat$barcode <- cell_names
seurat <- RenameCells(seurat, new.names=new_cell_names)

# #Combining VDJ and Seurat Object
seurat <- combineExpression(combined, seurat, proportion = TRUE, cloneCall='aa', group.by='Samples', chain = 'both')
# Renaming cells back to original barcodes
seurat <- RenameCells(seurat, new.names=seurat$barcode)
```

```{r}
seurat_Abau <- seurat
```

:::{.content-visible when-meta='processing'}

# Preprocessing Ftula MAIT dataset

:::

:::{.content-visible when-meta='verbose'}

## Read in data

:::

```{r}
#| results: hide
#| #Load the dataset from the cellranger outs
#| results: hide
#Load the dataset from the cellranger outs
scdata <- Read10X(data.dir = here("data/cluster_processing_2/Ftula_MAIT/outs/per_sample_outs/Ftula_MAIT/count/sample_filtered_feature_bc_matrix/"))

#Initialize the seurat object with the raw (non-normalized data)
seurat <- CreateSeuratObject(counts=scdata$'Gene Expression', min.cells = 3)

```

:::{.content-visible when-meta='processing'}
#### Confirming used HTOs 
:::

```{r}
print('Antibody Capture total reads:')
scdata$`Antibody Capture` |> as.matrix() |>
        rowSums()  |> as.data.frame() 
```

Hashtags Used: 1, 2, 3, 4, 5

```{r}
# Extracting HTO and biotin data
# scdata$'Antibody Capture' <-  scdata$'Antibody Capture'[-c(3), ]
# colnames(scdata$'HTO') <- colnames(scdata$'Antibody Capture')
# rownames(scdata$'HTO') <- rownames(scdata$'Antibody Capture')[-c(2, 3, 4, 5, 6, 7, 11, 17, 18, 19)]

# #Add HTO data as a new assay independent from RNA
HTO <- CreateAssayObject(counts = scdata$'Antibody Capture')
seurat[["HTO"]] <- HTO
```

:::{.content-visible when-meta='processing'}

## QC

:::

```{r}
#| layout-ncol: 2
#Quantifying percentage mitochondria
Idents(seurat) <- 'Tissue'
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, assay="RNA",pattern = "mt-")

VlnPlot(seurat, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol=3, pt.size = 0.01)

# Setting thresholds:
nCount_RNA_threshold <- 1500
nFeature_RNA_threshold <- 500
percent_mt_threshold <- 6


p1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nCount_RNA_threshold)

p2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nFeature_RNA_threshold)

p3 <- seurat@meta.data %>% 
  	ggplot(aes(x=percent.mt)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = percent_mt_threshold)

grid.arrange(p1, p2, p3)    

# # Visualize feature relationships
# # plot1 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "percent.mt")

# # plot2 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
# # plot1+plot2

cell_number <- list(nrow(seurat[[]]))
```

Filtering Low Quality Cells

```{r}
# Filter data

cell_number <- list(nrow(seurat[[]]))
seurat <- subset(seurat, subset = nFeature_RNA  > nFeature_RNA_threshold & percent.mt < percent_mt_threshold & nCount_RNA > nCount_RNA_threshold)
cell_number <- append(cell_number, nrow(seurat[[]]))

print("Cell number before and after filterings")
print(cell_number)
```

:::{.content-visible when-meta='verbose'}
## Data Normalization 
:::

```{r}
# Normalize RNA data with SCTransform
# seurat <- SCTransform(seurat, verbose = T)

# Normalize HTO data with CLR
seurat <- NormalizeData(seurat, assay = "HTO",normalization.method = "CLR")
```

:::{.content-visible when-meta='processing'}
## Hashtag Demultiplexing - MAIT
:::

```{r}
#| fig-width: 14
#| fig-height: 10
plot1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot1+plot2

seurat <- HTODemux(seurat, assay = "HTO", positive.quantile = 0.99)

table(seurat$hash.ID)
table(seurat$HTO_maxID)
table(seurat$HTO_classification.global)

#tSNE-visualization
seurat_subset <- subset(seurat, idents = "Negative", invert =TRUE)
DefaultAssay(seurat_subset) <- "HTO"
seurat_subset <- ScaleData(seurat_subset, assay = "HTO",  features = rownames(seurat_subset), verbose = FALSE )
seurat_subset <- RunPCA(seurat_subset, assay = "HTO", rownames(seurat_subset), approx = FALSE)
seurat_subset <- RunTSNE(seurat_subset, assay = "HTO", dims = 1:24,  perplexity = 100, check_duplicates=FALSE)
DimPlot(seurat_subset)
```

Hashtag staining and demultiplexing was very successful, with extremely clear separation between the hashtags.

```{r}
print("Number of Singlets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Singlet')))

print("Number of Doublets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Doublet')))

# Filtering singlets only
Idents(seurat) <- "HTO_classification.global"
seurat <-subset(seurat, idents = "Singlet")
```

:::{.content-visible when-meta='verbose'}
## Create Groups
:::

```{r}

seurat@meta.data <- seurat@meta.data |>
    mutate(Groups = 'Ftula') |>
    mutate(Samples = str_c(Groups, hash.ID, sep = '_')) |> 
    mutate(Samples = factor(Samples, levels = c(
            'Ftula_Mouse-1',
            'Ftula_Mouse-2',
            'Ftula_Mouse-3',
            'Ftula_Mouse-4',
            'Ftula_Mouse-5')))

```

:::{.content-visible when-meta='verbose'}
## Import TCR data
:::

```{r}
contigs <- read.csv(here('data/cluster_processing_2/Ftula_MAIT/outs/per_sample_outs/Ftula_MAIT/vdj_t/filtered_contig_annotations.csv'))

#Demultiplexing VDJ libraries
contig_list <- createHTOContigList(contigs, seurat, group.by = "Samples")
contig_list <- contig_list[c(levels(seurat$Samples))]

# Combining the Contigs
combined <- combineTCR(contig_list, removeNA = TRUE, samples=c(levels(seurat$Samples)), filterMulti = TRUE)

# # #Adding Samples as variables
combined <- addVariable(combined, variable.name='Samples', variables=c(levels(seurat$Samples)))                                    

## Integrating data with seurat object
cell_names <- Cells(seurat)
groups_merge <- pull(seurat@meta.data, Samples)

#Changing barcodes on the seurat object for the merge with VDJ data
new_cell_names <- paste(groups_merge, cell_names, sep='_')
seurat$barcode <- cell_names
seurat <- RenameCells(seurat, new.names=new_cell_names)

# #Combining VDJ and Seurat Object
seurat <- combineExpression(combined, seurat, proportion = TRUE, cloneCall='aa', group.by='Samples', chain = 'both')
# Renaming cells back to original barcodes
seurat <- RenameCells(seurat, new.names=seurat$barcode)
```

```{r}
seurat_Ftula <- seurat
```

:::{.content-visible when-meta='processing'}

# Preprocessing CT MAIT dataset

:::

:::{.content-visible when-meta='verbose'}

## Read in data

:::

```{r}
#| results: hide
#| #Load the dataset from the cellranger outs
#| results: hide
#Load the dataset from the cellranger outs
scdata <- Read10X(data.dir = here("data/cluster_processing_2/CT_MAIT/outs/per_sample_outs/CT_MAIT/count/sample_filtered_feature_bc_matrix/"))

#Initialize the seurat object with the raw (non-normalized data)
seurat <- CreateSeuratObject(counts=scdata$'Gene Expression', min.cells = 3)

```

:::{.content-visible when-meta='processing'}
#### Confirming used HTOs
:::

```{r}
print('Antibody Capture total reads:')
scdata$`Antibody Capture` |> as.matrix() |>
        rowSums()  |> as.data.frame() 
```

Hashtags Used: 1, 2, 4, 5

```{r}
# Extracting HTO and biotin data
scdata$'Antibody Capture' <-  scdata$'Antibody Capture'[-c(3), ]
# colnames(scdata$'HTO') <- colnames(scdata$'Antibody Capture')
# rownames(scdata$'HTO') <- rownames(scdata$'Antibody Capture')[-c(2, 3, 4, 5, 6, 7, 11, 17, 18, 19)]

# #Add HTO data as a new assay independent from RNA
HTO <- CreateAssayObject(counts = scdata$'Antibody Capture')
seurat[["HTO"]] <- HTO
```

:::{.content-visible when-meta='processing'}

## QC

:::

```{r}
#| layout-ncol: 2
#Quantifying percentage mitochondria
Idents(seurat) <- 'Tissue'
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, assay="RNA",pattern = "mt-")

VlnPlot(seurat, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol=3, pt.size = 0.01)

# Setting thresholds:
nCount_RNA_threshold <- 1500
nFeature_RNA_threshold <- 500
percent_mt_threshold <- 6


p1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nCount_RNA_threshold)

p2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nFeature_RNA_threshold)

p3 <- seurat@meta.data %>% 
  	ggplot(aes(x=percent.mt)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = percent_mt_threshold)

grid.arrange(p1, p2, p3)    

# # Visualize feature relationships
# # plot1 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "percent.mt")

# # plot2 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
# # plot1+plot2

cell_number <- list(nrow(seurat[[]]))
```

Filtering Low Quality Cells

```{r}
# Filter data

cell_number <- list(nrow(seurat[[]]))
seurat <- subset(seurat, subset = nFeature_RNA  > nFeature_RNA_threshold & percent.mt < percent_mt_threshold & nCount_RNA > nCount_RNA_threshold)
cell_number <- append(cell_number, nrow(seurat[[]]))

print("Cell number before and after filterings")
print(cell_number)
```

:::{.content-visible when-meta='verbose'}
## Data Normalization 
:::

```{r}
# Normalize RNA data with SCTransform
# seurat <- SCTransform(seurat, verbose = T)

# Normalize HTO data with CLR
seurat <- NormalizeData(seurat, assay = "HTO",normalization.method = "CLR")
```

:::{.content-visible when-meta='processing'}
## Hashtag Demultiplexing - MAIT
:::

```{r}
#| fig-width: 14
#| fig-height: 10
plot1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot1+plot2

seurat <- HTODemux(seurat, assay = "HTO", positive.quantile = 0.99)

table(seurat$hash.ID)
table(seurat$HTO_maxID)
table(seurat$HTO_classification.global)

#tSNE-visualization
seurat_subset <- subset(seurat, idents = "Negative", invert =TRUE)
DefaultAssay(seurat_subset) <- "HTO"
seurat_subset <- ScaleData(seurat_subset, assay = "HTO",  features = rownames(seurat_subset), verbose = FALSE )
seurat_subset <- RunPCA(seurat_subset, assay = "HTO", rownames(seurat_subset), approx = FALSE)
seurat_subset <- RunTSNE(seurat_subset, assay = "HTO", dims = 1:24,  perplexity = 100, check_duplicates=FALSE)
DimPlot(seurat_subset)
```

Hashtag staining and demultiplexing was very successful, with extremely clear separation between the hashtags.

```{r}
print("Number of Singlets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Singlet')))

print("Number of Doublets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Doublet')))

# Filtering singlets only
Idents(seurat) <- "HTO_classification.global"
seurat <-subset(seurat, idents = "Singlet")
```

:::{.content-visible when-meta='verbose'}
## Create Groups
:::

```{r}

seurat@meta.data <- seurat@meta.data |>
    mutate(Groups = 'CT') |>
    mutate(Samples = str_c(Groups, hash.ID, sep = '_')) |> 
    mutate(Samples=case_when(
            hash.ID == 'Mouse-4' ~ 'CT_Mouse-3',
            hash.ID == 'Mouse-5' ~ 'CT_Mouse-4',
            TRUE ~ Samples
    )) |>  
    mutate(Samples = factor(Samples, levels = c(
            'CT_Mouse-1',
            'CT_Mouse-2',
            'CT_Mouse-3',
            'CT_Mouse-4')))

```

:::{.content-visible when-meta='verbose'}
## Import TCR data
:::

```{r}
contigs <- read.csv(here('data/cluster_processing_2/CT_MAIT/outs/per_sample_outs/CT_MAIT/vdj_t/filtered_contig_annotations.csv'))

#Demultiplexing VDJ libraries
contig_list <- createHTOContigList(contigs, seurat, group.by = "Samples")
contig_list <- contig_list[c(levels(seurat$Samples))]

# Combining the Contigs
combined <- combineTCR(contig_list, removeNA = TRUE, samples=c(levels(seurat$Samples)), filterMulti = TRUE)

# # #Adding Samples as variables
combined <- addVariable(combined, variable.name='Samples', variables=c(levels(seurat$Samples)))                                    

## Integrating data with seurat object
cell_names <- Cells(seurat)
groups_merge <- pull(seurat@meta.data, Samples)

#Changing barcodes on the seurat object for the merge with VDJ data
new_cell_names <- paste(groups_merge, cell_names, sep='_')
seurat$barcode <- cell_names
seurat <- RenameCells(seurat, new.names=new_cell_names)

# #Combining VDJ and Seurat Object
seurat <- combineExpression(combined, seurat, proportion = TRUE, cloneCall='aa', group.by='Samples', chain = 'both')
# Renaming cells back to original barcodes
seurat <- RenameCells(seurat, new.names=seurat$barcode)
```

```{r}
seurat_CT <- seurat
```

# Merged processing

## Merging datasets

```{r}
data_list <- list(CT_MAIT = seurat_CT, Abau_MAIT = seurat_Abau, Ftula_MAIT = seurat_Ftula)
# Merging the datasets
seurat <- Merge_Seurat_List(data_list, 
                       add.cell.ids = names(data_list),
                       project = "MAIT_Abau_Ftula_CT_Lungs")
```

:::{.content-visible when-meta='verbose'}

### Save Object

:::

```{r}
# saveRDS(seurat, file = here('data/Merged_objects_seurat.rds'))
```

:::{.content-visible when-meta='verbose'}

## Data Normalization

:::

```{r}
# Normalize RNA data with SCTransform
seurat <- SCTransform(seurat, verbose = F) 
# seurat@assays$SCT
# seurat <- PrepSCTFindMarkers(seurat)
```

:::{.content-visible when-meta='processing'}

## Clustering

:::

```{r}
#| fig-width: 4
#| fig-height: 4
#| fig-cap: 'Elbow plot showing the percentage of variability represented by by each PC to select number of dimensions '

#Dimensionality reduction
seurat <- RunPCA(seurat,npcs = 100)

#Determining dimensionality of the dataset
ElbowPlot(seurat, ndims = 100)
```

:::{.content-visible when-meta='processing'}

Chosen Number of Dimensions:

:::

```{r}
# Describe number of dimensions

dimensions  <- 25 
dimensions
```

```{r}
#| fig-cap: UMAP plots showing the clustering results at different resolutions
#| fig-width: 10
#| fig-height: 10

seurat <- RunUMAP(seurat, dims=1:dimensions, verbose = F, seed.use = 3514L)
# resolutions <- c(0.25, 0.375, 0.5, 0.625, 0.75, 1, 1.25, 1.5, 1.75)
resolutions <- c(0.05, 0.1, 0.15, 0.25, 0.375, 0.5)
seurat <- FindNeighbors(seurat, dims = 1:dimensions, verbose = F)
seurat <- FindClusters(seurat, resolution = resolutions, verbose = F)

p <- list()
i <- 1
for (resolution in resolutions ) {
    Idents(seurat) <- paste0('SCT_snn_res.', resolution) 
    p2 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")    
    p[[i]] <- p2
    i <- i+1
}

plot <- grid.arrange(grobs = p)
# ggsave('initial_clustering_results_by_resolution.pdf', path = here('result/figures'), plot = plot)
```

:::{.content-visible when-meta='processing'}

Chosen resolution

:::

```{r}
resolution <- 0.1
resolution
```

:::{.content-visible when-meta='verbose'}

### Save Object

:::

```{r}
# saveRDS(seurat, file = here('data/Merged_objects_seurat.rds'))
```

```{r}
# seurat <- readRDS(here('data/Merged_objects_seurat.rds'))
object_annotations <- 'full_object_merged'
```

### Clustering Results

```{r}
#| layout-ncol: 2
Idents(seurat) <- paste0('SCT_snn_res.', resolution)
seurat[['seurat_clusters']]<- Idents(seurat)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Groups', ncol = 2) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_group_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
# DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Samples', ncol = 4) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
# ggsave(paste0('UMAP_clusters_by_sample_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = FALSE, group.by = 'Groups') + ggtitle(paste0('R ', resolution)) + theme(legend.position = "right")
```

### Top DEGs per cluster

```{r}
#| fig-width: 6
#| fig-height: 12
seurat <- PrepSCTFindMarkers(seurat)
plot1 <- top_genes_per_cluster(seurat, n_genes_to_plot=5, object_annotations, tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = F) 
print(plot1)
ggsave(paste0(figures_path, 'DotPlot_Top3_per_cluster', object_annotations, '.pdf'), width = 10, height = 16)
```

```{r}
#| layout-ncol: 2
seurat[["percent.ribo"]] <- PercentageFeatureSet(seurat, assay="RNA",pattern = "^Rp[sl]")
FeaturePlot_scCustom(seurat, features = c('nCount_RNA', 'nFeature_RNA', 'percent.mt', 'percent.ribo'), colors_use = sequential_palette)
FeaturePlot_scCustom(seurat, features = c('Zbtb16', '', 'Icos'), colors_use = sequential_palette)
VlnPlot(seurat, features = c('nCount_RNA', 'nFeature_RNA', 'percent.mt', 'percent.ribo'), group.by = 'seurat_clusters', pt.size = 0.1) + NoLegend()
VlnPlot(seurat, features = c('Zbtb16', 'Ubc', 'Icos'), group.by = 'seurat_clusters', pt.size = 0.1) + NoLegend()
FeatureScatter(seurat, feature1 = 'nCount_RNA', feature2 = 'nFeature_RNA', split.by = 'seurat_clusters', ncol = 3)
```

:::{.content-visible when-meta='processing'}

### Automatic cell type annotations

:::

```{r}
local_path <- paste0(figures_path, object_annotations,'_cell_type_annotations')
unlink(local_path,recursive = T)
dir.create(local_path)

# Normalize and scale data
seurat <- JoinLayers(seurat, assay = 'RNA')
seurat <- NormalizeData(seurat, assay = "RNA", normalization.method = "LogNormalize", scale.fct = 10000)
seurat <- ScaleData(seurat, assay = 'RNA')
```

:::{.content-visible when-meta='processing'}

#### SingleR annotations

:::

:::{.content-visible when-meta='processing'}

Annotations per Cluster

:::

```{r}
#| fig-cap: UMAP plot showing the clustering results with SingleR fine annotations on a per cluster basis
#| fig-width: 9
#| fig-height: 6
seurat <- annotate_seurat_with_SingleR_Eduard(seurat, local_path, database = 'ImmGen', annotation_basis = 'cluster_coarse', split_by_groups = FALSE)
```

:::{.content-visible when-meta='processing'}

Annotations per Cell

:::

```{r}
#| fig-cap: UMAP plot showing the clustering results with SingleR coarse annotations on a per cell basis

annotate_seurat_with_SingleR_Eduard(seurat, local_path, database = 'ImmGen', annotation_basis = 'cell_coarse', split_by_groups = FALSE)
```

### Plotting Genes of Interest for cluster annotation

```{r}
#| fig-cap: Feature plots showing the expression of T cell markers
#| fig-width: 8
#| fig-height: 10
FeaturePlot_scCustom(seurat, features = c('Cd3e', 'Cd4', 'Cd8a', 'Cd8b1', 'Trdc', 'Trac'), pt.size = 0.01, colors_use = sequential_palette) +
    theme(legend.position = 'none')
```

```{r}
#| fig-cap: Feature plots showing the expression of myeloid cell markers
#| fig-width: 8
#| fig-height: 10
FeaturePlot_scCustom(seurat, features = c('Sirpa', 'Xcr1', 'Itgae', 'Itgax', 'Csf1r', 'Csf2ra'), pt.size = 0.1, colors_use = sequential_palette) +
    theme(legend.position = 'none')
```
```{r}
#| fig-cap: Feature plots showing the expression of B cell an epithelial cell markers
#| #| fig-width: 8
#| fig-height: 13
FeaturePlot_scCustom(seurat, features = c('Ighm', 'Ighd', 'Igha', 'Igkc','Pax5', 'Prdm1', 'Epcam', 'Muc2'), pt.size = 0.1, colors_use = sequential_palette, num_columns = 2) 
```

### Sub clustering cluster 22

```{r}
#| results: hide 
Idents(seurat) <- 'seurat_clusters'
seurat <- FindSubCluster(seurat, graph.name = 'SCT_snn', cluster = 19, resolution = 0.3)
Idents(seurat) <- 'sub.cluster'
DimPlot(seurat, label = T)
```


### Cell Type Annotations

### Annotating clusters

```{r}
Idents(seurat) <- 'sub.cluster'
seurat <- RenameIdents(seurat
    , '0' = 'B cells'  
    , '1' = 'Epithelial cells'
    , '2' = 'Stromal cells'
    , '3' = 'Lymphocytes'
    , '4' = 'Epithelial cells'
    , '5' = 'Epithelial cells'
    , '6' = 'Lymphocytes'
    , '7' = 'Myeloid cells' 
    , '8' = 'Stromal cells'
    , '9' = 'B cells'
    , '10' = 'Stromal cells'
    , '11' = 'Lymphocytes'
    , '12' = 'Myeloid cells'
    , '13' = 'Stromal cells'
    , '14' = 'Stromal cells'
    , '15' = 'Myeloid cells'
    , '16' = 'Myeloid cells'
    , '17' = 'Stromal cells' 
    , '18' = 'Epithelial cells'
    , '19_0' = 'Lymphocytes'
    , '19_1' = 'B cells'
    , '19_2' = 'Lymphocytes'
    , '19_3' = 'Lymphocytes'
    , '19_4' = 'Lymphocytes'
    , '20' = 'Epithelial cells'
    , '21' = 'Epithelial cells'
    , '22' = 'Red blood cells'
    , '23' = 'Lymphocytes'
    , '24' = 'Stromal cells'
    , '25' = 'Endothelial cells'

)
seurat[['major_cell_types']] <- Idents(seurat)
```

```{r}
#| layout-ncol: 2
#| fig-cap: UMAP plot showing the cell type annotations
Idents(seurat) <- 'major_cell_types'
DimPlot(seurat, label=T, ncol = 2) 
ggsave(paste0('UMAP_major_cell_types_', object_annotations, '.pdf'), path = figures_path, width = 7, height = 5)

DimPlot(seurat, label=F, split.by = 'Groups', ncol = 2)
ggsave(paste0('UMAP_major_cell_types_by_group_', object_annotations, '.pdf'), path = figures_path, width = 9, height = 5)
```

### Cell frequency per condition

```{r}
extract_cell_counts(seurat, major_cell_types, figures_path, tables_path, object_annotations=object_annotations)
```

:::{.content-visible when-meta='verbose'}

## Save Object

:::

```{r}
saveRDS(seurat, file = here('data/E2_object.rds'))
```

## Differential Gene Expression Analysis

Given that we don't have biological replicates, I am performing a Wilcoxon test on the existing single-cell dataset, comparing cross-fostering and control conditions for each cell type. Please note that this test is usually not recommended since it treats each cell as a biological replicate, thus leading to a lot of false positives. In general, it is recommended to include biological replicates in the experiment using different hashtags for each mouse, if possible. 

```{r}
seurat <- readRDS(here('data/E2_object_lymphocytes.rds'))

gProfiler2 <- TRUE 
```

### DEG analysis - Cross-fostered vs Control - fine cell type annotations

```{r}
#| fig-width: 7
#| fig-height: 7
#| layout-ncol: 2
#| results: hide

path <- paste0(results_path, 'DEG_Wilcox_analysis_Lymphocytes_Cross-fostered_vs_Control_cell_types/')
unlink(path,recursive = T)
dir.create(path)
Idents(seurat) <- 'cell_types'

DEG_counts <- data.frame(matrix(ncol=3, nrow=0))
colnames(DEG_counts) <- c('DEG_count', 'DEG_UP_count', 'DEG_DOWN_count')
rnames <- c()

DEG_counts <- DEG_FindMarkers_SCT_assay(seurat, comparison='Condition', group1='Control', group2='Cross-fostered', cluster= 'all_lymphocytes' , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = F)
rnames <- c(rnames, 'all_lymphocytes')

for (x in levels(seurat$cell_types)) {
    # cat(paste0("Cell type: ", x, "\n"))
    seurat_small <- subset(seurat, subset = cell_types == x )
    counts <- DEG_FindMarkers_SCT_assay(seurat_small, comparison='Condition', group1='Control', group2='Cross-fostered', cluster= x , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = T)
    rm(seurat_small)
    DEG_counts <- rbind(DEG_counts, counts)
    rnames <- c(rnames, x)
}
rownames(DEG_counts) <- rnames
write.csv(DEG_counts, file = file.path(tables_path, "DEG_counts.csv"))
DEG_counts
```


### DEG analysis - Cross-fostered vs Control - major cell type annotations

```{r}
#| fig-width: 7
#| fig-height: 7
#| layout-ncol: 2
#| results: hide

path <- paste0(results_path, 'DEG_Wilcox_analysis_Lymphocytes_Cross-fostered_vs_Control_major_cell_types/')
unlink(path,recursive = T)
dir.create(path)
Idents(seurat) <- 'major_cell_types'

DEG_counts <- data.frame(matrix(ncol=3, nrow=0))
colnames(DEG_counts) <- c('DEG_count', 'DEG_UP_count', 'DEG_DOWN_count')
rnames <- c()

DEG_counts <- DEG_FindMarkers_SCT_assay(seurat, comparison='Condition', group1='Control', group2='Cross-fostered', cluster= 'all_lymphocytes' , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = F)
rnames <- c(rnames, 'all_lymphocytes')

for (x in levels(seurat$major_cell_types)) {
    # cat(paste0("Cell type: ", x, "\n"))
    seurat_small <- subset(seurat, subset = major_cell_types == x )
    counts <- DEG_FindMarkers_SCT_assay(seurat_small, comparison='Condition', group1='Control', group2='Cross-fostered', cluster= x , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = T)
    rm(seurat_small)
    DEG_counts <- rbind(DEG_counts, counts)
    rnames <- c(rnames, x)
}
rownames(DEG_counts) <- rnames
write.csv(DEG_counts, file = file.path(tables_path, "DEG_counts.csv"))
DEG_counts
```

## Conclusions - lymphocytes 

 


