---
title: Analysis of Abau and Ftul infected mice - lungs
author: Eduard Ansaldo
date: '`r Sys.Date()`'
verbose: false
engine: knitr
knitr:
  opts_chunk: 
    echo: false
    warning: false
    message: false
    error: false
    cache: false
    results: markdown
    code-fold: true
    cache-lazy: false
    
processing: true

execute: 
  cache: true


format:
  html:
    embed-resources: true
    fig-cap-location: top
    toc: true
    toc-depth: 3
    toc-expand: 1
    toc-location: left
    toc-title: Table of Contents
    toc-collapse: true
    number-sections: true
    page-layout: full
    fig-width: 6
    fig-height: 6
    fig-align: 'center'
---

::: {.content-visible when-meta="verbose"}

# Initialization

:::

```{r}
# General R and plotting 
library(conflicted)
library(tidyverse)
library(ggplot2)
library(scales)
library(patchwork)
library(cowplot)
library(gridExtra)
library(ggrepel)
library(stringr)
library(VennDiagram)
library(pheatmap)
library(viridis)
library(here)
library(readxl)

# Single Cell Analysis Packages
library(Seurat)
library(scRepertoire)
library(circlize)
library(scCustomize)
library(SingleR)
library(celldex)
library(UCell)
library(presto)
library(scDblFinder)

# DEG, pathway enrichment and visualization packages
library(DESeq2)
library(gprofiler2)
library(clusterProfiler)
library(DOSE)
library(pathview)
library(org.Mm.eg.db)
library(scRepertoire)
library(enrichplot)
library(msigdbr)
library(rlang)

# Python Version 3.9.6!!! 

knitr::dep_prev()
# peconflicts_scout()
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(base::setdiff)
conflicts_prefer(dplyr::rename)

```

```{r}
#| results: hide 
# Setting working directory and seed
# renv::update()
# renv::snapshot()
# renv::init()
# renv::status()
set.seed(3514)
i_am('scripts/Analysis.qmd')
here()
results_path <- here('results/')
dir.create(results_path)
results_path <- here('results/')
dir.create(results_path)
figures_path <- here('results/figures/')
dir.create(figures_path)
# data_path <- here('data/')
# dir.create(data_path)
tables_path <- here('results/tables/')
dir.create(tables_path)

# Loading custom functions
source(here('scripts/function_template.r'))
source(here('scripts/gProfiler2_functions.r'))

# Setting up color palettes
diverging_palette <- hcl.colors(n = 20,'Purple-Green',rev = T)
sequential_palette_dotplot <- hcl.colors(n = 20,'YlGn',rev = T)
sequential_palette <- hcl.colors(n = 20,'Mako',rev = T)
```

# Initial notes
Data was preprocessed in cellranger 9.0.1 Using the intron=true option, and aggregating all three samples to normalize read depth.

:::{.content-visible when-meta='processing'}

# Preprocessing aggregated dataset

:::

```{r}

#| results: hide
#| #Load the dataset from the cellranger outs
#| results: hide
#Load the dataset from the cellranger outs
scdata <- Read10X(data.dir = here("data/cluster_processing_2/aggr_all/outs/count/filtered_feature_bc_matrix"))

#Initialize the seurat object with the raw (non-normalized data)
seurat <- CreateSeuratObject(counts=scdata$'Gene Expression', min.cells = 3)

# #Add HTO data as a new assay independent from RNA
HTO <- CreateAssayObject(counts = scdata$'Antibody Capture')
seurat[["HTO"]] <- HTO

```

:::{.content-visible when-meta='processing'}

## Splitting dataset

:::

```{r}

seurat$barcode <- Cells(seurat)
seurat@meta.data <- seurat@meta.data |>
    mutate(GEM = str_extract(barcode, '\\d')) |>
    mutate(GEM = case_when(
            GEM == '1' ~ 'CT_MAIT',
            GEM == '2' ~ 'Ftula_MAIT',
            GEM == '3' ~ 'Abau_MAIT'
    )) 
# Splitting the dataset into the three groups
seurat_list <- SplitObject(seurat, split.by = 'GEM')

```

:::{.content-visible when-meta='processing'}

# Preprocessing Abau MAIT dataset

:::

:::{.content-visible when-meta='verbose'}

## Read in data

:::

```{r}
#| results: hide

seurat <- seurat_list$Abau_MAIT

```

:::{.content-visible when-meta='processing'}
#### Confirming used HTOs - MAIT
:::

```{r}
print('Antibody Capture total reads:')
seurat@assays$HTO$counts |> as.matrix() |>
        rowSums()  |> as.data.frame() 

HTO_data <- seurat@assays$HTO$counts

# Extracting HTO and biotin data
HTO_data <-  HTO_data[-c(3), ]
# colnames(scdata$'HTO') <- colnames(scdata$'Antibody Capture')
# rownames(scdata$'HTO') <- rownames(scdata$'Antibody Capture')[-c(2, 3, 4, 5, 6, 7, 11, 17, 18, 19)]

# #Add HTO data as a new assay independent from RNA
HTO <- CreateAssayObject(counts = HTO_data)
seurat[["HTO"]] <- HTO
```

Hashtags Used: 1, 2, 4, 5

:::{.content-visible when-meta='processing'}

## QC

:::

```{r}
#| layout-ncol: 2
#Quantifying percentage mitochondria
Idents(seurat) <- 'Tissue'
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, assay="RNA",pattern = "mt-")

VlnPlot(seurat, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol=3, pt.size = 0.01)

# Setting thresholds:
nCount_RNA_threshold <- 1500
nFeature_RNA_threshold <- 500
percent_mt_threshold <- 6


p1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nCount_RNA_threshold)

p2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nFeature_RNA_threshold)

p3 <- seurat@meta.data %>% 
  	ggplot(aes(x=percent.mt)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = percent_mt_threshold)

grid.arrange(p1, p2, p3)    

# # Visualize feature relationships
# # plot1 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "percent.mt")

# # plot2 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
# # plot1+plot2

cell_number <- list(nrow(seurat[[]]))
```

Filtering Low Quality Cells

```{r}
# Filter data

cell_number <- list(nrow(seurat[[]]))
seurat <- subset(seurat, subset = nFeature_RNA  > nFeature_RNA_threshold & percent.mt < percent_mt_threshold & nCount_RNA > nCount_RNA_threshold)
cell_number <- append(cell_number, nrow(seurat[[]]))

print("Cell number before and after filterings")
print(cell_number)
```

:::{.content-visible when-meta='verbose'}
## Data Normalization 
:::

```{r}
# Normalize RNA data with SCTransform
# seurat <- SCTransform(seurat, verbose = T)

# Normalize HTO data with CLR
seurat <- NormalizeData(seurat, assay = "HTO",normalization.method = "CLR")
```

:::{.content-visible when-meta='processing'}
## Hashtag Demultiplexing - MAIT
:::

```{r}
#| fig-width: 14
#| fig-height: 10
plot1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot1+plot2

seurat <- HTODemux(seurat, assay = "HTO", positive.quantile = 0.99)

table(seurat$hash.ID)
table(seurat$HTO_maxID)
table(seurat$HTO_classification.global)

#tSNE-visualization
seurat_subset <- subset(seurat, idents = "Negative", invert =TRUE)
DefaultAssay(seurat_subset) <- "HTO"
seurat_subset <- ScaleData(seurat_subset, assay = "HTO",  features = rownames(seurat_subset), verbose = FALSE )
seurat_subset <- RunPCA(seurat_subset, assay = "HTO", rownames(seurat_subset), approx = FALSE)
seurat_subset <- RunTSNE(seurat_subset, assay = "HTO", dims = 1:24,  perplexity = 100, check_duplicates=FALSE)
DimPlot(seurat_subset)
```

Hashtag staining and demultiplexing was very successful, with extremely clear separation between the hashtags.

```{r}
print("Number of Singlets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Singlet')))

print("Number of Doublets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Doublet')))

# Filtering singlets only
Idents(seurat) <- "HTO_classification.global"
seurat <-subset(seurat, idents = "Singlet")
```

:::{.content-visible when-meta='verbose'}
## Create Groups
:::

```{r}

seurat@meta.data <- seurat@meta.data |>
    mutate(Groups = 'Abau') |>
    mutate(Samples = str_c(Groups, hash.ID, sep = '_')) |> 
    mutate(Samples = case_when(
            hash.ID == 'Mouse-4' ~ 'Abau_Mouse-3',
            hash.ID == 'Mouse-5' ~ 'Abau_Mouse-4',
            TRUE ~ Samples
    )) |>
    mutate(Samples = factor(Samples, levels = c(
            'Abau_Mouse-1',
            'Abau_Mouse-2',
            'Abau_Mouse-3',
            'Abau_Mouse-4')))

```

```{r}
seurat_Abau <- seurat
```

:::{.content-visible when-meta='processing'}

# Preprocessing Ftula MAIT dataset

:::

:::{.content-visible when-meta='verbose'}

## Read in data

:::

```{r}
#| results: hide

seurat <- seurat_list$Ftula_MAIT

```

:::{.content-visible when-meta='processing'}
#### Confirming used HTOs 
:::

```{r}
print('Antibody Capture total reads:')
seurat@assays$HTO$counts |> as.matrix() |>
        rowSums()  |> as.data.frame() 

HTO_data <- seurat@assays$HTO$counts

# Extracting HTO and biotin data
# HTO_data <-  HTO_data[-c(3), ]
# colnames(scdata$'HTO') <- colnames(scdata$'Antibody Capture')
# rownames(scdata$'HTO') <- rownames(scdata$'Antibody Capture')[-c(2, 3, 4, 5, 6, 7, 11, 17, 18, 19)]

# #Add HTO data as a new assay independent from RNA
HTO <- CreateAssayObject(counts = HTO_data)
seurat[["HTO"]] <- HTO
```

Hashtags Used: 1, 2, 3, 4, 5

:::{.content-visible when-meta='processing'}

## QC

:::

```{r}
#| layout-ncol: 2
#Quantifying percentage mitochondria
Idents(seurat) <- 'Tissue'
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, assay="RNA",pattern = "mt-")

VlnPlot(seurat, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol=3, pt.size = 0.01)

# Setting thresholds:
nCount_RNA_threshold <- 1500
nFeature_RNA_threshold <- 500
percent_mt_threshold <- 6


p1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nCount_RNA_threshold)

p2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nFeature_RNA_threshold)

p3 <- seurat@meta.data %>% 
  	ggplot(aes(x=percent.mt)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = percent_mt_threshold)

grid.arrange(p1, p2, p3)    

# # Visualize feature relationships
# # plot1 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "percent.mt")

# # plot2 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
# # plot1+plot2

cell_number <- list(nrow(seurat[[]]))
```

Filtering Low Quality Cells

```{r}
# Filter data

cell_number <- list(nrow(seurat[[]]))
seurat <- subset(seurat, subset = nFeature_RNA  > nFeature_RNA_threshold & percent.mt < percent_mt_threshold & nCount_RNA > nCount_RNA_threshold)
cell_number <- append(cell_number, nrow(seurat[[]]))

print("Cell number before and after filterings")
print(cell_number)
```

:::{.content-visible when-meta='verbose'}
## Data Normalization 
:::

```{r}
# Normalize RNA data with SCTransform
# seurat <- SCTransform(seurat, verbose = T)

# Normalize HTO data with CLR
seurat <- NormalizeData(seurat, assay = "HTO",normalization.method = "CLR")
```

:::{.content-visible when-meta='processing'}
## Hashtag Demultiplexing - MAIT
:::

```{r}
#| fig-width: 14
#| fig-height: 10
plot1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot1+plot2

seurat <- HTODemux(seurat, assay = "HTO", positive.quantile = 0.99)

table(seurat$hash.ID)
table(seurat$HTO_maxID)
table(seurat$HTO_classification.global)

#tSNE-visualization
seurat_subset <- subset(seurat, idents = "Negative", invert =TRUE)
DefaultAssay(seurat_subset) <- "HTO"
seurat_subset <- ScaleData(seurat_subset, assay = "HTO",  features = rownames(seurat_subset), verbose = FALSE )
seurat_subset <- RunPCA(seurat_subset, assay = "HTO", rownames(seurat_subset), approx = FALSE)
seurat_subset <- RunTSNE(seurat_subset, assay = "HTO", dims = 1:24,  perplexity = 100, check_duplicates=FALSE)
DimPlot(seurat_subset)
```

Hashtag staining and demultiplexing was very successful, with extremely clear separation between the hashtags.

```{r}
print("Number of Singlets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Singlet')))

print("Number of Doublets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Doublet')))

# Filtering singlets only
Idents(seurat) <- "HTO_classification.global"
seurat <-subset(seurat, idents = "Singlet")
```

:::{.content-visible when-meta='verbose'}
## Create Groups
:::

```{r}

seurat@meta.data <- seurat@meta.data |>
    mutate(Groups = 'Ftula') |>
    mutate(Samples = str_c(Groups, hash.ID, sep = '_')) |> 
    mutate(Samples = factor(Samples, levels = c(
            'Ftula_Mouse-1',
            'Ftula_Mouse-2',
            'Ftula_Mouse-3',
            'Ftula_Mouse-4',
            'Ftula_Mouse-5')))

```

```{r}
seurat_Ftula <- seurat
```

:::{.content-visible when-meta='processing'}

# Preprocessing CT MAIT dataset

:::

:::{.content-visible when-meta='verbose'}

## Read in data

:::

```{r}
#| results: hide

seurat <- seurat_list$CT_MAIT

```

:::{.content-visible when-meta='processing'}
#### Confirming used HTOs
:::

```{r}
print('Antibody Capture total reads:')
seurat@assays$HTO$counts |> as.matrix() |>
        rowSums()  |> as.data.frame() 

HTO_data <- seurat@assays$HTO$counts

# Extracting HTO and biotin data
HTO_data <-  HTO_data[-c(3), ]
# colnames(scdata$'HTO') <- colnames(scdata$'Antibody Capture')
# rownames(scdata$'HTO') <- rownames(scdata$'Antibody Capture')[-c(2, 3, 4, 5, 6, 7, 11, 17, 18, 19)]

# #Add HTO data as a new assay independent from RNA
HTO <- CreateAssayObject(counts = HTO_data)
seurat[["HTO"]] <- HTO
```

Hashtags Used: 1, 2, 4, 5

:::{.content-visible when-meta='processing'}

## QC

:::

```{r}
#| layout-ncol: 2
#Quantifying percentage mitochondria
Idents(seurat) <- 'Tissue'
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, assay="RNA",pattern = "mt-")

VlnPlot(seurat, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol=3, pt.size = 0.01)

# Setting thresholds:
nCount_RNA_threshold <- 1500
nFeature_RNA_threshold <- 500
percent_mt_threshold <- 6


p1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nCount_RNA_threshold)

p2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nFeature_RNA_threshold)

p3 <- seurat@meta.data %>% 
  	ggplot(aes(x=percent.mt)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = percent_mt_threshold)

grid.arrange(p1, p2, p3)    

# # Visualize feature relationships
# # plot1 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "percent.mt")

# # plot2 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
# # plot1+plot2

cell_number <- list(nrow(seurat[[]]))
```

Filtering Low Quality Cells

```{r}
# Filter data

cell_number <- list(nrow(seurat[[]]))
seurat <- subset(seurat, subset = nFeature_RNA  > nFeature_RNA_threshold & percent.mt < percent_mt_threshold & nCount_RNA > nCount_RNA_threshold)
cell_number <- append(cell_number, nrow(seurat[[]]))

print("Cell number before and after filterings")
print(cell_number)
```

:::{.content-visible when-meta='verbose'}
## Data Normalization 
:::

```{r}
# Normalize RNA data with SCTransform
# seurat <- SCTransform(seurat, verbose = T)

# Normalize HTO data with CLR
seurat <- NormalizeData(seurat, assay = "HTO",normalization.method = "CLR")
```

:::{.content-visible when-meta='processing'}
## Hashtag Demultiplexing - MAIT
:::

```{r}
#| fig-width: 14
#| fig-height: 10
plot1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot1+plot2

seurat <- HTODemux(seurat, assay = "HTO", positive.quantile = 0.99)

table(seurat$hash.ID)
table(seurat$HTO_maxID)
table(seurat$HTO_classification.global)

#tSNE-visualization
seurat_subset <- subset(seurat, idents = "Negative", invert =TRUE)
DefaultAssay(seurat_subset) <- "HTO"
seurat_subset <- ScaleData(seurat_subset, assay = "HTO",  features = rownames(seurat_subset), verbose = FALSE )
seurat_subset <- RunPCA(seurat_subset, assay = "HTO", rownames(seurat_subset), approx = FALSE)
seurat_subset <- RunTSNE(seurat_subset, assay = "HTO", dims = 1:24,  perplexity = 100, check_duplicates=FALSE)
DimPlot(seurat_subset)
```

Hashtag staining and demultiplexing was very successful, with extremely clear separation between the hashtags.

```{r}
print("Number of Singlets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Singlet')))

print("Number of Doublets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Doublet')))

# Filtering singlets only
Idents(seurat) <- "HTO_classification.global"
seurat <-subset(seurat, idents = "Singlet")
```

:::{.content-visible when-meta='verbose'}
## Create Groups
:::

```{r}

seurat@meta.data <- seurat@meta.data |>
    mutate(Groups = 'CT') |>
    mutate(Samples = str_c(Groups, hash.ID, sep = '_')) |> 
    mutate(Samples=case_when(
            hash.ID == 'Mouse-4' ~ 'CT_Mouse-3',
            hash.ID == 'Mouse-5' ~ 'CT_Mouse-4',
            TRUE ~ Samples
    )) |>  
    mutate(Samples = factor(Samples, levels = c(
            'CT_Mouse-1',
            'CT_Mouse-2',
            'CT_Mouse-3',
            'CT_Mouse-4')))

```

```{r}
seurat_CT <- seurat
```

# Merged processing

## Re-merging datasets

```{r}
data_list <- list(CT_MAIT = seurat_CT, Abau_MAIT = seurat_Abau, Ftula_MAIT = seurat_Ftula)
# Merging the datasets
seurat <- Merge_Seurat_List(data_list, 
                       add.cell.ids = names(data_list),
                       project = "MAIT_Abau_Ftula_CT_Lungs")

seurat <- JoinLayers(seurat, assay = 'RNA')

```

:::{.content-visible when-meta='verbose'}

## Edit Groups

:::

```{r}

seurat@meta.data <- seurat@meta.data |>
    mutate(Groups = factor(Groups, levels = c('CT', 'Abau', 'Ftula'))) |>
    mutate(Samples = factor(Samples, levels = c(
            'CT_Mouse-1',
            'CT_Mouse-2',
            'CT_Mouse-3',
            'CT_Mouse-4',
            'Abau_Mouse-1',
            'Abau_Mouse-2',
            'Abau_Mouse-3',
            'Abau_Mouse-4',
            'Ftula_Mouse-1',
            'Ftula_Mouse-2',
            'Ftula_Mouse-3',
            'Ftula_Mouse-4',
            'Ftula_Mouse-5')))

seurat <- RenameCells(seurat, new.names=seurat$barcode)
```

:::{.content-visible when-meta='verbose'}

## Import TCR data

:::

```{r}
contigs <- read.csv(here('data/cluster_processing_2/aggr_all/outs/vdj_t/filtered_contig_annotations.csv'))

#Demultiplexing VDJ libraries
contig_list <- createHTOContigList(contigs, seurat, group.by = "Samples")
contig_list <- contig_list[c(levels(seurat$Samples))]
# Combining the Contigs
combined <- combineTCR(contig_list, removeNA = TRUE, samples=c(levels(seurat$Samples)), filterMulti = TRUE)

# # #Adding Samples as variables
combined <- addVariable(combined, variable.name='Samples', variables=c(unique(seurat$Samples)))                                    

## Integrating data with seurat object
cell_names <- Cells(seurat)
groups_merge <- pull(seurat@meta.data, Samples)

#Changing barcodes on the seurat object for the merge with VDJ data
new_cell_names <- paste(groups_merge, cell_names, sep='_')
seurat$barcode <- cell_names
seurat <- RenameCells(seurat, new.names=new_cell_names)

# #Combining VDJ and Seurat Object
seurat <- combineExpression(combined, seurat, proportion = TRUE, cloneCall='aa', group.by='Samples', chain = 'both')
# Renaming cells back to original barcodes
seurat <- RenameCells(seurat, new.names=seurat$barcode)
```

:::{.content-visible when-meta='verbose'}

## Data Normalization

:::

```{r}
# Normalize RNA data with SCTransform
seurat <- SCTransform(seurat, verbose = F) 
# seurat@assays$SCT
# seurat <- PrepSCTFindMarkers(seurat)
```

:::{.content-visible when-meta='processing'}

## Clustering

:::

```{r}
#| fig-width: 4
#| fig-height: 4
#| fig-cap: 'Elbow plot showing the percentage of variability represented by by each PC to select number of dimensions '

#Dimensionality reduction
seurat <- RunPCA(seurat,npcs = 100)

#Determining dimensionality of the dataset
ElbowPlot(seurat, ndims = 100)
```

:::{.content-visible when-meta='processing'}

Chosen Number of Dimensions:

:::

```{r}
# Describe number of dimensions

dimensions  <- 25 
dimensions
```

```{r}
#| fig-cap: UMAP plots showing the clustering results at different resolutions
#| fig-width: 10
#| fig-height: 10

seurat <- RunUMAP(seurat, dims=1:dimensions, verbose = F, seed.use = 3514L)
# resolutions <- c(0.25, 0.375, 0.5, 0.625, 0.75, 1, 1.25, 1.5, 1.75)
resolutions <- c(0.05, 0.1, 0.15, 0.25, 0.375, 0.5)
seurat <- FindNeighbors(seurat, dims = 1:dimensions, verbose = F)
seurat <- FindClusters(seurat, resolution = resolutions, verbose = F)

p <- list()
i <- 1
for (resolution in resolutions ) {
    Idents(seurat) <- paste0('SCT_snn_res.', resolution) 
    p2 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")    
    p[[i]] <- p2
    i <- i+1
}

plot <- grid.arrange(grobs = p)
# ggsave('initial_clustering_results_by_resolution.pdf', path = here('result/figures'), plot = plot)
```

:::{.content-visible when-meta='processing'}

Chosen resolution

:::

```{r}
resolution <- 0.375
resolution
```

```{r}
object_annotations <- 'full_object_merged_pre_filtering'
```

### Clustering Results

```{r}
#| layout-ncol: 2
Idents(seurat) <- paste0('SCT_snn_res.', resolution)
seurat[['seurat_clusters']]<- Idents(seurat)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Groups', ncol = 2) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_group_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
# DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Samples', ncol = 4) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
# ggsave(paste0('UMAP_clusters_by_sample_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = FALSE, group.by = 'Groups') + ggtitle(paste0('R ', resolution)) + theme(legend.position = "right")
```

### Top DEGs per cluster

```{r}
#| layout-ncol: 3
# seurat <- PrepSCTFindMarkers(seurat)
plot1 <- top_genes_per_cluster(seurat, n_genes_to_plot = 3, object_annotations = object_annotations, tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = F) 
```

```{r}
#| fig-width: 8
#| fig-height: 12
print(plot1)
ggsave(paste0(figures_path, 'DotPlot_Top3_per_cluster', object_annotations, '.pdf'), width = 8, height = 12)
```
:::{.content-visible when-meta='processing'}

### Automatic cell type annotations

:::

```{r}
local_path <- paste0(figures_path, object_annotations,'_cell_type_annotations')
unlink(local_path,recursive = T)
dir.create(local_path)

# Normalize and scale data
# seurat <- JoinLayers(seurat, assay = 'RNA')
seurat <- NormalizeData(seurat, assay = "RNA", normalization.method = "LogNormalize", scale.fct = 10000)
seurat <- ScaleData(seurat, assay = 'RNA')
```

:::{.content-visible when-meta='processing'}

#### SingleR annotations

:::

:::{.content-visible when-meta='processing'}

Annotations per Cluster

:::

```{r}
#| fig-cap: UMAP plot showing the clustering results with SingleR fine annotations on a per cluster basis
#| fig-width: 9
#| fig-height: 6
seurat <- annotate_seurat_with_SingleR_Eduard(seurat, local_path, database = 'ImmGen', annotation_basis = 'cluster_coarse', split_by_groups = FALSE)
```

:::{.content-visible when-meta='processing'}

Annotations per Cell

:::

```{r}
#| fig-cap: UMAP plot showing the clustering results with SingleR coarse annotations on a per cell basis

annotate_seurat_with_SingleR_Eduard(seurat, local_path, database = 'ImmGen', annotation_basis = 'cell_coarse', split_by_groups = FALSE)
```

### Plotting Genes of Interest for cluster annotation

```{r}
#| fig-cap: Feature plots showing the expression of T cell markers
#| fig-width: 8
#| fig-height: 10
FeaturePlot_scCustom(seurat, features = c('Cd3e', 'Cd4', 'Cd8a', 'Cd8b1', 'Trdc', 'Trac'), pt.size = 0.01, colors_use = sequential_palette) +
    theme(legend.position = 'none')
```

```{r}
#| fig-cap: Feature plots showing the expression of myeloid cell markers
#| fig-width: 8
#| fig-height: 10
FeaturePlot_scCustom(seurat, features = c('Sirpa', 'Xcr1', 'Itgae', 'Itgax', 'Csf1r', 'Csf2ra'), pt.size = 0.1, colors_use = sequential_palette) +
    theme(legend.position = 'none')
```

```{r}
#| fig-cap: Feature plots showing the expression of neuron Markers
#| #| fig-width: 8
#| fig-height: 13
FeaturePlot_scCustom(seurat, features = c('Tubb3', 'Snap25', 'Gad1', 'Slc17a7'), pt.size = 0.1, colors_use = sequential_palette, num_columns = 2) 
```

## Removing contaminating clusters and preprocessing

### Removing macrophage cluster

```{r}
Idents(seurat) <- 'seurat_clusters'
seurat <- subset(seurat, idents = c('10'), invert = TRUE)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
```

:::{.content-visible when-meta='verbose'}

## Data Normalization

:::

```{r}
# Normalize RNA data with SCTransform
seurat <- SCTransform(seurat, verbose = F) 
# seurat@assays$SCT
# seurat <- PrepSCTFindMarkers(seurat)
```

:::{.content-visible when-meta='processing'}

## Clustering
Removing VDJ genes from clustering

:::

```{r}
#| fig-width: 4
#| #| fig-height: 4
#| fig-cap: 'Elbow plot showing the percentage of variability represented by by each PC to select number of dimensions '

seurat <- quietVDJgenes(seurat)

#Dimensionality reduction
seurat <- RunPCA(seurat,npcs = 100)

#Determining dimensionality of the dataset
ElbowPlot(seurat, ndims = 100)
```

:::{.content-visible when-meta='processing'}

Chosen Number of Dimensions:

:::

```{r}
# Describe number of dimensions

dimensions  <- 25 
dimensions
```

```{r}
#| fig-cap: UMAP plots showing the clustering results at different resolutions
#| fig-width: 10
#| fig-height: 10

seurat <- RunUMAP(seurat, dims=1:dimensions, verbose = F, seed.use = 3514L)
# resolutions <- c(0.25, 0.375, 0.5, 0.625, 0.75, 1, 1.25, 1.5, 1.75)
resolutions <- c(0.05, 0.1, 0.15, 0.175, 0.2, 0.225, 0.25, 0.375, 0.5)
seurat <- FindNeighbors(seurat, dims = 1:dimensions, verbose = F)
seurat <- FindClusters(seurat, resolution = resolutions, verbose = F)

p <- list()
i <- 1
for (resolution in resolutions ) {
    Idents(seurat) <- paste0('SCT_snn_res.', resolution) 
    p2 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")    
    p[[i]] <- p2
    i <- i+1
}

plot <- grid.arrange(grobs = p)
# ggsave('initial_clustering_results_by_resolution.pdf', path = here('result/figures'), plot = plot)
```

:::{.content-visible when-meta='processing'}

Chosen resolution

:::

```{r}
resolution <- 0.15
resolution
```

:::{.content-visible when-meta='verbose'}

### Save Object

:::

```{r}
saveRDS(seurat, file = here('data/MAIT_Abau_Ftula_CT_Lungs.rds'))
```

# Analysis

```{r}
seurat <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs.rds'))
object_annotations <- 'full_object_filtered'
resolution <- 0.15
```

## Clustering Results

```{r}
#| layout-ncol: 2
Idents(seurat) <- paste0('SCT_snn_res.', resolution)
seurat[['seurat_clusters']]<- Idents(seurat)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Groups', ncol = 2) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_group_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Samples', ncol = 4) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_sample_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = FALSE, group.by = 'Groups') + ggtitle(paste0('R ', resolution)) + theme(legend.position = "right")
```

### Cluster frequency per condition

```{r}
extract_cell_counts(seurat, seurat_clusters, figures_path, tables_path, object_annotations=object_annotations)
```

## Top DEGs per cluster

```{r}
#| layout-ncol: 3
# seurat <- PrepSCTFindMarkers(seurat)
plot1 <- top_genes_per_cluster(seurat, n_genes_to_plot=5, object_annotations, tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = T) 
```

```{r}
#| fig-width: 8
#| fig-height: 12
print(plot1)
ggsave(paste0(figures_path, 'DotPlot_Top3_per_cluster', object_annotations, '.pdf'), width = 8, height = 12)
```

## Exploring cluster 3

```{r}
#| layout-ncol: 2
seurat[["percent.ribo"]] <- PercentageFeatureSet(seurat, assay="RNA",pattern = "^Rp[sl]")
FeaturePlot_scCustom(seurat, features = c('nCount_RNA', 'nFeature_RNA', 'percent.mt', 'percent.ribo'), colors_use = sequential_palette)
FeaturePlot_scCustom(seurat, features = c('Zbtb16', '', 'Icos'), colors_use = sequential_palette)
VlnPlot(seurat, features = c('nCount_RNA', 'nFeature_RNA', 'percent.mt', 'percent.ribo'), group.by = 'seurat_clusters', pt.size = 0.1) + NoLegend()
VlnPlot(seurat, features = c('Zbtb16', 'Ubc', 'Icos'), group.by = 'seurat_clusters', pt.size = 0.1) + NoLegend()
FeatureScatter(seurat, feature1 = 'nCount_RNA', feature2 = 'nFeature_RNA', split.by = 'seurat_clusters', ncol = 3)
```

Cluster 3 looks like a potential technical artifact. At first glance, the slightly higher mitochondrial gene count and lower RNA counts suggested lower quality cells we could filter out, however the very low ribosomal protein count. And the expression of other expected genes such as PLZF argues against that. If anything, they look higher quality by these metrics. I will test whether regressing out ribosomal protein content and percent mitochondrial genes merges this cluster with the rest of the clusters. If not, wine to figure out whether these are different somehow, perhaps location in the lungs, lymphoid structure, circulation etc. They do express the MAIT TCR, (see below).

## Regressing out ribosomal protein content and percent mitochondrial genes

:::

```{r}
#| fig-width: 4
#| #| fig-height: 4
#| fig-cap: 'Elbow plot showing the percentage of variability represented by by each PC to select number of dimensions '

# Regressing out ribosomal protein content
seurat <- SCTransform(seurat, vars.to.regress = c('percent.ribo', 'percent.mt'), verbose = T)
seurat <- quietVDJgenes(seurat)

#Dimensionality reduction
seurat <- RunPCA(seurat,npcs = 100)

#Determining dimensionality of the dataset
ElbowPlot(seurat, ndims = 100)
```

:::{.content-visible when-meta='processing'}

Chosen Number of Dimensions:

:::

```{r}
# Describe number of dimensions

dimensions  <- 25 
dimensions
```

```{r}
#| fig-cap: UMAP plots showing the clustering results at different resolutions
#| fig-width: 10
#| fig-height: 10

seurat <- RunUMAP(seurat, dims=1:dimensions, verbose = F, seed.use = 3514L)
# resolutions <- c(0.25, 0.375, 0.5, 0.625, 0.75, 1, 1.25, 1.5, 1.75)
resolutions <- c(0.05, 0.1, 0.15, 0.175, 0.2, 0.225, 0.25, 0.375, 0.5)
seurat <- FindNeighbors(seurat, dims = 1:dimensions, verbose = F)
seurat <- FindClusters(seurat, resolution = resolutions, verbose = F)

p <- list()
i <- 1
for (resolution in resolutions ) {
    Idents(seurat) <- paste0('SCT_snn_res.', resolution) 
    p2 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")    
    p[[i]] <- p2
    i <- i+1
}

plot <- grid.arrange(grobs = p)
# ggsave('initial_clustering_results_by_resolution.pdf', path = here('result/figures'), plot = plot)
```

:::{.content-visible when-meta='processing'}

Chosen resolution

:::

```{r}
resolution <- 0.15
resolution
```

:::{.content-visible when-meta='verbose'}

### Save Object

:::

```{r}
saveRDS(seurat, file = here('data/MAIT_Abau_Ftula_CT_Lungs.rds'))
```

# Analysis #2 - post ribosomal and mitochondrial gene regression

```{r}
seurat <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs.rds'))
object_annotations <- 'full_object_filtered'
resolution <- 0.15
```

## Clustering Results

```{r}
#| layout-ncol: 2
Idents(seurat) <- paste0('SCT_snn_res.', resolution)
seurat[['seurat_clusters']]<- Idents(seurat)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Groups', ncol = 2) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_group_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Samples', ncol = 4) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_sample_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = FALSE, group.by = 'Groups') + ggtitle(paste0('R ', resolution)) + theme(legend.position = "right")
```

### Cluster frequency per condition

```{r}
extract_cell_counts(seurat, seurat_clusters, figures_path, tables_path, object_annotations=object_annotations)
```

## Top DEGs per cluster

```{r}
#| layout-ncol: 3
#seurat <- PrepSCTFindMarkers(seurat)
plot1 <- top_genes_per_cluster(seurat, n_genes_to_plot=5, object_annotations, tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = T) 
```

```{r}
#| fig-width: 8
#| fig-height: 12
print(plot1)
ggsave(paste0(figures_path, 'DotPlot_Top3_per_cluster', object_annotations, '.pdf'), width = 8, height = 12)
```

After regressing out mitochondrial, and ribosomal gene content, cluster 3, which was very separate from the rest of the MAIT cell clusters, is now still a distinct cluster (cluster 4) but much closer to the rest of the clusters, suggesting that this is indeed a different cell state. They still express the MAIT TCR (see below).

## Signatures in manuscript

```{r}

DefaultAssay(seurat) <- 'SCT'
signatures_table <- read_excel(here('data/Hagoulet_signatures.xlsx'), skip = 1)
signatures_table <- signatures_table |>
    dplyr::filter(!(row_number() == 1)) |>
    dplyr::select(-Signature) 

vector <- signatures_table$ISG |> purrr::discard(is.na)

signatures <- map(.x = signatures_table, .f = ~ purrr::discard(.x, is.na))

# Removing genes that are not detected
blacklist_elements <- c('H2afz','Smsc2','Hist1h2ap','Flg','Flg2','Bmp2','Bmp4','Ngf','Fgf7','Fgf10','Fgf22','Ctgf','Mmp3','Mmp13','Defb1','Defb6','Epgn','Mmp2','Angpt1','Hgf','Cyr61','Inhbb','Btc','Ereg','Lep','Vegfc','Vegfd','Pdgfc','Cxcl12','Shh','Ihh','Dll3','Dll4','Wnt1','Wnt2','Wnt3a','Wnt5a','Wnt6','Wnt7a','Wnt7b','Wnt8a','Wnt8b','Wnt9a','Wnt9b','Wnt16','Chat','Thbs1','Fam46a','Kcna4','Fam129b','Tdpoz4','Ero1l','Cd244','A430078G23Rik','D1Ertd622e','Fam49a')

signatures <- map(.x = signatures, .f = ~ .x[!(.x %in% blacklist_elements)])


```

### Implementing signatures with UCell

```{r}

seurat <- AddModuleScore_UCell(seurat, features = signatures, name = NULL)

# Scale signatures (z-score) for comparison and visualization purposes
seurat@meta.data  <-  seurat@meta.data |>
    mutate(across(names(signatures), ~ scale(.x, center = T, scale = T), .names = '{col}_scaled'))
```

### Violin Plots of signatures
Violin plots showing the distribution of signature scores (z-scored) across the different conditions

```{r}
#| layout-ncol: 3

# Visualization
signature_violin_plot <- function (signature) {
    plot1 <- VlnPlot(seurat, features = paste0(signature, '_scaled'), group.by = 'Groups', pt.size = 0) + labs(title = signature) + NoLegend()
    print(plot1)
    ggsave(plot = plot1, filename = paste0('VlnPlot_', signature, '_by_group_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 4)
}
for (signature in names(signatures)) {
    signature_violin_plot(signature) 
}


```

Violin plots showing the distribution of signature scores (z-scored) across the different conditions

```{r}
#| layout-ncol: 3

# Visualization
signature_violin_plot <- function (signature) {
    plot1 <- VlnPlot(seurat, features = paste0(signature, '_scaled'), group.by = 'seurat_clusters', pt.size = 0) + NoLegend() + labs(title = signature)
    print(plot1)
    ggsave(plot = plot1, filename = paste0('VlnPlot_', signature, '_by_seurat_clusters_', object_annotations, '.pdf'), path = figures_path, width = 8, height = 4)
}
for (signature in names(signatures)) {
    signature_violin_plot(signature) 
}

```

### Dot Plots of Signature genes
Dot plots showing the expression of signature genes across the different Groups

```{r}
#| fig-width: 4
#| fig-height: 12
#| layout-ncol: 4 

signature_dot_plot <- function (signature) {
    plot1 <- DotPlot_scCustom(seurat, features = signatures[[signature]] |> rev() |> unique() , group.by = 'Groups', colors_use = sequential_palette_dotplot, flip_axes = T, scale = F, dot.scale = 6) + labs(title = signature)
    print(plot1)
    ggsave(plot = plot1, filename = paste0('DotPlot_', signature, '_by_groups_', object_annotations, '.pdf'), path = figures_path, width = 5, height = length(signatures[[signature]])/4 + 1)
}
for (signature in names(signatures)) {
    signature_dot_plot(signature) 
}

```

### FeaturePlots of Signatures 
FeaturePlots showing the expression of signatures across the UMAP

```{r}
#| layout-ncol: 2

diverging_palette_2 <- hcl.colors(n = 20,'RdBu',rev = T)
signature_feature_plot <- function (signature) {
    min_value  <-  min(seurat@meta.data |> pull(paste0(signature, '_scaled')))
    max_value  <-  max(seurat@meta.data |> pull(paste0(signature, '_scaled')))
    cut_off_value <- min(abs(min_value), abs(max_value))
    plot1 <- FeaturePlot_scCustom(seurat, features = paste0(signature, '_scaled'), colors_use = diverging_palette_2, na_cutoff=NULL, min.cutoff = -1*cut_off_value, max.cutoff = cut_off_value) + labs(title = signature)
    print(plot1)
    ggsave(plot = plot1, filename = paste0('FeaturePlot Underscore', signature, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
}
for (signature in names(signatures)) {
    signature_feature_plot(signature) 
}

FeaturePlot_scCustom(seurat, features = c('Sell', 'Lef1', 'S1pr1', 'Cd44'), colors_use = sequential_palette)
DotPlot_scCustom(seurat, features = c('Sell', 'Lef1', 'S1pr1', 'Cd44') , group.by = 'seurat_clusters', colors_use = sequential_palette_dotplot, flip_axes = T, scale = T, dot.scale = 6) + labs(title = 'LN MAIT cell markers')
```


## Differential gene expression analysis

Given that we don't have biological replicates, I am performing a Wilcoxon test on the existing single-cell dataset, comparing cross-fostering and control conditions for each cell type. Please note that this test is usually not recommended since it treats each cell as a biological replicate, thus leading to a lot of false positives. In general, it is recommended to include biological replicates in the experiment using different hashtags for each mouse, if possible.

Please note that more informative pathway enrichment analysis results can be obtained from Metascape.org

```{r}
#| fig-width: 7
#| fig-height: 7
#| layout-ncol: 2
#| results: hide

gProfiler2 <- TRUE 

path <- paste0(results_path, 'DEG_Wilcox_analysis_Lymphocytes_Abau_vs_Ftula_seurat_clusters')
unlink(path,recursive = T)
dir.create(path)
Idents(seurat) <- 'seurat_clusters'

DEG_counts <- data.frame(matrix(ncol=3, nrow=0))
colnames(DEG_counts) <- c('DEG_count', 'DEG_UP_count', 'DEG_DOWN_count')
rnames <- c()

DEG_counts <- DEG_FindMarkers_SCT_assay(seurat, comparison='Groups', group1='Ftula', group2='Abau', cluster= 'all_MAITs' , path=path, label_threshold = 10000, max_overlaps = 20,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = F)
rnames <- c(rnames, 'all_MAITs')

for (x in levels(seurat$seurat_clusters)) {
    # cat(paste0("Cell type: ", x, "\n"))
    seurat_small <- subset(seurat, subset = seurat_clusters == x )
    counts <- DEG_FindMarkers_SCT_assay(seurat_small, comparison='Groups', group1='Ftula', group2='Abau', cluster= x , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = T)
    rm(seurat_small)
    DEG_counts <- rbind(DEG_counts, counts)
    rnames <- c(rnames, x)
}
rownames(DEG_counts) <- rnames
write.csv(DEG_counts, file = file.path(path, "DEG_counts.csv"))
DEG_counts
```

### Metascape results for all MAITs, up in Abau

![](../results/Metascape_results_all_MAITs_up_in_Abau/Enrichment_heatmap/HeatmapSelectedGO.pdf)

### Metascape results for all MAITs, up in Ftula

![](../results/Metascape_results_all_MAITs_up_in_Ftula/Enrichment_heatmap/HeatmapSelectedGO.pdf)
 
## cNMF analysis


Work in progress

```{r}


# library(SeuratDisk)
# seurat_local <- seurat
# # DefaultAssay(seurat_local) <- 'RNA'
# # seurat_local <- DietSeurat(seurat_local, assays = c("RNA", 'SCT'), dimreducs = c("pca", "umap"), layers = c('counts'))
# SaveH5Seurat(seurat_local, filename = here('data', "seurat.h5Seurat"), overwrite = T)
# SeuratDisk::Convert(source = here('data', "seurat.h5Seurat"), dest = "h5ad", overwrite = T)

```

## MiloR Differential Abundance Analysis

Work in progress

# TCR repertoire Analysis

## Clonotype Frequency

```{r}
#| fig.width: 8
#| fig.height: 8
table(seurat$cloneSize)
#UMAP clonotype frequency
slot(seurat, "meta.data")$cloneSize <- factor(slot(seurat, "meta.data")$cloneSize, 
                levels = c(
                           "Small (1e-04 < X <= 0.001)",
                            'Medium (0.001 < X <= 0.01)', 
                            'Large (0.01 < X <= 0.1)',
                            'Hyperexpanded (0.1 < X <= 1)',
                                                        NA)) 
DimPlot_scCustom(seurat, group.by = "cloneSize", pt.size = 0.5, order=T, colors_use = viridis(length(levels(seurat$cloneSize)))) 
ggsave(filename = 'UMAP_VDJ_clone_frequencies.pdf', path = figures_path, width = 8, height = 8)
DimPlot_scCustom(seurat, group.by = "cloneSize", pt.size = 0.5, order=T, colors_use = viridis(length(levels(seurat$cloneSize))), split.by = 'Groups')
ggsave(filename = 'UMAP_VDJ_clone_frequencies_by_group.pdf', path = figures_path, width = 13, height = 8)  
```

```{r}
#| fig-width: 9
#| fig-height: 6
DimPlot_scCustom(seurat, group.by = "cloneSize", pt.size = 0.5, order=T, colors_use = viridis(length(levels(seurat$cloneSize))), split.by = 'Samples', num_columns = 4)& theme(text = element_text(size = 7))
ggsave(filename = 'UMAP_VDJ_clone_frequencies_by_sample.pdf', path = figures_path, width = 9, height = 6)
```

:::{.content-visible when-meta='verbose'}
## Parsing alpha and beta gene usage
:::

```{r}
#| fig.width: 8
#| fig.height: 8

# Checking for full TCR sequences
seurat@meta.data <- seurat@meta.data |>
    mutate(TCR_exists = ifelse(is.na(CTaa), 'NO', 'YES'),
        TCR_full = ifelse(str_detect(CTaa, '_NA') | str_detect(CTaa, 'NA_'), 'NO', 'YES'))
#scRNAseq_test <- subset(seurat, subset = TCR_full == 'YES')
print('Number of total cells: ')
seurat@meta.data  |> nrow()
print('Number of cells with either TCR sequence: ')
seurat@meta.data |> filter(TCR_exists == 'YES') |> nrow()
print('Number of cells with full TCR sequences detected: ')
seurat@meta.data |> filter(TCR_full == 'YES') |> nrow()

# Extracting the TCR alpha chain
local_dataframe <- seurat@meta.data |> dplyr::select(starts_with('CT')) 
local_dataframe <- local_dataframe |>
                            separate_wider_delim(CTgene,delim = '_',names = c('CT_alpha', 'CT_beta'), cols_remove = FALSE)  |>
                            separate_wider_delim(CT_alpha,delim = '.',names = c('CT_V_alpha', 'CT_J_alpha', NA), cols_remove = FALSE, too_many = 'drop')  |>
                            mutate(CT_alpha_final = str_c(CT_V_alpha, '_', CT_J_alpha))
seurat$CT_alpha_final <- pull(local_dataframe, CT_alpha_final)


# Extracting the TCR beta chain
local_dataframe <- seurat@meta.data |> dplyr::select(starts_with('CT')) 
local_dataframe <- local_dataframe |>
                            separate_wider_delim(CTgene,delim = '_',names = c('CT_alpha', 'CT_beta'), cols_remove = FALSE)  |>
                            separate_wider_delim(CT_beta,delim = '.',names = c('CT_V_beta', 'CT_J_beta', NA), cols_remove = FALSE, too_many = 'drop')  |>
                            mutate(CT_beta_final = str_c(CT_V_beta, '_', CT_J_beta))
seurat$CT_beta_final <- pull(local_dataframe, CT_beta_final)


# for (column in seurat@meta.data |> dplyr::select(starts_with('CT'))) {
#     print(colnames(column))
#     print(length(column))
# }
```

## Annotating known invariant TCRs

### iNKTs (TRAV11_TRAJ18)

```{r}
#| layout-ncol: 2
#| fig.width: 8
#| fig.height: 8

Idents(seurat) <- 'CT_alpha_final'
seurat@meta.data <- mutate(seurat@meta.data, highlight=ifelse(CT_alpha_final == "TRAV11_TRAJ18", 'TRAV11_TRAJ18', 'Other'))

seurat$highlight <- factor(seurat$highlight, levels=c('Other','TRAV11_TRAJ18'))

DimPlot_scCustom(seurat, group.by = "highlight", order = T, pt.size = 1, colors_use = hcl.colors(n = 2, palette = 'ag_GrnYl')) + ggtitle('iNKTs (TRAV11_TRAJ18)')
ggsave(filename = 'UMAP_VDJ_iNKTs.pdf', width = 6, height = 5, path = figures_path)
DimPlot_scCustom(seurat, group.by = "highlight", order = T, pt.size = 1, colors_use = hcl.colors(n = 2, palette = 'ag_GrnYl'), split.by = 'Groups') & labs(color = 'iNKT TCR')
ggsave(filename = 'UMAP_VDJ_iNKTs_by_group.pdf', width = 8, height = 8, path = figures_path)
```

### MAITs (TRAV1_TRAJ33)

```{r}
#| layout-ncol: 2
#| fig.width: 8
#| fig.height: 8

#MAITs
Idents(seurat) <- 'CT_alpha_final'
seurat@meta.data <- mutate(seurat@meta.data, highlight_MAIT=ifelse(CT_alpha_final == "TRAV1_TRAJ33", 'TRAV1_TRAJ33', 'Other'))

seurat$highlight_MAIT <- factor(seurat$highlight_MAIT, levels=c('Other','TRAV1_TRAJ33'))
DimPlot_scCustom(seurat, group.by = "highlight_MAIT", order = T, pt.size = 1, colors_use = hcl.colors(n = 2, palette = 'ag_GrnYl')) + ggtitle('MAIT TCR (TRAV1_TRAJ33)') + labs(color = 'MAIT TCR')
ggsave(filename = 'UMAP_VDJ_MAIT.pdf', width = 6, height = 5, path = figures_path)
DimPlot_scCustom(seurat, group.by = "highlight_MAIT", order = T, pt.size = 1, colors_use = hcl.colors(n = 2, palette = 'ag_GrnYl'), split.by = 'Groups') & labs(color = 'MAIT TCR')
ggsave(filename = 'UMAP_VDJ_MAIT_by_group.pdf', width = 8, height = 8, path = figures_path)  
```

```{r}
DimPlot_scCustom(seurat, group.by = "highlight_MAIT", order = T, pt.size = 1, colors_use = hcl.colors(n = 2, palette = 'ag_GrnYl'), split.by = 'Samples', num_columns = 4) & labs(color = 'MAIT TCR') & theme(text = element_text(size = 7))
ggsave(filename = 'UMAP_VDJ_MAIT_by_sample.pdf', width = 8, height = 8, path = figures_path)
```



### QFLs (TRAV9D-3_TRAJ21)

```{r}
#| fig.width: 8
#| fig.height: 8

#QFL
Idents(seurat) <- 'CT_alpha_final'
seurat@meta.data <- mutate(seurat@meta.data, highlight_QFL=ifelse(CT_alpha_final == "TRAV9D-3_TRAJ21", 'TRAV9D-3_TRAJ21', 'Other'))

seurat$highlight_QFL <- factor(seurat$highlight_QFL, levels=c('Other','TRAV9D-3_TRAJ21'))
DimPlot_scCustom(seurat, group.by = "highlight_QFL", order = T, pt.size = 1, colors_use = hcl.colors(n = 2, palette = 'ag_GrnYl')) + ggtitle('QFL TCR (TRAV9D-3_TRAJ21)')
ggsave(filename = 'UMAP_VDJ_QFL.pdf', width = 8, height = 5, path = figures_path)
DimPlot_scCustom(seurat, group.by = "highlight_QFL", order = T, pt.size = 1, colors_use = hcl.colors(n = 2, palette = 'ag_GrnYl'), split.by = 'Groups') & labs(color = 'QFL TCR')

ggsave(filename = 'UMAP_VDJ_QFL_by_group.pdf', width = 8, height = 8, path = figures_path)
```

## Repertoire Diversity

### D50 diversity measure

```{r}
#| output: false
#Extracting TCR data for clusters of interest
Idents(seurat) <- 'Groups'
combined2 <- scRepertoire:::.expression2List(seurat, split.by ='ident')

cell_types <- unique(seurat$Origin)

#Initiating results data frame
results <- as.data.frame(matrix(nrow = 0,ncol = length(cell_types)))
# colnames(results)
rnames <- c()

#Calculate D50
for (HTO in unique(seurat$Groups)) {
    
    result <- c()

    for (cell_type in cell_types) {
        
        #Extracting data for cell type and HTO
        cell_type_HTO_data <- combined2[[cell_type]]|>
                                    filter(Groups == HTO) |>
                                    dplyr::select(c('Groups', 'CTaa')) |>
                                    add_count(CTaa, sort=TRUE)

        #Calculating D50
        if (nrow(cell_type_HTO_data) < 20) {
            D50 <- NA 
        } else {
            L50 <- floor(nrow(cell_type_HTO_data)/2)
            number_unique_50 <- cell_type_HTO_data[1:L50,] %>% summarise(n_distinct(CTaa)) %>% as.numeric()
            number_unique_total <- cell_type_HTO_data[] %>% summarise(n_distinct(CTaa)) %>% as.numeric()
            D50 <- number_unique_50/number_unique_total
        }
        result <- c(result, D50)
        names <- c(names, cell_type)
    }
    results <- rbind(results, result)
    rnames <- c(rnames, HTO)
    print(HTO)

}
colnames(results) <- paste0(cell_types, '_D50')
results <- results %>% mutate(Mouse=rnames) %>% arrange(Mouse) |> relocate(Mouse)

write.csv(results, file=paste0(results_path, 'D50_per_major_cell_type.csv'), row.names=FALSE)
head(results)
```

```{r}
# Convert results to long format for plotting
results_long <- results %>%
    pivot_longer(-Mouse, names_to = "cell_type", values_to = "D50")

# Remove NA values for plotting
results_long <- results_long %>% filter(!is.na(D50))

# Plot: x axis is Mouse, show only dots (no columns)
ggplot(results_long, aes(x = Mouse, y = D50, color = cell_type)) +
    geom_jitter(width = 0.15, size = 2, show.legend = TRUE) +
    theme_minimal() +
    labs(x = "Mouse", y = "D50 Diversity", title = "D50 per Group by Origin") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 14),
          axis.line = element_line(colour = "black")) +
    scale_y_continuous(limits = c(0, 0.5))
ggsave(filename = paste0(figures_path, 'D50_per_origin.pdf'))
```

```{r}
#| output: false
#Extracting TCR data for clusters of interest
Idents(seurat) <- 'cell_types'
combined2 <- scRepertoire:::.expression2List(seurat, split.by ='ident')

cell_types <- levels(seurat$cell_types)

#Initiating results data frame
results <- as.data.frame(matrix(nrow = 0,ncol = length(cell_types)))
# colnames(results)
rnames <- c()

#Calculate D50
for (HTO in unique(seurat$Groups)) {
    
    result <- c()
    names <- c()

    for (cell_type in cell_types) {
        
        names <- c(names, cell_type)
        
        #Extracting data for cell type and HTO
        cell_type_HTO_data <- combined2[[cell_type]]|>
                                    filter(Groups == HTO) |>
                                    dplyr::select(c('Groups', 'CTaa')) |>
                                    add_count(CTaa, sort=TRUE)

        #Calculating D50
        if (nrow(cell_type_HTO_data) < 20) {
            D50 <- NA 
        } else {
            L50 <- floor(nrow(cell_type_HTO_data)/2)
            number_unique_50 <- cell_type_HTO_data[1:L50,] %>% summarise(n_distinct(CTaa)) %>% as.numeric()
            number_unique_total <- cell_type_HTO_data[] %>% summarise(n_distinct(CTaa)) %>% as.numeric()
            D50 <- number_unique_50/number_unique_total
        }
        result <- c(result, D50)
        names(result) <- names
    }
    results <- rbind(results, result)
    rnames <- c(rnames, HTO)
    print(HTO)

}

# print(rnames)
# nrow(results)
colnames(results) <- paste0(cell_types, '_D50')
# colnames(results)

results <- results %>% mutate(Mouse=rnames) %>% arrange(Mouse) |> relocate(Mouse)
results

write.csv(results, file=paste0(results_path, 'D50_per_cell_type.csv'), row.names=FALSE)
```

```{r}
# Convert results to long format for plotting
results_long <- results %>%
    pivot_longer(-Mouse, names_to = "cell_type", values_to = "D50")

# Remove NA values for plotting
results_long <- results_long %>% filter(!is.na(D50))

# Plot: x axis is Mouse, show only dots (no columns)
ggplot(results_long, aes(x = Mouse, y = D50, color = cell_type)) +
    geom_jitter(width = 0.15, size = 2, show.legend = TRUE) +
    theme_minimal() +
    labs(x = "Mouse", y = "D50 Diversity", title = "D50 per Group by Cell Type") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size = 14),
          axis.line = element_line(colour = "black")) +
    scale_y_continuous(limits = c(0, 0.5)) 
    # scale_color_manual(values = color_palette)
```

```{r}
#| include: false
#| vscode: {languageId: r}
# #Extracting TCR data for clusters of interest
# Idents(scRNAseq) <- 'cell_types'
# combined2 <- scRepertoire:::.expression2List(scRNAseq, split.by ='ident')[

# diversity_measure_results <- clonalDiversity(combined2, cloneCall = 'aa', exportTable = T, n.boots = 100)
# #diversity_measure_results <- diversity_measure_results |>
#     #separate_wider_delim(Group, '_', names = c('cell_type', 'Sample'))
# write.csv(diversity_measure_results, file=paste0(path, 'diversity_measure_results.csv'), row.names=T)
```


## Overlap Analysis 

### Morisita Overlap Index of all cells


```{r}
immunarch_table <- exportClones(seurat, format = 'immunarch', write.file = FALSE, group.by = 'Samples')
overlap <- repOverlap(immunarch_table$data, .method = 'morisita', .verbose=FALSE)
vis(overlap) + ggtitle('Morisita overlap index')
ggsave(filename = paste0(figures_path, 'morisita_overlap.pdf'))


combined2 <- scRepertoire:::.expression2List(seurat, split.by ='Samples')
morisita_table <- clonalOverlap(combined2, cloneCall = 'aa', chain = 'both',exportTable = T,method = 'morisita')
write.csv(morisita_table, file = paste0(results_path, 'morisita_table.csv'))
```

### Morisita Overlap Index of cell types

```{r}
#| fig-width: 10
#| fig-height: 8
immunarch_table <- exportClones(seurat, format = 'immunarch', write.file = FALSE, group.by = 'cell_types')
overlap <- repOverlap(immunarch_table$data, .method = 'morisita', .verbose=FALSE)
vis(overlap) + ggtitle('Morisita overlap index') + labs(x = NULL, y = NULL)
ggsave(filename = paste0(figures_path, 'morisita_overlap_cell_types.pdf'), width = 10, height = 8)

combined2 <- scRepertoire:::.expression2List(seurat, split.by ='cell_types')
morisita_table <- clonalOverlap(combined2, cloneCall = 'aa', chain = 'both',exportTable = T,method = 'morisita')
write.csv(morisita_table, file = paste0(results_path, 'morisita_table_cell_types.csv'))
```

########## Continue Here 08/21/25 ##########

### Overlap Coefficient Index of all cells

```{r}
immunarch_table <- exportClones(seurat, format = 'immunarch', write.file = FALSE, group.by = 'Samples')
overlap <- repOverlap(immunarch_table$data, .method = 'overlap', .verbose=FALSE)
vis(overlap) + ggtitle('Overlap Coefficient')
ggsave(filename = paste0(figures_path, 'overlap_coefficient.pdf'))

combined2 <- scRepertoire:::.expression2List(seurat, split.by ='Samples')
overlap_coefficient <- clonalOverlap(combined2, cloneCall = 'aa', chain = 'both',exportTable = T,method = 'overlap')
write.csv(overlap_coefficient, file = paste0(results_path, 'overlap_coefficient.csv'))
```

### Overlap Coefficient Index of cell types

```{r}
#| vscode: {languageId: r}
immunarch_table <- exportClones(seurat, format = 'immunarch', write.file = FALSE, group.by = 'cell_types')
overlap <- repOverlap(immunarch_table$data, .method = 'overlap', .verbose=FALSE)
vis(overlap) + ggtitle('Overlap coefficient')
ggsave(filename = paste0(figures_path, 'overlap_coefficient_cell_types.pdf'))


combined2 <- scRepertoire:::.expression2List(seurat, split.by ='Samples')
overlap_coefficient <- clonalOverlap(combined2, cloneCall = 'aa', chain = 'both',exportTable = T,method = 'overlap')
write.csv(overlap_coefficient, file = paste0(results_path, 'overlap_coefficient_overlap_coefficient_cell_types.csv'))
```

## Visualizing Clonal Overlap: Circos plots

### Overlap Between Samples

```{r}
#| results: hide
color_palette <-c(
        "MAIT-B2m-Cre--uninfected"  =  'skyblue1', 
        "MAIT-B2m-Cre+-uninfected"  =  'skyblue3', 
        "MAIT-B2m-Cre--Abau"  =  'slateblue1',       
        "MAIT-B2m-Cre+-Abau"  =  'slateblue3',       
        "abDN-B2m-Cre--uninfected"  =  'wheat1', 
        "abDN-B2m-Cre+-uninfected"  =  'wheat3', 
        "abDN-B2m-Cre--Abau"  =  'tan1',       
        "abDN-B2m-Cre+-Abau"  =  'tan3')        
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

grouping_variable <- 'Samples'
variables_to_color_by <- NULL

overlap_circos_and_tables(
    seurat,
    grouping_variable = 'Samples',
    results_path = results_path,
    variables_to_color_by = NULL,
    cell_types_column = 'cell_types',
    write_table = TRUE,
    figures_path = figures_path,
    circle_margin = 1,
    major_ticks = 200,
    cex = 0.6,
    samples = levels(seurat$Samples),
    groups = levels(seurat$Groups),
    color_palette = color_palette)
```

```{r}
#| fig-height: 6
path <- paste0(figures_path, 'Circos_clonotypes_per_', grouping_variable, '.png')
knitr::include_graphics(path, dpi = 100)
```

### Overlap Between cell types

```{r}
#| results: hide 
color_palette <-c(
        "MAIT-B2m-Cre--uninfected"  =  'skyblue1', 
        "MAIT-B2m-Cre+-uninfected"  =  'skyblue3', 
        "MAIT-B2m-Cre--Abau"  =  'slateblue1',       
        "MAIT-B2m-Cre+-Abau"  =  'slateblue3',       
        "abDN-B2m-Cre--uninfected"  =  'wheat1', 
        "abDN-B2m-Cre+-uninfected"  =  'wheat3', 
        "abDN-B2m-Cre--Abau"  =  'tan1',       
        "abDN-B2m-Cre+-Abau"  =  'tan3')        
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

grouping_variable <- 'cell_types'
variables_to_color_by <- NULL

TCR_data <- overlap_circos_and_tables(
    seurat,
    grouping_variable = grouping_variable,
    results_path = results_path,
    variables_to_color_by = NULL,
    cell_types_column = 'cell_types',
    write_table = TRUE,
    figures_path = figures_path,
    circle_margin = 1,
    major_ticks = 200,
    cex = 0.6,
    samples = levels(seurat$cell_types),
    groups = levels(seurat$cell_types),
    color_palette = NULL)
```

```{r}
#| fig-height: 6
path <- paste0(figures_path, 'Circos_clonotypes_per_', grouping_variable, '.png') 
knitr::include_graphics(path, dpi = 100)
```

## Gene usage analysis

### TRBV gene usage

```{r}
#| fig-width: 10
#| fig-height: 12
combined2 <- scRepertoire:::.expression2List(seurat, split.by ='cell_types')

vizGenes(combined2,
         x.axis = "TRBV",
         y.axis = NULL, # No specific y-axis variable, will group all samples
         plot = "barplot",
         summary.fun = "percent") + theme(text = element_text(size = 10))

ggsave(filename = paste0(figures_path, 'TRBV_gene_usage_per_cell_type.pdf'), width = 12, height = 18)

combined4 <- scRepertoire:::.expression2List(seurat, split.by ='Samples')

vizGenes(combined4,
         x.axis = "TRBV",
         y.axis = NULL, # No specific y-axis variable, will group all samples
         plot = "barplot",
         summary.fun = "percent") + theme(text = element_text(size = 10
         ))

ggsave(filename = paste0(figures_path, 'TRBV_gene_usage_per_samples.pdf'), width = 12, height = 18)
```

### TRBV TRBJ gene parings

```{r}
#| fig-width: 12
#| fig-height: 16
vizGenes(combined4,
         x.axis = "TRBV",
         y.axis = "TRBJ",
         plot = "heatmap",
         summary.fun = "percent") # Display percentages

ggsave(filename = paste0(figures_path, 'TRBV_TRBJ_gene_pairings.pdf'), width = 12, height = 16)
```

### TRAV gene usage

```{r}
#| fig-width: 16
#| fig-height: 14


vizGenes(combined4,
         x.axis = "TRAV",
         y.axis = NULL, # No specific y-axis variable, will group all samples
         plot = "barplot",
         summary.fun = "percent") 

ggsave(filename = paste0(figures_path, 'TRAV_gene_usage_barplot.pdf'), width = 22, height = 18)
```

### TRAV TRAJ gene parings

```{r}
#| fig-width: 20
#| fig-height: 30
#| fig.width: 12
vizGenes(combined4,
         x.axis = "TRAV",
         y.axis = "TRAJ",
         plot = "heatmap",
         summary.fun = "percent")

ggsave(filename = paste0(figures_path, 'TRAV_TRAJ_pairings_heatmap.pdf'), width = 20, height = 30)
```

## Expanded Clonotypes

### Top  expanded clonotypes

Please note the clonotypes are defined at the level of the Amino Acid CDR3 region only. 


```{r}
#| vscode: {languageId: r}
head(TCR_data, n = 30) |> knitr::kable()
```

# Conclusions