---
title: Analysis of Abau and Ftul infected mice - lungs
author: Eduard Ansaldo
date: '`r Sys.Date()`'
verbose: false
engine: knitr
knitr:
  opts_chunk: 
    echo: false
    warning: false
    message: false
    error: false
    cache: false
    results: markdown
    code-fold: true
    cache-lazy: false
    
processing: true

execute: 
  cache: true


format:
  html:
    embed-resources: true
    fig-cap-location: top
    toc: true
    toc-depth: 3
    toc-expand: 1
    toc-location: left
    toc-title: Table of Contents
    toc-collapse: true
    number-sections: true
    page-layout: full
    fig-width: 6
    fig-height: 6
    fig-align: 'center'
---

::: {.content-visible when-meta="verbose"}

# Initialization

:::

```{r}
# General R and plotting 
library(conflicted)
library(tidyverse)
library(ggplot2)
library(scales)
library(patchwork)
library(cowplot)
library(gridExtra)
library(ggrepel)
library(stringr)
library(VennDiagram)
library(pheatmap)
library(viridis)
library(here)
library(readxl)

# Single Cell Analysis Packages
library(Seurat)
library(scRepertoire)
library(circlize)
library(scCustomize)
library(SingleR)
library(celldex)
library(UCell)
library(presto)
library(scDblFinder)

# DEG, pathway enrichment and visualization packages
library(DESeq2)
library(gprofiler2)
library(clusterProfiler)
library(DOSE)
library(pathview)
library(org.Mm.eg.db)
library(scRepertoire)
library(enrichplot)
library(msigdbr)
library(rlang)

# Python Version 3.9.6!!! 

knitr::dep_prev()
conflict_scout()
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")

```

```{r}
#| results: hide 
# Setting working directory and seed
# renv::update()
# renv::snapshot()
# renv::init()
# renv::status()
set.seed(3514)
i_am('scripts/Analysis_aggregated.qmd')
here()
results_path <- here('results/')
dir.create(results_path)
results_path <- here('results/')
dir.create(results_path)
figures_path <- here('results/figures/')
dir.create(figures_path)
# data_path <- here('data/')
# dir.create(data_path)
tables_path <- here('results/tables/')
dir.create(tables_path)

# Loading custom functions
source(here('scripts/function_template.r'))
source(here('scripts/gProfiler2_functions.r'))

# Setting up color palettes
diverging_palette <- hcl.colors(n = 20,'Purple-Green',rev = T)
sequential_palette_dotplot <- hcl.colors(n = 20,'YlGn',rev = T)
sequential_palette <- hcl.colors(n = 20,'Mako',rev = T)
```

:::{.content-visible when-meta='processing'}

# Preprocessing Abau MAIT dataset

:::

:::{.content-visible when-meta='verbose'}

## Read in data

:::

```{r}
#| results: hide
#| #Load the dataset from the cellranger outs
#| results: hide
#Load the dataset from the cellranger outs
scdata <- Read10X(data.dir = here("data/cluster_processing_2/aggr_all/outs/count/filtered_feature_bc_matrix"))

#Initialize the seurat object with the raw (non-normalized data)
seurat <- CreateSeuratObject(counts=scdata$'Gene Expression', min.cells = 3)

# #Add HTO data as a new assay independent from RNA
HTO <- CreateAssayObject(counts = scdata$'Antibody Capture')
seurat[["HTO"]] <- HTO
```

:::{.content-visible when-meta='processing'}
#### Confirming used HTOs
:::

```{r}
print('Antibody Capture total reads:')
scdata$`Antibody Capture` |> as.matrix() |>
        rowSums()  |> as.data.frame() 
```

:::{.content-visible when-meta='processing'}

## QC

:::

```{r}
#| layout-ncol: 2
#Quantifying percentage mitochondria
Idents(seurat) <- 'Tissue'
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, assay="RNA",pattern = "mt-")

VlnPlot(seurat, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol=3, pt.size = 0.01)

# Setting thresholds:
nCount_RNA_threshold <- 1500
nFeature_RNA_threshold <- 500
percent_mt_threshold <- 6


p1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nCount_RNA_threshold)

p2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nFeature_RNA_threshold)

p3 <- seurat@meta.data %>% 
  	ggplot(aes(x=percent.mt)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = percent_mt_threshold)

grid.arrange(p1, p2, p3)    

# # Visualize feature relationships
# # plot1 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "percent.mt")

# # plot2 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
# # plot1+plot2

cell_number <- list(nrow(seurat[[]]))
```

Filtering Low Quality Cells

```{r}
# Filter data

cell_number <- list(nrow(seurat[[]]))
seurat <- subset(seurat, subset = nFeature_RNA  > nFeature_RNA_threshold & percent.mt < percent_mt_threshold & nCount_RNA > nCount_RNA_threshold)
cell_number <- append(cell_number, nrow(seurat[[]]))

print("Cell number before and after filterings")
print(cell_number)
```

:::{.content-visible when-meta='verbose'}
## Data Normalization 
:::

```{r}
# Normalize RNA data with SCTransform
# seurat <- SCTransform(seurat, verbose = T)

# Normalize HTO data with CLR
seurat <- NormalizeData(seurat, assay = "HTO",normalization.method = "CLR")
```

:::{.content-visible when-meta='processing'}
## Hashtag Demultiplexing - MAIT
:::

```{r}
#| fig-width: 14
#| fig-height: 10
plot1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot1+plot2

seurat <- HTODemux(seurat, assay = "HTO", positive.quantile = 0.99)

table(seurat$hash.ID)
table(seurat$HTO_maxID)
table(seurat$HTO_classification.global)

#tSNE-visualization
seurat_subset <- subset(seurat, idents = "Negative", invert =TRUE)
DefaultAssay(seurat_subset) <- "HTO"
seurat_subset <- ScaleData(seurat_subset, assay = "HTO",  features = rownames(seurat_subset), verbose = FALSE )
seurat_subset <- RunPCA(seurat_subset, assay = "HTO", rownames(seurat_subset), approx = FALSE)
seurat_subset <- RunTSNE(seurat_subset, assay = "HTO", dims = 1:24,  perplexity = 100, check_duplicates=FALSE)
DimPlot(seurat_subset)
```

Hashtag staining and demultiplexing was very successful, with extremely clear separation between the hashtags.

```{r}
print("Number of Singlets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Singlet')))

print("Number of Doublets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Doublet')))

# Filtering singlets only
Idents(seurat) <- "HTO_classification.global"
seurat <-subset(seurat, idents = "Singlet")
```

:::{.content-visible when-meta='verbose'}
## Create Groups
:::

```{r}
seurat$barcodes <- Cells(seurat)
seurat@meta.data <- seurat@meta.data |> 
    mutate(Groups = case_when(
        str_detect(barcodes,'1') ~ 'CT',
        str_detect(barcodes,'2') ~ 'Ftul',
        str_detect(barcodes,'3') ~ 'Abau',
        TRUE ~ 'Unknown'
    )) |>
    mutate(Samples = str_c(Groups, hash.ID, sep = '_'))  
unique(seurat$Groups)
unique(seurat$Samples)

```

:::{.content-visible when-meta='verbose'}
## Import TCR data
:::

```{r}
contigs <- read.csv(here('data/cluster_processing_2/aggr_all/outs/vdj_t/filtered_contig_annotations.csv'))


#Demultiplexing VDJ libraries
contig_list <- createHTOContigList(contigs, seurat, group.by = "Samples")
contig_list <- contig_list[c(unique(seurat$Samples))]

# Combining the Contigs
combined <- combineTCR(contig_list, removeNA = TRUE, samples=c(unique(seurat$Samples)), filterMulti = TRUE)

# # #Adding Samples as variables
combined <- addVariable(combined, variable.name='Samples', variables=c(unique(seurat$Samples)))                                    

## Integrating data with seurat object
cell_names <- Cells(seurat)
groups_merge <- pull(seurat@meta.data, Samples)

#Changing barcodes on the seurat object for the merge with VDJ data
new_cell_names <- paste(groups_merge, cell_names, sep='_')
seurat$barcode <- cell_names
seurat <- RenameCells(seurat, new.names=new_cell_names)

# #Combining VDJ and Seurat Object
seurat <- combineExpression(combined, seurat, proportion = TRUE, cloneCall='aa', group.by='Samples', chain = 'both')
# Renaming cells back to original barcodes
seurat <- RenameCells(seurat, new.names=seurat$barcode)
```

:::{.content-visible when-meta='verbose'}

## Data Normalization

:::

```{r}
# Normalize RNA data with SCTransform
seurat <- SCTransform(seurat, verbose = F) 
```

:::{.content-visible when-meta='processing'}

## Clustering

:::

```{r}
#| fig-width: 4
#| fig-height: 4
#| fig-cap: 'Elbow plot showing the percentage of variability represented by by each PC to select number of dimensions '

#Dimensionality reduction
seurat <- RunPCA(seurat,npcs = 100)

#Determining dimensionality of the dataset
ElbowPlot(seurat, ndims = 100)
```

:::{.content-visible when-meta='processing'}

Chosen Number of Dimensions:

:::

```{r}
# Describe number of dimensions

dimensions  <- 35 
dimensions
```

```{r}
#| results: hide
seurat <- RunUMAP(seurat, dims=1:dimensions, verbose = F, seed.use = 3514L)
resolutions <- c(0.25, 0.375, 0.5, 0.625, 0.75, 1, 1.25, 1.5, 1.75)
seurat <- FindNeighbors(seurat, dims = 1:dimensions, verbose = F)
seurat <- FindClusters(seurat, resolution = resolutions, verbose = F)
```

```{r}
#| fig-cap: UMAP plots showing the clustering results at different resolutions
#| fig-width: 10
#| fig-height: 10

p <- list()
i <- 1
for (resolution in resolutions ) {
    Idents(seurat) <- paste0('SCT_snn_res.', resolution) 
    p2 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")    
    p[[i]] <- p2
    i <- i+1
}

plot <- grid.arrange(grobs = p)
# ggsave('initial_clustering_results_by_resolution.pdf', path = here('result/figures'), plot = plot)
```

:::{.content-visible when-meta='processing'}

Chosen resolution

:::

```{r}
resolution <- 0.25
resolution
```

:::{.content-visible when-meta='verbose'}

### Save Object

:::

```{r}
saveRDS(seurat, file = here('data/aggregated_object.rds'))
```

```{r}
seurat <- readRDS(here('data/aggregated_object.rds'))
object_annotations <- 'full_object_merged'
```

### Clustering Results

```{r}
#| layout-ncol: 2
Idents(seurat) <- paste0('SCT_snn_res.', resolution)
seurat[['seurat_clusters']]<- Idents(seurat)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Groups', ncol = 2) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_group_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
# DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Samples', ncol = 4) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
# ggsave(paste0('UMAP_clusters_by_sample_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = FALSE, group.by = 'Groups') + ggtitle(paste0('R ', resolution)) + theme(legend.position = "right")
FeaturePlot_scCustom(seurat, features = c('nCount_RNA', 'nFeature_RNA'), colors_use = sequential_palette)
```

Mirrored clusters suggest batch effects, I will integrate the different lanes.

## Integration

```{r}
#| fig-cap: 'Elbow plot showing the percentage of variability represented by by each PC to select number of dimensions '

# Split RNA assay into the original Layers
#seurat[["RNA"]] <- split(seurat[["RNA"]], f = seurat$Origin) # Not doing this since the layers were not joined by PrepSCTFindMarkers

# Normalize RNA data with SCTransform, PCA, and integrate

# seurat <- SCTransform(seurat, verbose = F, assay = 'RNA')

# seurat <- RunPCA(seurat, verbose = T, assay = 'SCT', npcs=100)

seurat <- IntegrateLayers(seurat, method = HarmonyIntegration, 
                          normalization.method = 'SCT', 
                          orig.reduction = 'pca',
                          new.reduction = 'harmony',
                          verbose = T)

# #Determining dimensionality of the dataset
ElbowPlot(seurat, ndims = 100)
```


:::{.content-visible when-meta='processing'}

Chosen Number of Dimensions:

:::

```{r}
# Describe number of dimensions

dimensions  <- 40
dimensions
```

```{r}
#| results: hide
seurat <- RunUMAP(seurat, dims=1:dimensions, verbose = F, seed.use = 3514L, reduction = 'harmony')
resolutions <- c(0.25, 0.375, 0.5, 0.625, 0.75, 1, 1.25, 1.5, 1.75)
seurat <- FindNeighbors(seurat, dims = 1:dimensions, verbose = F, reduction = 'harmony')
seurat <- FindClusters(seurat, resolution = resolutions, verbose = F)
```

```{r}
#| fig-cap: UMAP plots showing the clustering results at different resolutions
#| fig-width: 10
#| fig-height: 10

p <- list()
i <- 1
for (resolution in resolutions ) {
    Idents(seurat) <- paste0('SCT_snn_res.', resolution) 
    p2 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")    
    p[[i]] <- p2
    i <- i+1
}

plot <- grid.arrange(grobs = p)
# ggsave('initial_clustering_results_by_resolution.pdf', path = here('result/figures'), plot = plot)
```

:::{.content-visible when-meta='processing'}

Chosen resolution

:::

```{r}
resolution <- 0.5
resolution
```

:::{.content-visible when-meta='verbose'}

### Save Object

:::

```{r}
saveRDS(seurat, file = here('data/E2_object.rds'))
```

## Clustering Results - Integrated

```{r}
seurat <- readRDS(here('data/E2_object.rds'))
object_annotations <- 'full_object_integrated'
resolution <- 0.5
```

```{r}
#| layout-ncol: 2
Idents(seurat) <- paste0('SCT_snn_res.', resolution)
seurat[['seurat_clusters']]<- Idents(seurat)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Groups', ncol = 2) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_group_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
# DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Samples', ncol = 4) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
# ggsave(paste0('UMAP_clusters_by_sample_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = FALSE, group.by = 'Groups') + ggtitle(paste0('R ', resolution)) + theme(legend.position = "right")
```

Integration corrected batch effects


### Top DEGs per cluster

```{r}
#| fig-width: 6
#| fig-height: 12
seurat <- PrepSCTFindMarkers(seurat)
plot1 <- top_genes_per_cluster(seurat, object_annotations, tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = F) 
print(plot1)
ggsave(paste0(figures_path, 'DotPlot_Top3_per_cluster', object_annotations, '.pdf'), width = 10, height = 16)
```

:::{.content-visible when-meta='processing'}

### Automatic cell type annotations

:::

```{r}
local_path <- paste0(figures_path, object_annotations,'_cell_type_annotations')
unlink(local_path,recursive = T)
dir.create(local_path)

# Normalize and scale data
seurat <- JoinLayers(seurat, assay = 'RNA')
seurat <- NormalizeData(seurat, assay = "RNA", normalization.method = "LogNormalize", scale.fct = 10000)
seurat <- ScaleData(seurat, assay = 'RNA')
```

:::{.content-visible when-meta='processing'}

#### SingleR annotations

:::

:::{.content-visible when-meta='processing'}

Annotations per Cluster

:::

```{r}
#| fig-cap: UMAP plot showing the clustering results with SingleR fine annotations on a per cluster basis
#| fig-width: 9
#| fig-height: 6
seurat <- annotate_seurat_with_SingleR_Eduard(seurat, local_path, database = 'ImmGen', annotation_basis = 'cluster_coarse', split_by_groups = FALSE)
```

:::{.content-visible when-meta='processing'}

Annotations per Cell

:::

```{r}
#| fig-cap: UMAP plot showing the clustering results with SingleR coarse annotations on a per cell basis
annotate_seurat_with_SingleR_Eduard(seurat, local_path, database = 'ImmGen', annotation_basis = 'cell_coarse', split_by_groups = FALSE)
```

### Plotting Genes of Interest for cluster annotation

```{r}
#| fig-cap: Feature plots showing the expression of T cell markers
#| fig-width: 8
#| fig-height: 10
FeaturePlot_scCustom(seurat, features = c('Cd3e', 'Cd4', 'Cd8a', 'Cd8b1', 'Trdc', 'Trac'), pt.size = 0.01, colors_use = sequential_palette) +
    theme(legend.position = 'none')
```

```{r}
#| fig-cap: Feature plots showing the expression of myeloid cell markers
#| fig-width: 8
#| fig-height: 10
FeaturePlot_scCustom(seurat, features = c('Sirpa', 'Xcr1', 'Itgae', 'Itgax', 'Csf1r', 'Csf2ra'), pt.size = 0.1, colors_use = sequential_palette) +
    theme(legend.position = 'none')
```
```{r}
#| fig-cap: Feature plots showing the expression of B cell an epithelial cell markers
#| #| fig-width: 8
#| fig-height: 13
FeaturePlot_scCustom(seurat, features = c('Ighm', 'Ighd', 'Igha', 'Igkc','Pax5', 'Prdm1', 'Epcam', 'Muc2'), pt.size = 0.1, colors_use = sequential_palette, num_columns = 2) 
```

### Sub clustering cluster 22

```{r}
#| results: hide 
Idents(seurat) <- 'seurat_clusters'
seurat <- FindSubCluster(seurat, graph.name = 'SCT_snn', cluster = 19, resolution = 0.3)
Idents(seurat) <- 'sub.cluster'
DimPlot(seurat, label = T)
```


### Cell Type Annotations

### Annotating clusters

```{r}
Idents(seurat) <- 'sub.cluster'
seurat <- RenameIdents(seurat
    , '0' = 'B cells'  
    , '1' = 'Epithelial cells'
    , '2' = 'Stromal cells'
    , '3' = 'Lymphocytes'
    , '4' = 'Epithelial cells'
    , '5' = 'Epithelial cells'
    , '6' = 'Lymphocytes'
    , '7' = 'Myeloid cells' 
    , '8' = 'Stromal cells'
    , '9' = 'B cells'
    , '10' = 'Stromal cells'
    , '11' = 'Lymphocytes'
    , '12' = 'Myeloid cells'
    , '13' = 'Stromal cells'
    , '14' = 'Stromal cells'
    , '15' = 'Myeloid cells'
    , '16' = 'Myeloid cells'
    , '17' = 'Stromal cells' 
    , '18' = 'Epithelial cells'
    , '19_0' = 'Lymphocytes'
    , '19_1' = 'B cells'
    , '19_2' = 'Lymphocytes'
    , '19_3' = 'Lymphocytes'
    , '19_4' = 'Lymphocytes'
    , '20' = 'Epithelial cells'
    , '21' = 'Epithelial cells'
    , '22' = 'Red blood cells'
    , '23' = 'Lymphocytes'
    , '24' = 'Stromal cells'
    , '25' = 'Endothelial cells'

)
seurat[['major_cell_types']] <- Idents(seurat)
```

```{r}
#| layout-ncol: 2
#| fig-cap: UMAP plot showing the cell type annotations
Idents(seurat) <- 'major_cell_types'
DimPlot(seurat, label=T, ncol = 2) 
ggsave(paste0('UMAP_major_cell_types_', object_annotations, '.pdf'), path = figures_path, width = 7, height = 5)

DimPlot(seurat, label=F, split.by = 'Groups', ncol = 2)
ggsave(paste0('UMAP_major_cell_types_by_group_', object_annotations, '.pdf'), path = figures_path, width = 9, height = 5)
```

### Cell frequency per condition

```{r}
extract_cell_counts(seurat, major_cell_types, figures_path, tables_path, object_annotations=object_annotations)
```

:::{.content-visible when-meta='verbose'}

## Save Object

:::

```{r}
saveRDS(seurat, file = here('data/E2_object.rds'))
```

# Preprocessing - subclustering Epithelial Cells

```{r}
seurat <- readRDS(here('data/E2_object.rds'))
object_annotations <- 'IECs'
# Subset the Seurat object
seurat <- subset(seurat, subset = major_cell_types == 'Epithelial cells')
DimPlot(seurat)
```

## Without integration

```{r}
#| fig-width: 4
#| fig-height: 4
#| fig-cap: 'Elbow plot showing the percentage of variability represented by by each PC to select number of dimensions '
# Split RNA assay into the original Layers

seurat[["RNA"]] <- split(seurat[["RNA"]], f = seurat$Groups) 

# Normalize RNA data with SCTransform, PCA, and integrate

seurat <- SCTransform(seurat, verbose = F, assay = 'RNA')

seurat <- RunPCA(seurat, verbose = T, assay = 'SCT', npcs=100)

# seurat <- IntegrateLayers(seurat, method = HarmonyIntegration, 
#                           normalization.method = 'SCT', 
#                           orig.reduction = 'pca',
#                           new.reduction = 'harmony',
#                           verbose = T)

# #Determining dimensionality of the dataset
ElbowPlot(seurat, ndims = 100)
```


:::{.content-visible when-meta='processing'}

Chosen Number of Dimensions:

:::

```{r}
# Describe number of dimensions

dimensions  <- 35
dimensions
```

```{r}
i#| results: hide
seurat <- RunUMAP(seurat, dims=1:dimensions, verbose = F, seed.use = 3514L, reduction = 'pca')
resolutions <- c(0.25, 0.375, 0.5, 0.625, 0.75, 1, 1.25, 1.5, 1.75)
seurat <- FindNeighbors(seurat, dims = 1:dimensions, verbose = F, reduction = 'pca')
seurat <- FindClusters(seurat, resolution = resolutions, verbose = F)
```

```{r}
#| fig-cap: UMAP plots showing the clustering results at different resolutions
#| fig-width: 10
#| fig-height: 10

p <- list()
i <- 1
for (resolution in resolutions ) {
    Idents(seurat) <- paste0('SCT_snn_res.', resolution) 
    p2 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")    
    p[[i]] <- p2
    i <- i+1
}

plot <- grid.arrange(grobs = p)
# ggsave('initial_clustering_results_by_resolution.pdf', path = here('result/figures'), plot = plot)
```

:::{.content-visible when-meta='processing'}

Chosen resolution

:::

```{r}
#| layout-ncol: 2
resolution <- 0.375
resolution

Idents(seurat) <- paste0('SCT_snn_res.', resolution)
seurat[['seurat_clusters']]<- Idents(seurat)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Groups', ncol = 2) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_group_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
# DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Samples', ncol = 4) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
# ggsave(paste0('UMAP_clusters_by_sample_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = FALSE, group.by = 'Groups') + ggtitle(paste0('R ', resolution)) + theme(legend.position = "right")
```

## With  integration

```{r}
seurat <- readRDS(here('data/E2_object.rds'))
object_annotations <- 'IECs'
# Subset the Seurat object
seurat <- subset(seurat, subset = major_cell_types == 'Epithelial cells')
DimPlot(seurat)
```

```{r}
#| fig-width: 4
#| fig-height: 4
#| fig-cap: 'Elbow plot showing the percentage of variability represented by by each PC to select number of dimensions '
# Split RNA assay into the original Layers

seurat[["RNA"]] <- split(seurat[["RNA"]], f = seurat$Groups) 

# Normalize RNA data with SCTransform, PCA, and integrate

seurat <- SCTransform(seurat, verbose = F, assay = 'RNA')

seurat <- RunPCA(seurat, verbose = T, assay = 'SCT', npcs=100)

seurat <- IntegrateLayers(seurat, method = HarmonyIntegration, 
                          normalization.method = 'SCT', 
                          orig.reduction = 'pca',
                          new.reduction = 'harmony',
                          verbose = T)

# #Determining dimensionality of the dataset
ElbowPlot(seurat, ndims = 100)
```


:::{.content-visible when-meta='processing'}

Chosen Number of Dimensions:

:::

```{r}
# Describe number of dimensions

dimensions  <- 35
dimensions
```

```{r}
#| results: hide
seurat <- RunUMAP(seurat, dims=1:dimensions, verbose = F, seed.use = 3514L, reduction = 'harmony')
resolutions <- c(0.25, 0.375, 0.5, 0.625, 0.75, 1, 1.25, 1.5, 1.75)
seurat <- FindNeighbors(seurat, dims = 1:dimensions, verbose = F, reduction = 'harmony')
seurat <- FindClusters(seurat, resolution = resolutions, verbose = F)
```

```{r}
#| fig-cap: UMAP plots showing the clustering results at different resolutions
#| fig-width: 10
#| fig-height: 10

p <- list()
i <- 1
for (resolution in resolutions ) {
    Idents(seurat) <- paste0('SCT_snn_res.', resolution) 
    p2 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")    
    p[[i]] <- p2
    i <- i+1
}

plot <- grid.arrange(grobs = p)
# ggsave('initial_clustering_results_by_resolution.pdf', path = here('result/figures'), plot = plot)
```

:::{.content-visible when-meta='processing'}

Chosen resolution

:::

```{r}
resolution <- 0.375
resolution
```

:::{.content-visible when-meta='verbose'}

### Save Object

:::

```{r}
saveRDS(seurat, file = here('data/E2_object_IECs.rds'))
```

# Analysis - Epithelial Cells

```{r}
seurat <- readRDS(here('data/E2_object_IECs.rds'))
object_annotations <- 'IECs'
resolution <- 0.375
```

```{r}
#| layout-ncol: 2
Idents(seurat) <- paste0('SCT_snn_res.', resolution)
seurat[['seurat_clusters']]<- Idents(seurat)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Groups', ncol = 2) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_group_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
# DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Samples', ncol = 4) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
# ggsave(paste0('UMAP_clusters_by_sample_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = FALSE, group.by = 'Groups') + ggtitle(paste0('R ', resolution)) + theme(legend.position = "right")
```

Integration was able to bring together similar cell types between the large intestine and the small intestine, but whether it got rid of some biological differences between small intestine and colon is hard to ascertain with the current experimental design. I may consider analyzing them separately

## Cluster Proportions

```{r}
extract_cell_counts(seurat, seurat_clusters, figures_path, tables_path, object_annotations=object_annotations)
```

## Top DEGs per cluster

```{r}
#| fig-width: 6
#| fig-height: 12
seurat <- PrepSCTFindMarkers(seurat)
plot1 <- top_genes_per_cluster(seurat, object_annotations, tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = F) 
print(plot1)
ggsave(paste0(figures_path, 'DotPlot_Top3_per_cluster', object_annotations, '.pdf'), width = 10, height = 16)
```

## 2017 Nature 'A single cell survey of the intestinal epithelium' signatures

```{r}
#| layout-ncol: 2
Goblet <- c('Agr2', 'Fcgbp', 'Tff3', 'Clca1', 'Zg16', 'Tpsg1', 'Muc2', 'Galnt12', 'Atoh1', 'Rep15', 'S100a6', 'Pdia5', 'Klk1', 'Pla2g10', 'Spdef', 'Lrrc26', 'Ccl9', 'Bace2', 'Bcas1', 'Slc12a8', 'Smim14', 'Tspan13', 'Txndc5', 'Creb3l4', 'C1galt1c1', 'Creb3l1', 'Qsox1', 'Guca2a', 'Scin', 'Ern2', 'AW112010', 'Fkbp11', 'Capn9', 'Stard3nl', 'Slc50a1', 'Sdf2l1', 'Hgfac', 'Galnt7', 'Hpd', 'Ttc39a', 'Tmed3', 'Pdia6', 'Uap1', 'Gcnt3', 'Tnfaip8', 'Dnajc10', 'Ergic1', 'Tsta3', 'Kdelr3', 'Foxa3', 'Tpd52', 'Tmed9', 'Spink4', 'Nans', 'Cmtm7', 'Creld2', 'Tm9sf3', 'Wars', 'Smim6', 'Manf', 'Oit1', 'Tram1', 'Kdelr2', 'Xbp1', 'Serp1', 'Vimp', 'Guk1', 'Sh3bgrl3', 'Cmpk1', 'Tmsb10', 'Dap', 'Ostc', 'Ssr4', 'Sec61b', 'Pdia3', 'Gale', 'Klf4', 'Krtcap2', 'Arf4', 'Sep15', 'Ssr2', 'Ramp1', 'Calr', 'Ddost')
 Paneth <- c('Gm15284','AY761184','Defa17','Gm14851','Defa22','Defa-rs1','Defa3','Defa24','Defa26','Defa21','Lyz1','Gm15292','Mptx2','Ang4')
 Tuft <- c('Alox5ap','Lrmp','Hck','Avil','Rgs13','Ltc4s','Trpm5','Dclk1','Spib','Fyb','Ptpn6','Matk','Snrnp25','Sh2d7','Ly6g6f','Kctd12','1810046K07Rik','Hpgds','Tuba1a','Pik3r5','Vav1','Tspan6','Skap2','Pygl','Ccdc109b','Ccdc28b','Plcg2','Ly6g6d','Alox5','Pou2f3','Gng13','Bmx','Ptpn18','Nebl','Limd2','Pea15a','Tmem176a','Smpx','Itpr2','Il13ra1','Siglecf','Ffar3','Rac2','Hmx2','Bpgm','Inpp5j','Ptgs1','Aldh2','Pik3cg','Cd24a','Ethe1','Inpp5d','Krt23','Gprc5c','Reep5','Csk','Bcl2l14','Tmem141','Coprs','Tmem176b','1110007C09Rik','Ildr1','Galk1','Zfp428','Rgs2','Inpp5b','Gnai2','Pla2g4a','Acot7','Rbm38','Gga2','Myo1b','Adh1','Bub3','Sec14l1','Asah1','Ppp3ca','Agt','Gimap1','Krt18','Pim3','2210016L21Rik','Tmem9','Lima1','Fam221a','Nt5c3','Atp2a3','Mlip','Vdac3','Ccdc23','Tmem45b','Cd47','Lect2','Pla2g16','Mocs2','Arpc5','Ndufaf3')
 Enteroendocrine <-   c('Chgb','Gfra3','Cck','Vwa5b2','Neurod1','Fev','Aplp1','Scgn','Neurog3','Resp18','Trp53i11','Bex2','Rph3al','Scg5','Pcsk1','Isl1','Maged1','Fabp5','Celf3','Pcsk1n','Fam183b','Prnp','Tac1','Gpx3','Cplx2','Nkx2-2','Olfm1','Vim','Rimbp2','Anxa6','Scg3','Ngfrap1','Insm1','Gng4','Pax6','Cnot6l','Cacna2d1','Tox3','Slc39a2','Riiad1')
Enterocyte_proximal <- c('Gsta1','Rbp2','Adh6a','Apoa4','Reg3a','Creb3l3','Cyp3a13','Cyp2d26','Ms4a10','Ace','Aldh1a1','Rdh7','H2-Q2','Hsd17b6','Gstm3','Gda','Apoc3','Gpd1','Fabp1','Slc5a1','Mme','Cox7a1','Gsta4','Lct','Khk','Mttp','Xdh','Sult1b1','Treh','Lpgat1','Dhrs1','Cyp2c66','Ephx2','Cyp2c65','Cyp3a25','Slc2a2','Ugdh','Gstm6','Retsat','Ppap2a','Acsl5','Cyb5r3','Cyb5b','Ckmt1','Aldob','Ckb','Scp2','Prap1')
Enterocyte_distal <- c('Tmigd1','Fabp6','Slc51b','Slc51a','Mep1a','Fam151a','Naaladl1','Slc34a2','Plb1','Nudt4','Dpep1','Pmp22','Xpnpep2','Muc3','Neu1','Clec2h','Phgr1','2200002D01Rik','Prss30','Cubn','Plec','Fgf15','Crip1','Krt20','Dhcr24','Myo15b','Amn','Enpep','Anpep','Slc7a9','Ocm','Anxa2','Aoc1','Ceacam20','Arf6','Abcb1a','Xpnpep1','Vnn1','Cndp2','Nostrin','Slc13a1','Aspa','Maf','Myh14')
Enteroendocrine <- c('Chgb','Chga','Gfra3','Cpe','Tac1','Fam183b','Hmgn3','Cck','Fev','Gch1','Pcsk1n','Bex2','Neurog3','Ngfrap1','Vwa5b2','Resp18','Sct','Aplp1','Scgn','Neurod1','Nkx2-2','Insm1','Vim','Rbp4','Isl1','Ddc','Mtch1','Tph1','Cldn4','Scg5','Maged1','Rprml','Cryba2','Rph3al','Celf3','Cacna1a','Trp53i11','Gpx3','Pcsk1','Fabp5','Fxyd6','Cplx2','Cdkn1c','Rundc3a','Pycr2','Myl7','Ffar2','Prnp','Rimbp2','Slc25a4','Bambi','Itm2c','Cacna2d1','Fgd2','Gadd45a','Cited2','Olfm1','Slc39a2','Ptov1','Rab3c','Tox3','Cdkn1a','Anxa6','Krt7','Btg2','Cnot6l','Riiad1','Marcksl1','Pax6','Wbp5','Scg3','Nisch','Gstz1','Bax','Gm43861','Slc18a1','Gng4')
Enterocyte_immature_distal <- c('Reg3g','Gsdmc4','Prss32','Krt8')     
Enterocyte_immature_proximal <- c('Casp6')
Enterocyte_mature_distal <- c('Tmigd1','Fabp6','Slc51b','Slc51a','Mep1a','Fam151a','Naaladl1','Slc34a2','Plb1','Nudt4','Dpep1','Pmp22','Xpnpep2','Muc3','Neu1','Clec2h','Phgr1','2200002D01Rik','Prss30','Cubn','Plec','Fgf15','Crip1','Krt20','Dhcr24','Myo15b','Amn','Enpep','Anpep','Slc7a9','Ocm','Anxa2','Aoc1','Ceacam20','Arf6','Abcb1a','Xpnpep1','Vnn1','Cndp2','Nostrin','Slc13a1','Aspa','Maf','Myh14')
Enterocyte_mature_proximal <-c('Apoa4','Fabp1','Apoc2','Rbp2','Apoc3','Leap2','Cyp2b10','Cyp3a11','Lct','Gsta1','Gstm1','Gstm3','Ephx2','Ms4a10','Fam213a','Cbr1','Adh6a','Cyb5r3','Dhrs1','Ifi27l2b','Cyb5a','Cyp3a25','Gda','Ckb','Prap1','Cgref1','Dnase1','Aldh1a1','Khk','Lpgat1','Treh','Reg3a','Acsl5','Ace','Aldob','H2-Q2','Rdh7','Ckmt1','Cyp3a13','P4hb','Mdh1','Ppap2a','Slc2a2','Cox7a1','Sec14l2','Gsta4','Mme','Retsat','Mttp','Creb3l3','Slc5a1','Sult1b1','Hsd17b6','Scp2','Cyb5b','Cyp2c65','Gpx4','Xdh','Cyp2d26','Ugdh','Gstm6','Ndufa1','Gpd1','Cyp2c66') 
Goblet <- c('Agr2','Spink4','Fcgbp','Tff3','Muc2','Zg16','Clca1','Ccl6','Klk1','Tpsg1','Ccl9','Txndc5','Smim14_ENSMUSG00000037822','Tspan13','Atoh1','Lrrc26','Ramp1','Galnt12','Mmp7','Qsox1','Fkbp11','Rep15','Tmsb10','Pla2g10','Tsta3','Pdia6','Sdf2l1','S100a6','Manf','Slc12a8','Creb3l1','Sh3bgrl3','Spdef','Tpd52','Pdia5','Cmpk1','Serp1','Tmed3','Selm','Creb3l4','Smim6','Krtcap2','Bace2','Stard3nl','Bcas1','Nans','C1galt1c1','Xbp1','Hpd','Slc50a1','Guk1','Tmed9','Ssr4','Hgfac','Ostc','Creld2','Sec61b','Gale','Kdelr2','Ssr2','Ern2','Ergic1','AW112010','Gcnt3','Guca2a','Klf4','Sep15','Galnt7','Uap1','Dnajc10','Ddost','Oit1','Foxa3','Tm9sf3','Cracr2b','Vimp','Capn9','Scin','Pdia3','Rnase1','Calr','Wars','Snhg18','Dap','Ttc39a','Dad1','Tnfaip8','Tram1','Kdelr3','Arf4','Cmtm7')
Paneth <-c('Gm15284','Gm14851','Defa21','Defa22','AY761184','Defa24','Defa17','Lyz1','Defa-rs1','Defa3','Mptx2','Ang4','Defa26','Gm15292') 
Stem <- c('Lgr5','Gkn3','Ascl2','Olfm4','Rgmb','Igfbp4','2210407C18Rik','Jun','Pdgfa','Soat1','Tnfrsf19','Cyp2e1','Fstl1','H2-Eb1','Ifitm3','Prelp','Scn2b','A930009A15Rik','H2-Ab1','Slc1a2','Cd74','Sp5','Noxa1','Rgcc','Sorbs2','Sectm1b','H2-Aa','Cdo1','Slc14a1','Clca2','Tifa','Pls3','Hmgcs2','Arid5b','Agr3','Slc12a2','Rassf5','Rnf43','Nrn1','Lamb3','Cd44','Axin2','Slc27a2','Afap1l1','Ccdc3','Lrig1','Noxo1','Cdk6','Amica1','Tgif1','Tns3','Nr2e3','Efna4','Rnf32','Prss23','2010009K17Rik','Smoc2','Mecom','Esrrg','Aqp1','Znrf3','Grb7','Phgdh','2410004N09Rik','Clca4','Aqp4','Lcp1','E030011O05Rik','Snhg1','BC064078','Car12','Zbtb38','Cdca7','Fam13a','Shisa2','Dtx4','Slc19a2','Fam115c','Mir703','Cd14','Mettl20','Myo9a','App','Clic6','Wee1','2410006H16Rik','Lancl1','1500012F01Rik','Casp12','Sh3rf1','Lrp4','Arhgef26','Etv6','1700024F13Rik','Cttnbp2','Slc16a13','Htr4','Pdxk','Immp2l','Rps15a-ps6','Rps15a-ps4','Nap1l1','Sdc4','Epn3','Sipa1l1','Wfdc15b','Zfp341','Ngef','Nrg4','Csad','Rpl34-ps1','Rin2','Cd81','Irf2bp2','Sesn3','Phlpp1','Yap1','Mfge8','Zfp825','Itga1','Pcdh8','Vdr','Kcnq1','Slc28a2','Zfp36l1','Urod','Rgs12','Nfib','Sdsl','Nfia')
TA <- c('Stmn1', 'Tubb5', 'Zfp808','Ctcf','Zfp101','Zfp652')
Tuft <- c('Lrmp','Alox5ap','Rgs13','Sh2d6','Ltc4s','Avil','Hck','Dclk1','Snrnp25','Cd24a','Trpm5','Kctd12','Aldh2','Il13ra1','Gng13','Tmem176a','Skap2','Ptpn6','Ly6g6f','Fyb','Adh1','Tmem176b','Hpgds','Reep5','Ptpn18','Spib','Bpgm','Galk1','Matk','Tuba1a','1810046K07Rik','Hmx2','Ccdc28b','Ethe1','Limd2','Sh2d7','Ccdc109b','Tspan6','Smpx','Vav1','Ly6g6d','Pik3r5','Nebl','Plcg2','Rbm38','Vdac3','Krt18','Asah1','Cd47','Krt23','Bcl2l14','Lima1','Pygl','Itpr2','Inpp5j','Pea15a','Rac2','Pou2f3','Atp2a3','Bmx','Acot7','Gnai2','Alox5','Ppp3ca','Ptgs1','Calm2','Zfp428','Tmem141','Myo1b','Siglecf','Pla2g4a','Inpp5b','Fam221a','Bub3','Arpc5','Pla2g16','1110007C09Rik','Gimap1','Coprs','Lect2','Nrgn','Agt','Ffar3','Tmem45b','Ccdc23','Rgs2','Mlip','Csk','2210016L21Rik','St6galnac2','Ildr1','Gprc5c','Mocs2','Nrep','Pik3cg','Malat1','Sec14l1','Ndufaf3','Inpp5d','Pim3','Tmem9','Gga2','Nt5c3')

signatures <- list(Enterocyte_immature_proximal, Enterocyte_immature_distal, Enterocyte_mature_proximal, Enterocyte_mature_distal, Goblet, Tuft, Paneth, Enteroendocrine, TA, Stem)

names(signatures) <- c('Enterocyte_immature_proximal', 'Enterocyte_immature_distal', 'Enterocyte_mature_proximal', 'Enterocyte_mature_distal', 'Goblet', 'Tuft', 'Paneth', 'Enteroendocrine', 'TA', 'Stem')

palette <- hcl.colors(10,palette = 'mako', rev = T)

#  signatures <- list(Enterocyte_proximal, Enterocyte_distal, Goblet, Paneth, Tuft, Enteroendocrine)

# #Calculate present expression and filter expressed in more than one percent of cells
# counts <- GetAssayData(seurat,layer = 'counts', assay = 'RNA')
# percent_expression <- rowMeans(counts > 0)*100 
# percent_expression <- percent_expression[percent_expression >= 1]


# length(percent_expression)
# Tissue_remodeling_filtered <- Tissue_remodeling[Tissue_remodeling %in% names(percent_expression)]
# Tissue_remodeling_released_factors_filtered <- Tissue_remodeling_released_factors[ Tissue_remodeling_released_factors %in% names(percent_expression)]
# Circulating_filtered <- Circulating[ Circulating %in% names(percent_expression)]
# Exhaustion_filtered <- Exhaustion[Exhaustion %in% names(percent_expression)]
# Stemness_filtered <- Stemness[Stemness %in% names(percent_expression)]
# TRM_filtered <- TRM[TRM %in% names(percent_expression)]
# TRM_neg_filtered <- TRM_neg[TRM_neg %in% names(percent_expression)]
# Exhaustion_down_filtered <- Exhaustion_down[Exhaustion_down %in% names(percent_expression)]
# Circulating_down_filtered <- Circulating_down[Circulating_down %in% names(percent_expression)]


# names(signatures) <- c('Enterocyte_proximal', 'Enterocyte_distal', 'Goblet', 'Paneth', 'Tuft', 'Enteroendocrine')
seurat <- AddModuleScore_UCell(seurat, features = signatures, name = NULL)


seurat@meta.data  <-  seurat@meta.data |>
    mutate(
        Enterocyte_immature_proximal = scale(Enterocyte_immature_proximal),
        Enterocyte_immature_distal = scale(Enterocyte_immature_distal),
        Enterocyte_mature_proximal = scale(Enterocyte_mature_proximal),
        Enterocyte_mature_distal = scale(Enterocyte_mature_distal),
        TA = scale(TA),
        Stem = scale(Stem),
        Goblet = scale(Goblet),
        Paneth = scale(Paneth),
        Tuft = scale(Tuft),
        Enteroendocrine = scale(Enteroendocrine),
    )    

for (signature in names(signatures)) {
    plot1 <- FeaturePlot_scCustom(seurat_object = seurat,  features =signature,colors_use= palette)
    ggsave(plot = plot1, filename = paste0('Signature_UCell_2017_atlas_', signature, '.pdf'), height = 5, width = 5, path = figures_path)
    print(plot1)
}
```

```{r}
#| fig-width: 7
#| fig-height: 7
#| fig-cap: Feature plots showing the expression of stem cell, goblet cell markers, and cluster specific enterocyte genes
FeaturePlot_scCustom(seurat, features = c('Clca4a', 'Clca4b', 'Lgr5', 'Muc2'), pt.size = 0.01, colors_use = sequential_palette) +
    theme(legend.position = 'none')
Idents(seurat) <- 'seurat_clusters'
DimPlot(seurat, label = T, pt.size = 0.5)
#
```

### Sub clustering cluster 5

```{r}
Idents(seurat) <- 'seurat_clusters'
seurat <- FindSubCluster(seurat, graph.name = 'SCT_snn', cluster = 6, resolution = 0.1)
DimPlot(seurat, group.by = 'sub.cluster', label = T)
```

## Annotating clusters

Here I annotate all the clusters based on the gene signatures from the 2017 Intestinal Atlas nature paper, phtop gene per cluster list. Clusters 12 and 3_2 are small immune cell contaminants that I will filter out. 

```{r}
Idents(seurat) <- 'sub.cluster'
seurat <- RenameIdents(seurat
    , '0' = 'Goblet cells 1'
    , '1' = 'Stem cells'
    , '2' = 'TA cells 1'
    , '3' = 'Enterocytes 1'
    , '4' = 'Goblet cells 2'
    , '5' = 'TA cells 2'
    , '6_0' = 'Enterocytes 1'
    , '6_1' = 'Enterocytes 5'
    , '7' = 'Enterocytes 2' 
    , '8' = 'Enterocytes 3'
    , '9' = 'Enterocytes 2'
    , '10' = 'Enteroendocrine cells 1'
    , '11' = 'Enterocytes 4'
    , '12' = 'Stem cells'
    , '13' = 'Enteroendocrine cells 2'
    , '14' = 'Goblet cells 3'
)
seurat[['cell_types']] <- Idents(seurat)
seurat@meta.data <- seurat@meta.data |> mutate(cell_types = factor(cell_types, levels = str_sort(unique(cell_types))))
if ('Other' %in% levels(seurat$cell_types)) {
    seurat@meta.data <- seurat@meta.data |> mutate(cell_types = fct_relevel(cell_types, 'Other', after = Inf))
}

Idents(seurat) <- 'sub.cluster'
seurat <- RenameIdents(seurat
    , '0' = 'Goblet cells'
    , '1' = 'Stem cells'
    , '2' = 'TA cells'
    , '3' = 'Enterocytes'
    , '4' = 'Goblet cells'
    , '5' = 'TA cells'
    , '6_0' = 'Enterocytes'
    , '6_1' = 'Enterocytes colonic'
    , '7' = 'Enterocytes SI' 
    , '8' = 'Enterocytes 2'
    , '9' = 'Enterocytes SI'
    , '10' = 'Enteroendocrine cells'
    , '11' = 'Enterocytes colonic'
    , '12' = 'Stem cells'
    , '13' = 'Enteroendocrine cells'
    , '14' = 'Goblet cells'
)
seurat[['major_cell_types']] <- Idents(seurat)
seurat@meta.data <- seurat@meta.data |> mutate(major_cell_types = factor( major_cell_types , levels = str_sort(unique(major_cell_types))))
if ('Other' %in% levels(seurat$major_cell_types)) {
    seurat@meta.data <- seurat@meta.data |> mutate(cell_types = fct_relevel(major_cell_types, 'Other', after = Inf))
}

```

```{r}
#| layout-ncol: 2
#| fig-cap: UMAP plot showing the cell type annotations
Idents(seurat) <- 'major_cell_types'
DimPlot(seurat, label=T, ncol = 2) 
ggsave(paste0('UMAP_major_cell_types_', object_annotations, '.pdf'), path = figures_path, width = 7, height = 5)
DimPlot(seurat, label=F, split.by = 'Groups', ncol = 2)
ggsave(paste0('UMAP_major_cell_types_by_group_', object_annotations, '.pdf'), path = figures_path, width = 9, height = 5)

Idents(seurat) <- 'cell_types'
DimPlot(seurat, label=T, ncol = 2) 
ggsave(paste0('UMAP_cell_types_', object_annotations, '.pdf'), path = figures_path, width = 7, height = 5)
DimPlot(seurat, label=F, split.by = 'Groups', ncol = 2)
ggsave(paste0('UMAP_cell_types_by_group_', object_annotations, '.pdf'), path = figures_path, width = 9, height = 5)


```

## Cell frequency per condition

```{r}
#| fig-width: 10
#| fig-height: 6
extract_cell_counts(seurat, grouping_var =  major_cell_types , figures_path, tables_path, object_annotations)
extract_cell_counts(seurat, grouping_var =  cell_types , figures_path, tables_path, object_annotations)
```

:::{.content-visible when-meta='verbose'}

## Save Object

:::

```{r}
saveRDS(seurat, file = here('data/E2_object_IECs.rds'))
```

## Differential Gene Expression Analysis

Given that we don't have biological replicates, I am performing a Wilcoxon test on the existing single-cell dataset, comparing cross-fostering and control conditions for each cell type. Please note that this test is usually not recommended since it treats each cell as a biological replicate, thus leading to a lot of false positives. In general, it is recommended to include biological replicates in the experiment using different hashtags for each mouse, if possible. 

```{r}
seurat <- readRDS(here('data/E2_object_IECs.rds'))

gProfiler2 <- TRUE 
```

### DEG analysis - Cross-fostered vs Control - fine cell type annotations

```{r}
#| fig-width: 7
#| fig-height: 7
#| layout-ncol: 2
#| results: hide

path <- paste0(results_path, 'DEG_Wilcox_analysis_IECs_Cross-fostered_vs_Control_cell_types/')
unlink(path,recursive = T)
dir.create(path)
Idents(seurat) <- 'cell_types'

DEG_counts <- data.frame(matrix(ncol=3, nrow=0))
colnames(DEG_counts) <- c('DEG_count', 'DEG_UP_count', 'DEG_DOWN_count')
rnames <- c()

DEG_counts <- DEG_FindMarkers_SCT_assay(seurat, comparison='Condition', group1='Control', group2='Cross-fostered', cluster= 'all_IECs' , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = F)
rnames <- c(rnames, 'all_IECs')

for (x in levels(seurat$cell_types)) {
    # cat(paste0("Cell type: ", x, "\n"))
    seurat_small <- subset(seurat, subset = cell_types == x )
    counts <- DEG_FindMarkers_SCT_assay(seurat_small, comparison='Condition', group1='Control', group2='Cross-fostered', cluster= x , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = T)
    rm(seurat_small)
    DEG_counts <- rbind(DEG_counts, counts)
    rnames <- c(rnames, x)
}
rownames(DEG_counts) <- rnames
write.csv(DEG_counts, file = file.path(tables_path, "DEG_counts.csv"))
DEG_counts
```

### DEG analysis - Cross-fostered vs Control - major cell type annotations

```{r}
#| fig-width: 7
#| fig-height: 7
#| layout-ncol: 2
#| results: hide

path <- paste0(results_path, 'DEG_Wilcox_analysis_IECs_Cross-fostered_vs_Control_major_cell_types/')
unlink(path,recursive = T)
dir.create(path)
Idents(seurat) <- 'major_cell_types'

DEG_counts <- data.frame(matrix(ncol=3, nrow=0))
colnames(DEG_counts) <- c('DEG_count', 'DEG_UP_count', 'DEG_DOWN_count')
rnames <- c()

DEG_counts <- DEG_FindMarkers_SCT_assay(seurat, comparison='Condition', group1='Control', group2='Cross-fostered', cluster= 'all_IECs' , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = F)
rnames <- c(rnames, 'all_IECs')

for (x in levels(seurat$major_cell_types)) {
    # cat(paste0("Cell type: ", x, "\n"))
    seurat_small <- subset(seurat, subset = major_cell_types == x )
    counts <- DEG_FindMarkers_SCT_assay(seurat_small, comparison='Condition', group1='Control', group2='Cross-fostered', cluster= x , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = T)
    rm(seurat_small)
    DEG_counts <- rbind(DEG_counts, counts)
    rnames <- c(rnames, x)
}
rownames(DEG_counts) <- rnames
write.csv(DEG_counts, file = file.path(tables_path, "DEG_counts.csv"))
DEG_counts
```

### DEG analysis - Goblet  cell cluster 1 versus 2 (Cluster 2 is reduced in cross-fostered mice)

```{r}
#| fig-width: 7
#| #| fig-height: 7
#| layout-ncol: 2
#| results: hide

path <- paste0(results_path, 'DEG_Wilcox_analysis_IECs_Goblet_cells_1_vs_2/')
unlink(path,recursive = T)
dir.create(path)
Idents(seurat) <- 'cell_types'

DEG_counts <- DEG_FindMarkers_SCT_assay(seurat, comparison='cell_types', group1='Goblet cells 1', group2='Goblet cells 2', cluster= 'Goblet cells' , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = F)
```

## Conclusions - Epithelial Cells

Overall, cross fostering induces an upregulation of lipid metabolic, transport, and bindings, functions in epithelial cells, as well as PPARg singling, all pointing towards increased fatty acid metabolism and, absorption. This signature is evident when looking at all epithelial cells together. Or when looking at stem cells and transit amplifying cells in particular.Furthermore, That is a reduction. On a specific goblet cell cluster. That is characterized by more mature or secretory gene programs, while a different goblet cell cluster (with perhaps more antimicrobial or induced functions) is not affected.

# Preprocessing - subclustering Stromal Cells

```{r}
seurat <- readRDS(here('data/E2_object.rds'))
object_annotations <- 'stromal_cells'
Idents(seurat) <- 'major_cell_types'
DimPlot(seurat, label = T)
# Subset the Seurat object
seurat <- subset(seurat, subset = major_cell_types %in% c('Stromal cells', 'Endothelial cells'))
DimPlot(seurat)
```

## Without integration

```{r}
#| fig-width: 4
#| fig-height: 4
#| fig-cap: 'Elbow plot showing the percentage of variability represented by by each PC to select number of dimensions '
# Split RNA assay into the original Layers

seurat[["RNA"]] <- split(seurat[["RNA"]], f = seurat$Groups) 

# Normalize RNA data with SCTransform, PCA, and integrate

seurat <- SCTransform(seurat, verbose = F, assay = 'RNA')

seurat <- RunPCA(seurat, verbose = T, assay = 'SCT', npcs=100)

# seurat <- IntegrateLayers(seurat, method = HarmonyIntegration, 
#                           normalization.method = 'SCT', 
#                           orig.reduction = 'pca',
#                           new.reduction = 'harmony',
#                           verbose = T)

# #Determining dimensionality of the dataset
ElbowPlot(seurat, ndims = 100)
```


:::{.content-visible when-meta='processing'}

Chosen Number of Dimensions:

:::

```{r}
# Describe number of dimensions

dimensions  <- 40
dimensions
```

```{r}
#| results: hide
seurat <- RunUMAP(seurat, dims=1:dimensions, verbose = F, seed.use = 3514L, reduction = 'pca')
resolutions <- c(0.25, 0.375, 0.5, 0.625, 0.75, 1, 1.25, 1.5, 1.75)
seurat <- FindNeighbors(seurat, dims = 1:dimensions, verbose = F, reduction = 'pca')
seurat <- FindClusters(seurat, resolution = resolutions, verbose = F)
```

```{r}
#| fig-cap: UMAP plots showing the clustering results at different resolutions
#| fig-width: 10
#| fig-height: 10

p <- list()
i <- 1
for (resolution in resolutions ) {
    Idents(seurat) <- paste0('SCT_snn_res.', resolution) 
    p2 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")    
    p[[i]] <- p2
    i <- i+1
}

plot <- grid.arrange(grobs = p)
# ggsave('initial_clustering_results_by_resolution.pdf', path = here('result/figures'), plot = plot)
```

:::{.content-visible when-meta='processing'}

Chosen resolution

:::

```{r}
#| layout-ncol: 2
resolution <- 0.375
resolution

Idents(seurat) <- paste0('SCT_snn_res.', resolution)
seurat[['seurat_clusters']]<- Idents(seurat)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Groups', ncol = 2) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_group_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
# DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Samples', ncol = 4) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
# ggsave(paste0('UMAP_clusters_by_sample_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = FALSE, group.by = 'Groups') + ggtitle(paste0('R ', resolution)) + theme(legend.position = "right")
colnames(seurat@meta.data)
unique(seurat$Condition)
```

Some mirroring of clusters, if more minor than in the epithelial cell compartment. I will integrate for consistency across analysis.

## With  integration

```{r}
seurat <- readRDS(here('data/E2_object.rds'))
object_annotations <- 'stromal_cells'
# DimPlot(seurat, label = T)
# Subset the Seurat object
seurat <- subset(seurat, subset = major_cell_types %in% c('Stromal cells', 'Endothelial cells'))
# DimPlot(seurat)
```

```{r}
#| fig-width: 4
#| fig-height: 4
#| fig-cap: 'Elbow plot showing the percentage of variability represented by by each PC to select number of dimensions '
# Split RNA assay into the original Layers

seurat[["RNA"]] <- split(seurat[["RNA"]], f = seurat$Groups) 

# Normalize RNA data with SCTransform, PCA, and integrate

seurat <- SCTransform(seurat, verbose = F, assay = 'RNA')

seurat <- RunPCA(seurat, verbose = T, assay = 'SCT', npcs=100)

seurat <- IntegrateLayers(seurat, method = HarmonyIntegration, 
                          normalization.method = 'SCT', 
                          orig.reduction = 'pca',
                          new.reduction = 'harmony',
                          verbose = T)

# #Determining dimensionality of the dataset
ElbowPlot(seurat, ndims = 100)
```


:::{.content-visible when-meta='processing'}

Chosen Number of Dimensions:

:::

```{r}
# Describe number of dimensions

dimensions  <- 35
dimensions
```

```{r}
#| results: hide
seurat <- RunUMAP(seurat, dims=1:dimensions, verbose = F, seed.use = 3514L, reduction = 'harmony')
resolutions <- c(0.25, 0.375, 0.5, 0.625, 0.75, 1, 1.25, 1.5, 1.75)
seurat <- FindNeighbors(seurat, dims = 1:dimensions, verbose = F, reduction = 'harmony')
seurat <- FindClusters(seurat, resolution = resolutions, verbose = F)
```

```{r}
#| fig-cap: UMAP plots showing the clustering results at different resolutions
#| fig-width: 10
#| fig-height: 10

p <- list()
i <- 1
for (resolution in resolutions ) {
    Idents(seurat) <- paste0('SCT_snn_res.', resolution) 
    p2 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")    
    p[[i]] <- p2
    i <- i+1
}

plot <- grid.arrange(grobs = p)
# ggsave('initial_clustering_results_by_resolution.pdf', path = here('result/figures'), plot = plot)
```

:::{.content-visible when-meta='processing'}

Chosen resolution

:::

```{r}
resolution <- 0.25
resolution
```

:::{.content-visible when-meta='verbose'}

### Save Object

:::

```{r}
saveRDS(seurat, file = here('data/E2_object_stromal.rds'))
```

# Analysis - Stromal Cells

```{r}
seurat <- readRDS(here('data/E2_object_stromal.rds'))
object_annotations <- 'stromal_cells'
resolution <- 0.25
```

```{r}
#| layout-ncol: 2
Idents(seurat) <- paste0('SCT_snn_res.', resolution)
seurat[['seurat_clusters']]<- Idents(seurat)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Groups', ncol = 2) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_group_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
# DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Samples', ncol = 4) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
# ggsave(paste0('UMAP_clusters_by_sample_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = FALSE, group.by = 'Groups') + ggtitle(paste0('R ', resolution)) + theme(legend.position = "right")
```

## Cluster Proportions

```{r}
extract_cell_counts(seurat, seurat_clusters, figures_path, tables_path, object_annotations=object_annotations)
```

## Top DEGs per cluster

```{r}
#| layout-ncol: 3
#| fig-width: 6
#| fig-height: 6
source(here('scripts/function_template.r'))
seurat <- PrepSCTFindMarkers(seurat)
plot1 <- top_genes_per_cluster(seurat, object_annotations, tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = T) 
```

```{r}
#| fig-width: 6
#| fig-height: 12
print(plot1)
ggsave(paste0(figures_path, 'DotPlot_Top3_per_cluster', object_annotations, '.pdf'), width = 10, height = 16)
```

:::{.content-visible when-meta='processing'}

## Automatic cell type annotations

:::

```{r}
local_path <- paste0(figures_path, object_annotations,'_cell_type_annotations')
unlink(local_path,recursive = T)
dir.create(local_path)

# Normalize and scale data
seurat <- JoinLayers(seurat, assay = 'RNA')
seurat <- NormalizeData(seurat, assay = "RNA", normalization.method = "LogNormalize", scale.fct = 10000)
seurat <- ScaleData(seurat, assay = 'RNA')
```


:::{.content-visible when-meta='processing'}

### SingleR annotations

:::

:::{.content-visible when-meta='processing'}

Annotations per Cluster

:::

```{r}
#| fig-cap: UMAP plot showing the clustering results with SingleR fine annotations on a per cluster basis
#| fig-width: 9
#| fig-height: 6
seurat <- annotate_seurat_with_SingleR_Eduard(seurat, local_path, database = 'ImmGen', annotation_basis = 'cluster_fine', split_by_groups = FALSE)
```

:::{.content-visible when-meta='processing'}

Annotations per Cell

:::

```{r}
#| fig-cap: UMAP plot showing the clustering results with SingleR coarse annotations on a per cell basis
annotate_seurat_with_SingleR_Eduard(seurat, local_path, database = 'ImmGen', annotation_basis = 'cell_coarse', split_by_groups = FALSE)
```
### Sub clustering cluster 5

```{r}
# Idents(seurat) <- 'seurat_clusters'
# seurat <- FindSubCluster(seurat, graph.name = 'SCT_snn', cluster = 6, resolution = 0.1)
# DimPlot(seurat, group.by = 'sub.cluster', label = T)
```

## Plotting Genes of Interest for cluster annotation

```{r}
#| fig-cap: Feature plots showing the expression of fibroblast markers
#| fig-width: 8
#| fig-height: 8
Idents(seurat) <- 'seurat_clusters'
fibroblast_markers <- c(
  "Col1a1", "Col1a2", "Col3a1",
  "Dcn", "Lum", "Vim",
  "Pdgfra", "Fap", "Thy1",
  "Acta2"  # myofibroblast marker
)

FeaturePlot_scCustom(seurat, features = c('Col1a1', 'Dcn', 'Pdgfra', 'Fap', 'Thy1', 'Acta2'), pt.size = 0.01, colors_use = sequential_palette) +
    theme(legend.position = 'none')

DotPlot_scCustom(seurat, features = fibroblast_markers, colors_use=sequential_palette_dotplot, x_lab_rotate = T)

```

```{r}
#| fig-cap: Feature plots showing the expression of various stromal cell markers
#| fig-width: 7
#| fig-height: 9 

fibroblast_markers <- c(
  "Col1a1", "Col1a2", "Col3a1",
  "Dcn", "Lum", "Vim",
  "Pdgfra", "Fap", "Thy1"
)
FeaturePlot_scCustom(seurat, features = fibroblast_markers, pt.size = 0.01, colors_use = sequential_palette) + plot_annotation(title = "Fibroblast markers", theme = theme(plot.title = element_text(size = 20)))

other_fibroblast_markers <- c(
  'Acta2', 'Aoc3', 'Fn1', 'Postn', 'Timp1'
)
FeaturePlot_scCustom(seurat, features = other_fibroblast_markers, pt.size = 0.01, colors_use = sequential_palette) + plot_annotation(title = "Fibroblast markers", theme = theme(plot.title = element_text(size = 20)))

# Endothelial markers
endothelial_markers <- c("Pecam1", "Cdh5", "Vwf")
FeaturePlot_scCustom(seurat, features = endothelial_markers, pt.size = 0.01, colors_use = sequential_palette) + plot_annotation(title = "Endothelial markers", theme = theme(plot.title = element_text(size = 20)))
    
# Myofibroblasts
myofibroblast_markers <- c("Acta2", "Tagln", "Myl9", "Myh11", "Tpm2", 'Nkx2-3')
FeaturePlot_scCustom(seurat, features = myofibroblast_markers, pt.size = 0.01, colors_use = sequential_palette) + plot_annotation(title = "Myofibroblast markers", theme = theme(plot.title = element_text(size = 20)))

# Epithelial markers
epithelial_markers <- c("Epcam", "Krt18", "Krt19")
FeaturePlot_scCustom(seurat, features = epithelial_markers, pt.size = 0.01, colors_use = sequential_palette) + plot_annotation(title = "Epithelial markers", theme = theme(plot.title = element_text(size = 20)))
    
# Immune markers
immune_markers <- c("Ptprc", "Cd45", "Cd3e", "Cd79a")
FeaturePlot_scCustom(seurat, features = immune_markers, pt.size = 0.01, colors_use = sequential_palette) + plot_annotation(title = "Immune markers", theme = theme(plot.title = element_text(size = 20)))

# Pericyte markers
pericyte_markers <- c("Pdgfrb", "Cspg4", "Rgs5", "Acta2")
FeaturePlot_scCustom(seurat, features = pericyte_markers, pt.size = 0.01, colors_use = sequential_palette) + plot_annotation(title = "Pericyte markers", theme = theme(plot.title = element_text(size = 20)))

# MSC (Mesenchymal stem cell) markers
msc_markers <- c("Ly6a", "Cd44", "Eng", "Alcam", "Lepr", "Nestin")
FeaturePlot_scCustom(seurat, features = msc_markers, pt.size = 0.01, colors_use = sequential_palette) + plot_annotation(title = "Mesenchymal stem cell markers", theme = theme(plot.title = element_text(size = 20)))

# Adipocyte-lineage cells
adipo_markers <- c("Pparg", "Fabp4", "Adipoq", "Cebpa", "Plin1")
FeaturePlot_scCustom(seurat, features = adipo_markers, pt.size = 0.01, colors_use = sequential_palette) + plot_annotation(title = "Adipocyte-lineage markers", theme = theme(plot.title = element_text(size = 20)))

# Chondrocytes / osteoblasts
cartilage_bone_markers <- c("Sox9", "Col2a1", "Acan", "Runx2", "Bglap", "Sp7")
FeaturePlot_scCustom(seurat, features = cartilage_bone_markers, pt.size = 0.01, colors_use = sequential_palette) + plot_annotation(title = "Chondrocyte / Osteoblast markers", theme = theme(plot.title = element_text(size = 20)))

# Smooth muscle markers
smooth_muscle_markers <- c(
  "Acta2", "Myh11", "Tagln"
)
FeaturePlot_scCustom(seurat, features = smooth_muscle_markers, pt.size = 0.01, colors_use = sequential_palette) + plot_annotation(title = "Smooth muscle markers", theme = theme(plot.title = element_text(size = 20)))

egc_markers <- c(
  "Sox10",     # Master TF for enteric glia and neural crest derivatives
  "Plp1",      # Myelin protein, expressed in EGCs
  "Gfap",      # Classic glial fibrillary acidic protein (some EGC subsets)
  "S100b",     # Calcium-binding protein, glial marker
  "Ptprz1",    # Protein tyrosine phosphatase receptor type Z1
  "Gpr17",     # GPCR, expressed in immature glia and EGCs
  "Fabp7",     # Fatty acid binding protein 7, radial glia/EGC
  "Erbb3",     # Receptor tyrosine kinase, crucial for EGC development
  "Nrcam",     # Neural cell adhesion molecule, glial marker
  "L1cam",     # Cell adhesion molecule
  "Chl1",      # Close homolog of L1CAM, neural adhesion
  "Gpr37l1",   # Glial enriched GPCR
  "Adgrg6",    # GPR126, required for glial development
  "Sh3tc2",    # Peripheral glial marker
  "Crym",      # Glial lineage marker
  "Gfra1",     # GDNF receptor, involved in ENS
  "Sox2",       # Neural progenitor/stem cell factor (in EGC progenitors)
  'Mbp',       # Myelin basic protein, expressed in SCHWANN cells 
  'Mpz'        # Myelin protein zero, SCHWANN cells
)
DotPlot_scCustom(seurat, features = egc_markers, colors_use = sequential_palette_dotplot, y_lab_rotate = F, flip_axes = T) + plot_annotation(title = "Enteric glial cell markers", theme = theme(plot.title = element_text(size = 20)))

vip_enteric_neuron_markers <- c(
  "Vip",       # Vasoactive intestinal peptide
  "Phox2a",    # Transcription factor for enteric neuron development
  "Ret",       # Receptor tyrosine kinase essential for ENS neurons
  "Nos1",      # Neuronal nitric oxide synthase (inhibitory neurons)
  "Chrna3",    # Nicotinic acetylcholine receptor subunit alpha 3
  "Chrnb4",    # Nicotinic acetylcholine receptor subunit beta 4
  "Chrna7",    # Nicotinic acetylcholine receptor subunit alpha 7
  "Snap25",    # Synaptic protein, pan-neuronal
  "Tubb3",     # Neuronal tubulin
  "Elavl3",    # Neuronal RNA-binding protein
  "Elavl4",    # Neuronal RNA-binding protein
  "Syt1",      # Synaptotagmin 1, synaptic vesicle release
  "Gal",       # Galanin neuropeptide (often co-expressed)
  "Kcnq3",     # Potassium channel subunit
  "Kcnj3",     # Potassium channel subunit
  "Cacna1b",   # Voltage-gated calcium channel alpha 1B subunit
  "Chga",      # Chromogranin A, neurosecretory vesicle protein
  "Chgb"       # Chromogranin B, neurosecretory vesicle protein
)
cholinergic_neuron_markers <- c(
  "Chat",      # Choline acetyltransferase (acetylcholine synthesis)
  "Slc18a3",   # Vesicular acetylcholine transporter
  "Vip",       # Sometimes co-expressed but less abundant here
  "Calb2",     # Calbindin 2, calcium-binding protein
  "Nos1"       # Can overlap with other inhibitory neurons
)
nitrergic_neuron_markers <- c(
  "Nos1",      # Nitric oxide synthase 1 (inhibitory motor neurons)
  "Calb1",     # Calbindin 1
  "Th",        # Tyrosine hydroxylase (sometimes)
  "Sox6"       # Transcription factor for this subtype
)

DotPlot_scCustom(seurat, features = unique(c(vip_enteric_neuron_markers, cholinergic_neuron_markers, nitrergic_neuron_markers)), colors_use = sequential_palette_dotplot, y_lab_rotate = F, flip_axes = T) + plot_annotation(title = "Enteric neuron markers", theme = theme(plot.title = element_text(size = 20)))


# Crypt-associated fibroblast markers (trophocytes, stem cell niche supporters)
crypt_fibroblast_markers <- c(
  "Pdgfra", "Cd34", "Grem1", "Foxl1", "Wnt2b", "Kitl", "Wnt5a", "Wnt5b",
  "Dlk1", "Spry1", "Spry4", "Ackr4", "Ly6h", "Adamts18", "Adamtsl3"
)

# Villus-tip fibroblast markers (ECM remodeling, epithelial differentiation)
villus_tip_fibroblast_markers <- c(
  "Adamdec1", "Postn", "Bmp2", "Bmp5", "Col18a1", "Serpine2", "Mmp16", "Fbln1"
)

# Interstitial fibroblast markers (general lamina propria fibroblasts)
interstitial_fibroblast_markers <- c(
  "Pdgfra", "Dpt", "Pi16", "Cd34", "Mgp"
)

# Inflammatory fibroblast markers (activated in inflammation)
inflammatory_fibroblast_markers <- c(
  "Il6", "Cxcl1", "Ccl2", "Ccl7"
)

all_fibroblast_markers <- rev(unique(c(
  crypt_fibroblast_markers,
  villus_tip_fibroblast_markers,
  interstitial_fibroblast_markers,
  inflammatory_fibroblast_markers
)))
DotPlot_scCustom(seurat, features = all_fibroblast_markers, colors_use = sequential_palette_dotplot, y_lab_rotate = F, flip_axes = T) + plot_annotation(title = "Fibroblast subset markers", theme = theme(plot.title = element_text(size = 20)))

developmental_fibroblast_markers <- c(
  # ECM & adhesion
  "Adam12", "Adamts12", "Thsd4", "Chsy3", "Fndc3b", "Slit3", "Mrc2", "Egfr",
  
  # Transcription factors
  "Tshz2", "Pbx1", "Meis1", "Nfia", "Zfpm2", "Gli2", "Tcf7l1", 
  "Ebf1", "Ebf2", "Ebf3", "Aff3", "Ncoa2", "Maml2", "Hivep2",
  
  # Signaling & guidance
  "Ror1", "Ntrk3", "Efna5", "Sema5a", "Adgrd1", "Adgrb3", "Pcdh9", "Rora",

  # Stromal/mesenchymal function
  "Dock1", "Pdzrn3", "Celf2", "Bnc2", "Dlc1", "Camk1d", "Bicc1"
)
DotPlot_scCustom(seurat, features = developmental_fibroblast_markers, colors_use = sequential_palette_dotplot, y_lab_rotate = F, flip_axes = T) + plot_annotation(title = "Developmentally patterned (EMT?) fibroblast markers", theme = theme(plot.title = element_text(size = 20)))
```

## Annotating clusters


```{r}
Idents(seurat) <- 'seurat_clusters'
DimPlot(seurat, label = T)
seurat <- RenameIdents(seurat
    , '0' = 'Fibroblasts'
    , '1' = 'Fibroblasts'
    , '2' = 'Perycites'
    , '3' = 'Enteric Glia'
    , '4' = 'Miofibroblasts'
    , '5' = 'Smooth muscle cells'
    , '6' = 'Smooth muscle cells'
    , '7' = 'Mitotic Fibroblasts' 
    , '8' = 'Fibroblasts'
    , '9' = 'Enteric neurons'
    , '10' = 'Blood endothelial cells'
    , '11' = 'Fibroblasts'
    , '12' = 'Enteric Glia'
    , '13' = 'Schwann cellsb'
    , '14' = 'Lymphatic endothelial cells'
    , '15' = 'Mesopthelial-like cells'
)
seurat[['cell_types']] <- Idents(seurat)
seurat@meta.data <- seurat@meta.data |> mutate(cell_types = factor(cell_types, levels = str_sort(unique(cell_types))))
if ('Other' %in% levels(seurat$cell_types)) {
    seurat@meta.data <- seurat@meta.data |> mutate(cell_types = fct_relevel(cell_types, 'Other', after = Inf))
}

```

```{r}
#| layout-ncol: 2
#| fig-cap: UMAP plot showing the cell type annotations

Idents(seurat) <- 'cell_types'
DimPlot(seurat, label=F, ncol = 2) 
ggsave(paste0('UMAP_cell_types_', object_annotations, '.pdf'), path = figures_path, width = 7, height = 5)
DimPlot(seurat, label=F, split.by = 'Groups', ncol = 2)
ggsave(paste0('UMAP_cell_types_by_group_', object_annotations, '.pdf'), path = figures_path, width = 9, height = 5)


```

## Cell frequency per condition

```{r}
#| fig-width: 10
#| fig-height: 6
extract_cell_counts(seurat, grouping_var =  major_cell_types , figures_path, tables_path, object_annotations)
extract_cell_counts(seurat, grouping_var =  cell_types , figures_path, tables_path, object_annotations)
```

:::{.content-visible when-meta='verbose'}

## Save Object

:::

```{r}
saveRDS(seurat, file = here('data/E2_object_stromal.rds'))
```

## Differential Gene Expression Analysis

Given that we don't have biological replicates, I am performing a Wilcoxon test on the existing single-cell dataset, comparing cross-fostering and control conditions for each cell type. Please note that this test is usually not recommended since it treats each cell as a biological replicate, thus leading to a lot of false positives. In general, it is recommended to include biological replicates in the experiment using different hashtags for each mouse, if possible. 

```{r}
seurat <- readRDS(here('data/E2_object_stromal.rds'))

gProfiler2 <- TRUE 
```

### DEG analysis - Cross-fostered vs Control - fine cell type annotations

```{r}
#| fig-width: 7
#| fig-height: 7
#| layout-ncol: 2
#| results: hide

path <- paste0(results_path, 'DEG_Wilcox_analysis_stromal_Cross-fostered_vs_Control_cell_types/')
unlink(path,recursive = T)
dir.create(path)
Idents(seurat) <- 'cell_types'

DEG_counts <- data.frame(matrix(ncol=3, nrow=0))
colnames(DEG_counts) <- c('DEG_count', 'DEG_UP_count', 'DEG_DOWN_count')
rnames <- c()

DEG_counts <- DEG_FindMarkers_SCT_assay(seurat, comparison='Condition', group1='Control', group2='Cross-fostered', cluster= 'all_stromal_cells' , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = F)
rnames <- c(rnames, 'all_stromal_cells')

for (x in levels(seurat$cell_types)) {
    # cat(paste0("Cell type: ", x, "\n"))
    seurat_small <- subset(seurat, subset = cell_types == x )
    counts <- DEG_FindMarkers_SCT_assay(seurat_small, comparison='Condition', group1='Control', group2='Cross-fostered', cluster= x , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = T)
    rm(seurat_small)
    DEG_counts <- rbind(DEG_counts, counts)
    rnames <- c(rnames, x)
}
rownames(DEG_counts) <- rnames
write.csv(DEG_counts, file = file.path(tables_path, "DEG_counts.csv"))
DEG_counts
```


## Conclusions - Stromal cells

Overall, there are very few changes in stromal cells upon cross fostering. Perhaps there is a small signature of lipid metabolism or transport, although that could be contaminating epithelial ambient RNA or leftover epithelial cells/doublets, as the UMI counts are low. Furthermore, the upregulation of fibroblast-related genes in enteric glia could be a cluster contamination as well. 

# Preprocessing - subclustering lymphoid cells

```{r}
seurat <- readRDS(here('data/E2_object.rds'))
object_annotations <- 'lymphoid_cells'
Idents(seurat) <- 'major_cell_types'
DimPlot(seurat, label = T)
# Subset the Seurat object
seurat <- subset(seurat, subset = major_cell_types %in% c('Lymphocytes'))
DimPlot(seurat)
```

## Without integration

```{r}
#| fig-width: 4
#| fig-height: 4
#| fig-cap: 'Elbow plot showing the percentage of variability represented by by each PC to select number of dimensions '
# Split RNA assay into the original Layers

seurat[["RNA"]] <- split(seurat[["RNA"]], f = seurat$Groups) 

# Normalize RNA data with SCTransform, PCA, and integrate

seurat <- SCTransform(seurat, verbose = F, assay = 'RNA')

seurat <- RunPCA(seurat, verbose = T, assay = 'SCT', npcs=100)

# seurat <- IntegrateLayers(seurat, method = HarmonyIntegration, 
#                           normalization.method = 'SCT', 
#                           orig.reduction = 'pca',
#                           new.reduction = 'harmony',
#                           verbose = T)

# #Determining dimensionality of the dataset
ElbowPlot(seurat, ndims = 100)
```


:::{.content-visible when-meta='processing'}

Chosen Number of Dimensions:

:::

```{r}
# Describe number of dimensions

dimensions  <- 40
dimensions
```

```{r}
#| results: hide
seurat <- RunUMAP(seurat, dims=1:dimensions, verbose = F, seed.use = 3514L, reduction = 'pca')
resolutions <- c(0.25, 0.375, 0.5, 0.625, 0.75, 1, 1.25, 1.5, 1.75)
seurat <- FindNeighbors(seurat, dims = 1:dimensions, verbose = F, reduction = 'pca')
seurat <- FindClusters(seurat, resolution = resolutions, verbose = F)
```

```{r}
#| fig-cap: UMAP plots showing the clustering results at different resolutions
#| fig-width: 10
#| fig-height: 10

p <- list()
i <- 1
for (resolution in resolutions ) {
    Idents(seurat) <- paste0('SCT_snn_res.', resolution) 
    p2 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")    
    p[[i]] <- p2
    i <- i+1
}

plot <- grid.arrange(grobs = p)
# ggsave('initial_clustering_results_by_resolution.pdf', path = here('result/figuresto
# '), plot = plot)
```

:::{.content-visible when-meta='processing'}

Chosen resolution

:::

```{r}
#| layout-ncol: 2
resolution <- 0.625
resolution

Idents(seurat) <- paste0('SCT_snn_res.', resolution)
seurat[['seurat_clusters']]<- Idents(seurat)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Groups', ncol = 2) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_group_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
# DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Samples', ncol = 4) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
# ggsave(paste0('UMAP_clusters_by_sample_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = FALSE, group.by = 'Groups') + ggtitle(paste0('R ', resolution)) + theme(legend.position = "right")
colnames(seurat@meta.data)
unique(seurat$Condition)
```

I will integrate for consistency across analysis.

## With  integration

```{r}
seurat <- readRDS(here('data/E2_object.rds'))
object_annotations <- 'lymphoid_cells'
Idents(seurat) <- 'major_cell_types'
# DimPlot(seurat, label = T)
# Subset the Seurat object
seurat <- subset(seurat, subset = major_cell_types %in% c('Lymphocytes'))
# DimPlot(seurat)
```

```{r}
#| fig-width: 4
#| fig-height: 4
#| fig-cap: 'Elbow plot showing the percentage of variability represented by by each PC to select number of dimensions '
# Split RNA assay into the original Layers

seurat[["RNA"]] <- split(seurat[["RNA"]], f = seurat$Groups) 

# Normalize RNA data with SCTransform, PCA, and integrate

seurat <- SCTransform(seurat, verbose = F, assay = 'RNA')

seurat <- RunPCA(seurat, verbose = T, assay = 'SCT', npcs=100)

seurat <- IntegrateLayers(seurat, method = HarmonyIntegration, 
                          normalization.method = 'SCT', 
                          orig.reduction = 'pca',
                          new.reduction = 'harmony',
                          verbose = T)

# #Determining dimensionality of the dataset
ElbowPlot(seurat, ndims = 100)
```


:::{.content-visible when-meta='processing'}

Chosen Number of Dimensions:

:::

```{r}
# Describe number of dimensions

dimensions  <- 35
dimensions
```

```{r}
#| results: hide
seurat <- RunUMAP(seurat, dims=1:dimensions, verbose = F, seed.use = 3514L, reduction = 'harmony')
resolutions <- c(0.25, 0.375, 0.5, 0.625, 0.75, 1, 1.25, 1.5, 1.75)
seurat <- FindNeighbors(seurat, dims = 1:dimensions, verbose = F, reduction = 'harmony')
seurat <- FindClusters(seurat, resolution = resolutions, verbose = F)
```

```{r}
#| fig-cap: UMAP plots showing the clustering results at different resolutions
#| fig-width: 10
#| fig-height: 10

p <- list()
i <- 1
for (resolution in resolutions ) {
    Idents(seurat) <- paste0('SCT_snn_res.', resolution) 
    p2 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")    
    p[[i]] <- p2
    i <- i+1
}

plot <- grid.arrange(grobs = p)
# ggsave('initial_clustering_results_by_resolution.pdf', path = here('result/figures'), plot = plot)
```

:::{.content-visible when-meta='processing'}

Chosen resolution

:::

```{r}
resolution <- 0.5
resolution
```

:::{.content-visible when-meta='verbose'}

### Save Object

:::

```{r}
saveRDS(seurat, file = here('data/E2_object_lymphocytes.rds'))
```

# Analysis - Lymphocytes

```{r}
seurat <- readRDS(here('data/E2_object_lymphocytes.rds'))
object_annotations <- 'lymphoid_cells'
resolution <- 0.5
```

```{r}
#| layout-ncol: 2
Idents(seurat) <- paste0('SCT_snn_res.', resolution)
seurat[['seurat_clusters']]<- Idents(seurat)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Groups', ncol = 2) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_group_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
# DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Samples', ncol = 4) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
# ggsave(paste0('UMAP_clusters_by_sample_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = FALSE, group.by = 'Groups') + ggtitle(paste0('R ', resolution)) + theme(legend.position = "right")
```

## Cluster Proportions

```{r}
extract_cell_counts(seurat, seurat_clusters, figures_path, tables_path, object_annotations=object_annotations)
```

## Top DEGs per cluster

```{r}
#| layout-ncol: 3
#| fig-width: 6
#| fig-height: 6
source(here('scripts/function_template.r'))
seurat <- PrepSCTFindMarkers(seurat)
plot1 <- top_genes_per_cluster(seurat, object_annotations = object_annotations, tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = T) 
```

```{r}
#| fig-width: 8
#| fig-height: 12
print(plot1)
ggsave(paste0(figures_path, 'DotPlot_Top3_per_cluster', object_annotations, '.pdf'), width = 10, height = 16)
```

:::{.content-visible when-meta='processing'}

## Automatic cell type annotations

:::

```{r}
local_path <- paste0(figures_path, object_annotations,'_cell_type_annotations')
unlink(local_path,recursive = T)
dir.create(local_path)

# Normalize and scale data
seurat <- JoinLayers(seurat, assay = 'RNA')
seurat <- NormalizeData(seurat, assay = "RNA", normalization.method = "LogNormalize", scale.fct = 10000)
seurat <- ScaleData(seurat, assay = 'RNA')
```


:::{.content-visible when-meta='processing'}

### SingleR annotations

:::

:::{.content-visible when-meta='processing'}

Annotations per Cluster

:::

```{r}
#| fig-cap: UMAP plot showing the clustering results with SingleR fine annotations on a per cluster basis
#| fig-width: 9
#| fig-height: 6
seurat <- annotate_seurat_with_SingleR_Eduard(seurat, local_path, database = 'ImmGen', annotation_basis = 'cluster_fine', split_by_groups = FALSE)
```

:::{.content-visible when-meta='processing'}

Annotations per Cell

:::

```{r}
#| fig-cap: UMAP plot showing the clustering results with SingleR coarse annotations on a per cell basis
annotate_seurat_with_SingleR_Eduard(seurat, local_path, database = 'ImmGen', annotation_basis = 'cell_coarse', split_by_groups = FALSE)
```

## Plotting Genes of Interest for cluster annotation

```{r}
#| fig-cap: Feature plots showing the expression of T cell markers
#| fig-width: 8
#| fig-height: 8
Idents(seurat) <- 'seurat_clusters'
T_cell_markers <- c('Trac', 'Cd3e','Cd4', 'Cd8b1', 'Cd8a', 'Trdc', 'Nkg7', 'Gzma', 'Gzmb', 'Gzmk', 'Prf1', 'Klrb1c', 'Klrc1', 'Klrd1', 'Klrk1', 'Lef1', 'Tcf7', 'Ccr7', 'Sell', 'Cd44', 'Cd69', 'Il7r', 'Cxcr6', 'Cxcr3', 'Ccr5', 'Pdcd1', 'Lag3', 'Tigit', 'Havcr2', 'Ctla4', 'Zbtb16', 'Eomes','Foxp3', 'Il2ra', 'Il2rb', 'Il10', 'Il17a', 'Rorc', 'Gata3', 'Il4', 'Il5', 'Il13', 'Ifng', 'Itga1', 'Itgae', 'H2-Aa', 'H2-Ab1')

FeaturePlot_scCustom(seurat, features = c('Trac', 'Cd3e','Cd4', 'Cd8b1', 'Cd8a', 'Trdc'), pt.size = 0.01, colors_use = sequential_palette) 
     
FeaturePlot_scCustom(seurat, features = c('Rorc', 'H2-Aa' ), pt.size = 0.01, colors_use = sequential_palette) 

gene_list_plot <- c('H2-Ab1', 'Rorc', 'Aire', 'Cd4', 'Il4', 'Itgb8', 'Cxcr6', 'Il7r', 'Cd274', 'Il2ra', 'Cd80', 'Cd86', 'Cd40', 'Ccr6')
DotPlot_scCustom(seurat, features = gene_list_plot |> rev(), colors_use=sequential_palette_dotplot, x_lab_rotate = T, flip_axes=T)

DotPlot_scCustom(seurat, features = T_cell_markers |> rev(), colors_use=sequential_palette_dotplot, x_lab_rotate = T, flip_axes=T)

FeaturePlot_scCustom(seurat, features = c('Rorc', 'H2-Aa', 'H2-Ab1', 'Aire' ), pt.size = 0.01, colors_use = sequential_palette)

```


## Subclustering cluster 3 (ILC3 MHCII+)

```{r}
#| results: hide 
Idents(seurat) <- 'seurat_clusters'
seurat <- FindSubCluster(seurat, graph.name = 'SCT_snn', cluster = 3, resolution = 0.3)
Idents(seurat) <- 'sub.cluster'
DimPlot(seurat, label = T)
```

```{r}
Idents(seurat) <- 'sub.cluster'
seurat_local <- subset(seurat, subset = sub.cluster %in% c('3_0', '3_1', '3_2', '7'))
gene_list_plot <- c('H2-Aa', 'H2-Ab1', 'Rorc', 'Aire', 'Cd4', 'Il4', 'Itgb8', 'Cxcr6', 'Il7r', 'Cd274', 'Il2ra', 'Cd80', 'Cd86', 'Cd40', 'Ccr6', 'Ncr1', 'Ncr', 'Itga4', 'Itgb7')
DotPlot_scCustom(seurat_local, features = gene_list_plot |> rev(), colors_use=sequential_palette_dotplot, x_lab_rotate = T, flip_axes=T)
rm(seurat_local)
```

Cluster 3_1 very clearly represents ILC3 MHCII+ cells. Cluster 7 represents CD4- ILC3s. Cluster 3_0 and 3_2 represent CD4+ ILC3s.  

## Annotating clusters


```{r}
Idents(seurat) <- 'sub.cluster'
seurat <- RenameIdents(seurat
    , '0' = 'naive CD4 T cells'
    , '1' = 'naive CD8 T cells'
    , '2' = 'naive CD4 T cells'
    , '3_0' = 'ILC3 CD4+'
    , '3_1' = 'ILC3 MHCII+'
    , '3_2' = 'ILC3 CD4+'    
    , '4' = 'ILC1'
    , '5' = 'proliferating T cells'
    , '6' = 'ILC2'
    , '7' = 'ILC3 CD4-' 
    , '8' = 'naive CD4 T cells'
    , '9' = 'gd T cells - IEL'
    , '10' = 'gd T cells - Type17'
    , '11' = 'DN IEL'
    , '12' = 'Tregs'
    , '13' = 'Mast cells'
    , '14' = 'naive CD4 T cells - ISGs'
    , '15' = 'Basophils'
)
seurat[['cell_types']] <- Idents(seurat)
seurat@meta.data <- seurat@meta.data |> mutate(cell_types = factor(cell_types, levels = str_sort(unique(cell_types))))
seurat@meta.data <- seurat@meta.data |> mutate(cell_types = fct_relevel(cell_types, 'DN IEL', after = Inf))
seurat@meta.data <- seurat@meta.data |> mutate(cell_types = fct_relevel(cell_types, 'Basophils', after = Inf))
seurat@meta.data <- seurat@meta.data |> mutate(cell_types = fct_relevel(cell_types, 'Mast cells', after = Inf))
if ('Other' %in% levels(seurat$cell_types)) {
    seurat@meta.data <- seurat@meta.data |> mutate(cell_types = fct_relevel(cell_types, 'Other', after = Inf))
}

Idents(seurat) <- 'sub.cluster'
DimPlot(seurat, label = T)
seurat <- RenameIdents(seurat
    , '0' = 'naive CD4 T cells'
    , '1' = 'naive CD8 T cells'
    , '2' = 'naive CD4 T cells'
    , '3_0' = 'ILC3'
    , '3_1' = 'ILC3'
    , '3_2' = 'ILC3'    
    , '4' = 'ILC1'
    , '5' = 'proliferating T cells'
    , '6' = 'ILC2'
    , '7' = 'ILC3' 
    , '8' = 'naive CD4 T cells'
    , '9' = 'gd T cells'
    , '10' = 'gd T cells'
    , '11' = 'DN IEL'
    , '12' = 'Tregs'
    , '13' = 'Mast cells'
    , '14' = 'naive CD4 T cells'
    , '15' = 'Basophils'
)
seurat[['major_cell_types']] <- Idents(seurat)
seurat@meta.data <- seurat@meta.data |> mutate(major_cell_types = factor(major_cell_types, levels = str_sort(unique(major_cell_types))))
seurat@meta.data <- seurat@meta.data |> mutate(major_cell_types = fct_relevel(major_cell_types, 'DN IEL', after = Inf))
seurat@meta.data <- seurat@meta.data |> mutate(major_cell_types = fct_relevel(major_cell_types, 'Basophils', after = Inf))
seurat@meta.data <- seurat@meta.data |> mutate(major_cell_types = fct_relevel(major_cell_types, 'Mast cells', after = Inf))
if ('Other' %in% levels(seurat$major_cell_types)) {
    seurat@meta.data <- seurat@meta.data |> mutate(major_cell_types = fct_relevel(major_cell_types, 'Other', after = Inf))
}

```

```{r}
#| layout-ncol: 2
#| fig-cap: UMAP plot showing the cell type annotations

#| layout-ncol: 2
#| fig-cap: UMAP plot showing the cell type annotations
Idents(seurat) <- 'major_cell_types'
DimPlot(seurat, label=F, ncol = 2) 
ggsave(paste0('UMAP_major_cell_types_', object_annotations, '.pdf'), path = figures_path, width = 7, height = 5)
DimPlot(seurat, label=F, split.by = 'Groups', ncol = 2)
ggsave(paste0('UMAP_major_cell_types_by_group_', object_annotations, '.pdf'), path = figures_path, width = 9, height = 5)

Idents(seurat) <- 'cell_types'
DimPlot(seurat, label=F, ncol = 2) 
ggsave(paste0('UMAP_cell_types_', object_annotations, '.pdf'), path = figures_path, width = 7, height = 5)
DimPlot(seurat, label=F, split.by = 'Groups', ncol = 2)
ggsave(paste0('UMAP_cell_types_by_group_', object_annotations, '.pdf'), path = figures_path, width = 9, height = 5)


```

## Cell frequency per condition

```{r}
#| fig-width: 10
#| fig-height: 6
extract_cell_counts(seurat, grouping_var =  major_cell_types , figures_path, tables_path, object_annotations)
extract_cell_counts(seurat, grouping_var =  cell_types , figures_path, tables_path, object_annotations)
```

:::{.content-visible when-meta='verbose'}

## Save Object

:::

```{r}
saveRDS(seurat, file = here('data/E2_object_lymphocytes.rds'))
```

## Differential Gene Expression Analysis

Given that we don't have biological replicates, I am performing a Wilcoxon test on the existing single-cell dataset, comparing cross-fostering and control conditions for each cell type. Please note that this test is usually not recommended since it treats each cell as a biological replicate, thus leading to a lot of false positives. In general, it is recommended to include biological replicates in the experiment using different hashtags for each mouse, if possible. 

```{r}
seurat <- readRDS(here('data/E2_object_lymphocytes.rds'))

gProfiler2 <- TRUE 
```

### DEG analysis - Cross-fostered vs Control - fine cell type annotations

```{r}
#| fig-width: 7
#| fig-height: 7
#| layout-ncol: 2
#| results: hide

path <- paste0(results_path, 'DEG_Wilcox_analysis_Lymphocytes_Cross-fostered_vs_Control_cell_types/')
unlink(path,recursive = T)
dir.create(path)
Idents(seurat) <- 'cell_types'

DEG_counts <- data.frame(matrix(ncol=3, nrow=0))
colnames(DEG_counts) <- c('DEG_count', 'DEG_UP_count', 'DEG_DOWN_count')
rnames <- c()

DEG_counts <- DEG_FindMarkers_SCT_assay(seurat, comparison='Condition', group1='Control', group2='Cross-fostered', cluster= 'all_lymphocytes' , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = F)
rnames <- c(rnames, 'all_lymphocytes')

for (x in levels(seurat$cell_types)) {
    # cat(paste0("Cell type: ", x, "\n"))
    seurat_small <- subset(seurat, subset = cell_types == x )
    counts <- DEG_FindMarkers_SCT_assay(seurat_small, comparison='Condition', group1='Control', group2='Cross-fostered', cluster= x , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = T)
    rm(seurat_small)
    DEG_counts <- rbind(DEG_counts, counts)
    rnames <- c(rnames, x)
}
rownames(DEG_counts) <- rnames
write.csv(DEG_counts, file = file.path(tables_path, "DEG_counts.csv"))
DEG_counts
```


### DEG analysis - Cross-fostered vs Control - major cell type annotations

```{r}
#| fig-width: 7
#| fig-height: 7
#| layout-ncol: 2
#| results: hide

path <- paste0(results_path, 'DEG_Wilcox_analysis_Lymphocytes_Cross-fostered_vs_Control_major_cell_types/')
unlink(path,recursive = T)
dir.create(path)
Idents(seurat) <- 'major_cell_types'

DEG_counts <- data.frame(matrix(ncol=3, nrow=0))
colnames(DEG_counts) <- c('DEG_count', 'DEG_UP_count', 'DEG_DOWN_count')
rnames <- c()

DEG_counts <- DEG_FindMarkers_SCT_assay(seurat, comparison='Condition', group1='Control', group2='Cross-fostered', cluster= 'all_lymphocytes' , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = F)
rnames <- c(rnames, 'all_lymphocytes')

for (x in levels(seurat$major_cell_types)) {
    # cat(paste0("Cell type: ", x, "\n"))
    seurat_small <- subset(seurat, subset = major_cell_types == x )
    counts <- DEG_FindMarkers_SCT_assay(seurat_small, comparison='Condition', group1='Control', group2='Cross-fostered', cluster= x , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2, is_integrated_subset = T)
    rm(seurat_small)
    DEG_counts <- rbind(DEG_counts, counts)
    rnames <- c(rnames, x)
}
rownames(DEG_counts) <- rnames
write.csv(DEG_counts, file = file.path(tables_path, "DEG_counts.csv"))
DEG_counts
```

## Conclusions - lymphocytes 

 


