---
title: Analysis of Abau and Ftul infected mice - lungs
author: Eduard Ansaldo
date: '`r Sys.Date()`'
verbose: false
engine: knitr
knitr:
  opts_chunk: 
    echo: false
    warning: false
    message: false
    error: false
    cache: true
    results: markdown
    code-fold: true
    cache-lazy: false
    
processing: true

execute: 
  cache: true


format:
  html:
    embed-resources: true
    fig-cap-location: top
    toc: true
    toc-depth: 3
    toc-expand: 1
    toc-location: left
    toc-title: Table of Contents
    toc-collapse: true
    number-sections: true
    page-layout: full
    fig-width: 6
    fig-height: 6
    fig-align: 'center'
---

::: {.content-visible when-meta="verbose"}

# Initialization

:::

```{r}
#| cache: false
# General R and plotting 
library(conflicted)
library(tidyverse)
library(ggplot2)
library(scales)
library(patchwork)
library(cowplot)
library(gridExtra)
library(ggrepel)
library(stringr)
library(VennDiagram)
library(pheatmap)
library(viridis)
library(here)
library(readxl)

# Single Cell Analysis Packages
library(Seurat)
library(scRepertoire)
library(circlize)
library(scCustomize)
library(SingleR)
library(celldex)
library(UCell)
library(presto)
library(scDblFinder)
library(immunarch)

# DEG, pathway enrichment and visualization packages
library(DESeq2)
library(gprofiler2)
library(clusterProfiler)
library(DOSE)
library(pathview)
library(org.Mm.eg.db)
library(scRepertoire)
library(enrichplot)
library(msigdbr)
library(rlang)

# Python Version 3.9.6!!! 

# peconflicts_scout()
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(base::setdiff)
conflicts_prefer(dplyr::rename)

```

```{r}
#| results: hide 
# Setting working directory and seed
# renv::update()
# renv::snapshot()
# renv::init()
# renv::status()
set.seed(3514)
i_am('scripts/Analysis_comparing_integration.qmd')
here()
results_path <- here('results/')
dir.create(results_path)
results_path <- here('results/')
dir.create(results_path)
figures_path <- here('results/figures/')
dir.create(figures_path)
# data_path <- here('data/')
# dir.create(data_path)
tables_path <- here('results/tables/')
dir.create(tables_path)

# Loading custom functions
source(here('scripts/function_template.r'))
source(here('scripts/gProfiler2_functions.r'))
source(here('scripts/circlize_functions.r'))

# Setting up color palettes
diverging_palette <- hcl.colors(n = 20,'Purple-Green',rev = T)
sequential_palette_dotplot <- hcl.colors(n = 20,'YlGn',rev = T)
sequential_palette <- hcl.colors(n = 20,'Mako',rev = T)

knitr::dep_prev()
```

# Initial notes
Data was preprocessed in cellranger 9.0.1 Using the intron=true option, and aggregating all three samples to normalize read depth.

:::{.content-visible when-meta='processing'}

# Preprocessing aggregated dataset

:::

```{r}

#| results: hide
#| #Load the dataset from the cellranger outs
#| results: hide
#Load the dataset from the cellranger outs
scdata <- Read10X(data.dir = here("data/cluster_processing_2/aggr_all/outs/count/filtered_feature_bc_matrix"))

#Initialize the seurat object with the raw (non-normalized data)
seurat <- CreateSeuratObject(counts=scdata$'Gene Expression', min.cells = 3)


# #Add HTO data as a new assay independent from RNA
HTO <- CreateAssayObject(counts = scdata$'Antibody Capture')
seurat[["HTO"]] <- HTO

```

:::{.content-visible when-meta='processing'}

## Splitting dataset

:::

```{r}

seurat$barcode <- Cells(seurat)
seurat@meta.data <- seurat@meta.data |>
    mutate(GEM = str_extract(barcode, '\\d')) |>
    mutate(GEM = case_when(
            GEM == '1' ~ 'CT_MAIT',
            GEM == '2' ~ 'Ftula_MAIT',
            GEM == '3' ~ 'Abau_MAIT'
    )) 
# Splitting the dataset into the three groups
seurat_list <- SplitObject(seurat, split.by = 'GEM')

```

:::{.content-visible when-meta='processing'}

# Preprocessing Abau MAIT dataset

:::

:::{.content-visible when-meta='verbose'}

## Read in data

:::

```{r}
#| results: hide

seurat <- seurat_list$Abau_MAIT

```

:::{.content-visible when-meta='processing'}
#### Confirming used HTOs - MAIT
:::

```{r}
print('Antibody Capture total reads:')
seurat@assays$HTO$counts |> as.matrix() |>
        rowSums()  |> as.data.frame() 

HTO_data <- seurat@assays$HTO$counts

# Extracting HTO and biotin data
HTO_data <-  HTO_data[-c(3), ]
# colnames(scdata$'HTO') <- colnames(scdata$'Antibody Capture')
# rownames(scdata$'HTO') <- rownames(scdata$'Antibody Capture')[-c(2, 3, 4, 5, 6, 7, 11, 17, 18, 19)]

# #Add HTO data as a new assay independent from RNA
HTO <- CreateAssayObject(counts = HTO_data)
seurat[["HTO"]] <- HTO
```

Hashtags Used: 1, 2, 4, 5

:::{.content-visible when-meta='processing'}

## QC

:::

```{r}
#| layout-ncol: 2
#Quantifying percentage mitochondria
Idents(seurat) <- 'Tissue'
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, assay="RNA",pattern = "mt-")

VlnPlot(seurat, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol=3, pt.size = 0.01)

# Setting thresholds:
nCount_RNA_threshold <- 1500
nFeature_RNA_threshold <- 500
percent_mt_threshold <- 6


p1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nCount_RNA_threshold)

p2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nFeature_RNA_threshold)

p3 <- seurat@meta.data %>% 
  	ggplot(aes(x=percent.mt)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = percent_mt_threshold)

grid.arrange(p1, p2, p3)    

# # Visualize feature relationships
# # plot1 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "percent.mt")

# # plot2 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
# # plot1+plot2

cell_number <- list(nrow(seurat[[]]))
```

Filtering Low Quality Cells

```{r}
# Filter data

cell_number <- list(nrow(seurat[[]]))
seurat <- subset(seurat, subset = nFeature_RNA  > nFeature_RNA_threshold & percent.mt < percent_mt_threshold & nCount_RNA > nCount_RNA_threshold)
cell_number <- append(cell_number, nrow(seurat[[]]))

print("Cell number before and after filterings")
print(cell_number)
```

:::{.content-visible when-meta='verbose'}
## Data Normalization 
:::

```{r}
# Normalize RNA data with SCTransform
# seurat <- SCTransform(seurat, verbose = T)

# Normalize HTO data with CLR
seurat <- NormalizeData(seurat, assay = "HTO",normalization.method = "CLR")
```

:::{.content-visible when-meta='processing'}
## Hashtag Demultiplexing - MAIT
:::

```{r}
#| fig-width: 14
#| fig-height: 10
plot1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot1+plot2

seurat <- HTODemux(seurat, assay = "HTO", positive.quantile = 0.99)

table(seurat$HTO_classification.global)

#tSNE-visualization
seurat_subset <- subset(seurat, idents = "Negative", invert =TRUE)
DefaultAssay(seurat_subset) <- "HTO"
seurat_subset <- ScaleData(seurat_subset, assay = "HTO",  features = rownames(seurat_subset), verbose = FALSE )
seurat_subset <- RunPCA(seurat_subset, assay = "HTO", rownames(seurat_subset), approx = FALSE)
seurat_subset <- RunTSNE(seurat_subset, assay = "HTO", dims = 1:24,  perplexity = 100, check_duplicates=FALSE)
DimPlot(seurat_subset)
```


```{r}
print("Number of Singlets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Singlet')))

print("Number of Doublets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Doublet')))

# Filtering singlets only
Idents(seurat) <- "HTO_classification.global"
seurat <-subset(seurat, idents = "Singlet")
```

:::{.content-visible when-meta='verbose'}
## Create Groups
:::

```{r}

seurat@meta.data <- seurat@meta.data |>
    mutate(Groups = 'Abau') |>
    mutate(Samples = str_c(Groups, hash.ID, sep = '_')) |> 
    mutate(Samples = case_when(
            hash.ID == 'Mouse-4' ~ 'Abau_Mouse-3',
            hash.ID == 'Mouse-5' ~ 'Abau_Mouse-4',
            TRUE ~ Samples
    )) |>
    mutate(Samples = factor(Samples, levels = c(
            'Abau_Mouse-1',
            'Abau_Mouse-2',
            'Abau_Mouse-3',
            'Abau_Mouse-4')))

```

```{r}
seurat_Abau <- seurat
```

:::{.content-visible when-meta='processing'}

# Preprocessing Ftula MAIT dataset

:::

:::{.content-visible when-meta='verbose'}

## Read in data

:::

```{r}
#| results: hide

seurat <- seurat_list$Ftula_MAIT

```

:::{.content-visible when-meta='processing'}
#### Confirming used HTOs 
:::

```{r}
print('Antibody Capture total reads:')
seurat@assays$HTO$counts |> as.matrix() |>
        rowSums()  |> as.data.frame() 

HTO_data <- seurat@assays$HTO$counts

# Extracting HTO and biotin data
# HTO_data <-  HTO_data[-c(3), ]
# colnames(scdata$'HTO') <- colnames(scdata$'Antibody Capture')
# rownames(scdata$'HTO') <- rownames(scdata$'Antibody Capture')[-c(2, 3, 4, 5, 6, 7, 11, 17, 18, 19)]

# #Add HTO data as a new assay independent from RNA
HTO <- CreateAssayObject(counts = HTO_data)
seurat[["HTO"]] <- HTO
```

Hashtags Used: 1, 2, 3, 4, 5

:::{.content-visible when-meta='processing'}

## QC

:::

```{r}
#| layout-ncol: 2
#Quantifying percentage mitochondria
Idents(seurat) <- 'Tissue'
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, assay="RNA",pattern = "mt-")

VlnPlot(seurat, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol=3, pt.size = 0.01)

# Setting thresholds:
nCount_RNA_threshold <- 1500
nFeature_RNA_threshold <- 500
percent_mt_threshold <- 6


p1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nCount_RNA_threshold)

p2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nFeature_RNA_threshold)

p3 <- seurat@meta.data %>% 
  	ggplot(aes(x=percent.mt)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = percent_mt_threshold)

grid.arrange(p1, p2, p3)    

# # Visualize feature relationships
# # plot1 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "percent.mt")

# # plot2 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
# # plot1+plot2

cell_number <- list(nrow(seurat[[]]))
```

Filtering Low Quality Cells

```{r}
# Filter data

cell_number <- list(nrow(seurat[[]]))
seurat <- subset(seurat, subset = nFeature_RNA  > nFeature_RNA_threshold & percent.mt < percent_mt_threshold & nCount_RNA > nCount_RNA_threshold)
cell_number <- append(cell_number, nrow(seurat[[]]))

print("Cell number before and after filterings")
print(cell_number)
```

:::{.content-visible when-meta='verbose'}
## Data Normalization 
:::

```{r}
# Normalize RNA data with SCTransform
# seurat <- SCTransform(seurat, verbose = T)

# Normalize HTO data with CLR
seurat <- NormalizeData(seurat, assay = "HTO",normalization.method = "CLR")
```

:::{.content-visible when-meta='processing'}
## Hashtag Demultiplexing - MAIT
:::

```{r}
#| fig-width: 14
#| fig-height: 10
plot1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot1+plot2

seurat <- HTODemux(seurat, assay = "HTO", positive.quantile = 0.97)

table(seurat$hash.ID)
table(seurat$HTO_classification.global)

#tSNE-visualization
seurat_subset <- subset(seurat, idents = "Negative", invert =TRUE)
DefaultAssay(seurat_subset) <- "HTO"
seurat_subset <- ScaleData(seurat_subset, assay = "HTO",  features = rownames(seurat_subset), verbose = FALSE )
seurat_subset <- RunPCA(seurat_subset, assay = "HTO", rownames(seurat_subset), approx = FALSE)
seurat_subset <- RunTSNE(seurat_subset, assay = "HTO", dims = 1:24,  perplexity = 100, check_duplicates=FALSE)
DimPlot(seurat_subset)
```


```{r}
print("Number of Singlets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Singlet')))

print("Number of Doublets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Doublet')))

# Filtering singlets only
Idents(seurat) <- "HTO_classification.global"
seurat <-subset(seurat, idents = "Singlet")
```

:::{.content-visible when-meta='verbose'}
## Create Groups
:::

```{r}

seurat@meta.data <- seurat@meta.data |>
    mutate(Groups = 'Ftula') |>
    mutate(Samples = str_c(Groups, hash.ID, sep = '_')) |> 
    mutate(Samples = factor(Samples, levels = c(
            'Ftula_Mouse-1',
            'Ftula_Mouse-2',
            'Ftula_Mouse-3',
            'Ftula_Mouse-4',
            'Ftula_Mouse-5')))

```

```{r}
seurat_Ftula <- seurat
```

:::{.content-visible when-meta='processing'}

# Preprocessing CT MAIT dataset

:::

:::{.content-visible when-meta='verbose'}

## Read in data

:::

```{r}
#| results: hide

seurat <- seurat_list$CT_MAIT

```

:::{.content-visible when-meta='processing'}
#### Confirming used HTOs
:::

```{r}
print('Antibody Capture total reads:')
seurat@assays$HTO$counts |> as.matrix() |>
        rowSums()  |> as.data.frame() 

HTO_data <- seurat@assays$HTO$counts

# Extracting HTO and biotin data
HTO_data <-  HTO_data[-c(3), ]
# colnames(scdata$'HTO') <- colnames(scdata$'Antibody Capture')
# rownames(scdata$'HTO') <- rownames(scdata$'Antibody Capture')[-c(2, 3, 4, 5, 6, 7, 11, 17, 18, 19)]

# #Add HTO data as a new assay independent from RNA
HTO <- CreateAssayObject(counts = HTO_data)
seurat[["HTO"]] <- HTO
```

Hashtags Used: 1, 2, 4, 5

:::{.content-visible when-meta='processing'}

## QC

:::

```{r}
#| layout-ncol: 2
#Quantifying percentage mitochondria
Idents(seurat) <- 'Tissue'
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, assay="RNA",pattern = "mt-")

VlnPlot(seurat, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol=3, pt.size = 0.01)

# Setting thresholds:
nCount_RNA_threshold <- 1500
nFeature_RNA_threshold <- 500
percent_mt_threshold <- 6


p1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nCount_RNA_threshold)

p2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nFeature_RNA_threshold)

p3 <- seurat@meta.data %>% 
  	ggplot(aes(x=percent.mt)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = percent_mt_threshold)

grid.arrange(p1, p2, p3)    

# # Visualize feature relationships
# # plot1 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "percent.mt")

# # plot2 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
# # plot1+plot2

cell_number <- list(nrow(seurat[[]]))
```

Filtering Low Quality Cells

```{r}
# Filter data

cell_number <- list(nrow(seurat[[]]))
seurat <- subset(seurat, subset = nFeature_RNA  > nFeature_RNA_threshold & percent.mt < percent_mt_threshold & nCount_RNA > nCount_RNA_threshold)
cell_number <- append(cell_number, nrow(seurat[[]]))

print("Cell number before and after filterings")
print(cell_number)
```

:::{.content-visible when-meta='verbose'}
## Data Normalization 
:::

```{r}
# Normalize RNA data with SCTransform
# seurat <- SCTransform(seurat, verbose = T)

# Normalize HTO data with CLR
seurat <- NormalizeData(seurat, assay = "HTO",normalization.method = "CLR")
```

:::{.content-visible when-meta='processing'}
## Hashtag Demultiplexing - MAIT
:::

```{r}
#| fig-width: 14
#| fig-height: 10
plot1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_HTO)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") 

plot1+plot2

seurat <- HTODemux(seurat, assay = "HTO", positive.quantile = 0.99)

table(seurat$hash.ID)
table(seurat$HTO_classification.global)

#tSNE-visualization
seurat_subset <- subset(seurat, idents = "Negative", invert =TRUE)
DefaultAssay(seurat_subset) <- "HTO"
seurat_subset <- ScaleData(seurat_subset, assay = "HTO",  features = rownames(seurat_subset), verbose = FALSE )
seurat_subset <- RunPCA(seurat_subset, assay = "HTO", rownames(seurat_subset), approx = FALSE)
seurat_subset <- RunTSNE(seurat_subset, assay = "HTO", dims = 1:24,  perplexity = 100, check_duplicates=FALSE)
DimPlot(seurat_subset)
```


```{r}
print("Number of Singlets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Singlet')))

print("Number of Doublets")
print(nrow(seurat@meta.data |> filter( HTO_classification.global  == 'Doublet')))

# Filtering singlets only
Idents(seurat) <- "HTO_classification.global"
seurat <-subset(seurat, idents = "Singlet")
```

:::{.content-visible when-meta='verbose'}
## Create Groups
:::

```{r}

seurat@meta.data <- seurat@meta.data |>
    mutate(Groups = 'CT') |>
    mutate(Samples = str_c(Groups, hash.ID, sep = '_')) |> 
    mutate(Samples=case_when(
            hash.ID == 'Mouse-4' ~ 'CT_Mouse-3',
            hash.ID == 'Mouse-5' ~ 'CT_Mouse-4',
            TRUE ~ Samples
    )) |>  
    mutate(Samples = factor(Samples, levels = c(
            'CT_Mouse-1',
            'CT_Mouse-2',
            'CT_Mouse-3',
            'CT_Mouse-4')))

```

```{r}
seurat_CT <- seurat
```

# Merged processing

## Re-merging datasets

```{r}
data_list <- list(CT_MAIT = seurat_CT, Abau_MAIT = seurat_Abau, Ftula_MAIT = seurat_Ftula)
# Merging the datasets
seurat <- Merge_Seurat_List(data_list, 
                       add.cell.ids = names(data_list),
                       project = "MAIT_Abau_Ftula_CT_Lungs")

seurat <- JoinLayers(seurat, assay = 'RNA')

```

:::{.content-visible when-meta='verbose'}

## Edit Groups

:::

```{r}

seurat@meta.data <- seurat@meta.data |>
    mutate(Groups = factor(Groups, levels = c('CT', 'Abau', 'Ftula'))) |>
    mutate(Samples = factor(Samples, levels = c(
            'CT_Mouse-1',
            'CT_Mouse-2',
            'CT_Mouse-3',
            'CT_Mouse-4',
            'Abau_Mouse-1',
            'Abau_Mouse-2',
            'Abau_Mouse-3',
            'Abau_Mouse-4',
            'Ftula_Mouse-1',
            'Ftula_Mouse-2',
            'Ftula_Mouse-3',
            'Ftula_Mouse-4',
            'Ftula_Mouse-5')))

seurat <- RenameCells(seurat, new.names=seurat$barcode)
```

:::{.content-visible when-meta='verbose'}

## Import TCR data

:::

```{r}
contigs <- read.csv(here('data/cluster_processing_2/aggr_all/outs/vdj_t/filtered_contig_annotations.csv'))

#Demultiplexing VDJ libraries
contig_list <- createHTOContigList(contigs, seurat, group.by = "Samples")
contig_list <- contig_list[c(levels(seurat$Samples))]
# Combining the Contigs
combined <- combineTCR(contig_list, removeNA = TRUE, samples=c(levels(seurat$Samples)), filterMulti = TRUE)

# # #Adding Samples as variables
combined <- addVariable(combined, variable.name='Samples', variables=c(unique(seurat$Samples)))                                    

## Integrating data with seurat object
cell_names <- Cells(seurat)
groups_merge <- pull(seurat@meta.data, Samples)

#Changing barcodes on the seurat object for the merge with VDJ data
new_cell_names <- paste(groups_merge, cell_names, sep='_')
seurat$barcode <- cell_names
seurat <- RenameCells(seurat, new.names=new_cell_names)

# #Combining VDJ and Seurat Object
seurat <- combineExpression(combined, seurat, proportion = TRUE, cloneCall='aa', group.by='Samples', chain = 'both')
# Renaming cells back to original barcodes
seurat <- RenameCells(seurat, new.names=seurat$barcode)
```

:::{.content-visible when-meta='verbose'}

## Data Normalization

:::

```{r}
# Normalize RNA data with SCTransform
seurat <- SCTransform(seurat, verbose = F) 
# seurat@assays$SCT
# seurat <- PrepSCTFindMarkers(seurat)
```

:::{.content-visible when-meta='processing'}

## Clustering

:::

```{r}
#| fig-width: 4
#| fig-height: 4
#| fig-cap: 'Elbow plot showing the percentage of variability represented by by each PC to select number of dimensions '

#Dimensionality reduction
seurat <- RunPCA(seurat,npcs = 100)

#Determining dimensionality of the dataset
ElbowPlot(seurat, ndims = 100)
```

:::{.content-visible when-meta='processing'}

Chosen Number of Dimensions:

:::

```{r}
# Describe number of dimensions

dimensions  <- 25 
dimensions
```

```{r}
#| fig-cap: UMAP plots showing the clustering results at different resolutions
#| fig-width: 10
#| fig-height: 10

seurat <- RunUMAP(seurat, dims=1:dimensions, verbose = F, seed.use = 3514L)
# resolutions <- c(0.25, 0.375, 0.5, 0.625, 0.75, 1, 1.25, 1.5, 1.75)
resolutions <- c(0.05, 0.1, 0.15, 0.25, 0.375, 0.5)
seurat <- FindNeighbors(seurat, dims = 1:dimensions, verbose = F)
seurat <- FindClusters(seurat, resolution = resolutions, verbose = F)

p <- list()
i <- 1
for (resolution in resolutions ) {
    Idents(seurat) <- paste0('SCT_snn_res.', resolution) 
    p2 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")    
    p[[i]] <- p2
    i <- i+1
}

plot <- grid.arrange(grobs = p)
# ggsave('initial_clustering_results_by_resolution.pdf', path = here('result/figures'), plot = plot)
```

:::{.content-visible when-meta='processing'}

Chosen resolution

:::

```{r}
resolution <- 0.375
resolution
```

```{r}
object_annotations <- 'full_object_merged_pre_filtering'
```

### Clustering Results

```{r}
#| layout-ncol: 2
Idents(seurat) <- paste0('SCT_snn_res.', resolution)
seurat[['seurat_clusters']]<- Idents(seurat)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Groups', ncol = 2) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_group_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
# DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Samples', ncol = 4) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
# ggsave(paste0('UMAP_clusters_by_sample_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = FALSE, group.by = 'Groups') + ggtitle(paste0('R ', resolution)) + theme(legend.position = "right")
```

### Top DEGs per cluster

```{r}
#| layout-ncol: 3
# seurat <- PrepSCTFindMarkers(seurat)
plot1 <- top_genes_per_cluster(seurat, n_genes_to_plot = 3, object_annotations = object_annotations, tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = F) 
```

```{r}
#| fig-width: 8
#| fig-height: 12
print(plot1)
ggsave(paste0(figures_path, 'DotPlot_Top3_per_cluster', object_annotations, '.pdf'), width = 8, height = 12)
```
:::{.content-visible when-meta='processing'}

### Automatic cell type annotations

:::

```{r}
local_path <- paste0(figures_path, object_annotations,'_cell_type_annotations')
unlink(local_path,recursive = T)
dir.create(local_path)

# Normalize and scale data
# seurat <- JoinLayers(seurat, assay = 'RNA')
seurat <- NormalizeData(seurat, assay = "RNA", normalization.method = "LogNormalize", scale.fct = 10000)
seurat <- ScaleData(seurat, assay = 'RNA')
```

:::{.content-visible when-meta='processing'}

#### SingleR annotations

:::

:::{.content-visible when-meta='processing'}

Annotations per Cluster

:::

```{r}
#| fig-cap: UMAP plot showing the clustering results with SingleR fine annotations on a per cluster basis
#| fig-width: 9
#| fig-height: 6
seurat <- annotate_seurat_with_SingleR_Eduard(seurat, local_path, database = 'ImmGen', annotation_basis = 'cluster_coarse', split_by_groups = FALSE)
```

:::{.content-visible when-meta='processing'}

Annotations per Cell

:::

```{r}
#| fig-cap: UMAP plot showing the clustering results with SingleR coarse annotations on a per cell basis

annotate_seurat_with_SingleR_Eduard(seurat, local_path, database = 'ImmGen', annotation_basis = 'cell_coarse', split_by_groups = FALSE)
```

### Plotting Genes of Interest for cluster annotation

```{r}
#| fig-cap: Feature plots showing the expression of T cell markers
#| fig-width: 8
#| fig-height: 10
FeaturePlot_scCustom(seurat, features = c('Cd3e', 'Cd4', 'Cd8a', 'Cd8b1', 'Trdc', 'Trac'), pt.size = 0.01, colors_use = sequential_palette) +
    theme(legend.position = 'none')
```

```{r}
#| fig-cap: Feature plots showing the expression of myeloid cell markers
#| fig-width: 8
#| fig-height: 10
FeaturePlot_scCustom(seurat, features = c('Sirpa', 'Xcr1', 'Itgae', 'Itgax', 'Csf1r', 'Csf2ra'), pt.size = 0.1, colors_use = sequential_palette) +
    theme(legend.position = 'none')
```

```{r}
#| fig-cap: Feature plots showing the expression of neuron Markers
#| #| fig-width: 8
#| fig-height: 13
FeaturePlot_scCustom(seurat, features = c('Tubb3', 'Snap25', 'Gad1', 'Slc17a7'), pt.size = 0.1, colors_use = sequential_palette, num_columns = 2) 
```

## Removing contaminating clusters and preprocessing

### Removing macrophage cluster

```{r}
Idents(seurat) <- 'seurat_clusters'
seurat <- subset(seurat, idents = c('10'), invert = TRUE)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
```

:::{.content-visible when-meta='verbose'}

## Data Normalization

:::

```{r}
# Normalize RNA data with SCTransform
seurat <- SCTransform(seurat, verbose = F) 
# seurat@assays$SCT
# seurat <- PrepSCTFindMarkers(seurat)
```

:::{.content-visible when-meta='processing'}

## Clustering
Removing VDJ genes from clustering

:::

```{r}
#| fig-width: 4
#| #| fig-height: 4
#| fig-cap: 'Elbow plot showing the percentage of variability represented by by each PC to select number of dimensions '

seurat <- quietVDJgenes(seurat)

#Dimensionality reduction
seurat <- RunPCA(seurat,npcs = 100)

#Determining dimensionality of the dataset
ElbowPlot(seurat, ndims = 100)
```

:::{.content-visible when-meta='processing'}

Chosen Number of Dimensions:

:::

```{r}
# Describe number of dimensions

dimensions  <- 25 
dimensions
```

```{r}
#| fig-cap: UMAP plots showing the clustering results at different resolutions
#| fig-width: 10
#| fig-height: 10

seurat <- RunUMAP(seurat, dims=1:dimensions, verbose = F, seed.use = 3514L)
# resolutions <- c(0.25, 0.375, 0.5, 0.625, 0.75, 1, 1.25, 1.5, 1.75)
resolutions <- c(0.05, 0.1, 0.15, 0.175, 0.2, 0.225, 0.25, 0.375, 0.5)
seurat <- FindNeighbors(seurat, dims = 1:dimensions, verbose = F)
seurat <- FindClusters(seurat, resolution = resolutions, verbose = F)

p <- list()
i <- 1
for (resolution in resolutions ) {
    Idents(seurat) <- paste0('SCT_snn_res.', resolution) 
    p2 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")    
    p[[i]] <- p2
    i <- i+1
}

plot <- grid.arrange(grobs = p)
# ggsave('initial_clustering_results_by_resolution.pdf', path = here('result/figures'), plot = plot)
```

:::{.content-visible when-meta='processing'}

Chosen resolution

:::

```{r}
resolution <- 0.15
resolution
```

```{r}
object_annotations <- 'full_object_filtered'
resolution <- 0.15
```

## Clustering Results

```{r}
#| layout-ncol: 2
Idents(seurat) <- paste0('SCT_snn_res.', resolution)
seurat[['seurat_clusters']]<- Idents(seurat)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Groups', ncol = 2) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_group_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Samples', ncol = 4) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_sample_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = FALSE, group.by = 'Groups') + ggtitle(paste0('R ', resolution)) + theme(legend.position = "right")
```

### Cluster frequency per condition

```{r}
extract_cell_counts(seurat, seurat_clusters, figures_path, tables_path, object_annotations=object_annotations)
```

## Top DEGs per cluster

```{r}
#| layout-ncol: 3
# seurat <- PrepSCTFindMarkers(seurat)
plot1 <- top_genes_per_cluster(seurat, n_genes_to_plot=5, object_annotations, tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = T) 
```

```{r}
#| fig-width: 8
#| fig-height: 12
print(plot1)
ggsave(paste0(figures_path, 'DotPlot_Top3_per_cluster', object_annotations, '.pdf'), width = 8, height = 12)
```

## Exploring cluster 3

```{r}
#| layout-ncol: 2
seurat[["percent.ribo"]] <- PercentageFeatureSet(seurat, assay="RNA",pattern = "^Rp[sl]")
FeaturePlot_scCustom(seurat, features = c('nCount_RNA', 'nFeature_RNA', 'percent.mt', 'percent.ribo'), colors_use = sequential_palette)
FeaturePlot_scCustom(seurat, features = c('Zbtb16', '', 'Icos'), colors_use = sequential_palette)
VlnPlot(seurat, features = c('nCount_RNA', 'nFeature_RNA', 'percent.mt', 'percent.ribo'), group.by = 'seurat_clusters', pt.size = 0.1) + NoLegend()
VlnPlot(seurat, features = c('Zbtb16', 'Ubc', 'Icos'), group.by = 'seurat_clusters', pt.size = 0.1) + NoLegend()
FeatureScatter(seurat, feature1 = 'nCount_RNA', feature2 = 'nFeature_RNA', split.by = 'seurat_clusters', ncol = 3)
```

Cluster 3 looks like a potential technical artifact. At first glance, the slightly higher mitochondrial gene count and lower RNA counts suggested lower quality cells we could filter out. These also have a much lower ribosomal protein gene content, suggestive of damaged cells. Furthermore, the differential genes separating them from other clusters are not  informative and some are not expected to be present in this cell type (potential ambient RNA from the tissue). However the expression of other expected genes such as PLZF argues against that. One possibility is that these are damaged cells that have lost cytoplasmic integrity, leading to a loss of ribosomal protein genes and overall lower gene counts, while preserving nuclear transcripts. I will test this hypothesis by running DropletQC to quantify the fraction of nuclear (intronic) RNA, which we expect to be higher for damaged cells (Muskovic et al. Genome Biology 2021).

```{r}
saveRDS(seurat, file = here('data/MAIT_Abau_Ftula_CT_Lungs.rds'))
```

## Implementing DropletQC

```{r}
# renv::install('powellgenomicslab/DropletQC')
# devtools::install_github('powellgenomicslab/DropletQC')
# renv::snapshot()
library(DropletQC)
seurat <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs.rds'))
```

```{r}
# Ftula
# barcodes <- seurat |> subset(subset = Groups == 'Ftula') |> Cells() |> str_replace('-2', '-1')
nuclear_fraction_Ftula <- nuclear_fraction_tags(
   bam = here('data/cluster_processing_2/Ftula_MAIT/outs/per_sample_outs/Ftula_MAIT/count/sample_alignments.bam'),
   barcodes = here('data/cluster_processing_2/Ftula_MAIT/outs/per_sample_outs/Ftula_MAIT/count/sample_filtered_feature_bc_matrix/barcodes.tsv.gz'),
   tiles = 100,
   cores = 8,
   verbose = FALSE
)
# nuclear_fraction_Ftula |> head()
nuclear_fraction_Ftula <- nuclear_fraction_Ftula |> rownames_to_column('barcode') |> mutate(barcode = str_replace(barcode, '-1', '-2'))

# Control
# barcodes <- seurat |> subset(subset = Groups == 'CT') |> Cells()
nuclear_fraction_Control <- nuclear_fraction_tags(
   bam = here('data/cluster_processing_2/CT_MAIT/outs/per_sample_outs/CT_MAIT/count/sample_alignments.bam'),
   barcodes = here('data/cluster_processing_2/CT_MAIT/outs/per_sample_outs/CT_MAIT/count/sample_filtered_feature_bc_matrix/barcodes.tsv.gz'),
   tiles = 100,
   cores = 8,
   verbose = FALSE
)
# nuclear_fraction_Control |> head()
nuclear_fraction_Control <- nuclear_fraction_Control |> rownames_to_column('barcode')

# Abau
# barcodes <- seurat |> subset(subset = Groups == 'Abau') |> Cells() |> str_replace('-3', '-1')
nuclear_fraction_Abau <- nuclear_fraction_tags(
   bam = here('data/cluster_processing_2/Abau_MAIT/outs/per_sample_outs/Abau_MAIT/count/sample_alignments.bam'),
   barcodes = here('data/cluster_processing_2/Abau_MAIT/outs/per_sample_outs/Abau_MAIT/count/sample_filtered_feature_bc_matrix/barcodes.tsv.gz'),
   tiles = 100,
   cores = 8,
   verbose = FALSE
)

# Merging results
nuclear_fraction <- bind_rows(nuclear_fraction_Ftula, nuclear_fraction_Control, nuclear_fraction_Abau)

seurat@meta.data <- seurat@meta.data |>
    select(-starts_with('barcode')) |> 
    rownames_to_column('barcode') |>
    left_join(nuclear_fraction, by = 'barcode') |>
    column_to_rownames('barcode')

FeaturePlot_scCustom(seurat, features = c('nuclear_fraction'), colors_use = sequential_palette)

```

```{r}
#| layout-ncol: 2
VlnPlot_scCustom(seurat, features = c('nuclear_fraction')) & NoLegend()

ggplot(data = seurat@meta.data,
          aes(x = nuclear_fraction, group = seurat_clusters, fill = seurat_clusters)) +
   geom_density(adjust = 1.5, alpha = .4) + ggtitle("Nuclear fraction score")

seurat@meta.data <- seurat@meta.data |> mutate(nCount_RNA_log = log10(nCount_RNA))
FeatureScatter(seurat, feature1 = 'nuclear_fraction', feature2 = 'nCount_RNA_log', split.by = 'seurat_clusters', ncol = 2)
FeatureScatter(seurat, feature1 = 'nuclear_fraction', feature2 = 'percent.ribo', split.by = 'seurat_clusters', ncol = 2) + geom_hline(yintercept = 7, linetype = 'dashed', color = 'gray')

```

Very clearly Cluster number 4 is comprised of low quality, damaged cells with very low present ribosomal protein genes and high nuclear fraction. I will establish a threshold of 7% percent ribosomal protein genes to remove low-quality cells from all clusters and remove most cells from this cluster.

## Filtering out damaged cells

```{r}
#| layout-ncol: 2
#Quantifying percentage mitochondria
Idents(seurat) <- 'Tissue'
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, assay="RNA",pattern = "mt-")

VlnPlot(seurat, features = c("nCount_RNA", "nFeature_RNA", "percent.mt"), ncol=3, pt.size = 0.01)

# Setting thresholds:
nCount_RNA_threshold <- 1500
nFeature_RNA_threshold <- 500
percent_mt_threshold <- 6
ribosomal_threshold <- 7


p1 <- seurat@meta.data %>% 
  	ggplot(aes(x=nCount_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nCount_RNA_threshold)

p2 <- seurat@meta.data %>% 
  	ggplot(aes(x=nFeature_RNA)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = nFeature_RNA_threshold)

p3 <- seurat@meta.data %>% 
  	ggplot(aes(x=percent.mt)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = percent_mt_threshold)

p4 <- seurat@meta.data %>% 
  	ggplot(aes(x = percent.ribo)) + 
  	geom_density() + 
  	scale_x_log10() + 
  	theme_classic() +
  	ylab("Cell density") +
  	geom_vline(xintercept = ribosomal_threshold)

grid.arrange(p1, p2, p3, p4)    

# # Visualize feature relationships
# # plot1 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "percent.mt")

# # plot2 <- FeatureScatter(seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
# # plot1+plot2

cell_number <- list(nrow(seurat[[]]))
```

```{r}
# Filter data

cell_number <- list(nrow(seurat[[]]))
seurat <- subset(seurat, subset = nFeature_RNA  > nFeature_RNA_threshold & percent.mt < percent_mt_threshold & nCount_RNA > nCount_RNA_threshold & percent.ribo > ribosomal_threshold)
cell_number <- append(cell_number, nrow(seurat[[]]))

print("Cell number before and after filterings")
print(cell_number)
```

## Re Clustering

```{r}
#| fig-width: 4
#| #| fig-height: 4
#| fig-cap: 'Elbow plot showing the percentage of variability represented by by each PC to select number of dimensions '

# Regressing out ribosomal protein content
seurat <- SCTransform(seurat)
seurat <- quietVDJgenes(seurat)

#Dimensionality reduction
seurat <- RunPCA(seurat,npcs = 100)

#Determining dimensionality of the dataset
ElbowPlot(seurat, ndims = 100)
```

:::{.content-visible when-meta='processing'}

Chosen Number of Dimensions:

:::

```{r}
# Describe number of dimensions

dimensions  <- 25  
dimensions
```

```{r}
#| fig-cap: UMAP plots showing the clustering results at different resolutions
#| fig-width: 10
#| fig-height: 10

seurat <- RunUMAP(seurat, dims=1:dimensions, verbose = F, seed.use = 3514L, n.neighbors = 30, min.dist = 0.3)
# resolutions <- c(0.25, 0.375, 0.5, 0.625, 0.75, 1, 1.25, 1.5, 1.75)
resolutions <- c(0.05, 0.1, 0.2, 0.25, 0.375, 0.5, 0.625, 0.75, 1)
# resolutions <- c(0.375)
seurat <- FindNeighbors(seurat, dims = 1:dimensions, verbose = F)
seurat <- FindClusters(seurat, resolution = resolutions, verbose = F, algorithm = 4, random.seed = 3514L)

p <- list()
i <- 1
for (resolution in resolutions ) {
    Idents(seurat) <- paste0('SCT_snn_res.', resolution) 
    p2 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")    
    p[[i]] <- p2
    i <- i+1
}

plot <- grid.arrange(grobs = p)
# ggsave('initial_clustering_results_by_resolution.pdf', path = here('result/figures'), plot = plot)
```

```{r}
#renv::install('clustree')
library(clustree)
conflicts_prefer(base::setdiff)
clustree(seurat, prefix = "SCT_snn_res.")
```

:::{.content-visible when-meta='processing'}

Chosen resolution

:::

```{r}
resolution <- 0.625
resolution
```

:::{.content-visible when-meta='verbose'}

### Save Object

:::

```{r}
saveRDS(seurat, file = here('data/MAIT_Abau_Ftula_CT_Lungs_unintegrated.rds'))
```

## Clustering with integration - RPCA

```{r}
seurat <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs_unintegrated.rds'))
```

:::{.content-visible when-meta='verbose'}
## Normalize and Integrate
:::

```{r}
# Split layers by lane
seurat[["RNA"]] <- split(seurat[["RNA"]], f = seurat$Groups)

# Normalize RNA data with SCTransform
seurat <- SCTransform(seurat, verbose = T, assay = 'RNA')
seurat <- quietVDJgenes(seurat)

seurat <- RunPCA(seurat, verbose = T, assay = 'SCT', npcs=100)

# Integrate With RPCA (Reciprocal PCA - More appropriate for lanes from the same experiment with less overlap)
seurat <- IntegrateLayers(seurat, method = RPCAIntegration, 
                          normalization.method = 'SCT', 
                          orig.reduction = 'pca',
                          new.reduction = 'RPCA',
                          verbose = T)
```

:::{.content-visible when-meta='processing'}
## Clustering
:::

```{r}
#| fig-cap: 'Elbow plot showing the percentage of variability represented by by each PC to select number of dimensions '

#Determining dimensionality of the dataset
ElbowPlot(seurat, ndims = 100)
```

:::{.content-visible when-meta='processing'}
Chosen Number of Dimensions:
:::

```{r}
# Describe number of dimensions
dimensions  <- 30  
dimensions
```

```{r}
#| results: hide
seurat@meta.data <- seurat@meta.data |> select(-starts_with('SCT_snn_res'))
seurat <- RunUMAP(seurat, dims=1:dimensions, verbose = F, seed.use = 3514L, reduction = 'RPCA')
resolutions <- c(0.05, 0.1, 0.2, 0.25, 0.375, 0.5, 0.625, 0.75, 1)
seurat <- FindNeighbors(seurat, dims = 1:dimensions, verbose = F, reduction = 'RPCA')
seurat <- FindClusters(seurat, resolution = resolutions, verbose = F, algorithm = 4, random.seed = 3514L)
```

```{r}
#| layout-ncol: 2
#| fig-cap: UMAP plots showing the clustering results at different resolutions
#| fig-width: 10
#| fig-height: 10

p <- list()
i <- 1
for (resolution in resolutions ) {
    Idents(seurat) <- paste0('SCT_snn_res.', resolution) 
    p2 <- DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")    
    p[[i]] <- p2
    i <- i+1
}

plot <- grid.arrange(grobs = p)

library(clustree)
clustree(seurat, prefix = "SCT_snn_res.")
```

:::{.content-visible when-meta='processing'}
Chosen resolution
:::

```{r}
resolution <- 0.375
resolution
```

:::{.content-visible when-meta='verbose'}
### Save Object
:::

```{r}
saveRDS(seurat, file = here('data/MAIT_Abau_Ftula_CT_Lungs_integrated.rds'))
```


# Analysis - Unintegrated

```{r}
seurat <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs_unintegrated.rds'))
resolution <- 0.625
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
object_annotations <- 'unintegrated_object'
resolution <- 0.625
results_path <- here('results', 'unintegrated_analysis')
dir.create(results_path)
figures_path <- here('results', 'unintegrated_analysis', 'figures')
dir.create(figures_path)
tables_path <- here('results', 'unintegrated_analysis', 'tables')
dir.create(tables_path)
```

## Clustering Results

```{r}
#| layout-ncol: 2
Idents(seurat) <- paste0('SCT_snn_res.', resolution)
seurat[['seurat_clusters']]<- Idents(seurat)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Groups', ncol = 2) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_group_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Samples', ncol = 4) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_sample_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = FALSE, group.by = 'Groups') + ggtitle(paste0('R ', resolution)) + theme(legend.position = "right")
```

### Cluster frequency per condition

```{r}
extract_cell_counts(seurat, seurat_clusters, figures_path, tables_path, object_annotations=object_annotations)
```

## Top DEGs per cluster

```{r}
#| layout-ncol: 3
seurat <- PrepSCTFindMarkers(seurat)

plot1 <- top_genes_per_cluster(seurat, n_genes_to_plot=5, object_annotations, tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = T) 
```

```{r}
#| fig-width: 8
#| fig-height: 12
print(plot1)
ggsave(paste0(figures_path, 'DotPlot_Top3_per_cluster', object_annotations, '.pdf'), width = 8, height = 12)
```

## Plotting Genes of Interest from Wang et al 2024

```{r}
FeaturePlot_scCustom(seurat, features = c('Rorc', 'Il17a', 'Il17f', 'Icos'), colors_use = sequential_palette)
FeaturePlot_scCustom(seurat, features = c('Tbx21', 'Ifng', 'Gzmb', 'Cxcr3'), colors_use = sequential_palette)
FeaturePlot_scCustom(seurat, features = c('Ramp3', 'Sdc1', 'Gzma', 'Gzmk'), colors_use = sequential_palette)
FeaturePlot_scCustom(seurat, features = c('Klrg1', 'Slamf7', 'Ly6c', 'Itga1'), colors_use = sequential_palette)
FeaturePlot_scCustom(seurat, features = c('Ccr6', 'Il7r', 'Sdc1', 'Icos'), colors_use = sequential_palette)
FeaturePlot_scCustom(seurat, features = c('Pdcd1', 'Ctla4', 'Ly6a', 'S1pr5'), colors_use = sequential_palette)

```

## Looking at tissue dissociation stress markers

```{r}
seurat[['percent.hsp']] <- PercentageFeatureSet(seurat, assay="RNA",pattern = "^Hsp")
FeaturePlot_scCustom(seurat, features = c('percent.hsp'), colors_use = sequential_palette)
FeaturePlot_scCustom(seurat, features = c('Fos', 'Jun', 'percent.hsp'), colors_use = sequential_palette)
VlnPlot(seurat, features = c('percent.hsp', 'Jun'), group.by = 'seurat_clusters', pt.size = 0.1) + NoLegend()
```


## Signatures in manuscript

```{r}

DefaultAssay(seurat) <- 'SCT'
signatures_table <- read_excel(here('data/Hagoulet_signatures.xlsx'), skip = 1)
signatures_table <- signatures_table |>
    dplyr::filter(!(row_number() == 1)) |>
    dplyr::select(-Signature) 

vector <- signatures_table$ISG |> purrr::discard(is.na)

signatures <- map(.x = signatures_table, .f = ~ purrr::discard(.x, is.na))

# Removing genes that are not detected
blacklist_elements <- c('H2afz','Smsc2','Hist1h2ap','Flg','Flg2','Bmp2','Bmp4','Ngf','Fgf7','Fgf10','Fgf22','Igf1','Ctgf','Mmp3','Mmp13','Defb1','Defb6','Epgn','Mmp2','Angpt1','Hgf','Cyr61','Inhbb','Btc','Ereg','Lep','Vegfc','Vegfd','Pdgfc','Cxcl12','Shh','Dhh','Ihh','Dll3','Dll4','Wnt1','Wnt2','Wnt3a','Wnt5a','Wnt6','Wnt7a','Wnt7b','Wnt8a','Wnt8b','Wnt9a','Wnt9b','Wnt10a','Wnt16','Chat','Thbs1','Fam46a','Kcna4','Fam129b','Tdpoz4','Rassf6','Ero1l','Cd244','A430078G23Rik','D1Ertd622e','Fam49a')

signatures <- map(.x = signatures, .f = ~ .x[!(.x %in% blacklist_elements)])


```

### Implementing signatures with UCell

```{r}

seurat <- AddModuleScore_UCell(seurat, features = signatures, name = NULL)

# Scale signatures (z-score) for comparison and visualization purposes
seurat@meta.data  <-  seurat@meta.data |>
    mutate(across(names(signatures), ~ scale(.x, center = T, scale = T), .names = '{col}_scaled'))
```

### Violin Plots of signatures
Violin plots showing the distribution of signature scores (z-scored) across the different conditions

```{r}
#| layout-ncol: 3

# Visualization
signature_violin_plot <- function (signature) {
    plot1 <- VlnPlot(seurat, features = paste0(signature, '_scaled'), group.by = 'Groups', pt.size = 0) + labs(title = signature) + NoLegend()
    print(plot1)
    ggsave(plot = plot1, filename = paste0('VlnPlot_', signature, '_by_group_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 4)
}
for (signature in names(signatures)) {
    signature_violin_plot(signature) 
}


```

Violin plots showing the distribution of signature scores (z-scored) across the different conditions

```{r}
#| layout-ncol: 3

# Visualization
signature_violin_plot <- function (signature) {
    plot1 <- VlnPlot(seurat, features = paste0(signature, '_scaled'), group.by = 'seurat_clusters', pt.size = 0) + NoLegend() + labs(title = signature)
    print(plot1)
    ggsave(plot = plot1, filename = paste0('VlnPlot_', signature, '_by_seurat_clusters_', object_annotations, '.pdf'), path = figures_path, width = 8, height = 4)
}
for (signature in names(signatures)) {
    signature_violin_plot(signature) 
}

```

### Dot Plots of Signature genes
Dot plots showing the expression of signature genes across the different Groups

```{r}
#| fig-width: 4
#| fig-height: 12
#| layout-ncol: 4 

signature_dot_plot <- function (signature) {
    plot1 <- DotPlot_scCustom(seurat, features = signatures[[signature]] |> rev() |> unique() , group.by = 'Groups', colors_use = sequential_palette_dotplot, flip_axes = T, scale = F, dot.scale = 6) + labs(title = signature)
    print(plot1)
    ggsave(plot = plot1, filename = paste0('DotPlot_', signature, '_by_groups_', object_annotations, '.pdf'), path = figures_path, width = 5, height = length(signatures[[signature]])/4 + 1)
}
for (signature in names(signatures)) {
    signature_dot_plot(signature) 
}

```

### FeaturePlots of Signatures 
FeaturePlots showing the expression of signatures across the UMAP

```{r}
#| layout-ncol: 2

diverging_palette_2 <- hcl.colors(n = 20,'RdBu',rev = T)
signature_feature_plot <- function (signature) {
    min_value  <-  min(seurat@meta.data |> pull(paste0(signature, '_scaled')))
    max_value  <-  max(seurat@meta.data |> pull(paste0(signature, '_scaled')))
    cut_off_value <- min(abs(min_value), abs(max_value))
    plot1 <- FeaturePlot_scCustom(seurat, features = paste0(signature, '_scaled'), colors_use = diverging_palette_2, na_cutoff=NULL, min.cutoff = -1*cut_off_value, max.cutoff = cut_off_value) + labs(title = signature)
    print(plot1)
    ggsave(plot = plot1, filename = paste0('FeaturePlot Underscore', signature, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
}
for (signature in names(signatures)) {
    signature_feature_plot(signature) 
}

FeaturePlot_scCustom(seurat, features = c('Sell', 'Lef1', 'S1pr1', 'Cd44'), colors_use = sequential_palette)
DotPlot_scCustom(seurat, features = c('Sell', 'Lef1', 'S1pr1', 'Cd44') , group.by = 'seurat_clusters', colors_use = sequential_palette_dotplot, flip_axes = T, scale = T, dot.scale = 6) + labs(title = 'LN MAIT cell markers')
```

:::{.content-visible when-meta='verbose'}

### Save Object

:::

```{r}
saveRDS(seurat, file = here('data/MAIT_Abau_Ftula_CT_Lungs_unintegrated.rds'))
```

## Differential gene expression analysis 

Here I implement pseudobulk analysis utilizing the biological replicates of the experiment (this is one of the preferred approaches in the literature, and statistically more sound than using Wilcoxon test considering each cell as a replicate)

Please note that more informative pathway enrichment analysis results can be obtained from Metascape.org

```{r}
seurat <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs_unintegrated.rds'))
```

```{r}
#| fig-width: 7
#| fig-height: 7
#| layout-ncol: 2
#| results: hide

gProfiler2 <- F 
source(here('scripts/function_template.r'))
path <- here(results_path, 'DEG_pseudobulk_analysis_Lymphocytes_Abau_vs_Ftula_seurat_clusters')
dir.create(path)
Idents(seurat) <- 'seurat_clusters'

DEG_counts <- data.frame(matrix(ncol=3, nrow=0))
colnames(DEG_counts) <- c('DEG_count', 'DEG_UP_count', 'DEG_DOWN_count')
rnames <- c()

DEG_counts <- pseudobulk(seurat, comparison='Samples', group1='Ftula', group2='Abau', cluster= 'all_MAITs' , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2)
rnames <- c(rnames, 'all_MAITs')

for (x in levels(seurat$seurat_clusters)) {
    # cat(paste0("Cell type: ", x, "\n"))
    seurat_small <- subset(seurat, subset = seurat_clusters == x )
    counts <- pseudobulk(seurat_small, comparison='Samples', group1='Ftula', group2='Abau', cluster= x , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2)
    rm(seurat_small)
    DEG_counts <- rbind(DEG_counts, counts)
    rnames <- c(rnames, x)
}
rownames(DEG_counts) <- rnames
write.csv(DEG_counts, file = file.path(path, "DEG_counts.csv"))
DEG_counts
```

# Analysis - Integrated

```{r}
seurat <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs_integrated.rds'))
object_annotations <- 'full_object_filtered'
resolution <- 0.375
results_path <- here('results', 'integrated_analysis')
dir.create(results_path)
figures_path <- here('results', 'integrated_analysis', 'figures')
dir.create(figures_path)
tables_path <- here('results', 'integrated_analysis', 'tables')
dir.create(tables_path)
```

## Clustering Results

```{r}
#| layout-ncol: 2
Idents(seurat) <- paste0('SCT_snn_res.', resolution)
seurat[['seurat_clusters']]<- Idents(seurat)
DimPlot(seurat, reduction = "umap", label = TRUE) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Groups', ncol = 2) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_group_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
DimPlot(seurat, reduction = "umap", label = TRUE, split.by = 'Samples', ncol = 4) + ggtitle(paste0('R ', resolution)) + theme(legend.position = "none")
ggsave(paste0('UMAP_clusters_by_sample_R_', resolution, '_', object_annotations, '.pdf'), path = figures_path)
DimPlot(seurat, reduction = "umap", label = FALSE, group.by = 'Groups') + ggtitle(paste0('R ', resolution)) + theme(legend.position = "right")
```

### Cluster frequency per condition

```{r}
extract_cell_counts(seurat, seurat_clusters, figures_path, tables_path, object_annotations=object_annotations)
```

## Top DEGs per cluster

```{r}
#| layout-ncol: 3
seurat <- PrepSCTFindMarkers(seurat)

plot1 <- top_genes_per_cluster(seurat, n_genes_to_plot=5, object_annotations, tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = T) 
```

```{r}
#| fig-width: 8
#| fig-height: 12
print(plot1)
ggsave(paste0(figures_path, 'DotPlot_Top3_per_cluster', object_annotations, '.pdf'), width = 8, height = 12)
```

After regressing out mitochondrial, and ribosomal gene content, cluster 3, which was very separate from the rest of the MAIT cell clusters, is now still a distinct cluster (cluster 4) but much closer to the rest of the clusters, suggesting that this is indeed a different cell state. They still express the MAIT TCR (see below).


## Plotting Genes of Interest from Wang et al 2024

```{r}
FeaturePlot_scCustom(seurat, features = c('Rorc', 'Il17a', 'Il17f', 'Icos'), colors_use = sequential_palette)
FeaturePlot_scCustom(seurat, features = c('Tbx21', 'Ifng', 'Gzmb', 'Cxcr3'), colors_use = sequential_palette)
FeaturePlot_scCustom(seurat, features = c('Ramp3', 'Sdc1', 'Gzma', 'Gzmk'), colors_use = sequential_palette)
FeaturePlot_scCustom(seurat, features = c('Klrg1', 'Slamf7', 'Ly6c', 'Itga1'), colors_use = sequential_palette)
FeaturePlot_scCustom(seurat, features = c('Ccr6', 'Il7r', 'Sdc1', 'Icos'), colors_use = sequential_palette)
FeaturePlot_scCustom(seurat, features = c('Pdcd1', 'Ctla4', 'Ly6a', 'S1pr5'), colors_use = sequential_palette)

```

## Looking at tissue dissociation stress markers

```{r}
seurat[['percent.hsp']] <- PercentageFeatureSet(seurat, assay="RNA",pattern = "^Hsp")
FeaturePlot_scCustom(seurat, features = c('percent.hsp'), colors_use = sequential_palette)
FeaturePlot_scCustom(seurat, features = c('Fos', 'Jun', 'percent.hsp'), colors_use = sequential_palette)
VlnPlot(seurat, features = c('percent.hsp', 'Jun'), group.by = 'seurat_clusters', pt.size = 0.1) + NoLegend()
```



## Signatures in manuscript

```{r}

DefaultAssay(seurat) <- 'SCT'
signatures_table <- read_excel(here('data/Hagoulet_signatures.xlsx'), skip = 1)
signatures_table <- signatures_table |>
    dplyr::filter(!(row_number() == 1)) |>
    dplyr::select(-Signature) 

vector <- signatures_table$ISG |> purrr::discard(is.na)

signatures <- map(.x = signatures_table, .f = ~ purrr::discard(.x, is.na))

# Removing genes that are not detected
blacklist_elements <- c('H2afz','Smsc2','Hist1h2ap','Flg','Flg2','Bmp2','Bmp4','Ngf','Fgf7','Fgf10','Fgf22','Igf1','Ctgf','Mmp3','Mmp13','Defb1','Defb6','Epgn','Mmp2','Angpt1','Hgf','Cyr61','Inhbb','Btc','Ereg','Lep','Vegfc','Vegfd','Pdgfc','Cxcl12','Shh','Dhh','Ihh','Dll3','Dll4','Wnt1','Wnt2','Wnt3a','Wnt5a','Wnt6','Wnt7a','Wnt7b','Wnt8a','Wnt8b','Wnt9a','Wnt9b','Wnt10a','Wnt16','Chat','Thbs1','Fam46a','Kcna4','Fam129b','Tdpoz4','Rassf6','Ero1l','Cd244','A430078G23Rik','D1Ertd622e','Fam49a')

signatures <- map(.x = signatures, .f = ~ .x[!(.x %in% blacklist_elements)])


```

### Implementing signatures with UCell

```{r}

seurat <- AddModuleScore_UCell(seurat, features = signatures, name = NULL)

# Scale signatures (z-score) for comparison and visualization purposes
seurat@meta.data  <-  seurat@meta.data |>
    mutate(across(names(signatures), ~ scale(.x, center = T, scale = T), .names = '{col}_scaled'))
```

### Violin Plots of signatures
Violin plots showing the distribution of signature scores (z-scored) across the different conditions

```{r}
#| layout-ncol: 3

# Visualization
signature_violin_plot <- function (signature) {
    plot1 <- VlnPlot(seurat, features = paste0(signature, '_scaled'), group.by = 'Groups', pt.size = 0) + labs(title = signature) + NoLegend()
    print(plot1)
    ggsave(plot = plot1, filename = paste0('VlnPlot_', signature, '_by_group_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 4)
}
for (signature in names(signatures)) {
    signature_violin_plot(signature) 
}


```

Violin plots showing the distribution of signature scores (z-scored) across the different conditions

```{r}
#| layout-ncol: 3

# Visualization
signature_violin_plot <- function (signature) {
    plot1 <- VlnPlot(seurat, features = paste0(signature, '_scaled'), group.by = 'seurat_clusters', pt.size = 0) + NoLegend() + labs(title = signature)
    print(plot1)
    ggsave(plot = plot1, filename = paste0('VlnPlot_', signature, '_by_seurat_clusters_', object_annotations, '.pdf'), path = figures_path, width = 8, height = 4)
}
for (signature in names(signatures)) {
    signature_violin_plot(signature) 
}

```

### Dot Plots of Signature genes
Dot plots showing the expression of signature genes across the different Groups

```{r}
#| fig-width: 4
#| fig-height: 12
#| layout-ncol: 4 

signature_dot_plot <- function (signature) {
    plot1 <- DotPlot_scCustom(seurat, features = signatures[[signature]] |> rev() |> unique() , group.by = 'Groups', colors_use = sequential_palette_dotplot, flip_axes = T, scale = F, dot.scale = 6) + labs(title = signature)
    print(plot1)
    ggsave(plot = plot1, filename = paste0('DotPlot_', signature, '_by_groups_', object_annotations, '.pdf'), path = figures_path, width = 5, height = length(signatures[[signature]])/4 + 1)
}
for (signature in names(signatures)) {
    signature_dot_plot(signature) 
}

```

### FeaturePlots of Signatures 
FeaturePlots showing the expression of signatures across the UMAP

```{r}
#| layout-ncol: 2

diverging_palette_2 <- hcl.colors(n = 20,'RdBu',rev = T)
signature_feature_plot <- function (signature) {
    min_value  <-  min(seurat@meta.data |> pull(paste0(signature, '_scaled')))
    max_value  <-  max(seurat@meta.data |> pull(paste0(signature, '_scaled')))
    cut_off_value <- min(abs(min_value), abs(max_value))
    plot1 <- FeaturePlot_scCustom(seurat, features = paste0(signature, '_scaled'), colors_use = diverging_palette_2, na_cutoff=NULL, min.cutoff = -1*cut_off_value, max.cutoff = cut_off_value) + labs(title = signature)
    print(plot1)
    ggsave(plot = plot1, filename = paste0('FeaturePlot Underscore', signature, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
}
for (signature in names(signatures)) {
    signature_feature_plot(signature) 
}

FeaturePlot_scCustom(seurat, features = c('Sell', 'Lef1', 'S1pr1', 'Cd44'), colors_use = sequential_palette)
DotPlot_scCustom(seurat, features = c('Sell', 'Lef1', 'S1pr1', 'Cd44') , group.by = 'seurat_clusters', colors_use = sequential_palette_dotplot, flip_axes = T, scale = T, dot.scale = 6) + labs(title = 'LN MAIT cell markers')
```

:::{.content-visible when-meta='verbose'}

### Save Object

:::

```{r}
saveRDS(seurat, file = here('data/MAIT_Abau_Ftula_CT_Lungs_integrated.rds'))
```

## Differential gene expression analysis 

Here I implement pseudobulk analysis utilizing the biological replicates of the experiment (this is one of the preferred approaches in the literature, and statistically more sound than using Wilcoxon test considering each cell as a replicate)

Please note that more informative pathway enrichment analysis results can be obtained from Metascape.org

```{r}
seurat <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs_integrated.rds'))
```

```{r}
#| fig-width: 7
#| fig-height: 7
#| layout-ncol: 2
#| results: hide

gProfiler2 <- TRUE 

path <- here(results_path, 'DEG_pseudobulk_analysis_Lymphocytes_Abau_vs_Ftula_seurat_clusters')
dir.create(path)
Idents(seurat) <- 'seurat_clusters'

DEG_counts <- data.frame(matrix(ncol=3, nrow=0))
colnames(DEG_counts) <- c('DEG_count', 'DEG_UP_count', 'DEG_DOWN_count')
rnames <- c()

DEG_counts <- pseudobulk(seurat, comparison='Samples', group1='Ftula', group2='Abau', cluster= 'all_MAITs' , path=path, label_threshold = 10000, max_overlaps = 20,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2)
rnames <- c(rnames, 'all_MAITs')

for (x in levels(seurat$seurat_clusters)) {
    # cat(paste0("Cell type: ", x, "\n"))
    seurat_small <- subset(seurat, subset = seurat_clusters == x )
    counts <- pseudobulk(seurat_small, comparison='Samples', group1='Ftula', group2='Abau', cluster= x , path=path, label_threshold = 10000, max_overlaps = 15,gene_lists_to_plot = NA, FC_threshold = 1,p_value_threshold = 0.05, run_gProfiler2 = gProfiler2)
    rm(seurat_small)
    DEG_counts <- rbind(DEG_counts, counts)
    rnames <- c(rnames, x)
}
rownames(DEG_counts) <- rnames
write.csv(DEG_counts, file = file.path(path, "DEG_counts.csv"))
DEG_counts
```

