---
title: Analysis of Abau and Ftul infected mice - lungs
author: Eduard Ansaldo
date: '`r Sys.Date()`'
verbose: false
engine: knitr
knitr:
  opts_chunk: 
    echo: false
    warning: false
    message: false
    error: false
    cache: true
    results: markdown
    code-fold: true
    cache-lazy: false
    
processing: true

execute: 
  cache: true


format:
  html:
    embed-resources: true
    fig-cap-location: top
    toc: true
    toc-depth: 3

    toc-expand: 1
    toc-location: left
    toc-title: Table of Contents
    toc-collapse: true
    number-sections: true
    page-layout: full
    fig-width: 6
    fig-height: 6
    fig-align: 'center'
---

# Initialization

```{r}
#| cache: false
# General R and plotting 
library(conflicted)
library(tidyverse)
library(ggplot2)
library(scales)
library(patchwork)
library(cowplot)
library(gridExtra)
library(ggrepel)
library(stringr)
library(VennDiagram)
library(pheatmap)
library(viridis)
library(here)
library(readxl)
library(readxl)
library(ggbeeswarm)
library(tidyplots)
library(extrafont)
library(ggtext)

# Single Cell Analysis Packages
library(Seurat)
library(scRepertoire)
library(circlize)
library(scCustomize)
library(SingleR)
library(celldex)
library(UCell)
library(presto)
library(scDblFinder)
library(immunarch)

# DEG, pathway enrichment and visualization packages
library(DESeq2)
# library(gprofiler2)
library(clusterProfiler)
library(DOSE)
library(pathview)
library(org.Mm.eg.db)
library(scRepertoire)
library(enrichplot)
library(msigdbr)
library(rlang)

# Python Version 3.9.6!!! 

# peconflicts_scout()
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::count)
conflicts_prefer(dplyr::select)
conflicts_prefer(base::setdiff)
conflicts_prefer(dplyr::rename)
conflicts_prefer(base::unname)
conflicts_prefer(tidyr::expand)
conflicts_prefer(tidyplots::save_plot)

# Loading custom functions
source(here('scripts/function_template.r'))
source(here('scripts/metascape_functions.r'))
source(here('scripts/circlize_functions.r'))


```

```{r}
#| results: hide 
#| cache: false 
# Setting working directory and seed
# renv::update()
# renv::snapshot()
# renv::init()
# renv::status()
set.seed(3514)
# i_am('scripts/Analysis.qmd')
here()
results_path <- here('results/')
dir.create(results_path)
results_path <- here('results/')
dir.create(results_path)
figures_path <- here('results/figures/')
dir.create(figures_path)
# data_path <- here('data/')
# dir.create(data_path)
tables_path <- here('results/tables/')
dir.create(tables_path)

# Loading custom functions
source(here('scripts/function_template.r'))
source(here('scripts/metascape_functions.r'))
source(here('scripts/circlize_functions.r'))
source(here('scripts/ClusterProfiler_functions.r'))

# Setting up color palettes
diverging_palette <- hcl.colors(n = 20,'Purple-Green',rev = T)
sequential_palette_dotplot <- hcl.colors(n = 20,'YlGn',rev = T)
sequential_palette <- hcl.colors(n = 20,'Mako',rev = T)

pathway_enrichment <- c('ClusterProfiler')

knitr::dep_prev()
```


# Initialization

```{r}
seurat_original <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs.rds'))
seurat_cell_cycle <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs_cell_cycle.rds'))
seurat_Abau <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs_Abau_only.rds'))
seurat_Ftula <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs_Ftula_only.rds'))
results_path <- here('results', 'Manuscript_figures')
dir.create(results_path, recursive = T, showWarnings = F)
# unlink(results_path, recursive = T)
figures_path <- here('results/Manuscript_figures/figures/')
dir.create(figures_path, recursive = T, showWarnings = F)
tables_path <- here('results/Manuscript_figures/tables/')
dir.create(tables_path, recursive = T, showWarnings = F)
object_annotations <- 'manuscript_figures'

rename_samples_and_groups <- function(seurat_obj) {
    seurat_obj@meta.data <- seurat_obj@meta.data |>
        mutate(Samples = str_remove(Samples, '_Mouse')) |> 
        mutate(Samples = str_replace(Samples, 'CT-', 'Uninfected ')) |>
        mutate(Samples = str_replace(Samples, 'Abau-', 'A. bau ')) |>
        mutate(Samples = str_replace(Samples, 'Ftula-', 'F. tula ')) |>
        mutate(Groups = str_replace(Groups, 'CT', 'Uninfected')) |>
        mutate(Groups = str_replace(Groups, 'Abau', 'A. bau')) |>
        mutate(Groups = str_replace(Groups, 'Ftula', 'F. tula')) |>
        mutate(
            Samples = factor(Samples, levels = c(
                'Uninfected 1', 'Uninfected 2', 'Uninfected 3', 'Uninfected 4',
                'A. bau 1', 'A. bau 2', 'A. bau 3', 'A. bau 4',
                'F. tula 1', 'F. tula 2', 'F. tula 3', 'F. tula 4', 'F. tula 5'
            )),
            Groups = factor(Groups, levels = c('Uninfected', 'A. bau', 'F. tula'))
        )
    
    return(seurat_obj)
}

# Apply to all objects
seurat_original <- rename_samples_and_groups(seurat_original)
seurat_cell_cycle <- rename_samples_and_groups(seurat_cell_cycle)
seurat_Abau <- rename_samples_and_groups(seurat_Abau)
seurat_Ftula <- rename_samples_and_groups(seurat_Ftula)

```

## General parameters

```{r}
group_colors <- c(
    'Uninfected' = '#009e73',
    'A. bau' = '#0072b2', 
    'F. tula' = '#d55e10'
)

# Setting up color palettes
diverging_palette <- hcl.colors(n = 20,'Purple-Green',rev = T)
sequential_palette_dotplot <- hcl.colors(n = 20,'YlGn',rev = T)
sequential_palette <- hcl.colors(n = 20,'Mako',rev = T)
diverging_palette_2 <- hcl.colors(n = 20,'RdBu',rev = T)
# diverging_palette_3 <- hcl.colors(n = 20,'Geyser',rev = F)
diverging_palette_3 <- diverging_palette_2

UMAP_dot_size <- 0.4
rasterize <- F
raster_dpi <- c(3000, 3000)
alpha_exp <- 0.3

full_object_original <- list()
full_object_regressed <- list()
Abau_object <- list()
Ftula_object <- list()
VDJ_figures <- list()
```

# Generate original object plots

```{r}
seurat <- seurat_original
```

## DimPlots and cluster frequency

 ```{r}
Idents(seurat) <- 'seurat_clusters'

total_cell_number <- Cells(seurat) |> length()
Abau_cell_number <- Cells(seurat_Abau) |> length()
Ftula_cell_number <- Cells(seurat_Ftula) |> length()
Uninfected_cell_number <- Cells(seurat |> subset(subset = Groups == 'Uninfected')) |> length()

# Clusters UMAP plot

full_object_original$DimPlot <- DimPlot(seurat, group.by = 'seurat_clusters', pt.size = UMAP_dot_size, label = F, raster = rasterize) +
     labs(title = 'MAIT cell Leiden clusters', 
        #   subtitle = 'Lung single-cell RNA-seq', 
        #   caption = 'n = 4 Abau, n = 5 Ftula, n = 4 CT',
     x = 'UMAP 1', y = 'UMAP 2')   
full_object_original$DimPlot <- full_object_original$DimPlot  |> LabelClusters(id = 'seurat_clusters', size = 3, color = 'black', repel = F, fontface = 'bold') 

# Clusters split by group plot
full_object_original$DimPlot_groups <- DimPlot(seurat, group.by = 'seurat_clusters', pt.size = UMAP_dot_size, label = F, split.by = 'Groups', raster = rasterize) +
     labs( x = 'UMAP 1', y = 'UMAP 2', title = NULL)    +
     theme(strip.text = element_text(size = 14, face = 'italic'))



# Cluster frequency plot
# seurat_local <- subset(seurat, subset = Groups  != 'Uninfected')
cell_counts <- FetchData(seurat, vars = c("seurat_clusters", "Samples", "Groups"))
cell_counts <- arrange(cell_counts, Samples)
counts <- cell_counts %>% 
    add_count(Groups, name='total_cell_count_by_group')  |>
    mutate(weights = Ftula_cell_number/total_cell_count_by_group)  |>
    add_count(seurat_clusters, name='total_cell_count_by_cluster', wt =  weights)  |>
    count(seurat_clusters, Samples, Groups, total_cell_count_by_cluster , name='sample_cluster_count', wt = weights)  |> 
    mutate(frequency_within_cluster=sample_cluster_count*100/total_cell_count_by_cluster)  |> 
    mutate(Samples = as.character(Samples)) |> 
    arrange(Samples, desc(Samples)) 
frequency_table <- counts |> 
    arrange(Samples) |>     
    pivot_wider(id_cols = seurat_clusters,  names_from =  Samples , values_from = frequency_within_cluster, values_fill = 0)
# write.csv(frequency_table,file=here(tables_path, paste0( "frequency per roll }} per condition " , ".csv")), row.names=F) 

full_object_original$cluster_frequency <- counts |>
    ggplot(aes(x=seurat_clusters, fill=Groups, y = frequency_within_cluster)) +
    geom_col(aes(), position = 'fill') +
    # geom_bar(position=position_fill()) +
    theme_classic() +
    labs(x = 'Cluster', y = 'Frequency (%)', title = 'MAIT normalized cell cluster\nfrequency by sample')+
    scale_y_continuous(expand = expansion(mult = c(0, 0)), breaks = c(0, 0.25, 0.5, 0.75, 1), labels = c(0, 25, 50, 75, 100))+
    scale_fill_discrete(palette = group_colors) 

# full_object_original$DimPlot_groups

```

## Feature plots

```{r}

#  cluster Markers dot plot
seurat <- PrepSCTFindMarkers(seurat)
top_genes_per_cluster_results <- top_genes_per_cluster(seurat, n_genes_to_plot = 3, object_annotations = '', tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = F) 
full_object_original$cluster_dotplot <- top_genes_per_cluster_results$plot
ggsave(plot = full_object_original$cluster_dotplot$plot, filename = here(figures_path,  paste0('DotPlot_Top3_per_cluster', '.pdf')), width = 8, height = 12)


type_17_markers <- c('Rorc', 'Il17a', 'Sdc1', 'Il7r', 'Il1r1', 'Ccr6', 'Sdc4')
type_1_markers <- c('Tbx21', 'Ifng', 'I12rb2', 'Cxcr3', 'Gzmb', 'Gzmk', 'Klrg1', 'Klrd1', 'Ly6c2')
naive_markers <- c('S1pr1', 'Ccr7', 'Sell', 'Tcf7', 'Lef1')
regulatory_markers <- c('Arg2', 'Il10', 'Il21')
exhaustion_markers <- c('Ctla4', 'Pdcd1', 'Lag3', 'Tigit', 'Havcr2')
tissue_repair_markers <- c('Areg', 'Ramp3','Il18r1', 'Tgfb1', 'Tnf', 'Mmp25', 'Cxcl10', 'Rankl', 'Ifrd1')
transcription_factors <- c('Tbx21', 'Rorc', 'Gata3', 'Foxp3', 'Ahr', 'Maf', 'Prdm1', 'Bcl6', 'Zbtb16')
tissue_residency <- c('Itgae', 'Klrg1', 'Itga1', 'Itga2', 'Ccr9', 'Ccr10', 'Itga4', 'Itgb7')
activation_markers <- c('Cd69', 'Cd44', 'Nr4a1')
cytokines <- c('Ifng', 'Il17a', 'Il17f', 'Il4', 'Il10', 'Tgfb1', 'Tnf', 'Il2', 'Gzma', 'Gzmb', 'Nkg7', 'Ccl5', 'Il27ra', 'Ifgnr1')
costimulatory_receptors <- c('Ctla4', 'Lag3', 'Tigit', 'Havcr2', 'Cd28', 'Cd40lg', 'Pdcd1', 'Tnfsf4', 'Cd27', 'Tnfrsf9')

final_list <- c(type_17_markers, type_1_markers, naive_markers, regulatory_markers, exhaustion_markers, tissue_repair_markers, transcription_factors, tissue_residency, activation_markers, cytokines, costimulatory_receptors) |> unique() |> rev()

# final_gene_list <- 

# Idents(seurat) <- 'seurat_clusters'
# DotPlot_scCustom(seurat, features = final_list, colors_use=sequential_palette_dotplot, flip_axes = T, scale = T, col.min = 0, col.max = 2, dot.scale = 8, dot.min = 0, scale.min = 0, scale.max = 60, x_lab_rotate = T, y_lab_rotate = F)  + theme(axis.text.x = element_text(size = 14), axis.text.y = element_text(size = 14), legend.title = element_text(size = 18))
# ggsave(filename = here(figures_path, 'Dotplot_markers.pdf'), width = 25,height = 5)
# Idents(seurat) <- 'Groups'
# DotPlot_scCustom(seurat, features = final_list, colors_use=sequential_palette_dotplot, flip_axes = F, scale = F, col.min = 0, col.max = 2, dot.scale = 8, dot.min = 0, scale.min = 0, scale.max = 60, x_lab_rotate = T, y_lab_rotate = F)  + theme(axis.text.x = element_text(size = 14), axis.text.y = element_text(size = 14), legend.title = element_text(size = 18))


feature_plot <- function (gene) {
    # # Get the range of expression values for this gene
    # expr_range <- range(FetchData(seurat, vars = gene, assay = 'SCT', layer = 'data'), na.rm = TRUE)
    # expr_max <- ceiling(expr_range[2])
    # # Determine appropriate number of breaks based on range
    # n_breaks <- min(5, expr_max + 1)  # Maximum 5 breaks, or fewer if range is small
    # breaks <- if (expr_max <= 5) {
    #     seq(0, expr_max, by = 1)
    # } else {
    #     floor(pretty(c(0, expr_max), n = n_breaks))
    # }
    # # Ensure breaks are unique integers
    # breaks <- unique(as.integer(breaks))
    # print(breaks)
    plot1 <- FeaturePlot_scCustom(seurat, features = gene, colors_use = sequential_palette, pt.size = UMAP_dot_size, alpha_exp = alpha_exp) + 
        labs(title = paste0("*", gene, "*"), x = 'UMAP 1', y = 'UMAP 2') + 
        theme(plot.title = element_text(face = "italic")) 
        # scale_color_continuous(breaks = breaks)
    ggsave(plot = plot1, filename = paste0('FeaturePlot', gene, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
    return(plot1)
}
genes_to_plot <- c('Rorc', 'Il17a', 'Sdc1', 'Il7r', 'Il1r1', 'Ccr6', 'Tbx21', 'Ifng', 'Il12rb2', 'Cxcr3', 'Gzmb', 'Gzmk', 'Gzma', 'Gzmm', 'Klrg1', 'Klrd1', 'Ly6c2', 'Ramp3', 'Slamf7', 'Areg')
full_object_original$feature_plots <- genes_to_plot |> 
    set_names() |>
    map(feature_plot)

```

## Signature scores

```{r}

DefaultAssay(seurat) <- 'SCT'
signatures <- read_excel(here('data/Hagoulet_signatures.xlsx'), skip = 1)
signatures <- signatures |>
    dplyr::filter(!(row_number() == 1)) |>
    dplyr::select(-all_of(c('Signature', 'Tissue repair 1', 'Tissue repair 2', 'Tissue repair 3', 'Tissue repair 4')))
    
signatures <- map(.x = signatures, .f = ~ purrr::discard(.x, is.na)) |> as.list()
names(signatures) <- names(signatures) |> str_replace_all('_', ' ') 


tissue_repair_linehan <- c('Angpt1', 'Angpt2', 'Angpt4', 'Areg', 'Bmp2', 'Bmp4', 'Bmp7', 'Btc', 'Ccl2', 'Ccl3', 'Chat', 'Csf1', 'Csf2', 'Ctgf', 'Cxcl1', 'Cxcl10', 'Cxcl12', 'Cxcl15', 'Cxcl2', 'Cyr61', 'Defb1', 'Defb6', 'Dhh', 'Disp1', 'Dll1', 'Dll3', 'Dll4', 'Egf', 'Epgn', 'Ereg', 'Fgf1', 'Fgf10', 'Fgf22', 'Fgf7', 'Flg', 'Flg2', 'Furin', 'Hbegf', 'Hgf', 'Hif1a', 'Hmgb1', 'Igf1', 'Igf2', 'Ihh', 'Il1a', 'Il1b', 'Il6', 'Inhba', 'Inhbb', 'Jag1', 'Jag2', 'Lep', 'Mmp10', 'Mmp13', 'Mmp15', 'Mmp2', 'Mmp25', 'Mmp28', 'Mmp3', 'Ngf', 'Nrg1', 'Nrg2', 'Nrg3', 'Nrg4', 'Pdgfa', 'Pdgfb', 'Pdgfc', 'Pgf', 'Ptges2', 'Shh', 'Tgfa', 'Tgfb1', 'Tgfb2', 'Tgfb3', 'Thbs1', 'Tnf', 'Vegfa', 'Vegfb', 'Vegfc', 'Vegfd', 'Wnt1', 'Wnt10a', 'Wnt10b', 'Wnt11', 'Wnt16', 'Wnt2', 'Wnt2b', 'Wnt3', 'Wnt3a', 'Wnt4', 'Wnt5a', 'Wnt5b', 'Wnt6', 'Wnt7a', 'Wnt7b', 'Wnt8a', 'Wnt8b', 'Wnt9a', 'Wnt9b')

signatures$`Tissue repair` <- tissue_repair_linehan
signatures$MAIT17 <- signatures$MAIT17[!(signatures$MAIT17 %in% c('1700040D17Rik'))]

# Removing genes that are not detected
blacklist_elements <- c('H2afz','Smsc2','Hist1h2ap','Flg','Flg2','Bmp2','Bmp4','Ngf','Fgf7','Fgf10','Fgf22','Igf1','Ctgf','Mmp3','Mmp13','Defb1','Defb6','Epgn','Mmp2','Angpt1','Hgf','Cyr61','Inhbb','Btc','Ereg','Lep','Vegfc','Vegfd','Pdgfc','Cxcl12','Shh','Dhh','Ihh','Dll3','Dll4','Wnt1','Wnt2','Wnt3a','Wnt5a','Wnt6','Wnt7a','Wnt7b','Wnt8a','Wnt8b','Wnt9a','Wnt9b','Wnt10a','Wnt16','Chat','Thbs1','Fam46a','Kcna4','Fam129b','Tdpoz4','Rassf6','Ero1l','Cd244','A430078G23Rik','D1Ertd622e','Fam49a')

signatures <- map(.x = signatures, .f = ~ .x[!(.x %in% blacklist_elements)])

seurat <- AddModuleScore_UCell(seurat, features = signatures, name = NULL)

# Scale signatures (z-score) for comparison and visualization purposes
seurat@meta.data  <-  seurat@meta.data |>
    mutate(across(names(signatures), ~ scale(.x, center = T, scale = T), .names = '{col}_scaled'))
 # Visualization
signature_violin_plot <- function (signature) {
    max_y <- max(seurat@meta.data |> pull(signature))
    plot1 <- VlnPlot(seurat, features = signature, group.by = 'Groups', pt.size = 0, cols = group_colors) +
                labs(title = signature, x = NULL, y = 'Signature score') + NoLegend() +
                ylim(0, max_y*1.65) +
                scale_y_continuous(expand = expansion(mult = c(0, 0))) +
                theme(axis.text.x = element_text(face = 'italic')) +
                ggpubr::stat_compare_means(comparisons = list(c('A. bau', 'F. tula'), c('A. bau', 'Uninfected'), c('F. tula', 'Uninfected')), label = 'p.signif', method = 'wilcox.test',  hide.ns = F, label.y = max_y*1.1 , step.increase = 0.18, bracket.size = 0.7, size = 3, fontfamily = 'Arial') 
    ggsave(plot = plot1, filename = paste0('VlnPlot_', signature, '_by_group_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 4)
    return(plot1)
}

signature_dot_plot <- function (signature) {
    plot1 <- DotPlot_scCustom(seurat, features = signatures[[signature]] |> rev() |> unique() , group.by = 'Groups', colors_use = sequential_palette_dotplot, flip_axes = T, scale = F, dot.scale = 6) + labs(title = signature)
    ggsave(plot = plot1, filename = paste0('DotPlot_', signature, '_by_groups_', object_annotations, '.pdf'), path = figures_path, width = 5, height = length(signatures[[signature]])/4 + 1)
    return(plot1)
}
signature_feature_plot <- function (signature) {
    min_value  <-  min(seurat@meta.data |> pull(paste0(signature, '_scaled')))
    max_value  <-  max(seurat@meta.data |> pull(paste0(signature, '_scaled')))
    cut_off_value <- min(abs(min_value), abs(max_value))
    plot1 <- FeaturePlot_scCustom(seurat, features = paste0(signature, '_scaled'), colors_use = diverging_palette_3, na_cutoff=NULL, min.cutoff = -1*cut_off_value, max.cutoff = cut_off_value, pt.size = UMAP_dot_size, alpha_exp = alpha_exp) + labs(title = signature, x = 'UMAP 1', y = 'UMAP 2')
    ggsave(plot = plot1, filename = paste0('FeaturePlot_', signature, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
    return(plot1)
}

```

```{r}
full_object_original$signature_violin_plots <- map(names(signatures), signature_violin_plot)
names(full_object_original$signature_violin_plots) <- names(signatures)

full_object_original$signature_dot_plots <- map(names(signatures), signature_dot_plot)
names(full_object_original$signature_dot_plots) <- names(signatures)

full_object_original$signature_feature_plots <- map(names(signatures), signature_feature_plot)
names(full_object_original$signature_feature_plots) <- names(signatures)

```

# Generate cell cycle regressed object plots

```{r}
seurat <- seurat_cell_cycle
```

## DimPlots and cluster frequency

```{r}
Idents(seurat) <- 'seurat_clusters'

total_cell_number <- Cells(seurat) |> length()
Abau_cell_number <- Cells(seurat_Abau) |> length()
Ftula_cell_number <- Cells(seurat_Ftula) |> length()
Uninfected_cell_number <- Cells(seurat |> subset(subset = Groups == 'Uninfected')) |> length()

# Clusters UMAP plot
full_object_regressed$DimPlot <- DimPlot(seurat, group.by = 'seurat_clusters', pt.size = UMAP_dot_size, label = F, raster = rasterize) +
     labs(title = 'MAIT cell Leiden clusters', 
        #   subtitle = 'Lung single-cell RNA-seq', 
        #   caption = 'n = 4 Abau, n = 5 Ftula, n = 4 CT',
     x = 'UMAP 1', y = 'UMAP 2')  
full_object_regressed$DimPlot <- full_object_regressed$DimPlot  |> LabelClusters(id = 'seurat_clusters', size = 3, color = 'black', repel = F, fontface = 'bold') 

# Clusters split by group plot
full_object_regressed$DimPlot_groups <- DimPlot(seurat, group.by = 'seurat_clusters', pt.size = UMAP_dot_size, label = F, split.by = 'Groups', raster = rasterize) +
     labs( x = 'UMAP 1', y = 'UMAP 2', title = NULL)   +
     theme(strip.text = element_text(size = 14, face = 'italic'))

# Cluster frequency plot
# seurat_local <- subset(seurat, subset = Groups != 'CT')
cell_counts <- FetchData(seurat, vars = c("seurat_clusters", "Samples", "Groups"))
cell_counts <- arrange(cell_counts, Samples)
counts <- cell_counts %>% 
    add_count(Groups, name='total_cell_count_by_group')  |>
    mutate(weights = Abau_cell_number/total_cell_count_by_group)  |>
    add_count(seurat_clusters, name='total_cell_count_by_cluster', wt =  weights)  |>
    count(seurat_clusters, Samples, Groups, total_cell_count_by_cluster , name='sample_cluster_count', wt = weights)  |> 
    mutate(frequency_within_cluster=sample_cluster_count*100/total_cell_count_by_cluster)  |> 
    mutate(Samples = as.character(Samples)) |> 
    arrange(Samples, desc(Samples)) 
frequency_table <- counts |> 
    arrange(Samples) |>     
    pivot_wider(id_cols = seurat_clusters,  names_from =  Samples , values_from = frequency_within_cluster, values_fill = 0)
# write.csv(frequency_table,file=here(tables_path, paste0( "frequency per roll }} per condition " , ".csv")), row.names=F) 

full_object_regressed$cluster_frequency <- counts |>
    ggplot(aes(x=seurat_clusters, fill=Groups, y = frequency_within_cluster)) +
    geom_col(aes(), position = 'fill') +
    # geom_bar(position=position_fill()) +
    theme_classic() +
    labs(x = 'Cluster', y = 'Frequency (%)', title = 'MAIT normalized cell cluster\nfrequency by sample')+
    scale_y_continuous(expand = expansion(mult = c(0, 0)), breaks = c(0, 0.25, 0.5, 0.75, 1), labels = c(0, 25, 50, 75, 100))+
    scale_fill_discrete(palette = group_colors) 

```

## Top genes cluster and feature plots

```{r}
seurat <- PrepSCTFindMarkers(seurat)
top_genes_per_cluster_results <- top_genes_per_cluster(seurat, n_genes_to_plot = 3, object_annotations = '', tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = 'ClusterProfiler', nterms_to_plot = 3 ) 

# Extracting a third gene for the last cluster
top_genes_per_cluster_results2 <- top_genes_per_cluster(seurat, n_genes_to_plot = 4, object_annotations = '', tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = F )

additional_gene1 <- top_genes_per_cluster_results2$topn |> 
    filter(cluster == '9') |> 
    pull(gene) |> 
    pluck(4)
additional_gene1

full_object_regressed$cluster_dotplot <- top_genes_per_cluster_results$plot 
ggsave(plot = full_object_regressed$cluster_dotplot, filename = here(figures_path,  paste0('DotPlot_Top3_per_cluster', '.pdf')), width = 8, height = 12)
top_genes_per_cluster_results$ClusterProfiler_results

full_object_regressed$cluster_heatmap <- top_genes_per_cluster_results$topn |> pull(gene) |> c(additional_gene1) |> unique() |> plot_pathways_heatmap(seurat, pathway_name = 'top_3_per_cluster', color_palette = diverging_palette_2, grouping_var = 'seurat_clusters')
full_object_regressed$cluster_signatures <- top_genes_per_cluster_results$ClusterProfiler_results


type_17_markers <- c('Rorc', 'Il17a', 'Sdc1', 'Il7r', 'Il1r1', 'Ccr6', 'Sdc4')
type_1_markers <- c('Tbx21', 'Ifng', 'I12rb2', 'Cxcr3', 'Gzmb', 'Gzmk', 'Klrg1', 'Klrd1', 'Ly6c2')
naive_markers <- c('S1pr1', 'Ccr7', 'Sell', 'Tcf7', 'Lef1')
regulatory_markers <- c('Arg2', 'Il10', 'Il21')
exhaustion_markers <- c('Ctla4', 'Pdcd1', 'Lag3', 'Tigit', 'Havcr2')
tissue_repair_markers <- c('Areg', 'Ramp3','Il18r1', 'Tgfb1', 'Tnf', 'Mmp25', 'Cxcl10', 'Rankl', 'Ifrd1')
transcription_factors <- c('Tbx21', 'Rorc', 'Gata3', 'Foxp3', 'Ahr', 'Maf', 'Prdm1', 'Bcl6', 'Zbtb16')
tissue_residency <- c('Itgae', 'Klrg1', 'Itga1', 'Itga2', 'Ccr9', 'Ccr10', 'Itga4', 'Itgb7')
activation_markers <- c('Cd69', 'Cd44', 'Nr4a1')
cytokines <- c('Ifng', 'Il17a', 'Il17f', 'Il4', 'Il10', 'Tgfb1', 'Tnf', 'Il2', 'Gzma', 'Gzmb', 'Nkg7', 'Ccl5', 'Il27ra', 'Ifgnr1')
costimulatory_receptors <- c('Ctla4', 'Lag3', 'Tigit', 'Havcr2', 'Cd28', 'Cd40lg', 'Pdcd1', 'Tnfsf4', 'Cd27', 'Tnfrsf9')

final_list <- c(type_17_markers, type_1_markers, naive_markers, regulatory_markers, exhaustion_markers, tissue_repair_markers, transcription_factors, tissue_residency, activation_markers, cytokines, costimulatory_receptors) |> unique() |> rev()

genes_to_plot <- c('Rorc', 'Il17a', 'Sdc1', 'Il7r', 'Il1r1', 'Ccr6', 'Tbx21', 'Ifng', 'Il12rb2', 'Cxcr3', 'Gzmb', 'Gzmk', 'Gzma', 'Gzmm', 'Klrg1', 'Klrd1', 'Ly6c2', 'Ramp3', 'Slamf7', 'Areg')
full_object_regressed$feature_plots <- genes_to_plot |> 
    set_names() |>
    map(feature_plot )

```

## Signature scores

```{r}
DefaultAssay(seurat) <- 'SCT'
signatures <- read_excel(here('data/Hagoulet_signatures.xlsx'), skip = 1)
signatures <- signatures |>
    dplyr::filter(!(row_number() == 1)) |>
    dplyr::select(-all_of(c('Signature', 'Tissue repair 1', 'Tissue repair 2', 'Tissue repair 3', 'Tissue repair 4')))
    
signatures <- map(.x = signatures, .f = ~ purrr::discard(.x, is.na)) |> as.list()
names(signatures) <- names(signatures) |> str_replace_all('_', ' ')


tissue_repair_linehan <- c('Angpt1', 'Angpt2', 'Angpt4', 'Areg', 'Bmp2', 'Bmp4', 'Bmp7', 'Btc', 'Ccl2', 'Ccl3', 'Chat', 'Csf1', 'Csf2', 'Ctgf', 'Cxcl1', 'Cxcl10', 'Cxcl12', 'Cxcl15', 'Cxcl2', 'Cyr61', 'Defb1', 'Defb6', 'Dhh', 'Disp1', 'Dll1', 'Dll3', 'Dll4', 'Egf', 'Epgn', 'Ereg', 'Fgf1', 'Fgf10', 'Fgf22', 'Fgf7', 'Flg', 'Flg2', 'Furin', 'Hbegf', 'Hgf', 'Hif1a', 'Hmgb1', 'Igf1', 'Igf2', 'Ihh', 'Il1a', 'Il1b', 'Il6', 'Inhba', 'Inhbb', 'Jag1', 'Jag2', 'Lep', 'Mmp10', 'Mmp13', 'Mmp15', 'Mmp2', 'Mmp25', 'Mmp28', 'Mmp3', 'Ngf', 'Nrg1', 'Nrg2', 'Nrg3', 'Nrg4', 'Pdgfa', 'Pdgfb', 'Pdgfc', 'Pgf', 'Ptges2', 'Shh', 'Tgfa', 'Tgfb1', 'Tgfb2', 'Tgfb3', 'Thbs1', 'Tnf', 'Vegfa', 'Vegfb', 'Vegfc', 'Vegfd', 'Wnt1', 'Wnt10a', 'Wnt10b', 'Wnt11', 'Wnt16', 'Wnt2', 'Wnt2b', 'Wnt3', 'Wnt3a', 'Wnt4', 'Wnt5a', 'Wnt5b', 'Wnt6', 'Wnt7a', 'Wnt7b', 'Wnt8a', 'Wnt8b', 'Wnt9a', 'Wnt9b')

signatures$`Tissue repair` <- tissue_repair_linehan

signatures$MAIT17 <- signatures$MAIT17[!(signatures$MAIT17 %in% c('1700040D17Rik'))]

# Removing genes that are not detected
blacklist_elements <- c('H2afz','Smsc2','Hist1h2ap','Flg','Flg2','Bmp2','Bmp4','Ngf','Fgf7','Fgf10','Fgf22','Igf1','Ctgf','Mmp3','Mmp13','Defb1','Defb6','Epgn','Mmp2','Angpt1','Hgf','Cyr61','Inhbb','Btc','Ereg','Lep','Vegfc','Vegfd','Pdgfc','Cxcl12','Shh','Dhh','Ihh','Dll3','Dll4','Wnt1','Wnt2','Wnt3a','Wnt5a','Wnt6','Wnt7a','Wnt7b','Wnt8a','Wnt8b','Wnt9a','Wnt9b','Wnt10a','Wnt16','Chat','Thbs1','Fam46a','Kcna4','Fam129b','Tdpoz4','Rassf6','Ero1l','Cd244','A430078G23Rik','D1Ertd622e','Fam49a')

signatures <- map(.x = signatures, .f = ~ .x[!(.x %in% blacklist_elements)])

seurat <- AddModuleScore_UCell(seurat, features = signatures, name = NULL)

# Scale signatures (z-score) for comparison and visualization purposes
seurat@meta.data  <-  seurat@meta.data |>
    mutate(across(names(signatures), ~ scale(.x, center = T, scale = T), .names = '{col}_scaled'))
 # Visualization
signature_heatmap <- function (signature) {
    plot1 <- signatures[[signature]] |> unique() |> plot_pathways_heatmap(seurat, pathway_name = signature, color_palette = diverging_palette_2, grouping_var =  'Samples' ) +
     labs(title = signature)
    ggsave(plot = plot1, filename = paste0('Heatmap_', signature, '_by_samples_', object_annotations, '.pdf'), path = figures_path, width = 8, height = 6)
    return(plot1)
}
signature_heatmap2 <- function (signature) {
    plot1 <- signatures[[signature]] |> unique() |> plot_pathways_heatmap2(seurat, pathway_name = signature, color_palette = diverging_palette_2, grouping_var =  'Samples' ) +
     labs(title = signature)
    ggsave(plot = plot1, filename = paste0('Heatmap_', signature, '_by_samples_', object_annotations, '.pdf'), path = figures_path, width = 8, height = 6)
    return(plot1)
}


```

```{r}

full_object_regressed$signature_violin_plots <- map(names(signatures), signature_violin_plot)
names(full_object_regressed$signature_violin_plots) <- names(signatures)

full_object_regressed$signature_dot_plots <- map(names(signatures), signature_dot_plot)
names(full_object_regressed$signature_dot_plots) <- names(signatures)

full_object_regressed$signature_feature_plots <- map(names(signatures), signature_feature_plot)
names(full_object_regressed$signature_feature_plots) <- names(signatures)

full_object_regressed$signature_heatmaps <- map(names(signatures), signature_heatmap)
names(full_object_regressed$signature_heatmaps) <- names(signatures)

full_object_regressed$signature_heatmaps2 <- map(names(signatures), signature_heatmap2)
names(full_object_regressed$signature_heatmaps2) <- names(signatures)

# full_object_regressed$signature_heatmaps2 |> walk(print)
```


## Milo

```{r}
library(miloR)
library(SingleCellExperiment)
library(scran)
library(scater)

seurat <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs_cell_cycle.rds'))

# Convert to SingleCellExperiment
Idents(seurat) <- 'Samples'
DefaultAssay(seurat) <- 'RNA'
seurat_sce <- as.SingleCellExperiment(seurat)
reducedDim(seurat_sce, 'PCA') <- Embeddings(seurat, 'pca')
reducedDim(seurat_sce, 'UMAP') <- Embeddings(seurat, 'umap')
seurat_milo <- Milo(seurat_sce)
k_used <- 30
dimensions <- 30

# Running MiloR
trajectory_milo <- buildGraph(seurat_milo, k = k_used, d = dimensions, reduced.dim = 'PCA')
trajectory_milo <- makeNhoods(trajectory_milo, prop = 0.1, k = k_used, d = dimensions, refined = T, reduced_dims = 'PCA')

# Count cells in neighborhoods per sample
trajectory_milo <- countCells(trajectory_milo, sample = 'Samples', meta.data = data.frame(colData(seurat_milo)))

# DA testing preparation
trajectory_design <- data.frame(colData(trajectory_milo))[, c('Samples', 'Groups')]
trajectory_design <- distinct(trajectory_design)
rownames(trajectory_design) <- trajectory_design$Samples
trajectory_design <- trajectory_design[colnames(nhoodCounts(trajectory_milo)), , drop = FALSE ]
trajectory_milo <- calcNhoodDistance(trajectory_milo, d = dimensions, reduced.dim = 'PCA')
# trajectory_milo <- buildNhoodGraph(trajectory_milo)


# Pairwise DA testing using contrasts Abau vs Ftula
contrast <- c('GroupsAbau - GroupsFtula')
da_results_Abau_vs_Ftula <- testNhoods(trajectory_milo, design = ~ 0 + Groups, design.df = trajectory_design, model.contrasts = contrast, fdr.weighting = 'graph-overlap', norm.method = 'TMM')
da_results_Abau_vs_Ftula |> arrange(SpatialFDR) |> head()

da_results_Abau_vs_Ftula <- annotateNhoods(trajectory_milo, da_results_Abau_vs_Ftula, coldata_col = 'seurat_clusters')
da_results_Abau_vs_Ftula <- da_results_Abau_vs_Ftula |> mutate(seurat_clusters = factor(seurat_clusters, levels = levels(seurat$seurat_clusters) |> rev()))

### Neighborhood differential gene expression analysis
da_results_Abau_vs_Ftula <- annotateNhoods(trajectory_milo, da_results_Abau_vs_Ftula, coldata_col = 'seurat_clusters')
trajectory_milo <- logNormCounts(trajectory_milo)

# # manual grouping of neighborhoods
# FCutoff <- 3
# da_results_Abau_vs_Ftula |> 
#     ggplot(aes(x = logFC)) +
#     geom_histogram(bins = 100) +
#     geom_density() +
#     theme_classic()
# da_results_Abau_vs_Ftula <- da_results_Abau_vs_Ftula |>
#     mutate(NhoodGroup = case_when(logFC > 0 & logFC < FCutoff & SpatialFDR < 0.05 ~ 'Abau_enriched_log2FC_0-5',
#                             logFC >= FCutoff  & SpatialFDR < 0.05 ~ 'Abau_enriched_log2FC_5',
#                             logFC < 0 & logFC > -FCutoff & SpatialFDR < 0.05 ~ 'Ftula_enriched_log2FC_0-5',
#                             logFC <= -FCutoff  & SpatialFDR < 0.05 ~ 'Ftula_enriched_log2FC_5',
#                             TRUE ~ 'Not_significant')) 
# # plotDAbeeswarm(da_results_Abau_vs_Ftula, group.by = 'NhoodGroup', alpha = 0.05)                            
# # plotNhoodGroups(trajectory_milo, da_results_Abau_vs_Ftula, layout = 'UMAP', alpha = 0.05)
# nhood_markers <- findNhoodGroupMarkers(trajectory_milo, da_results_Abau_vs_Ftula, aggregate.samples = F, sample_col = 'Samples')
# nhood_markers |> pivot_longer(cols = starts_with('adj'), names_to = 'neighborhood', values_to = 'adj_pval', names_prefix = 'adj.P.Val_')   |>
#     pivot_longer(cols = starts_with('log'), names_to = 'neighborhood2', values_to = 'logFC', names_prefix = 'logFC_')    |> filter(neighborhood == neighborhood2)
# nhood_markers <- nhood_markers |>
#     as_tibble() |>
#     pivot_longer(cols = starts_with('adj'), names_to = 'neighborhood', values_to = 'adj_pval', names_prefix = 'adj.P.Val_')  |>
#     pivot_longer(cols = starts_with('log'), names_to = 'neighborhood2', values_to = 'logFC', names_prefix = 'logFC_')   |>
#     filter(neighborhood == neighborhood2)     |>
#     select(-neighborhood2) |>
#     filter(adj_pval < 0.05 & logFC > 0.2) |> 
#     group_by(neighborhood) |>
#     slice_max(order_by = logFC, n = 25) |>
#     ungroup() |>
 #     arrange(neighborhood, desc(logFC))
# #Topn markers
# nhood_markers |> 
#     group_by(neighborhood)  |> 
#     arrange(desc(logFC)) |>
#     slice_head(n = 10) -> topn

trajectory_milo <- buildNhoodGraph(trajectory_milo)

da_results_Abau_vs_Ftula <- groupNhoods(trajectory_milo, da_results_Abau_vs_Ftula, max.lfc.delta = NULL, overlap = 2, da.fdr = 0.05, compute.new = T, original.behaviour = F )
plotDAbeeswarm(da_results_Abau_vs_Ftula, group.by = 'NhoodGroup', alpha = 0.05)
plotNhoodGroups(trajectory_milo, da_results_Abau_vs_Ftula, layout = 'UMAP', alpha = 0.05) 

# saveRDS(da_results_Abau_vs_Ftula, file = here(results_path, 'MiloR_da_results_Abau_vs_Ftula_neighborhood_groups.rds'))
# saveRDS(trajectory_milo, file = here(results_path, 'MiloR_trajectory_milo_neighborhood_groups.rds'))

da_results_Abau_vs_Ftula <- readRDS(file = here('data', 'MiloR_da_results_Abau_vs_Ftula_neighborhood_groups.rds'))
trajectory_milo <- readRDS(file = here('data', 'MiloR_trajectory_milo_neighborhood_groups.rds'))

nhood_markers <- findNhoodGroupMarkers(trajectory_milo, da_results_Abau_vs_Ftula, aggregate.samples = F)
nhood_markers |> pivot_longer(cols = starts_with('adj'), names_to = 'neighborhood', values_to = 'adj_pval', names_prefix = 'adj.P.Val_')   |>
    pivot_longer(cols = starts_with('log'), names_to = 'neighborhood2', values_to = 'logFC', names_prefix = 'logFC_')    |> filter(neighborhood == neighborhood2)
nhood_markers <- nhood_markers |>
    as_tibble() |>
    pivot_longer(cols = starts_with('adj'), names_to = 'neighborhood', values_to = 'adj_pval', names_prefix = 'adj.P.Val_')  |>
    pivot_longer(cols = starts_with('log'), names_to = 'neighborhood2', values_to = 'logFC', names_prefix = 'logFC_')   |>
    filter(neighborhood == neighborhood2)     |>
    select(-neighborhood2) |>
    filter(adj_pval < 0.05 & logFC > 0.2) |> 
    group_by(neighborhood) |>
    slice_max(order_by = logFC, n = 100) |> ## How many genes to include is setup here!!!
    ungroup() |>
    arrange(neighborhood, desc(logFC)) |>
    mutate(neighborhood = fct_inseq(neighborhood))
#Topn markers
nhood_markers |> 
    group_by(neighborhood)  |> 
    arrange(desc(logFC), .by_group = TRUE) |>
    slice_head(n = 4) -> topn


```

```{r}
# pathway_enrichment <- c('Metascape', 'ClusterProfiler')
# Run pathway enrichment analysis
local_path_pathway_enrichment <- here(results_path, paste0('MiloR', '_neighborhood_identification_functional_analysis'))
unlink(local_path_pathway_enrichment, recursive = T)
dir.create(local_path_pathway_enrichment)
    if (!is.null(pathway_enrichment)) {
        if ('Metascape' %in% pathway_enrichment) {
            plot_list <- list()
            for (neighborhood in levels(nhood_markers$neighborhood)) {
                local_path_2 <- here(local_path_pathway_enrichment, neighborhood)
                dir.create(local_path_2, recursive = TRUE, showWarnings = FALSE)
                name <- paste0('Neighborhood_', neighborhood)
                significant_results <- nhood_markers |> 
                    filter(neighborhood == {{neighborhood}}) |>
                    pull(GeneID) |>
                    unique()
                    if (length(significant_results) < 3) {
                    next
                        }
                plot_list[[neighborhood]] <- Metascape_overrepresentation_analysis(significant_results,
                            local_path = local_path_2,
                            group = neighborhood,
                            filename = name,
                            grouping_var = 'Neighborhood',
                            nterms_to_plot_metascape = 5)
                }
            plots <- wrap_plots(plot_list) & 
            theme(plot.title = element_text(size = 9, hjust = 0.5),
                axis.text.y = element_text(size = 8),
                axis.text.x = element_text(size = 8),
                axis.title.x  = element_text(size = 8),
            axis.title.y  = element_text(size = 7))
        print(plots)
            }
            colnames(nhood_markers)
if ('ClusterProfiler' %in% pathway_enrichment) {
            all_genes <- Features(seurat[['RNA']]) |>unique()
            gene_lists <- nhood_markers %>%
                group_by(neighborhood) %>%
                arrange(desc(logFC), .by_group = TRUE) |>
                slice_head(n = 100)  |>
                ungroup() |>
                select(neighborhood, GeneID)  |>
                group_by(neighborhood) |>
                mutate(id = row_number()) |>
                ungroup() |>
                pivot_wider(names_from = neighborhood, values_from = GeneID)  |>
                select(-id) |>
                as.list() |>
                map(~ .x[!is.na(.x)])
            overrepresentation_results <- GO_overrepresentation_analysis_multiple_lists(gene_lists, all_genes, local_path_pathway_enrichment, ontology = 'ALL', minGSSize = 5, maxGSSize = 400, filename = '',  drop_levels = F, levels_to_drop = c(1), simplify_function = min, simplify_by = 'p.adjust', simplify_terms = F, run_network = F, network_n_terms = 100, nterms_to_plot = 3, font_size = 10, grouping_var = 'Neighborhood')
  }  }
full_object_regressed$Milo_figures <- list()
full_object_regressed$Milo_figures$pathways <- overrepresentation_results$plot

```

```{r}
da_results_Abau_vs_Ftula <- da_results_Abau_vs_Ftula |>
    mutate(NhoodGroup = fct_inseq(NhoodGroup))
    # mutate(NhoodGroup = as.character(NhoodGroup))
da_results_Abau_vs_Ftula |> colnames()
full_object_regressed$Milo_figures$pathways_genes <-plotNhoodExpressionGroups(trajectory_milo, da_results_Abau_vs_Ftula,
                          features = topn$GeneID |> unique() |> rev(),
                          scale = T,
                        grid.space = 'fixed'
                          )
da_results_Abau_vs_Ftula <- da_results_Abau_vs_Ftula |>
    # mutate(NhoodGroup = fct_inseq(NhoodGroup))
    mutate(NhoodGroup = as.character(NhoodGroup))
full_object_regressed$Milo_figures$pathways_genes
    
```

### Milo figures

```{r}

#Visualize Neighborhoods
# full_object_regressed$Milo_figures <- list()
full_object_regressed$Milo_figures[['neighborhood_graph']] <- plotNhoodGraphDA(trajectory_milo, da_results_Abau_vs_Ftula, layout = 'UMAP', alpha = 0.1) + 
     scale_fill_continuous(palette = diverging_palette_2) +
     theme_classic() + 
     labs(title = 'MiloR: *A. bau* vs *F. tula*', fill = 'log2FC<br>(*A. bau*/*F. tula*)', size = 'Neighborhood size', x = 'UMAP 1', y = 'UMAP 2')+ 
     theme(legend.title = element_text()) +
    guides(edge_width = 'none', size = 'none')
ggsave(plot = full_object_regressed$Milo_figures[['neighborhood_graph']], filename = 'MiloR_Neighborhoods_DA_Abau_vs_Ftula.pdf', path = figures_path, width = 5, height = 5)
full_object_regressed$Milo_figures[['neighborhood_graph']]

full_object_regressed$Milo_figures[['beeswarm_graph']] <- plotDAbeeswarm(da_results_Abau_vs_Ftula, group.by = 'seurat_clusters', alpha = 0.05)
da.res <- full_object_regressed$Milo_figures[['beeswarm_graph']]@data
n_groups <- length(levels(da.res$group_by))
full_object_regressed$Milo_figures[['beeswarm_graph']] <- da.res |> 
    ggplot(aes(pos_x, pos_y, color=logFC_color)) +
    scale_color_gradient2() +
    guides(color="none") +
    xlab('seurat_clusters') + ylab("Log Fold Change") +
    scale_x_continuous(
      breaks = seq(1,n_groups),
      labels = setNames(levels(da.res$group_by), seq(1,n_groups))
      ) +
    geom_point(size = .5) +
    coord_flip() +
    theme_bw(base_size=22) +
    theme(strip.text.y =  element_text(angle=0)) +
    scale_color_continuous(palette = diverging_palette_2) +
    labs(x = 'Leiden clusters', y = 'log2FC(*A. bau*/*F. tula*)', color = 'log2FC\n(*A. bau*/*F. tula*)') +
    theme(strip.text.y =  element_text(angle=0), axis.title.x = element_text())
ggsave(plot = full_object_regressed$Milo_figures[['beeswarm_graph']], filename = 'MiloR_Beeswarm_DA_Abau_vs_Ftula.pdf', path = figures_path, width = 7, height = 5)

#Visualize Neighborhoods
full_object_regressed$Milo_figures[['neighborhood_cluster_graph']] <- plotNhoodGroups(trajectory_milo, da_results_Abau_vs_Ftula, layout = 'UMAP', alpha = 0.1) + 
    #  scale_fill_continuous(palette = diverging_palette_2) +
     theme_classic() + 
     labs(title = 'MiloR: *A. bau* vs *F. tula*', size = 'Neighborhood size', x = 'UMAP 1', y = 'UMAP 2')+ 
    guides(edge_width = 'none', size = 'none')
ggsave(plot = full_object_regressed$Milo_figures[['neighborhood_cluster_graph']], filename = 'MiloR_Neighborhoods_Clusters_DA_Abau_vs_Ftula.pdf', path = figures_path, width = 5, height = 5)

da_results_Abau_vs_Ftula <- da_results_Abau_vs_Ftula |>
    mutate(NhoodGroup = fct_inseq(NhoodGroup))
full_object_regressed$Milo_figures[['beeswarm_graph_neighborhood_cluster']] <- plotDAbeeswarm(da_results_Abau_vs_Ftula, group.by = 'NhoodGroup',  alpha = 0.05)
da.res <- full_object_regressed$Milo_figures[['beeswarm_graph_neighborhood_cluster']]@data
n_groups <- length(levels(da.res$group_by))
full_object_regressed$Milo_figures[['beeswarm_graph_neighborhood_cluster']] <- da.res |> 
    ggplot(aes(pos_x, pos_y, color=logFC_color)) +
    scale_color_gradient2() +
    guides(color="none") +
    xlab('seurat_clusters') + ylab("Log Fold Change") +
    scale_x_continuous(
      breaks = seq(1,n_groups),
      labels = setNames(levels(da.res$group_by), seq(1,n_groups))
      ) +
    geom_point(size = .5) +
    coord_flip() +
    theme_bw(base_size=22) +
    scale_color_continuous(palette = diverging_palette_2) +
    labs(x = 'Neighborhood clusters', y = 'log2FC(*A. bau*/*F. tula*)', color = 'log2FC\n(*A. bau*/*F. tula*)') +
    theme(strip.text.y =  element_text(angle=0), axis.title.x = element_text()) 
ggsave(plot = full_object_regressed$Milo_figures[['beeswarm_graph_neighborhood_cluster']], filename = 'MiloR_Beeswarm_NhoodClusters_DA_Abau_vs_Ftula.pdf', path = figures_path, width = 7, height = 5)


# gene_list_plot <- topn |> pull(GeneID) |> unique() |> rev()
# Idents(seurat) <- 'seurat_clusters'
# DotPlot_scCustom(seurat,
#                 features = gene_list_plot,
#                 colors_use=sequential_palette_dotplot,
#                 flip_axes = T,
#                 dot.scale = 8,
#                 dot.min = 0,
#                 scale.min = 0,
#                 scale.max = 80,
#                 x_lab_rotate = T,
#                 y_lab_rotate = F) +
#     theme(axis.text.x = element_text(size = 14),
#         axis.text.y = element_text(size = 14),
#         legend.title = element_text(size = 18))

# local_path <- here(results_path, pcaste0('MiloR', '_neighborhood_identification_functional_analysis'))
# unlink(local_path, recursive = T)
# dir.create(local_path)
# for (neighborhood in unique(nhood_markers$neighborhood)) {
#      name <- paste0('Neighborhood_', neighborhood)
#      significant_results <- nhood_markers |> 
#           filter(neighborhood == {{neighborhood}}) |>
#           pull(GeneID) |>
#           unique()
#         if (length(significant_results) < 3) {
#           next
#               }
#   Metascape_overrepresentation_analysis(significant_genes_FC_ordered = significant_results, path = local_path, group = '', filename =  name,  grouping_var =  neighborhood )
#   }

full_object_regressed$Milo_figures |> walk(print)


```


## Differential gene expression analysis

```{r}
#| fig-width: 7
#| #| fig-height: 7
#| layout-ncol: 2
#| results: hide
seurat <- seurat_original
source(here('scripts/metascape_functions.r'))
# source(here('scripts/gProfiler2_functions.r'))
source(here('scripts/function_template.r'))
source(here('scripts/ClusterProfiler_functions.r'))

path <- here(results_path, 'DEG_pseudobulk_analysis_Abau_vs_Ftula')
unlink(path, recursive = T)
dir.create(path)
Idents(seurat) <- 'seurat_clusters'

seurat@meta.data <- seurat@meta.data |>
    mutate(Samples = str_remove(Samples, '_Mouse'))
# levels(seurat$Samples)
     
genes_to_label <- c('Sdc1', 'Zbtb7c', 'Acvr2a', 'Mmp25', 'Il7r', 'Cd164', 'Nr1d1', 'Ccr6', 'Hlf', 'Rorc', 'Ramp3', 'Mgat', 'Il1r1', 'Itgae', 'Tbx21', 'Gzma', 'Klrg1', 'Ifng', 'Gzmk', 'Lag3', 'Il1rl1', 'Plac8', 'Ifitm3', 'Batf3', 'Irf8', 'Bcat1', 'Mgat5', 'Palld', 'Ccr6', 'Itgae', 'Il17re', 'Ablim3', 'Igf1r', 'Irf8', 'Zbtb32', 'Bcat1', 'Tnfrsf8', 'Ppa1', 'Mif') |> unique() 
genes_not_to_label <- c('Bcat1', 'Lag3', 'Plac8', 'IL1rl1', 'Ifitm3', 'Batf3', 'Il1r1', 'Nr1d1', 'Cd164', 'Acvr2a', 'Mmp25', 'Zbtb7c', 'Hif', 'Ablim3', 'Palld', 'Mif', 'Zbtb32', 'Tnfrsf8')
genes_to_label <- genes_to_label[!(genes_to_label %in% genes_not_to_label)]
# genes_to_label <- list('Genes_of_interest' = genes_to_label)
# scRNAseq, comparison, group1, group2, cluster = 'all_clusters', 
#                        path = './', FC_threshold = 0.3, p_value_threshold = 0.05, genes_to_plot = NULL, 
#                        expression_threshold_for_gene_list = 20, minimum_cell_number = 10, 
#                        run_pathway_enrichment = NULL, genes_to_exclude = c(), ...

output <- pseudobulk(seurat, comparison='Samples', group1='F. tula', group2='A. bau', cluster= 'all_MAITs' , path=path, label_threshold = 1000, distance_from_diagonal_threshold=0.5, max_overlaps = 15, my_colors =c('#d55e10','#0072b2', 'grey'), genes_to_plot = NULL, FC_threshold = 1, p_value_threshold = 0.05, run_pathway_enrichment = pathway_enrichment, minGSSize = 4, maxGSSize = , drop_levels = F, levels_to_drop = c(1), ontology = 'BP', simplify_terms=T, run_network=T, network_n_terms=70)


source(here('scripts/function_template.r'))
full_object_original$volcano_plot <- volcano_plot(output$results, comparison='Samples', group1='F. tula', group2='A. bau', cluster= 'all_MAITs' , local_figures_path=path, label_size = 3, label_threshold = 1000, distance_from_diagonal_threshold=0.5, max_overlaps = 4, my_colors =c('#d55e10','#0072b2', 'grey'), genes_to_plot = genes_to_label, FC_threshold = 1, p_value_threshold = 0.05, run_pathway_enrichment = NULL, minGSSize = 4, maxGSSize = , drop_levels = F, levels_to_drop = c(1), ontology = 'BP', simplify_terms=T, run_network=T, network_n_terms=70, test_type = 'Pseudobulk', pt_size = 1) 
# full_object_original$volcano_plot

full_object_original$scatter_plot <- scatterplot(output$results, comparison='Samples', group1='F. tula', group2='A. bau', cluster= 'all_MAITs' , local_figures_path=path, label_size = 3, label_threshold = 1000, distance_from_diagonal_threshold=0.5, max_overlaps = 4, my_colors =c('#d55e10','#0072b2', 'grey'), genes_to_plot = genes_to_label, FC_threshold = 1, p_value_threshold = 0.05, run_pathway_enrichment = NULL, minGSSize = 4, maxGSSize = , drop_levels = F, levels_to_drop = c(1), ontology = 'BP', simplify_terms=T, run_network=T, network_n_terms=70, test_type = 'Pseudobulk', pt_size = 1)
# full_object_original$scatter_plot

results <- output$results
GO_functional_analysis(results,  grouping_var = 'all_MAITs', path= path, FC_threshold = 1, p_value_threshold = 0.05, group1 = 'Ftula', group2 = 'Abau', minGSSize = 4, maxGSSize = 400, drop_levels = F, levels_to_drop = c(), ontology = 'BP', simplify_terms=T, run_network=T, network_n_terms=200)

# UP in Abau all MAITs
GO_results <- read_csv(here('results/Manuscript_figures/DEG_pseudobulk_analysis_Abau_vs_Ftula/GO_functional_analysis_UP_in_Abau_all_MAITs/GO_OverRepresentation_analysis_results_filtered_BP.csv')) |> select(-...1) |> mutate(GeneRatio = as.numeric(str_split_fixed(GeneRatio, '/', 2)[,1]) / as.numeric(str_split_fixed(GeneRatio, '/', 2)[,2]))
pathways_to_plot <- GO_results |>
    arrange(desc(GeneRatio)) |>
    slice_head(n = 10) |> 
    pull(Description)
pathways_to_replace <- c('regulation of hemopoiesis' = 'alpha-beta T cell differentiation')
pathways_to_plot <- pathways_to_plot |> recode(!!!pathways_to_replace)
pathways_to_plot
full_object_original$pathways_Abau <-  GO_results |>
    filter(Description %in% pathways_to_plot) |>
    ggplot(aes(fct_reorder(Description, -log10(p.adjust)), -log10(p.adjust) , fill = -log10(p.adjust))) +
    geom_col(width = 0.5) +
    scale_fill_binned(palette = hcl.colors(n = 8, 'YlOrRd', rev = T)[3:8], breaks = c(2, 5, 10, 20, 30)) +
    labs(title = 'Representative top enriched<br>GO terms in *A. bau* MAITs', x = NULL, y = '-log10(Adjusted p-value)') +
    guides(fill = 'none') +
    theme_classic() +
    scale_y_continuous(expand = expansion(mult = c(0.01, 0.1)))+ 
    # scale_x_discrete(labels = label_wrap(22))+
    coord_flip() 
full_object_original$pathways_Abau

t_cell_activation_terms <- c('alpha-beta T cell differentiation', 'positive regulation of cell adhesion')
wound_repair_terms <- c('response to wounding', 'morphogenesis of an epithelium')

# t_cell_activation_terms
source(here('scripts/function_template.r'))
genes_to_plot <- GO_results |>
    filter(Description %in% t_cell_activation_terms) |>
    pull(geneID) |>
    str_split(pattern = '/') |>
    unlist() |>
    unique() |>
    head(20) 
full_object_original$pathways_Abau_Tcelldifferentiation <- plot_pathways_heatmap(genes_to_plot, seurat |> subset(subset = Groups %in% c('Uninfected'), invert = T), 'T cell differentiation terms', color_palette = diverging_palette_2)
full_object_original$pathways_Abau_Tcelldifferentiation

# wound_repair_terms
bgenes_to_plot <- GO_results |>
    filter(Description %in% wound_repair_terms) |>
    pull(geneID) |>
    str_split(pattern = '/') |>
    unlist() |>
    unique() |>
    head(19)  |>
    c('Areg')
full_object_original$pathways_Abau_woundrepair <- plot_pathways_heatmap(genes_to_plot, seurat |> subset(subset = Groups %in% c('Uninfected'), invert = T), 'Wound repair terms', color_palette = diverging_palette_2)
full_object_original$pathways_Abau_woundrepair


# UP in Ftula all MAITs
GO_results <- read_csv(here('results/Manuscript_figures/DEG_pseudobulk_analysis_Abau_vs_Ftula/GO_functional_analysis_UP_in_Ftula_all_MAITs/GO_OverRepresentation_analysis_results_filtered_BP.csv')) |> select(-...1) |> mutate(GeneRatio = as.numeric(str_split_fixed(GeneRatio, '/', 2)[,1]) / as.numeric(str_split_fixed(GeneRatio, '/', 2)[,2]))

pathways_to_plot <- GO_results |>
    arrange(desc(GeneRatio)) |>
    slice_head(n = 20) |> 
    pull(Description)
pathways_to_remove <- c('nuclear chromosome segregation', 'nuclear division', 'DNA recombination', 'meiotic cell cycle process', 'mononuclear cell proliferation', 'defense response to virus', 'positive regulation of response to biotic stimulus', 'regulation of chromosome organization', 'cell cycle G1/S phase transition')
pathways_to_plot <- setdiff(pathways_to_plot, pathways_to_remove) |> c('cell killing')
full_object_original$pathways_Ftula <-  GO_results |>
    filter(Description %in% pathways_to_plot) |>
    ggplot(aes(fct_reorder(Description,  -log10(p.adjust) ),  -log10(p.adjust) , fill = -log10(p.adjust))) +
    geom_col(width = 0.5) +
    scale_fill_binned(palette = hcl.colors(n = 8, 'YlOrRd', rev = T)[3:8], breaks = c(2, 5, 10, 20, 30)) +
    labs(title = 'Representative top enriched<br>GO terms in *F. tula* MAITs', x = NULL, y = '-log10(Adjusted p-value)') +
    guides(fill = 'none') +
    theme_classic() +
    scale_y_continuous(expand = expansion(mult = c(0.01, 0.1)))+ 
    # scale_x_discrete(labels = label_wrap(22))+
    coord_flip() 
full_object_original$pathways_Ftula

cytotoxicity_terms <- c('cell killing', 'lymphocyte mediated immunity', 'response to virus')

#cytotoxicity_terms
genes_to_plot <- GO_results |>
    filter(Description %in% cytotoxicity_terms) |>
    pull(geneID) |>
    str_split(pattern = '/') |>
    unlist() |>
    unique() |>
    head(20) 
full_object_original$pathways_Ftula_cytotoxicity <- plot_pathways_heatmap(genes_to_plot,  seurat |> subset(subset = Groups %in% c('Uninfected'), invert = T), 'Cytotoxicity / Response to virus terms', color_palette = diverging_palette_2) 
full_object_original$pathways_Ftula_cytotoxicity

```

# Generate plots for Ftula object

```{r}
seurat <- seurat_Ftula
Idents(seurat) <- 'seurat_clusters'

# Clusters UMAP plot
Ftula_object$DimPlot <- DimPlot(seurat, group.by = 'seurat_clusters', pt.size = UMAP_dot_size, label = F, raster = rasterize) +
     labs(title = 'MAIT cell Leiden clusters', 
        #   subtitle = 'Lung single-cell RNA-seq', 
        #   caption = 'n = 4 Abau, n = 5 Ftula, n = 4 CT',
     x = 'UMAP 1', y = 'UMAP 2')  
Ftula_object$DimPlot <- Ftula_object$DimPlot  |> LabelClusters(id = 'seurat_clusters', size = 3, color = 'black', repel = F, fontface = 'bold')

seurat <- PrepSCTFindMarkers(seurat)
top_genes_per_cluster_results <- top_genes_per_cluster(seurat, n_genes_to_plot = 3, object_annotations = '', tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = F, nterms_to_plot = 3 ) 
Ftula_object$cluster_dotplot <- top_genes_per_cluster_results$plot 
ggsave(plot = Ftula_object$cluster_dotplot, filename = here(figures_path,  paste0('DotPlot_Top3_per_cluster', '.pdf')), width = 8, height = 12)
top_genes_per_cluster_results$ClusterProfiler_results

Ftula_object$cluster_heatmap <- top_genes_per_cluster_results$topn |> pull(gene) |> unique() |> plot_pathways_heatmap(seurat, pathway_name = 'top_3_per_cluster', color_palette = diverging_palette_2, grouping_var = 'seurat_clusters')
Ftula_object$cluster_signatures <- top_genes_per_cluster_results$ClusterProfiler_results

genes_to_plot <- c('Rorc', 'Il17a', 'Sdc1', 'Il7r', 'Il1r1', 'Ccr6', 'Tbx21', 'Ifng', 'Il12rb2', 'Cxcr3', 'Gzmb', 'Gzmk', 'Gzma', 'Gzmm', 'Klrg1', 'Klrd1', 'Ly6c2', 'Ramp3', 'Slamf7', 'Areg', 'Mki67')
Ftula_object$feature_plots <- genes_to_plot |> 
    set_names() |>
    map(feature_plot )

Ftula_object$feature_plots$Tbx21_density <-  Plot_Density_Custom(seurat_object = seurat, features = c("Tbx21"), custom_palette = sequential_palette)
Ftula_object$feature_plots$Rorc_density <- Plot_Density_Custom(seurat_object = seurat, features = c("Rorc"), custom_palette = sequential_palette)
Ftula_object$feature_plots$Rorc_Tbx21_density <- Plot_Density_Joint_Only(seurat_object = seurat, features = c("Tbx21", "Rorc"), custom_palette = sequential_palette)

```

# Generate plots for Abau object

```{r}
seurat <- seurat_Abau
Idents(seurat) <- 'seurat_clusters'

# Clusters UMAP plot
Abau_object$DimPlot <- DimPlot(seurat, group.by = 'seurat_clusters', pt.size = UMAP_dot_size, label = F, raster = rasterize) +
     labs(title = 'MAIT cell Leiden clusters', 
        #   subtitle = 'Lung single-cell RNA-seq', 
        #   caption = 'n = 4 Abau, n = 5 Ftula, n = 4 CT',
     x = 'UMAP 1', y = 'UMAP 2')  
Abau_object$DimPlot <- Abau_object$DimPlot  |> LabelClusters(id = 'seurat_clusters', size = 3, color = 'black', repel = F, fontface = 'bold')

seurat <- PrepSCTFindMarkers(seurat)
top_genes_per_cluster_results <- top_genes_per_cluster(seurat, n_genes_to_plot = 3, object_annotations = '', tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = FALSE, nterms_to_plot = 3 ) 
Abau_object$cluster_dotplot <- top_genes_per_cluster_results$plot 
ggsave(plot = Abau_object$cluster_dotplot, filename = here(figures_path,  paste0('DotPlot_Top3_per_cluster', '.pdf')), width = 8, height = 12)
top_genes_per_cluster_results$ClusterProfiler_results

Abau_object$cluster_heatmap <- top_genes_per_cluster_results$topn |> pull(gene) |> unique() |> plot_pathways_heatmap(seurat, pathway_name = 'top_3_per_cluster', color_palette = diverging_palette_2, grouping_var = 'seurat_clusters')
Abau_object$cluster_signatures <- top_genes_per_cluster_results$ClusterProfiler_results
Abau_object$cluster_heatmap

genes_to_plot <- c('Rorc', 'Il17a', 'Sdc1', 'Il7r', 'Il1r1', 'Ccr6', 'Tbx21', 'Ifng', 'Il12rb2', 'Cxcr3', 'Gzmb', 'Gzmk', 'Gzma', 'Gzmm', 'Klrg1', 'Klrd1', 'Ly6c2', 'Ramp3', 'Slamf7', 'Areg', 'Mki67')
Abau_object$feature_plots <- genes_to_plot |> 
    set_names() |>
    map(feature_plot )

Abau_object$feature_plots$Tbx21_density <-  Plot_Density_Custom(seurat_object = seurat, features = c("Tbx21"), custom_palette = sequential_palette)
Abau_object$feature_plots$Rorc_density <- Plot_Density_Custom(seurat_object = seurat, features = c("Rorc"), custom_palette = sequential_palette)
Abau_object$feature_plots$Rorc_Tbx21_density <- Plot_Density_Joint_Only(seurat_object = seurat, features = c("Tbx21", "Rorc"), custom_palette = sequential_palette)

```

# cNMF results

```{r}
seurat <- seurat_cell_cycle
```

```{r}
#| fig-width: 10
#| fig-height: 10
library(Matrix)
source(here('scripts/function_template.r'))
local_path <- here('results', 'cNMF')
dir.create(local_path)
local_path_input <- here('results', 'cNMF','input')
dir.create(local_path_input)


filtered_dir  <- local_path_input
data_dir <- local_path
runname = "cNMF_run"
cnmf_results <- run_cnmf_results(
    seurat,                    # Seurat object
    data_dir,                  # path to directory containing cNMF run folder (string)
    runname,                   # cNMF run name (string)
    k_used = 15,
    local_density_threshold = 0.1,
    output_path = here::here('results', 'cNMF', runname, 'outputs'),
    do_cnmf = TRUE,            # whether to run system cnmf command
    sequential_palette = sequential_palette_dotplot, # palette for FeaturePlot_scCustom
    sequential_palette_dotplot = sequential_palette_dotplot, # palette for DotPlot_scCustom
    color_palette = diverging_palette_2,
    object_annotations = '',
    top_n = 75,                # top genes per program to extract
    topn_plot = 75,             # top genes used in # blog
    run_pathway_enrichment = NULL       # run Metascape_overrepresentation_analysis
)

#  k_to_test <- c(6, 9, 15, 18)
#  walk(k_to_test, function(k){
#      cnmf_results_k <- run_cnmf_results(
#          seurat,                    # Seurat object
#          data_dir,                  # path to directory containing cNMF run folder (string)
#          runname,                   # cNMF run name (string)
#          k_used = k,
#          local_density_threshold = 0.1,
#          output_path = here::here('results', 'cNMF', paste0(runname, '_k', k), 'outputs'),
#          do_cnmf = TRUE,            # whether to run system cnmf command
#          sequential_palette = sequential_palette, # palette for FeaturePlot_scCustom
#          sequential_palette_dotplot = sequential_palette_dotplot, # palette for DotPlot_scCustom
#          color_palette = diverging_palette_2,
#          object_annotations = '',
#          top_n = 75,                # top genes per program to extract
#          [] = 5,             # top genes used in dotplot
#          run_pathway_enrichment = pathway_enrichment       # run Metascape_overrepresentation_analysis
#  )})

```

```{r}
k_used <- 15
dt_str <- '0_1'
runname = "cNMF_run"
data_dir <- here('results', 'cNMF')

seurat <- cnmf_results$seurat
GEP_to_plot_numeric <- seq(1,12) |> as.character()
GEP_to_plot <- str_c(c('cNMF_program_'), seq(1,k_used))
GEP_annotations <- c(
    'GEP 1:<br>Actin filament<br>organization' = 'cNMF_program_1',
    'GEP 2:<br>T cell activation' = 'cNMF_program_2',
    'GEP 3:<br>Oxidative phosphorylation' = 'cNMF_program_3',
    'GEP 4:<br>T cell differentiation' = 'cNMF_program_4',
    'GEP 5:<br>Stress response' = 'cNMF_program_5',
    'GEP 6:<br>Translation and respiration' = 'cNMF_program_6',
    'GEP 7:<br>Translation 1' = 'cNMF_program_7',
    'GEP 8:<br>Cell migration' = 'cNMF_program_8',
    'GEP 9:<br>Translation 2' = 'cNMF_program_9',
    'GEP 10:<br>Cytotoxicity' = 'cNMF_program_10',
    'GEP 11:<br>Cell division' = 'cNMF_program_11',
    'GEP 12:<br>Interferon signaling' = 'cNMF_program_12'
)

GEP_to_plot <- c('GEP 1:<br>Actin filament<br>organization', 
                 'GEP 2:<br>T cell activation', 
                 'GEP 3:<br>Oxidative phosphorylation', 
                 'GEP 4:<br>T cell differentiation', 
                 'GEP 5:<br>Stress response', 
                 'GEP 6:<br>Translation and respiration',
                 'GEP 7:<br>Translation 1',
                 'GEP 8:<br>Cell migration',
                 'GEP 9:<br>Translation 2',
                 'GEP 10:<br>Cytotoxicity',
                 'GEP 11:<br>Cell division',
                 'GEP 12:<br>Interferon signaling'                
                 )

GEP_to_plot_heatmap <- c('GEP 1: Actin filament organization', 
                 'GEP 2: T cell activation', 
                 'GEP 3: Oxidative phosphorylation', 
                 'GEP 4: T cell differentiation', 
                 'GEP 5: Stress response', 
                 'GEP 6: Translation and respiration',
                 'GEP 7: Translation 1',
                 'GEP 8: Cell migration',
                 'GEP 9: Translation 2',
                 'GEP 10: Cytotoxicity',
                 'GEP 11: Cell division',
                 'GEP 12: Interferon signaling'                
                 )


seurat@meta.data <- seurat@meta.data |>
    rename(GEP_annotations)

full_object_regressed$cNMF_GEP_violin_plots <- map(GEP_to_plot, signature_violin_plot)
names(full_object_regressed$cNMF_GEP_violin_plots) <- GEP_to_plot

# Feature Plots
feature_plot_GEP <- function (gene) {
    plot1 <- FeaturePlot(seurat, features = gene, order = T, pt.size = UMAP_dot_size, alpha = 1, min.cutoff = 0, max.cutoff = 1) + labs(title = gene, x = 'UMAP 1', y = 'UMAP 2') +
        # labs(title = str_replace(gene, '\\n', '<br>')) +
        theme(plot.title = element_text(), legend.title = element_text() ) +
        scale_color_gradientn(colors = sequential_palette_dotplot, name = 'GEP<br>usage', breaks = c(0, 0.5, 1), limits = c(0, 1)) +
        theme(legend.position = 'right')
    ggsave(plot = plot1, filename = paste0('FeaturePlot', gene, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
    return(plot1)
}
full_object_regressed$cNMF_GEP_feature_plots <- map(GEP_to_plot, feature_plot_GEP)
names(full_object_regressed$cNMF_GEP_feature_plots) <- GEP_to_plot
full_object_regressed$cNMF_GEP_feature_plots

all_genes <- Features(seurat[['RNA']]) |>unique()
gene_list <- cnmf_results$top_colnames |> select(1:12) |> slice_head(n = 75) |> as.list() |>  map(~ .x[!is.na(.x)])
full_object_regressed$cNMF_GEP_pathways <- GO_overrepresentation_analysis_multiple_lists(gene_list, all_genes, local_path_pathway_enrichment, ontology = 'ALL', minGSSize = 5, maxGSSize = 400, filename = '',  drop_levels = F, levels_to_drop = c(), simplify_function = min, simplify_by = 'p.adjust', simplify_terms = F, run_network = F, network_n_terms = 100, nterms_to_plot = 4, font_size = 8) 
full_object_regressed$cNMF_GEP_pathways$plot
# full_object_regressed$cNMF_GEP_pathways$plot  <- full_object_regressed$cNMF_GEP_pathways$plot + labs(title = 'Pathway over-representation analysis \ntop 75 genes per program') + scale_fill_continuous(palette = sequential_palette, trans = 'reverse')


full_object_regressed$cNMF_GEP_pathways <- full_object_regressed$cNMF_GEP_pathways + labs(title = 'Pathway over-representation analysis \ntop 75 genes per program') 
full_object_regressed$cNMF_GEP_pathways

cnmf_results$top_colnames_long |> filter(gene_expression_program == '1') |> slice_head(n = 10)


# Heatmap of top genes per program
# 8.2) Extracting the top genes per program for heatmap plotting
spectra_score <- cnmf_results$spectra_score
genes_score <- spectra_score |>
    as.matrix() |> 
    t() |> 
    as.data.frame() |>
    rownames_to_column('gene')  |> 
    pivot_longer(-gene, names_to = 'GEP', values_to = 'loading')  |>
    filter( GEP %in% GEP_to_plot_numeric) |>
    mutate(GEP = fct_inseq(GEP)) |>
    group_by(GEP) |>
    # Z-score normalize the loadings within each GEP
    mutate(loading_zscore = loading) |>
    arrange(desc(loading_zscore), .by_group = TRUE) #|>
    # slice_max(order_by = loading_zscore, n = topn_plot, with_ties = FALSE)
genes_to_plot <- genes_score |>
    group_by(GEP) |> 
    slice_max(order_by = loading_zscore, n = 4, with_ties = FALSE) |> 
    ungroup() |> 
    distinct(gene) |> 
    pull(gene)
genes_score <- genes_score  |> 
    filter(gene %in% genes_to_plot)  |>
    mutate(gene = factor(gene, levels = genes_to_plot))   |>
    mutate(GEP = fct_rev(GEP)) 
print(genes_score |> head(8))
    # Calculate symmetric limits for the color scale based on z-scores
max_abs <- max(abs(genes_score$loading_zscore), na.rm = TRUE)

cnmf_results$heatmap_plot <- ggplot(genes_score, aes(x = gene, y = GEP, fill = loading_zscore)) +
    geom_tile(color = "grey60", linewidth = 0.3) +
    scale_fill_gradientn(
        colors = diverging_palette_2, 
        name = "z-score\nloading",
        limits = c(-max_abs, max_abs)
    ) +
    scale_y_discrete(labels = GEP_to_plot_heatmap |> rev()) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1, face = "italic"),
        axis.text.y = element_text(),
        plot.title = element_text(hjust = 0.5, size = 14)
    ) +
    labs(x = NULL, y = NULL, title = 'Top genes per\nGene Expression Program')
    print(cnmf_results$heatmap_plot)

# HEAT Map of GEP  across cells
seurat[['cNMF_programs']] <- CreateAssayObject(data = t(x = FetchData(object = seurat, vars = GEP_to_plot)))
seurat[['cNMF_programs']] |> rownames()
cnmf_results$heatmap_plot_cells <- DoHeatmap(object = seurat, features = GEP_to_plot |> str_replace_all('_', '-'), assay = 'cNMF_programs', slot = 'data') &
    scale_fill_gradientn(colors = sequential_palette, name = 'GEP\nloading') &
    theme(axis.text.y = element_text(size = 8))

```


# TCR repertoire analysis

## General repertoire analysis

```{r}
#| layout-ncol: 2
combined <- scRepertoire:::.expression2List(seurat, split.by ='Samples')
combined <- combined[c(10, 11, 12, 13, 1, 2, 3, 4, 5, 6, 7, 8, 9)]
combined <- addVariable(combined, variable.name = 'Groups', variables = c('Uninfected', 'Uninfected', 'Uninfected', 'Uninfected', 'A. bau', 'A. bau', 'A. bau', 'A. bau', 'F. tula', 'F. tula', 'F. tula', 'F. tula', 'F. tula'))

# Comparing clonal abundance between Abau and Ftula
VDJ_figures$clonalAbundance <-  clonalAbundance(seurat, group.by = 'Groups', scale = T) +
    labs(title = 'Clonal abundance by sample') +
    theme(plot.title = element_text(hjust = 0.5, size = 16))
ggsave(plot = VDJ_figures$clonalAbundance, filename = here(figures_path,  paste0('Clonal_Abundance_by_Group', '.pdf')), width = 7, height = 5)

# Comparing CDR3 length between Abau and Ftula
VDJ_figures$clonalLength_TRA <- clonalLength(seurat, group.by = 'Groups', scale = F, chain = 'TRA') +
    scale_fill_discrete(group_colors) +
    labs(title = 'Clonal length TRA by Group') +
    theme(plot.title = element_text(hjust = 0.5, size = 16))
ggsave(plot = VDJ_figures$clonalLength_TRA, filename = here(figures_path,  paste0('Clonal_Length_TRA_by_Group', '.pdf')), width = 7, height = 5)

VDJ_figures$clonalLength_TRB <- clonalLength(seurat, group.by = 'Groups', scale = F, chain = 'TRB') +
    scale_fill_discrete(group_colors) +
    labs(title = 'Clonal length TRB by Group') +
    theme(plot.title = element_text(hjust = 0.5, size = 16))
ggsave(plot = VDJ_figures$clonalLength_TRB, filename = here(figures_path,  paste0('Clonal_Length_TRB_by_Group', '.pdf')), width = 7, height = 5)

# Comparing clonal space between Abau and Ftula
VDJ_figures$clonalScatter <- clonalScatter(seurat,  
              cloneCall ="aa", 
              group.by = 'Groups',
              x.axis = "A. bau", 
              y.axis = "F. tula",
              dot.size = "total",
              graph = "proportion")
ggsave(plot = VDJ_figures$clonalScatter, filename = here(figures_path,  paste0('Clonal_Scatter_by_Group', '.pdf')), width = 7, height = 5)

# Analyzing percent AA usage
VDJ_figures$percentAA <- percentAA(seurat, chain = 'TRB', group.by = 'Groups', aa.length = 19) +
    labs(title = 'Percent of AA usage across CDR3-beta by Group') +
    theme(plot.title = element_text(hjust = 0.5, size = 16))
ggsave(plot = VDJ_figures$percentAA, filename = here(figures_path,  paste0('Percent_AA_usage_CDR3_TRB_by_Group', '.pdf')), width = 7, height = 5)

# Analyzing positional entropy
VDJ_figures$positionalEntropy <-  positionalEntropy(seurat, chain = 'TRB', group.by = 'Groups') +
    labs(title = 'Positional entropy of CDR3-beta by Group') +
    theme(plot.title = element_text(hjust = 0.5, size = 16))
ggsave(plot = VDJ_figures$positionalEntropy, filename = here(figures_path,  paste0('Positional_entropy_CDR3_TRB_by_Group', '.pdf')), width = 7, height = 5)

# Analyzing chemical properties
VDJ_figures$positionalProperty <- positionalProperty(seurat, 
                   chain = 'TRB', 
                   group.by = 'Groups', 
                   method = 'atchleyFactors') +
    labs(title = 'Positional properties of CDR3-beta by Group') +
    theme(plot.title = element_text(hjust = 0.5, size = 16))
ggsave(plot = VDJ_figures$positionalProperty, filename = here(figures_path,  paste0('Positional_properties_CDR3_TRB_by_Group', '.pdf')), width = 7, height = 5)

# Analyzing V-J gene pairings
VDJ_figures$percentVJ <- percentVJ(seurat, 
         chain = "TRB",
         group.by = 'Groups') +
    labs(title = 'Percent TRAV-TRAJ gene pairings by Groups') +
    theme(plot.title = element_text(hjust = 0.5, size = 16))
ggsave(plot = VDJ_figures$percentVJ, filename = here(figures_path,  paste0('Percent_TRBV_TRBJ_gene_pairings_by_Group', '.pdf')), width = 7, height = 5)

seurat_small <- seurat |> subset(subset = Groups %in% 'Uninfected', invert = T) # Removing uninfected samples for this analysis
df.vj <- percentVJ(seurat_small, 
                   chain = "TRB",
                   group.by = 'Samples',
                   exportTable = TRUE,
                   summary.fun = "proportion") # Export proportions for PCA
# Performing PCA on the V-J pairing matrix
pc.vj <- prcomp(t(df.vj)) 
# Getting data frame to plot from
df_plot_vj <- as.data.frame(cbind(pc.vj$x[,1:2], colnames(df.vj)))
colnames(df_plot_vj) <- c("PC1", "PC2", "Sample")
df_plot_vj$PC1 <- as.numeric(df_plot_vj$PC1)
df_plot_vj$PC2 <- as.numeric(df_plot_vj$PC2)
# Plotting the PCA results
VDJ_figures$PCA_TRB_gene_paring <- ggplot(df_plot_vj, aes(x = PC1, y = PC2)) +
  geom_point(aes(fill = Sample), shape = 21, size = 5) +
  guides(fill=guide_legend(title="Samples")) +
  scale_fill_manual(values = c('blue', 'blue1', 'blue2', 'blue3', 'darkorange', 'darkorange1', 'darkorange2', 'darkorange3', 'darkorange4')) +
  theme_classic() +
  labs(title = "PCA of TRBV-TRBJ Gene Pairings")
ggsave(plot = VDJ_figures$PCA_TRB_gene_paring, filename = here(figures_path,  paste0('PCA_TRBV_TRBJ_gene_pairings_by_Sample', '.pdf')), width = 7, height = 5)

# vizGenes(seurat, x.axis = 'TRAV', plot = 'barplot')
# vizGenes(seurat, x.axis = 'TRAJ', plot = 'barplot')



```

## V-J gene pairing analysis

```{r}

# Analyzing V-J gene pairings
TRBV_J_table <- percentVJ(seurat, 
         chain = "TRB",
         group.by = 'Groups',
         exportTable = T)

# Analyzing V-J gene pairings
TRAV_J_table <- percentVJ(seurat, 
         chain = "TRA",
         group.by = 'Groups',
         exportTable = T)


source(here('scripts/circlize_functions.r'))

VDJ_figures$TRB_circos_Uninfected <-  create_V_J_chord_diagram(TRBV_J_table, figures_path = figures_path,group_name = 'Uninfected')
# ggsave(plot = VDJ_figures$TRB_circos_Uninfected, filename = here(figures_path,  paste0('Circos_TRBV_TRBJ_Uninfected', '.pdf')), width = 7, height = 7)

VDJ_figures$TRB_circos_Abau <-  create_V_J_chord_diagram(TRBV_J_table, figures_path = figures_path, group_name = 'A. bau')
# ggsave(plot = VDJ_figures$TRB_circos_Abau, filename = here(figures_path,  paste0('Circos_TRBV_TRBJ_Abau', '.pdf')), width = 7, height = 7)

VDJ_figures$TRB_circos_Ftula <-  create_V_J_chord_diagram(TRBV_J_table, figures_path = figures_path,group_name = 'F. tula')
# ggsave(plot = VDJ_figures$TRB_circos_Ftula, filename = here(figures_path,  paste0('Circos_TRBV_TRBJ_Ftula', '.pdf')), width = 7, height = 7)

# create_V_J_chord_diagram(TRAV_J_table, figures_path = figures_path, group_name = 'A. bau', chain_a = 'TRAV', chain_b = 'TRAJ')
# create_V_J_chord_diagram(TRAV_J_table, figures_path = figures_path,group_name = 'F. tula', chain_a = 'TRAV', chain_b = 'TRAJ')
# create_V_J_chord_diagram(TRAV_J_table, figures_path = figures_path,group_name = 'Uninfected', chain_a = 'TRAV', chain_b = 'TRAJ')
```


## Diversity analysis: D50 

```{r}
source(here('scripts/function_template.r'))
plot <- calculate_D50(seurat, cell_grouping_var = orig.ident, replicate_var = Samples, replicate_group_var = Groups, results_path = results_path, figures_path = figures_path, tables_path = tables_path, colors= group_colors) 

VDJ_figures$D50 <- plot + 
    scale_y_continuous(limits = c(0, 0.5), expand = expansion(mult = c(0, 0))) +
    theme(plot.margin = margin(4, 4, 4, 4)) +
    ggpubr::stat_compare_means(comparisons = list(c('A. bau', 'F. tula'), c('A. bau', 'Uninfected'), c('F. tula', 'Uninfected')), label = 'p.signif', method = 'wilcox.test',  hide.ns = F, label.y = 0.40 , step.increase = 0.06, bracket.size = 0.7, size = 3, fontfamily = 'Arial') +
    coord_cartesian(clip = 'off') 
ggsave(plot = VDJ_figures$D50, filename = here(figures_path,  paste0('D50_by_Group', '.pdf')), width = 7, height = 5)
VDJ_figures$D50


```

## Repertoire overlap analysis: Overlap coefficient

```{r}

# immunarch_table <- exportClones(seurat, format = 'immunarch', write.file = FALSE, group.by = 'Samples')
# overlap <- repOverlap(immunarch_table$data, .method = 'overlap', .verbose=FALSE)
# vis(overlap) + ggtitle('Overlap Coefficient')
# ggsave(filename = paste0(figures_path, 'overlap_coefficient.pdf'))
combined2 <- scRepertoire:::.expression2List(seurat, split.by ='Samples')
overlap_coefficient <- clonalOverlap(combined2, cloneCall = 'aa', chain = 'both',exportTable = T,method = 'overlap')

# clonalOverlap(combined2, cloneCall = 'aa', chain = 'both',method = 'overlap') +
#     theme(axis.text.x = element_text(angle = 45, hjust = *1*)) +
#     labs(title = 'TCR repertoire overlap coefficient across samples') +
#     theme(plot.title = element_text(hjust = 0.5, size = 16))

comparison_colors <- c(
    'A. bau vs A. bau' = '#0072b2',
    'F. tula vs F. tula' = '#d55e10',
    'A. bau vs F. tula' = 'purple',
    'A. bau vs Uninfected' = 'grey50',
    'F. tula vs Uninfected' = 'grey50',
    'Uninfected vs Uninfected' = 'grey80'
)

VDJ_figures$overlap_coefficient <-  overlap_coefficient |> 
    rownames_to_column('Sample') |>
    pivot_longer(cols = -c(1), names_to = 'Sample_2', values_to = 'Morisita_index', values_drop_na = TRUE)   |>
    mutate(Comparison = case_when(
        str_detect(Sample, 'A. bau') & str_detect(Sample_2, 'A. bau') ~ 'A. bau vs A. bau',
        str_detect(Sample, 'F. tula') & str_detect(Sample_2, 'F. tula') ~ 'F. tula vs F. tula',
        str_detect(Sample, 'Uninfected') & str_detect(Sample_2, 'Uninfected') ~ 'Uninfected vs Uninfected',
        str_detect(Sample, 'A. bau') & str_detect(Sample_2, 'F. tula') ~ 'A. bau vs F. tula',
        str_detect(Sample, 'A. bau') & str_detect(Sample_2, 'Uninfected') ~ 'A. bau vs Uninfected',
        str_detect(Sample, 'F. tula') & str_detect(Sample_2, 'Uninfected') ~ 'F. tula vs Uninfected',
        str_detect(Sample, 'F. tula') & str_detect(Sample_2, 'A. bau') ~ 'A. bau vs F. tula',
        str_detect(Sample, 'Uninfected') & str_detect(Sample_2, 'A. bau') ~ 'A. bau vs Uninfected',
        str_detect(Sample, 'Uninfected') & str_detect(Sample_2, 'F. tula') ~ 'F. tula vs Uninfected',
        TRUE ~ 'Other'  
    ))  |>
    filter(Comparison %in% c('A. bau vs A. bau', 'F. tula vs F. tula', 'A. bau vs F. tula')) |>
    ggplot(aes(x = Comparison, y = Morisita_index, color = Comparison)) +
        geom_quasirandom(size = 1.5) +
        stat_summary(fun = mean, geom = "bar", position = position_dodge(width = 0.7), width = 0.5, aes(fill = Comparison, alpha = 0.5)) +
        stat_summary(fun.data = mean_se, fun.args = list(mult = 1),
            geom = "errorbar", width = 0.2, position = position_dodge(width = 0.7)) +
        theme_classic() +
        labs(x = 'Pairwise comparison', y = "Overlap coefficient", title = 'TCR repertoire overlap coefficient across groups') +
        guides(alpha = 'none')+
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
            text = element_text(size = 14),
            axis.line = element_line(colour = "black"))+
        scale_color_manual(values = comparison_colors) + scale_fill_manual(values = comparison_colors) +
        scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
        theme(plot.margin = margin(4, 4, 4, 4)) +
        ggpubr::stat_compare_means(comparisons = list(c('A. bau vs A. bau', 'A. bau vs F. tula'), c('F. tula vs F. tula', 'A. bau vs F. tula'), c('A. bau vs A. bau', 'F. tula vs F. tula')), label = 'p.signif', method = 'wilcox.test',  hide.ns = F , step.increase = 0.1, bracket.size = 0.7, size = 3, fontfamily = 'Arial') +
        coord_cartesian(clip = 'off')
ggsave(plot = VDJ_figures$overlap_coefficient, filename = here(figures_path,  paste0('Overlap_coefficient_by_Group', '.pdf')), width = 7, height = 5)




```

## Repertoire clonal dynamics

```{r}

seurat <- seurat_cell_cycle

# contigs <- read.csv(here('data/cluster_processing_2/aggr_all/outs/vdj_t/filtered_contig_annotations.csv'))
# #Demultiplexing VDJ libraries
# contig_list <- createHTOContigList(contigs, seurat, group.by = "Samples")
# contig_list <- contig_list[c(levels(seurat$Samples))]
# # Combining the Contigs
# combined <- combineTCR(contig_list, removeNA = TRUE, samples=c(levels(seurat$Samples)), filterMulti = TRUE)
# # # #Adding Samples as variables
# combined <- addVariable(combined, variable.name='Samples', variables=c(unique(seurat$Samples)))                                    
# ## Integrating data with seurat object
# cell_names <- Cells(seurat)
# groups_merge <- pull(seurat@meta.data, Samples)
# #Changing barcodes on the seurat object for the merge with VDJ data
# new_cell_names <- paste(groups_merge, cell_names, sep='_')
# seurat$barcode <- cell_names
# seurat <- RenameCells(seurat, new.names=new_cell_names)
# # #Combining VDJ and Seurat Object
# seurat <- combineExpression(combined, seurat, proportion = FALSE, cloneCall='aa', group.by='Samples', chain = 'both', cloneSize = c(Single=1, Double=2, Medium=10, Large=1000))
# # Renaming cells back to original barcodes
# seurat <- RenameCells(seurat, new.names=seurat$barcode)
# seurat@meta.data <- seurat@meta.data |> mutate(
#     cloneSize = case_when(
#         str_detect(cloneSize, 'Single') ~ 'Single (1)',
#         str_detect(cloneSize, 'Double') ~ 'Double (2)',
#         str_detect(cloneSize, 'Medium') ~ 'Medium (3-10)',
#         str_detect(cloneSize, 'Large') ~ 'Large (>10)',
#         is.na(cloneSize) ~ 'No TCR detected'
#     ) ) |>
#      mutate(cloneSize = factor(cloneSize, levels = c('Single (1)', 'Double (2)', 'Medium (3-10)', 'Large (>10)')))


clonotype_size_table <- seurat@meta.data |>
    dplyr::select(c('Samples', 'cloneSize')) |>
    table() |>
    as_tibble() |>
    mutate(Groups = case_when(
        str_detect(Samples, 'A. bau') ~ 'A. bau',
        str_detect(Samples, 'F. tula') ~ 'F. tula',
        str_detect(Samples, 'Uninfected') ~ 'Uninfected',
    ),
        # cloneSize = factor(cloneSize, levels = c('Single (1)', 'Double (2)', 'Medium (3-10)', 'Large (>10)') |> rev())
    )  |> 
    mutate(
        Groups = factor(Groups, levels = c('Uninfected', 'A. bau', 'F. tula')),
        Samples = factor(Samples, levels = levels(seurat$Samples))
        ) |>
    arrange((Samples))

clonotype_size_table <- clonotype_size_table |>
    group_by(Samples) |>
    mutate(frequency_within_sample = n/sum(n)*100) |>
    ungroup()  |>
    group_by(Groups) |>
    mutate(frequency_within_group = n/sum(n)*100) |>
    ungroup() 
head(clonotype_size_table)

VDJ_figures$clonotype_size_groups <-  clonotype_size_table |> 
    ggplot(aes(x=Groups, y = frequency_within_group, fill=cloneSize))  +
    geom_col() +
    theme_classic() +
    labs(x = 'Group', y = 'Percent (%)', title = 'Clonotype size by group')+
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    scale_fill_viridis(discrete = T, na.value = 'grey90', direction = -1)
ggsave(plot = VDJ_figures$clonotype_size_groups, filename = here(figures_path,  paste0('clonotype_size_by_group_', '.pdf')), width = 8, height = 6)

VDJ_figures$clonotype_size_groups2 <-  clonotype_size_table |> 
    filter(Groups != 'Uninfected') |>
    ggplot(aes(x=Groups, y = frequency_within_group, fill=cloneSize))  +
    geom_col() +
    theme_classic() +
    labs(x = '', y = 'Percent (%)', title = 'Clonotype size by group')+
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    scale_fill_viridis(discrete = T, na.value = 'grey90', direction = -1)
ggsave(plot = VDJ_figures$clonotype_size_groups2, filename = here(figures_path,  paste0('clonotype_size_by_group2_', '.pdf')), width = 8, height = 6)

VDJ_figures$clonotype_size_samples <- clonotype_size_table |> 
    ggplot(aes(x=Samples, y = n, fill=cloneSize))  +
    geom_col() +
    theme_classic() +
    labs(x = 'Group', y = 'Cell count', title = 'Clonotype size by group')+
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
    scale_fill_viridis(discrete = T, na.value = 'grey90', direction = -1)
ggsave(plot = VDJ_figures$clonotype_size_samples, filename = here(figures_path,  paste0('clonotype_size_by_sample_', '.pdf')), width = 10, height = 6)


```


```{r}
#| fig-cap: "Clonal repertoire representation per group, clonotypes are colored by their classification (Small, Medium, Large, Hyper Expanded as a proportion) on A PER SAMPLE basis, which is why a clonotype with the same number of sequences can have a different classification."
library( packcircles )
source(here('scripts/function_template.r'))
# Usage example:
plot_clonal_repertoire(seurat, 'A. bau', figures_path)
plot_clonal_repertoire(seurat, 'F. tula', figures_path)
plot_clonal_repertoire(seurat, 'Uninfected', figures_path)

```

## Motif analysis

```{r}
CDR3_length <- function (seurat, chain = 'beta', group_by = Groups, group = 'Control') {

    chain_column <- ifelse(chain == 'beta', 'CTaa_beta', 'CTaa_alpha')
    seurat@meta.data <- seurat@meta.data |>
        filter({{group_by}} == group) |>
        select(-any_of(c('CTaa_alpha', 'CTaa_beta'))) |>
        rownames_to_column(var = 'cell_barcode') |>
        separate_wider_delim(CTaa,delim = '_',names = c('CTaa_alpha', 'CTaa_beta'), cols_remove = FALSE)  |>
        column_to_rownames(var = 'cell_barcode')


    # Calculating median CDR3 length
    print(paste0('Mean CDR3', chain, ' length in ', group, ':'))
    mean_length <-seurat@meta.data |>
    filter(!is.na(!!as.name(chain_column))) |>
    pull(!!as.name(chain_column)) |>
    map_int(nchar) |>
    mean()
    print(mean_length)
    cat('\n')

    # Calculating median CDR3 length 
    print(paste0('Median CDR3', chain, ' length in ', group, ':'))
    median_length <-seurat@meta.data |>
    filter(!is.na(!!as.name(chain_column))) |>
    pull(!!as.name(chain_column)) |>
    map_int(nchar) |>
    median()
    print(median_length)
    cat('\n')
    cat('\n')

}

groups <- c('A. bau', 'F. tula', 'Uninfected')
walk(groups, \(group) CDR3_length(seurat, chain = 'alpha', group_by = Groups, group = group))
walk(groups, \(group) CDR3_length(seurat, chain = 'beta', group_by = Groups, group = group))


```


```{r}
#| layout-ncol: 3
# renv::install('ggseqlogo')
# renv::install('msa')
library(msa)
library(ggseqlogo)


seurat@meta.data <- seurat@meta.data |>
    rownames_to_column(var = 'cell_barcode') |>
    select(-any_of(c('CTaa_alpha', 'CTaa_beta'))) |>
    separate_wider_delim(CTaa,delim = '_',names = c('CTaa_alpha', 'CTaa_beta'), cols_remove = FALSE)  |>
    column_to_rownames(var = 'cell_barcode')


multiple_sequence_alignment_seurat <- function (seurat, chain, group_by = Groups, group = 'Control') {
        align <- seurat@meta.data |> 
            filter(!is.na({{chain}}) & {{group_by}} == group) |>
            pull({{chain}}) |> 
            msa(type = "protein", maxiters = 30)  |>
            msaConvert(type="seqinr::alignment") |>
            pluck('seq')
        return(align)
    }

sequence_motif_alignment <- function (seurat, chain = 'beta', group_by = Groups) {
        
    chain_column <- ifelse(chain == 'beta', 'CTaa_beta', 'CTaa_alpha')
    alignments = list()
    alignments <- map(seurat@meta.data |> pull({{group_by}}) |> levels(), \(group) multiple_sequence_alignment_seurat(seurat, chain = !!as.name(chain_column), group_by = {{group_by}},  group = group))
    names(alignments) <- seurat@meta.data |> pull({{group_by}}) |> levels()
    plot <- ggseqlogo(alignments, method = 'probability', ncol = 1) +
        labs(title = paste0('MAIT cell TCR ', chain, ' chain sequence motif by group')) +
        theme(plot.title = element_text(hjust = 0.5, size = 16))
    print(plot)
    return(list(plot = plot, alignments = alignments) )
}

sequence_motif_fixed_length <- function (seurat, chain = 'beta', group_by = Groups, group = 'Control', length = 14) {
        
    chain_column <- ifelse(chain == 'beta', 'CTaa_beta', 'CTaa_alpha')
    plot <- seurat@meta.data |>
        filter(!is.na(!!as.name(chain_column)) & {{group_by}} == group) |>
        filter(nchar(!!as.name(chain_column)) == length) |>
        pull(!!as.name(chain_column))  |>
    ggseqlogo(alignments, method = 'probability', ncol = 1) +
        labs(title = paste0('MAIT cell TCR ', chain, ' chain sequence motif in ', group, ' (CDR3 length = ', length, ')')) +
        theme(plot.title = element_text(hjust = 0.5, size = 14))
    print(plot)
    return(plot)
}
    

# # motif_results_beta <- sequence_motif(seurat, chain = 'beta', group_by = Groups)
# saveRDS(motif_results_beta, here(results_path, 'TCR_motif_results_TRB.rds'))
# motif_results_beta <- readRDS(here(results_path, 'TCR_motif_results_TRB.rds'))
# print(motif_results_beta$plot)

# # motif_results_alpha <- sequence_motif(seurat, chain = 'alpha', group_by = Groups)
# saveRDS(motif_results_alpha, here(results_path, 'TCR_motif_results_TRA.rds'))
# motif_results_alpha <- readRDS(here(results_path, 'TCR_motif_results_TRA.rds'))
# print(motif_results_alpha$plot)

groups <- c('A. bau', 'F. tula', 'Uninfected')
VDJ_figures$TRA_motifs <- map(groups, \(group) sequence_motif_fixed_length(seurat, chain = 'alpha', group_by = Groups, group = group, length = 12))
VDJ_figures$TRB_motifs_length_14 <- map(groups, \(group) sequence_motif_fixed_length(seurat, chain = 'beta', group_by = Groups, group = group, length = 14))
VDJ_figures$TRB_motifs_length_15 <- map(groups, \(group) sequence_motif_fixed_length(seurat, chain = 'beta', group_by = Groups, group = group, length = 15))

```

# Assembling figures

```{r}

figures_path <- here('Manuscript_figures')
unlink(figures_path, recursive = T)
dir.create(figures_path, recursive = T, showWarnings = F)

axis_thickness <- 0.4

```

```{r}
apply_figure_theme <- function(plot, markdown_elements = NULL, axis_thickness = NULL, remove_legend = FALSE, font_family = 'Arial', plot_margins = margin(3, 3, 3, 3), aspect_ratio = NULL) {
    # Base theme with all element_text
    theme_figures <- plot$theme + theme(
        base_family = font_family,
        text = element_text(size = 8, family = font_family),
        plot.title = element_text(size = 8, margin = margin(3, 3, 3, 3), hjust = 0.5, family = font_family),
        axis.text = element_text(size = 7, family = font_family),
        axis.text.x = element_text(size = 7, family = font_family),
        axis.text.y = element_text(size = 7, family = font_family),
        axis.title = element_text(size = 8, family = font_family),
        legend.title = element_text(size = 8, family = font_family),
        legend.text = element_text(size = 8, family = font_family),
        legend.key.size = unit(0.2, 'cm'),
        plot.margin = plot_margins,
        strip.text.x = element_text(size = 8, margin = margin(0, 0, 0, 0), family = font_family),
        legend.box.spacing = unit(0.05, 'cm'),
        plot.tag = element_text(size = 12, face = 'bold', family = font_family),

    )
    
    # Add axis line and tick thickness if specified
    if (!is.null(axis_thickness)) {
        axis_theme <- theme(
            axis.line = element_line(size = axis_thickness),
            axis.ticks = element_line(size = axis_thickness)
        )
        theme_figures <- theme_figures + axis_theme
    }
    
    # Remove legend if specified
    if (remove_legend) {
        legend_theme <- theme(legend.position = "none")
        theme_figures <- theme_figures + legend_theme
    }
    
    # Fix aspect ratio if specified
    if (!is.null(aspect_ratio)) {
        aspect_theme <- theme(aspect.ratio = aspect_ratio)
        theme_figures <- theme_figures + aspect_theme
    }
    
    # If markdown_elements specified, override with element_markdown
    if (!is.null(markdown_elements)) {                
        for (elem in markdown_elements) {
            theme_figures <- theme_figures +
              theme(!!as.name(elem) := element_markdown())                                  
            }
        return(plot + theme_figures)
    } else {
        return(plot + theme_figures)
    }}

```

## Figure 2

```{r}
# p1 <- textGrob('Figure 2', gp = gpar(fontsize = 20, fontface = 'bold'))

full_object_regressed$cluster_heatmap2 <- full_object_regressed$cluster_heatmap +
    scale_y_discrete(labels = c('g1' = '1', 'g2' = '2', 'g3' = '3', 'g4' = '4', 'g5' = '5', 'g6' = '6', 'g7' = '7', 'g8' = '8', 'g9' = '9')) +    
     labs(x = 'Genes', y = 'Leiden clusters', title = 'Top 3 genes per cluster')

full_object_regressed[['cluster_frequency']] <- full_object_regressed[['cluster_frequency']] +
    labs(x = 'Leiden clusters', title = 'Group frequency per cluster') +
    scale_fill_discrete(palette = group_colors, labels = c('Uninfected', '*A. bau*', '*F. tula*')) 

row_one <- plot_grid(
    (full_object_regressed[['DimPlot']] + labs(title = 'Leiden clusters'))|> apply_figure_theme(markdown_elements = NULL, axis_thickness = axis_thickness, remove_legend = TRUE),
    full_object_regressed[['DimPlot_groups']] |> apply_figure_theme(markdown_elements = NULL, remove_legend = TRUE),
    get_legend(full_object_regressed[['DimPlot']] |> apply_figure_theme(markdown_elements = NULL, remove_legend=FALSE) + theme(legend.box.margin = margin(0,12,0,12))),
    full_object_regressed[['cluster_frequency']] |> apply_figure_theme(markdown_elements = c('legend.text'), axis_thickness = axis_thickness),
    labels = c('A', 'B', 'C'),
    label_fontface = 'bold',
    label_size = 12,
    nrow = 1,
    align = 'h',
    axis = 'bt',
    rel_widths = c(1.2, 3, 0.4, 2)
)

row_two <- plot_grid(
    full_object_regressed$cluster_heatmap2 |> apply_figure_theme(markdown_elements = c('legend.title')),
    full_object_regressed$Milo_figures[['neighborhood_graph']] |> apply_figure_theme(markdown_elements = c('legend.title', 'plot.title'),axis_thickness = axis_thickness),
    full_object_regressed$Milo_figures[['beeswarm_graph']] |> apply_figure_theme(markdown_elements = c('legend.title', 'axis.title.x'), axis_thickness = axis_thickness),
    labels = c('D', 'E', 'F'),
    label_fontface = 'bold',
    label_size = 12,
    nrow = 1,
    align = 'h',
    axis = 'bt',
    rel_widths = c(3, 1.5, 1.5)
)


legend_1 <- get_legend(full_object_regressed$feature_plots$Rorc |> apply_figure_theme(markdown_elements = c('plot.title'), remove_legend=FALSE) + theme(legend.box.margin = margin(0,1,0,1)))
legend_2 <- get_legend(full_object_regressed$feature_plots$Il17a |> apply_figure_theme(markdown_elements = c('plot.title'), remove_legend=FALSE) + theme(legend.box.margin = margin(0,1,0,1)))
legend_3 <- get_legend(full_object_regressed$cNMF_GEP_feature_plots$`GEP 1:<br>Actin filament<br>organization` |> apply_figure_theme(markdown_elements = c('plot.title', 'legend.title'), remove_legend=FALSE) + theme(legend.box.margin = margin(0,1,0,1)))

margins_3 <- margin(0, 0, 0, 0)

no_y_axis <- theme(axis.line.y = element_blank(), axis.ticks.y = element_blank(), axis.text.y = element_blank(), axis.title.y = element_blank(), plot.margin = margin(0, 0, 0, 0))
no_x_axis <- theme(axis.line.x = element_blank(), axis.ticks.x = element_blank(), axis.text.x = element_blank(), axis.title.x = element_blank(), plot.margin = margin(0, 0, 0, 0))
# full_object_regressed$feature_plots$Rorc + no_y_axis

featured_plots <- plot_grid(
        full_object_regressed$feature_plots$Rorc |> apply_figure_theme(markdown_elements = c('plot.title'), axis_thickness = axis_thickness, remove_legend = TRUE,plot_margins = margin(0, 0, 0, 0)) + no_x_axis,
    full_object_regressed$feature_plots$Tbx21  |> apply_figure_theme(markdown_elements = c('plot.title'), axis_thickness = axis_thickness, remove_legend = TRUE,plot_margins = margin(0, 0, 0, 0)) + NoAxes(),
    legend_1,
    full_object_regressed$feature_plots$Il17a  |> apply_figure_theme(markdown_elements = c('plot.title'), axis_thickness = axis_thickness, remove_legend = TRUE,plot_margins = margin(0, 0, 0, 0)),
    full_object_regressed$feature_plots$Ifng |> apply_figure_theme(markdown_elements = c('plot.title'), axis_thickness = axis_thickness, remove_legend = TRUE,plot_margins = margin(0, 0, 0, 0)) + no_y_axis,
    legend_2,
    labels = c('G'),
    label_fontface = 'bold',
    label_size = 12,
    nrow = 2,
    align = 'hv',
    axis = 'btlr',
    rel_widths = c(1, 1, 0.4),
    rel_heights = c(1, 1)
)
featured_plots




GEP_plots <- plot_grid(
    full_object_regressed$cNMF_GEP_feature_plots$`GEP 1:<br>Actin filament<br>organization`  |> apply_figure_theme(markdown_elements = c('plot.title', 'legend.title'), axis_thickness = axis_thickness, remove_legend = TRUE,plot_margins = margin(0, 0, 0, 0)) + no_x_axis,
    full_object_regressed$cNMF_GEP_feature_plots$`GEP 2:<br>T cell activation` |> apply_figure_theme(markdown_elements = c('plot.title', 'legend.title'), axis_thickness = axis_thickness, remove_legend = TRUE,plot_margins = margin(0, 0, 0, 0)) + NoAxes(),
    legend_3,
    full_object_regressed$cNMF_GEP_feature_plots$`GEP 4:<br>T cell differentiation` |> apply_figure_theme(markdown_elements = c('plot.title', 'legend.title'), axis_thickness = axis_thickness, remove_legend = TRUE,plot_margins = margin(0, 0, 0, 0)) + no_x_axis,    
    full_object_regressed$cNMF_GEP_feature_plots$`GEP 8:<br>Cell migration` |> apply_figure_theme(markdown_elements = c('plot.title', 'legend.title'), axis_thickness = axis_thickness, remove_legend = TRUE,plot_margins = margin(0, 0, 0, 0)) + NoAxes(),
    legend_3,
    full_object_regressed$cNMF_GEP_feature_plots$`GEP 10:<br>Cytotoxicity` |> apply_figure_theme(markdown_elements = c('plot.title', 'legend.title'), axis_thickness = axis_thickness, remove_legend = TRUE,plot_margins = margin(0, 0, 0, 0)),
    full_object_regressed$cNMF_GEP_feature_plots$`GEP 12:<br>Interferon signaling` |> apply_figure_theme(markdown_elements = c('plot.title', 'legend.title'), axis_thickness = axis_thickness, remove_legend = TRUE,plot_margins = margin(0, 0, 0, 0)) + no_y_axis,    
    legend_3,
    labels = c('I'),
    label_fontface = 'bold',
    label_size = 12,
    nrow = 3,
    align = 'hv',
    axis = 'btlr',
    rel_widths = c(1, 1, 0.4),
    rel_heights = c(1, 1, 1)
)
GEP_plots

signature_plots <- plot_grid(
    full_object_regressed$signature_violin_plots$MAIT1 |> apply_figure_theme(markdown_elements = c('plot.title'), axis_thickness = axis_thickness, remove_legend = TRUE,plot_margins = margin(0, 0, 0, 0)),
    full_object_regressed$signature_violin_plots$MAIT17 |> apply_figure_theme(markdown_elements = c('plot.title'), axis_thickness = axis_thickness, remove_legend = TRUE,plot_margins = margin(0, 0, 0, 0)),
    labels = c('H'),
    label_fontface = 'bold',
    label_size = 12,
    nrow = 1
    # align = 'hv',
    # axis = 'bt'
)

row_three <- plot_grid(
    featured_plots,
    signature_plots,
    nrow = 2,
    align = 'v',
    # axis = 'b',
    rel_heights = c(2, 1)
)


motif_legend <- get_legend(VDJ_figures$TRB_motifs_length_14[[1]] |> apply_figure_theme(axis_thickness = axis_thickness, remove_legend=FALSE) + theme(legend.box.margin = margin(0,12,0,12), legend.position = 'right'))
bar_legend <- get_legend(VDJ_figures$D50 |> apply_figure_theme(axis_thickness = axis_thickness, remove_legend=FALSE) + theme(legend.box.margin = margin(0,12,0,12), legend.position = 'right'))

motifs <- plot_grid(
    (VDJ_figures$TRB_motifs_length_14[[3]] + labs(title = 'TCR-beta in Uninfected')) |> apply_figure_theme(markdown_elements = c('plot.title'), axis_thickness = axis_thickness, remove_legend = TRUE)+ no_x_axis,
    (VDJ_figures$TRB_motifs_length_14[[1]] + labs(title = 'TCR-beta in *A. bau*')) |>  apply_figure_theme(markdown_elements = c('plot.title'), axis_thickness = axis_thickness, remove_legend = TRUE) +no_x_axis,
    (VDJ_figures$TRB_motifs_length_14[[2]] + labs(title = 'TCR-beta in *F. tula*')) |>  apply_figure_theme(markdown_elements = c('plot.title'), axis_thickness = axis_thickness, remove_legend = TRUE),
    ncol = 1,
    # align = 'v',
    rel_heights = c(1, 1, 1.5)
)
motifs_with_legend <- plot_grid(
    (VDJ_figures$D50 + labs(title = 'TCR repertoire diversity', x = '') + 
        scale_color_discrete(palette = group_colors, labels = c('Uninfected', '*A. bau*', '*F. tula*')) +
        scale_x_discrete(labels = c('Uninfected', '*A. bau*', '*F. tula*')) +
        scale_fill_discrete(palette = group_colors, labels = c('Uninfected', '*A. bau*', '*F. tula*'))) |>
        apply_figure_theme(markdown_elements = c('legend.text', 'axis.text.x'), axis_thickness = axis_thickness, remove_legend = TRUE),
    NULL,
    bar_legend,
    motifs,
    NULL,
    motif_legend,
    ncol = 3,
    labels = c('J', '', '', 'L'),
    label_fontface = 'bold',
    label_size = 12,
    rel_widths = c(2, 0.1, 0.4),
    rel_heights = c(1.3, 2)
)


bar_plots <- plot_grid(
    # (VDJ_figures$clonotype_size_groups2 +
    #     labs(title = 'Clonotype size\ndistribution', x = '', fill = '')+
    #     theme(axis.text.x = element_text(fontface = 'italic'))) |>
    #     apply_figure_theme(axis_thickness = axis_thickness),
    (VDJ_figures$overlap_coefficient +
        labs(title = 'TCR repertoire\noverlap coefficient', x = '') +
        scale_fill_discrete(palette = comparison_colors, labels = c('*A. bau* vs *A. bau*', '*A. bau* vs *F. tula*', '*F. tula* vs *F. tula*')) +
        scale_color_discrete(palette = comparison_colors, labels = c('*A. bau* vs *A. bau*', '*A. bau* vs *F. tula*', '*F. tula* vs *F. tula*')) +
        scale_x_discrete(labels = c('*A. bau* vs *A. bau*', '*A. bau* vs *F. tula*', '*F. tula* vs *F. tula*')) +
        theme(legend.position = 'none')) |> 
        apply_figure_theme(markdown_elements = c('legend.text', 'axis.text.x'), axis_thickness = axis_thickness),
    NULL,
    NULL,
    labels = c('K'),
    label_fontface = 'bold',
    label_size = 12,
    nrow = 3,
    align = 'vh',
    axis = 'bt',
    rel_heights = c(3, 1, 1)
)


row_five <- plot_grid(
    row_three,
    GEP_plots,
    motifs_with_legend,
    bar_plots,
    ncol = 4,
    # align = 'vh',
    # axis = 'btlr',
    rel_widths = c(1, 1, 1, 0.7)
) 
row_five

figure_2 <- plot_grid(
    row_one,
    row_two,
    row_five,
    ncol = 1,
    align = 'hv',
    axis = 'tblr',
    labels = c('', ''),
    rel_heights = c(1, 1, 2, 1.2)
)
figure_2

ggsave(plot = figure_2, filename = 'Figure_2.pdf', path = figures_path, width = 10, height = 10, units = 'in', device = cairo_pdf)

 

```

## Figure S2

```{r}

row_one <- plot_grid(
    full_object_original[['DimPlot']] |> apply_figure_theme(axis_thickness = axis_thickness, remove_legend = TRUE) + labs(title = 'Leiden clusters'),
    full_object_original[['DimPlot_groups']] |> apply_figure_theme(markdown_elements = NULL, axis_thickness = axis_thickness, remove_legend = TRUE),
    get_legend(full_object_original[['DimPlot']] |> apply_figure_theme(remove_legend=FALSE) + theme(legend.box.margin = margin(0,12,0,12))),
    (full_object_original$signature_feature_plots$`Cycling G2M`+scale_color_continuous(palette = diverging_palette_3, breaks = c(-1, 0, 1)))  |> apply_figure_theme(markdown_elements = c('plot.title'), axis_thickness = axis_thickness, remove_legend = TRUE),
    (full_object_original$signature_feature_plots$`Cycling S` ) |> apply_figure_theme(markdown_elements = c('plot.title'), axis_thickness = axis_thickness, remove_legend = TRUE),
    get_legend((full_object_original$signature_feature_plots$`Cycling G2M`+scale_color_continuous(palette = diverging_palette_3, breaks = c(-1, 0, 1)))  |> apply_figure_theme(remove_legend=FALSE) + theme(legend.box.margin = margin(0,12,0,12))),
    labels = c('A', 'B', '', 'C', ''),
    label_fontface = 'bold',
    label_size = 12,
    nrow = 1,
    align = 'h',
    axis = 'bt',
    rel_widths = c(1.2, 3, 0.4, 1.2, 1.2, 0.4)
)
row_one

pathways_plots <- plot_grid(
    (full_object_original$pathways_Abau  +
        scale_x_discrete(labels = label_wrap(24))) |> 
        apply_figure_theme(markdown_elements = c('plot.title'), axis_thickness = axis_thickness, remove_legend = TRUE),
    (full_object_original$pathways_Ftula +
        scale_x_discrete(labels = label_wrap(24))) |> 
        apply_figure_theme(markdown_elements = c('plot.title'), axis_thickness = axis_thickness, remove_legend = TRUE),  
    labels = c('E', 'F'),
    label_fontface = 'bold',
    label_size = 12,
    nrow = 1,
    align = 'h'
)
pathways_plots

column_one_bottom <- plot_grid(
    (full_object_original$volcano_plot +
        labs(x = 'Average logFC (*A. bau*/*F. tula*)')) |> 
        apply_figure_theme(markdown_elements = c('axis.title.x'), axis_thickness = axis_thickness),
    pathways_plots,
    ncol = 1,
    align = 'v',
    rel_heights = c(2, 2),
    labels = c('D', '' ),
    label_fontface = 'bold',
    label_size = 12
)


column_two_bottom <- plot_grid(
    full_object_original$pathways_Abau_Tcelldifferentiation |> apply_figure_theme(markdown_elements = c('legend.title')) + theme(axis.text.y = element_text(fontface = 'italic')), 
    full_object_original$pathways_Abau_woundrepair |> apply_figure_theme(markdown_elements = c('legend.title')) + theme(axis.text.y = element_text(fontface = 'italic')),
    full_object_original$pathways_Ftula_cytotoxicity |> apply_figure_theme(markdown_elements = c('legend.title')) + theme(axis.text.y = element_text(fontface = 'italic')),
    nrow = 3,
    align = 'v',
    labels = c('G', 'H', 'I'),
    label_fontface = 'bold',
    label_size = 12,
    rel_heights = c(1, 1, 1)
)

row_two <- plot_grid(
    column_one_bottom,
    column_two_bottom,
    ncol = 2,
    align = 'h',
    axis = 'bt',
    rel_widths = c(1, 1)
    )

figure_S2 <- plot_grid(
    row_one,
    row_two,
    ncol = 1,
    align = 'hv',
    axis = 'tblr',
    rel_heights = c(1, 2)
)
figure_S2 
ggsave(plot = figure_S2, filename = 'Figure_S2.pdf', path = figures_path, width = 10, height = 10, units = 'in')

```