---
title: Analysis of Abau and Ftul infected mice - lungs
author: Eduard Ansaldo
date: '`r Sys.Date()`'
verbose: false
engine: knitr
knitr:
  opts_chunk: 
    echo: false
    warning: false
    message: false
    error: false
    cache: true
    results: markdown
    code-fold: true
    cache-lazy: false
    
processing: true

execute: 
  cache: true


format:
  html:
    embed-resources: true
    fig-cap-location: top
    toc: true
    toc-depth: 3

    toc-expand: 1
    toc-location: left
    toc-title: Table of Contents
    toc-collapse: true
    number-sections: true
    page-layout: full
    fig-width: 6
    fig-height: 6
    fig-align: 'center'
---

::: {.content-visible when-meta="verbose"}

# Initialization

:::

```{r}
#| cache: false
# General R and plotting 
library(conflicted)
library(tidyverse)
library(ggplot2)
library(scales)
library(patchwork)
library(cowplot)
library(gridExtra)
library(ggrepel)
library(stringr)
library(VennDiagram)
library(pheatmap)
library(viridis)
library(here)
library(readxl)
library(readxl)
library(ggbeeswarm)
library(tidyplots)
library(extrafont)

# Single Cell Analysis Packages
library(Seurat)
library(scRepertoire)
library(circlize)
library(scCustomize)
library(SingleR)
library(celldex)
library(UCell)
library(presto)
library(scDblFinder)
library(immunarch)

# DEG, pathway enrichment and visualization packages
library(DESeq2)
# library(gprofiler2)
library(clusterProfiler)
library(DOSE)
library(pathview)
library(org.Mm.eg.db)
library(scRepertoire)
library(enrichplot)
library(msigdbr)
library(rlang)

# Python Version 3.9.6!!! 

# peconflicts_scout()
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::count)
conflicts_prefer(dplyr::select)
conflicts_prefer(base::setdiff)
conflicts_prefer(dplyr::rename)
conflicts_prefer(base::unname)
conflicts_prefer(tidyr::expand)
conflicts_prefer(tidyplots::save_plot)

# Loading custom functions
source(here('scripts/function_template.r'))
source(here('scripts/metascape_functions.r'))
source(here('scripts/circlize_functions.r'))


```

```{r}
#| results: hide 
#| cache: false 
# Setting working directory and seed
# renv::update()
# renv::snapshot()
# renv::init()
# renv::status()
set.seed(3514)
# i_am('scripts/Analysis.qmd')
here()
results_path <- here('results/')
dir.create(results_path)
results_path <- here('results/')
dir.create(results_path)
figures_path <- here('results/figures/')
dir.create(figures_path)
# data_path <- here('data/')
# dir.create(data_path)
tables_path <- here('results/tables/')
dir.create(tables_path)

# Loading custom functions
source(here('scripts/function_template.r'))
source(here('scripts/metascape_functions.r'))
source(here('scripts/circlize_functions.r'))
source(here('scripts/ClusterProfiler_functions.r'))

# Setting up color palettes
diverging_palette <- hcl.colors(n = 20,'Purple-Green',rev = T)
sequential_palette_dotplot <- hcl.colors(n = 20,'YlGn',rev = T)
sequential_palette <- hcl.colors(n = 20,'Mako',rev = T)

pathway_enrichment <- c('ClusterProfiler')

knitr::dep_prev()
```


# Initialization

```{r}
seurat_original <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs.rds'))
seurat_cell_cycle <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs_cell_cycle.rds'))
seurat_Abau <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs_Abau_only.rds'))
seurat_Ftula <- readRDS(here('data/MAIT_Abau_Ftula_CT_Lungs_Ftula_only.rds'))
results_path <- here('results', 'Manuscript_figures')
dir.create(results_path, recursive = T, showWarnings = F)
unlink(results_path, recursive = T)
figures_path <- here('results/Manuscript_figures/figures/')
dir.create(figures_path, recursive = T, showWarnings = F)
tables_path <- here('results/Manuscript_figures/tables/')
dir.create(tables_path, recursive = T, showWarnings = F)
object_annotations <- 'manuscript_figures'




```

```{r}
group_colors <- c(
    'Abau' = '#0072b2', 
    'Ftula' = '#d55e10', 
    'CT' = '#009e73'
)

# Setting up color palettes
diverging_palette <- hcl.colors(n = 20,'Purple-Green',rev = T)
sequential_palette_dotplot <- hcl.colors(n = 20,'YlGn',rev = T)
sequential_palette <- hcl.colors(n = 20,'Mako',rev = T)
diverging_palette_2 <- hcl.colors(n = 20,'RdBu',rev = T)
diverging_palette_3 <- hcl.colors(n = 20,'Geyser',rev = F)

UMAP_dot_size <- 0.2

full_object_original <- list()
full_object_regressed <- list()
Abau_object <- list()
Ftula_object <- list()
Milo_figures <- list()
```

# Generate original object plots

```{r}
seurat <- seurat_original
seurat@meta.data <- seurat@meta.data |>
    mutate(Samples = str_remove(Samples, '_Mouse'))
# levels(seurat$Samples)
```

## DimPlots and cluster frequency

```{r}
Idents(seurat) <- 'seurat_clusters'

total_cell_number <- Cells(seurat) |> length()
Abau_cell_number <- Cells(seurat_Abau) |> length()
Ftula_cell_number <- Cells(seurat_Ftula) |> length()

# Clusters UMAP plot
full_object_original$DimPlot <- DimPlot(seurat, group.by = 'seurat_clusters', pt.size = UMAP_dot_size, label = F) +
     labs(title = 'MAIT cell Leiden clusters', 
        #   subtitle = 'Lung single-cell RNA-seq', 
        #   caption = 'n = 4 Abau, n = 5 Ftula, n = 4 CT',
     x = 'UMAP 1', y = 'UMAP 2')  
full_object_original$DimPlot <- full_object_original$DimPlot  |> LabelClusters(id = 'seurat_clusters', size = 3, color = 'black', repel = F, fontface = 'bold') 

# Clusters split by group plot
full_object_original$DimPlot_groups <- DimPlot(seurat, group.by = 'seurat_clusters', pt.size = UMAP_dot_size, label = F, split.by = 'Groups') +
     labs( x = 'UMAP 1', y = 'UMAP 2', title = NULL)  

# Cluster frequency plot
seurat_local <- subset(seurat, subset = Groups != 'CT')
cell_counts <- FetchData(seurat_local, vars = c("seurat_clusters", "Samples", "Groups"))
cell_counts <- arrange(cell_counts, Samples)
counts <- cell_counts %>% 
    add_count(Groups, name='total_cell_count_by_group')  |>
    mutate(weights = total_cell_count_by_group/Abau_cell_number)  |>
    add_count(seurat_clusters, name='total_cell_count_by_cluster', wt =  weights)  |>
    count(seurat_clusters, Samples, Groups, total_cell_count_by_cluster , name='sample_cluster_count', wt = weights)  |> 
    mutate(frequency_within_cluster=sample_cluster_count*100/total_cell_count_by_cluster)  |> 
    mutate(Samples = as.character(Samples)) |> 
    arrange(Samples, desc(Samples)) 
frequency_table <- counts |> 
    arrange(Samples) |>     
    pivot_wider(id_cols = seurat_clusters,  names_from =  Samples , values_from = frequency_within_cluster, values_fill = 0)
# write.csv(frequency_table,file=here(tables_path, paste0( "frequency per roll }} per condition " , ".csv")), row.names=F) 

full_object_original$cluster_frequency <- counts |>
    ggplot(aes(x=seurat_clusters, fill=Groups, y = frequency_within_cluster)) +
    geom_col(aes(), position = 'fill') +
    # geom_bar(position=position_fill()) +
    theme_classic() +
    labs(x = 'Cluster', y = 'Frequency (%)', title = 'MAIT normalized cell cluster\nfrequency by sample')+
    scale_y_continuous(expand = expansion(mult = c(0, 0)), breaks = c(0, 0.25, 0.5, 0.75, 1), labels = c(0, 25, 50, 75, 100))+
    scale_fill_discrete(palette = group_colors) 

```

## Feature plots

```{r}

#  cluster Markers dot plot
seurat <- PrepSCTFindMarkers(seurat)
top_genes_per_cluster_results <- top_genes_per_cluster(seurat, n_genes_to_plot = 3, object_annotations = '', tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = F) 
full_object_original$cluster_dotplot <- top_genes_per_cluster_results$plot
ggsave(plot = full_object_original$cluster_dotplot$plot, filename = here(figures_path,  paste0('DotPlot_Top3_per_cluster', '.pdf')), width = 8, height = 12)


type_17_markers <- c('Rorc', 'Il17a', 'Sdc1', 'Il7r', 'Il1r1', 'Ccr6', 'Sdc4')
type_1_markers <- c('Tbx21', 'Ifng', 'I12rb2', 'Cxcr3', 'Gzmb', 'Gzmk', 'Klrg1', 'Klrd1', 'Ly6c2')
naive_markers <- c('S1pr1', 'Ccr7', 'Sell', 'Tcf7', 'Lef1')
regulatory_markers <- c('Arg2', 'Il10', 'Il21')
exhaustion_markers <- c('Ctla4', 'Pdcd1', 'Lag3', 'Tigit', 'Havcr2')
tissue_repair_markers <- c('Areg', 'Ramp3','Il18r1', 'Tgfb1', 'Tnf', 'Mmp25', 'Cxcl10', 'Rankl', 'Ifrd1')
transcription_factors <- c('Tbx21', 'Rorc', 'Gata3', 'Foxp3', 'Ahr', 'Maf', 'Prdm1', 'Bcl6', 'Zbtb16')
tissue_residency <- c('Itgae', 'Klrg1', 'Itga1', 'Itga2', 'Ccr9', 'Ccr10', 'Itga4', 'Itgb7')
activation_markers <- c('Cd69', 'Cd44', 'Nr4a1')
cytokines <- c('Ifng', 'Il17a', 'Il17f', 'Il4', 'Il10', 'Tgfb1', 'Tnf', 'Il2', 'Gzma', 'Gzmb', 'Nkg7', 'Ccl5', 'Il27ra', 'Ifgnr1')
costimulatory_receptors <- c('Ctla4', 'Lag3', 'Tigit', 'Havcr2', 'Cd28', 'Cd40lg', 'Pdcd1', 'Tnfsf4', 'Cd27', 'Tnfrsf9')

final_list <- c(type_17_markers, type_1_markers, naive_markers, regulatory_markers, exhaustion_markers, tissue_repair_markers, transcription_factors, tissue_residency, activation_markers, cytokines, costimulatory_receptors) |> unique() |> rev()

# final_gene_list <- 

# Idents(seurat) <- 'seurat_clusters'
# DotPlot_scCustom(seurat, features = final_list, colors_use=sequential_palette_dotplot, flip_axes = T, scale = T, col.min = 0, col.max = 2, dot.scale = 8, dot.min = 0, scale.min = 0, scale.max = 60, x_lab_rotate = T, y_lab_rotate = F)  + theme(axis.text.x = element_text(size = 14), axis.text.y = element_text(size = 14), legend.title = element_text(size = 18))
# ggsave(filename = here(figures_path, 'Dotplot_markers.pdf'), width = 25,height = 5)
# Idents(seurat) <- 'Groups'
# DotPlot_scCustom(seurat, features = final_list, colors_use=sequential_palette_dotplot, flip_axes = F, scale = F, col.min = 0, col.max = 2, dot.scale = 8, dot.min = 0, scale.min = 0, scale.max = 60, x_lab_rotate = T, y_lab_rotate = F)  + theme(axis.text.x = element_text(size = 14), axis.text.y = element_text(size = 14), legend.title = element_text(size = 18))

feature_plot <- function (gene) {
    plot1 <- FeaturePlot_scCustom(seurat, features = gene,  colors_use = sequential_palette, pt.size = UMAP_dot_size, alpha_exp = 0.5) + labs(title = gene, x = 'UMAP 1', y = 'UMAP 2')
    ggsave(plot = plot1, filename = paste0('FeaturePlot', gene, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
    return(plot1)
}
genes_to_plot <- c('Rorc', 'Il17a', 'Sdc1', 'Il7r', 'Il1r1', 'Ccr6', 'Tbx21', 'Ifng', 'Il12rb2', 'Cxcr3', 'Gzmb', 'Gzmk', 'Gzma', 'Gzmm', 'Klrg1', 'Klrd1', 'Ly6c2', 'Ramp3', 'Slamf7', 'Areg')
full_object_original$feature_plots <- genes_to_plot |> 
    set_names() |>
    map(feature_plot )

```

## Signature scores

```{r}

DefaultAssay(seurat) <- 'SCT'
signatures <- read_excel(here('data/Hagoulet_signatures.xlsx'), skip = 1)
signatures <- signatures |>
    dplyr::filter(!(row_number() == 1)) |>
    dplyr::select(-all_of(c('Signature', 'Tissue repair 1', 'Tissue repair 2', 'Tissue repair 3', 'Tissue repair 4')))
    
signatures <- map(.x = signatures, .f = ~ purrr::discard(.x, is.na)) |> as.list()


tissue_repair_linehan <- c('Angpt1', 'Angpt2', 'Angpt4', 'Areg', 'Bmp2', 'Bmp4', 'Bmp7', 'Btc', 'Ccl2', 'Ccl3', 'Chat', 'Csf1', 'Csf2', 'Ctgf', 'Cxcl1', 'Cxcl10', 'Cxcl12', 'Cxcl15', 'Cxcl2', 'Cyr61', 'Defb1', 'Defb6', 'Dhh', 'Disp1', 'Dll1', 'Dll3', 'Dll4', 'Egf', 'Epgn', 'Ereg', 'Fgf1', 'Fgf10', 'Fgf22', 'Fgf7', 'Flg', 'Flg2', 'Furin', 'Hbegf', 'Hgf', 'Hif1a', 'Hmgb1', 'Igf1', 'Igf2', 'Ihh', 'Il1a', 'Il1b', 'Il6', 'Inhba', 'Inhbb', 'Jag1', 'Jag2', 'Lep', 'Mmp10', 'Mmp13', 'Mmp15', 'Mmp2', 'Mmp25', 'Mmp28', 'Mmp3', 'Ngf', 'Nrg1', 'Nrg2', 'Nrg3', 'Nrg4', 'Pdgfa', 'Pdgfb', 'Pdgfc', 'Pgf', 'Ptges2', 'Shh', 'Tgfa', 'Tgfb1', 'Tgfb2', 'Tgfb3', 'Thbs1', 'Tnf', 'Vegfa', 'Vegfb', 'Vegfc', 'Vegfd', 'Wnt1', 'Wnt10a', 'Wnt10b', 'Wnt11', 'Wnt16', 'Wnt2', 'Wnt2b', 'Wnt3', 'Wnt3a', 'Wnt4', 'Wnt5a', 'Wnt5b', 'Wnt6', 'Wnt7a', 'Wnt7b', 'Wnt8a', 'Wnt8b', 'Wnt9a', 'Wnt9b')

signatures$tissue_repair <- tissue_repair_linehan
signatures$MAIT17 <- signatures$MAIT17[!(signatures$MAIT17 %in% c('1700040D17Rik'))]

# Removing genes that are not detected
blacklist_elements <- c('H2afz','Smsc2','Hist1h2ap','Flg','Flg2','Bmp2','Bmp4','Ngf','Fgf7','Fgf10','Fgf22','Igf1','Ctgf','Mmp3','Mmp13','Defb1','Defb6','Epgn','Mmp2','Angpt1','Hgf','Cyr61','Inhbb','Btc','Ereg','Lep','Vegfc','Vegfd','Pdgfc','Cxcl12','Shh','Dhh','Ihh','Dll3','Dll4','Wnt1','Wnt2','Wnt3a','Wnt5a','Wnt6','Wnt7a','Wnt7b','Wnt8a','Wnt8b','Wnt9a','Wnt9b','Wnt10a','Wnt16','Chat','Thbs1','Fam46a','Kcna4','Fam129b','Tdpoz4','Rassf6','Ero1l','Cd244','A430078G23Rik','D1Ertd622e','Fam49a')

signatures <- map(.x = signatures, .f = ~ .x[!(.x %in% blacklist_elements)])

seurat <- AddModuleScore_UCell(seurat, features = signatures, name = NULL)

# Scale signatures (z-score) for comparison and visualization purposes
seurat@meta.data  <-  seurat@meta.data |>
    mutate(across(names(signatures), ~ scale(.x, center = T, scale = T), .names = '{col}_scaled'))
 # Visualization
signature_violin_plot <- function (signature) {
    max_y <- max(seurat@meta.data |> pull(signature))
    plot1 <- VlnPlot(seurat, features = signature, group.by = 'Groups', pt.size = 0, cols = group_colors) +
                labs(title = signature, x = NULL) + NoLegend() +
                ylim(0, max_y*1.65) +
                ggpubr::stat_compare_means(comparisons = list(c('Abau', 'Ftula'), c('Abau', 'CT'), c('Ftula', 'CT')), label = 'p.signif', method = 'wilcox.test',  hide.ns = F, label.y = max_y*1.1 , step.increase = 0.18, bracket.size = 0.7, size = 3) 
    ggsave(plot = plot1, filename = paste0('VlnPlot_', signature, '_by_group_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 4)
    return(plot1)
}

signature_dot_plot <- function (signature) {
    plot1 <- DotPlot_scCustom(seurat, features = signatures[[signature]] |> rev() |> unique() , group.by = 'Groups', colors_use = sequential_palette_dotplot, flip_axes = T, scale = F, dot.scale = 6) + labs(title = signature)
    ggsave(plot = plot1, filename = paste0('DotPlot_', signature, '_by_groups_', object_annotations, '.pdf'), path = figures_path, width = 5, height = length(signatures[[signature]])/4 + 1)
    return(plot1)
}
signature_feature_plot <- function (signature) {
    min_value  <-  min(seurat@meta.data |> pull(paste0(signature, '_scaled')))
    max_value  <-  max(seurat@meta.data |> pull(paste0(signature, '_scaled')))
    cut_off_value <- min(abs(min_value), abs(max_value))
    plot1 <- FeaturePlot_scCustom(seurat, features = paste0(signature, '_scaled'), colors_use = diverging_palette_3, na_cutoff=NULL, min.cutoff = -1*cut_off_value, max.cutoff = cut_off_value, pt.size = 0.1, alpha_exp = 0.2) + labs(title = signature, x = 'UMAP 1', y = 'UMAP 2')
    ggsave(plot = plot1, filename = paste0('FeaturePlot_', signature, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
    return(plot1)
}


```

```{r}
full_object_original$signature_violin_plots <- map(names(signatures), signature_violin_plot)
names(full_object_original$signature_violin_plots) <- names(signatures)

full_object_original$signature_dot_plots <- map(names(signatures), signature_dot_plot)
names(full_object_original$signature_dot_plots) <- names(signatures)

full_object_original$signature_feature_plots <- map(names(signatures), signature_feature_plot)
names(full_object_original$signature_feature_plots) <- names(signatures)

cut_off_value <- 1.5
signature_cycling_features <- list()
signature_cycling_features[[1]] <-
FeaturePlot_scCustom(seurat, features = paste0('Cycling_G2M', '_scaled'), colors_use = diverging_palette_2, na_cutoff=NULL, min.cutoff = -1*cut_off_value, max.cutoff = cut_off_value, pt.size = 0.1, alpha_exp = 0.3) + labs(title = 'Cycling_G2M')
signature_cycling_features[[2]] <- FeaturePlot_scCustom(seurat, features = paste0('Cycling_S', '_scaled'), colors_use = diverging_palette_2, na_cutoff=NULL, min.cutoff = -1*cut_off_value, max.cutoff = cut_off_value, pt.size = 0.1, alpha_exp = 0.3) + labs(title = 'Cycling_S')
```

# Generate cell cycle regressed object plots

```{r}
seurat <- seurat_cell_cycle
seurat@meta.data <- seurat@meta.data |>
    mutate(Samples = str_remove(Samples, '_Mouse'))
# levels(seurat$Samples)
```

## DimPlots and cluster frequency

```{r}
Idents(seurat) <- 'seurat_clusters'

total_cell_number <- Cells(seurat) |> length()
Abau_cell_number <- Cells(seurat_Abau) |> length()
Ftula_cell_number <- Cells(seurat_Ftula) |> length()

# Clusters UMAP plot
full_object_regressed$DimPlot <- DimPlot(seurat, group.by = 'seurat_clusters', pt.size = UMAP_dot_size, label = F) +
     labs(title = 'MAIT cell Leiden clusters', 
        #   subtitle = 'Lung single-cell RNA-seq', 
        #   caption = 'n = 4 Abau, n = 5 Ftula, n = 4 CT',
     x = 'UMAP 1', y = 'UMAP 2')  
full_object_regressed$DimPlot <- full_object_regressed$DimPlot  |> LabelClusters(id = 'seurat_clusters', size = 3, color = 'black', repel = F, fontface = 'bold') 

# Clusters split by group plot
full_object_regressed$DimPlot_groups <- DimPlot(seurat, group.by = 'seurat_clusters', pt.size = UMAP_dot_size, label = F, split.by = 'Groups') +
     labs( x = 'UMAP 1', y = 'UMAP 2', title = NULL)  

# Cluster frequency plot
seurat_local <- subset(seurat, subset = Groups != 'CT')
cell_counts <- FetchData(seurat_local, vars = c("seurat_clusters", "Samples", "Groups"))
cell_counts <- arrange(cell_counts, Samples)
counts <- cell_counts %>% 
    add_count(Groups, name='total_cell_count_by_group')  |>
    mutate(weights = total_cell_count_by_group/Abau_cell_number)  |>
    add_count(seurat_clusters, name='total_cell_count_by_cluster', wt =  weights)  |>
    count(seurat_clusters, Samples, Groups, total_cell_count_by_cluster , name='sample_cluster_count', wt = weights)  |> 
    mutate(frequency_within_cluster=sample_cluster_count*100/total_cell_count_by_cluster)  |> 
    mutate(Samples = as.character(Samples)) |> 
    arrange(Samples, desc(Samples)) 
frequency_table <- counts |> 
    arrange(Samples) |>     
    pivot_wider(id_cols = seurat_clusters,  names_from =  Samples , values_from = frequency_within_cluster, values_fill = 0)
# write.csv(frequency_table,file=here(tables_path, paste0( "frequency per roll }} per condition " , ".csv")), row.names=F) 

full_object_regressed$cluster_frequency <- counts |>
    ggplot(aes(x=seurat_clusters, fill=Groups, y = frequency_within_cluster)) +
    geom_col(aes(), position = 'fill') +
    # geom_bar(position=position_fill()) +
    theme_classic() +
    labs(x = 'Cluster', y = 'Frequency (%)', title = 'MAIT normalized cell cluster\nfrequency by sample')+
    scale_y_continuous(expand = expansion(mult = c(0, 0)), breaks = c(0, 0.25, 0.5, 0.75, 1), labels = c(0, 25, 50, 75, 100))+
    scale_fill_discrete(palette = group_colors) 

```

## Top genes cluster and feature plots

```{r}
seurat <- PrepSCTFindMarkers(seurat)
top_genes_per_cluster_results <- top_genes_per_cluster(seurat, n_genes_to_plot = 3, object_annotations = '', tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = 'ClusterProfiler', nterms_to_plot = 3 ) 
full_object_regressed$cluster_dotplot <- top_genes_per_cluster_results$plot 
ggsave(plot = full_object_regressed$cluster_dotplot, filename = here(figures_path,  paste0('DotPlot_Top3_per_cluster', '.pdf')), width = 8, height = 12)
top_genes_per_cluster_results$ClusterProfiler_results

full_object_regressed$cluster_heatmap <- top_genes_per_cluster_results$topn |> pull(gene) |> unique() |> plot_pathways_heatmap(seurat, pathway_name = 'top_3_per_cluster', color_palette = diverging_palette_2, grouping_var = 'seurat_clusters')
full_object_regressed$cluster_signatures <- top_genes_per_cluster_results$ClusterProfiler_results


type_17_markers <- c('Rorc', 'Il17a', 'Sdc1', 'Il7r', 'Il1r1', 'Ccr6', 'Sdc4')
type_1_markers <- c('Tbx21', 'Ifng', 'I12rb2', 'Cxcr3', 'Gzmb', 'Gzmk', 'Klrg1', 'Klrd1', 'Ly6c2')
naive_markers <- c('S1pr1', 'Ccr7', 'Sell', 'Tcf7', 'Lef1')
regulatory_markers <- c('Arg2', 'Il10', 'Il21')
exhaustion_markers <- c('Ctla4', 'Pdcd1', 'Lag3', 'Tigit', 'Havcr2')
tissue_repair_markers <- c('Areg', 'Ramp3','Il18r1', 'Tgfb1', 'Tnf', 'Mmp25', 'Cxcl10', 'Rankl', 'Ifrd1')
transcription_factors <- c('Tbx21', 'Rorc', 'Gata3', 'Foxp3', 'Ahr', 'Maf', 'Prdm1', 'Bcl6', 'Zbtb16')
tissue_residency <- c('Itgae', 'Klrg1', 'Itga1', 'Itga2', 'Ccr9', 'Ccr10', 'Itga4', 'Itgb7')
activation_markers <- c('Cd69', 'Cd44', 'Nr4a1')
cytokines <- c('Ifng', 'Il17a', 'Il17f', 'Il4', 'Il10', 'Tgfb1', 'Tnf', 'Il2', 'Gzma', 'Gzmb', 'Nkg7', 'Ccl5', 'Il27ra', 'Ifgnr1')
costimulatory_receptors <- c('Ctla4', 'Lag3', 'Tigit', 'Havcr2', 'Cd28', 'Cd40lg', 'Pdcd1', 'Tnfsf4', 'Cd27', 'Tnfrsf9')

final_list <- c(type_17_markers, type_1_markers, naive_markers, regulatory_markers, exhaustion_markers, tissue_repair_markers, transcription_factors, tissue_residency, activation_markers, cytokines, costimulatory_receptors) |> unique() |> rev()

# final_gene_list <- 

# Idents(seurat) <- 'seurat_clusters'
# DotPlot_scCustom(seurat, features = final_list, colors_use=sequential_palette_dotplot, flip_axes = T, scale = T, col.min = 0, col.max = 2, dot.scale = 8, dot.min = 0, scale.min = 0, scale.max = 60, x_lab_rotate = T, y_lab_rotate = F)  + theme(axis.text.x = element_text(size = 14), axis.text.y = element_text(size = 14), legend.title = element_text(size = 18))
# ggsave(filename = here(figures_path, 'Dotplot_markers.pdf'), width = 25,height = 5)
# Idents(seurat) <- 'Groups'
# DotPlot_scCustom(seurat, features = final_list, colors_use=sequential_palette_dotplot, flip_axes = F, scale = F, col.min = 0, col.max = 2, dot.scale = 8, dot.min = 0, scale.min = 0, scale.max = 60, x_lab_rotate = T, y_lab_rotate = F)  + theme(axis.text.x = element_text(size = 14), axis.text.y = element_text(size = 14), legend.title = element_text(size = 18))

genes_to_plot <- c('Rorc', 'Il17a', 'Sdc1', 'Il7r', 'Il1r1', 'Ccr6', 'Tbx21', 'Ifng', 'Il12rb2', 'Cxcr3', 'Gzmb', 'Gzmk', 'Gzma', 'Gzmm', 'Klrg1', 'Klrd1', 'Ly6c2', 'Ramp3', 'Slamf7', 'Areg')
full_object_regressed$feature_plots <- genes_to_plot |> 
    set_names() |>
    map(feature_plot )

```

## Signature scores

```{r}
DefaultAssay(seurat) <- 'SCT'
signatures <- read_excel(here('data/Hagoulet_signatures.xlsx'), skip = 1)
signatures <- signatures |>
    dplyr::filter(!(row_number() == 1)) |>
    dplyr::select(-all_of(c('Signature', 'Tissue repair 1', 'Tissue repair 2', 'Tissue repair 3', 'Tissue repair 4')))
    
signatures <- map(.x = signatures, .f = ~ purrr::discard(.x, is.na)) |> as.list()


tissue_repair_linehan <- c('Angpt1', 'Angpt2', 'Angpt4', 'Areg', 'Bmp2', 'Bmp4', 'Bmp7', 'Btc', 'Ccl2', 'Ccl3', 'Chat', 'Csf1', 'Csf2', 'Ctgf', 'Cxcl1', 'Cxcl10', 'Cxcl12', 'Cxcl15', 'Cxcl2', 'Cyr61', 'Defb1', 'Defb6', 'Dhh', 'Disp1', 'Dll1', 'Dll3', 'Dll4', 'Egf', 'Epgn', 'Ereg', 'Fgf1', 'Fgf10', 'Fgf22', 'Fgf7', 'Flg', 'Flg2', 'Furin', 'Hbegf', 'Hgf', 'Hif1a', 'Hmgb1', 'Igf1', 'Igf2', 'Ihh', 'Il1a', 'Il1b', 'Il6', 'Inhba', 'Inhbb', 'Jag1', 'Jag2', 'Lep', 'Mmp10', 'Mmp13', 'Mmp15', 'Mmp2', 'Mmp25', 'Mmp28', 'Mmp3', 'Ngf', 'Nrg1', 'Nrg2', 'Nrg3', 'Nrg4', 'Pdgfa', 'Pdgfb', 'Pdgfc', 'Pgf', 'Ptges2', 'Shh', 'Tgfa', 'Tgfb1', 'Tgfb2', 'Tgfb3', 'Thbs1', 'Tnf', 'Vegfa', 'Vegfb', 'Vegfc', 'Vegfd', 'Wnt1', 'Wnt10a', 'Wnt10b', 'Wnt11', 'Wnt16', 'Wnt2', 'Wnt2b', 'Wnt3', 'Wnt3a', 'Wnt4', 'Wnt5a', 'Wnt5b', 'Wnt6', 'Wnt7a', 'Wnt7b', 'Wnt8a', 'Wnt8b', 'Wnt9a', 'Wnt9b')

signatures$`Tissue repair` <- tissue_repair_linehan

signatures$MAIT17 <- signatures$MAIT17[!(signatures$MAIT17 %in% c('1700040D17Rik'))]

# Removing genes that are not detected
blacklist_elements <- c('H2afz','Smsc2','Hist1h2ap','Flg','Flg2','Bmp2','Bmp4','Ngf','Fgf7','Fgf10','Fgf22','Igf1','Ctgf','Mmp3','Mmp13','Defb1','Defb6','Epgn','Mmp2','Angpt1','Hgf','Cyr61','Inhbb','Btc','Ereg','Lep','Vegfc','Vegfd','Pdgfc','Cxcl12','Shh','Dhh','Ihh','Dll3','Dll4','Wnt1','Wnt2','Wnt3a','Wnt5a','Wnt6','Wnt7a','Wnt7b','Wnt8a','Wnt8b','Wnt9a','Wnt9b','Wnt10a','Wnt16','Chat','Thbs1','Fam46a','Kcna4','Fam129b','Tdpoz4','Rassf6','Ero1l','Cd244','A430078G23Rik','D1Ertd622e','Fam49a')

signatures <- map(.x = signatures, .f = ~ .x[!(.x %in% blacklist_elements)])

seurat <- AddModuleScore_UCell(seurat, features = signatures, name = NULL)

# Scale signatures (z-score) for comparison and visualization purposes
seurat@meta.data  <-  seurat@meta.data |>
    mutate(across(names(signatures), ~ scale(.x, center = T, scale = T), .names = '{col}_scaled'))
 # Visualization
signature_violin_plot <- function (signature) {
    max_y <- max(seurat@meta.data |> pull(signature))
    plot1 <- VlnPlot(seurat, features = signature, group.by = 'Groups', pt.size = 0, cols = group_colors) +
                labs(title = signature, x = NULL) + NoLegend() +
                ylim(0, max_y*1.65) +
                ggpubr::stat_compare_means(comparisons = list(c('Abau', 'Ftula'), c('Abau', 'CT'), c('Ftula', 'CT')), label = 'p.signif', method = 'wilcox.test',  hide.ns = F, label.y = max_y*1.1 , step.increase = 0.18, bracket.size = 0.7, size = 3) 
    ggsave(plot = plot1, filename = paste0('VlnPlot_', signature, '_by_group_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 4)
    return(plot1)
}

signature_dot_plot <- function (signature) {
    plot1 <- DotPlot_scCustom(seurat, features = signatures[[signature]] |> rev() |> unique() , group.by = 'Groups', colors_use = sequential_palette_dotplot, flip_axes = F, scale = F, dot.scale = 10, x_lab_rotate = T) + labs(title = signature)
    ggsave(plot = plot1, filename = paste0('DotPlot_', signature, '_by_groups_', object_annotations, '.pdf'), path = figures_path, width = 5, height = length(signatures[[signature]])/4 + 1)
    return(plot1)
}
signature_feature_plot <- function (signature) {
    min_value  <-  min(seurat@meta.data |> pull(paste0(signature, '_scaled')))
    max_value  <-  max(seurat@meta.data |> pull(paste0(signature, '_scaled')))
    cut_off_value <- min(abs(min_value), abs(max_value))
    plot1 <- FeaturePlot_scCustom(seurat, features = paste0(signature, '_scaled'), colors_use = diverging_palette_3, na_cutoff=NULL, min.cutoff = -1*cut_off_value, max.cutoff = cut_off_value, pt.size = 0.1, alpha_exp = 0.2) + labs(title = signature, x = 'UMAP 1', y = 'UMAP 2')
    ggsave(plot = plot1, filename = paste0('FeaturePlot_', signature, '_', object_annotations, '.pdf'), path = figures_path, width = 5, height = 5)
    return(plot1)
}
signature_heatmap <- function (signature) {
    plot1 <- signatures[[signature]] |> unique() |> plot_pathways_heatmap(seurat, pathway_name = signature, color_palette = diverging_palette_2, grouping_var =  'Samples' ) +
     labs(title = signature)
    ggsave(plot = plot1, filename = paste0('Heatmap_', signature, '_by_samples_', object_annotations, '.pdf'), path = figures_path, width = 8, height = 6)
    return(plot1)
}
signature_heatmap2 <- function (signature) {
    plot1 <- signatures[[signature]] |> unique() |> plot_pathways_heatmap2(seurat, pathway_name = signature, color_palette = diverging_palette_2, grouping_var =  'Samples' ) +
     labs(title = signature)
    ggsave(plot = plot1, filename = paste0('Heatmap_', signature, '_by_samples_', object_annotations, '.pdf'), path = figures_path, width = 8, height = 6)
    return(plot1)
}


```

```{r}

full_object_regressed$signature_violin_plots <- map(names(signatures), signature_violin_plot)
names(full_object_regressed$signature_violin_plots) <- names(signatures)

full_object_regressed$signature_dot_plots <- map(names(signatures), signature_dot_plot)
names(full_object_regressed$signature_dot_plots) <- names(signatures)

full_object_regressed$signature_feature_plots <- map(names(signatures), signature_feature_plot)
names(full_object_regressed$signature_feature_plots) <- names(signatures)

full_object_regressed$signature_heatmaps <- map(names(signatures), signature_heatmap)
names(full_object_regressed$signature_heatmaps) <- names(signatures)

full_object_regressed$signature_heatmaps2 <- map(names(signatures), signature_heatmap2)
names(full_object_regressed$signature_heatmaps2) <- names(signatures)

full_object_regressed$signature_heatmaps2 |> walk(print)

cut_off_value <- 1.5
signature_cycling_features <- list()
signature_cycling_features[[1]] <-
FeaturePlot_scCustom(seurat, features = paste0('Cycling_G2M', '_scaled'), colors_use = diverging_palette_2, na_cutoff=NULL, min.cutoff = -1*cut_off_value, max.cutoff = cut_off_value, pt.size = 0.1, alpha_exp = 0.3) + labs(title = 'Cycling_G2M')
signature_cycling_features[[2]] <- FeaturePlot_scCustom(seurat, features = paste0('Cycling_S', '_scaled'), colors_use = diverging_palette_2, na_cutoff=NULL, min.cutoff = -1*cut_off_value, max.cutoff = cut_off_value, pt.size = 0.1, alpha_exp = 0.3) + labs(title = 'Cycling_S')


```


## Milo

```{r}
library(miloR)
library(SingleCellExperiment)
library(scran)
library(scater)

seurat <- seurat_cell_cycle

# Convert to SingleCellExperiment
Idents(seurat) <- 'Samples'
DefaultAssay(seurat) <- 'RNA'
seurat_sce <- as.SingleCellExperiment(seurat)
reducedDim(seurat_sce, 'PCA') <- Embeddings(seurat, 'pca')
reducedDim(seurat_sce, 'UMAP') <- Embeddings(seurat, 'umap')
seurat_milo <- Milo(seurat_sce)
k_used <- 30
dimensions <- 30

# Running MiloR
trajectory_milo <- buildGraph(seurat_milo, k = k_used, d = dimensions, reduced.dim = 'PCA')
trajectory_milo <- makeNhoods(trajectory_milo, prop = 0.1, k = k_used, d = dimensions, refined = T, reduced_dims = 'PCA')

# Count cells in neighborhoods per sample
trajectory_milo <- countCells(trajectory_milo, sample = 'Samples', meta.data = data.frame(colData(seurat_milo)))

# DA testing preparation
trajectory_design <- data.frame(colData(trajectory_milo))[, c('Samples', 'Groups')]
trajectory_design <- distinct(trajectory_design)
rownames(trajectory_design) <- trajectory_design$Samples
trajectory_design <- trajectory_design[colnames(nhoodCounts(trajectory_milo)), , drop = FALSE ]
trajectory_milo <- calcNhoodDistance(trajectory_milo, d = dimensions, reduced.dim = 'PCA')
# trajectory_milo <- buildNhoodGraph(trajectory_milo)


# Pairwise DA testing using contrasts Abau vs Ftula
contrast <- c('GroupsAbau - GroupsFtula')
da_results_Abau_vs_Ftula <- testNhoods(trajectory_milo, design = ~ 0 + Groups, design.df = trajectory_design, model.contrasts = contrast, fdr.weighting = 'graph-overlap', norm.method = 'TMM')
da_results_Abau_vs_Ftula |> arrange(SpatialFDR) |> head()

da_results_Abau_vs_Ftula <- annotateNhoods(trajectory_milo, da_results_Abau_vs_Ftula, coldata_col = 'seurat_clusters')
da_results_Abau_vs_Ftula <- da_results_Abau_vs_Ftula |> mutate(seurat_clusters = factor(seurat_clusters, levels = levels(seurat$seurat_clusters) |> rev()))

### Neighborhood differential gene expression analysis
da_results_Abau_vs_Ftula <- annotateNhoods(trajectory_milo, da_results_Abau_vs_Ftula, coldata_col = 'seurat_clusters')
trajectory_milo <- logNormCounts(trajectory_milo)

# # manual grouping of neighborhoods
# FCutoff <- 3
# da_results_Abau_vs_Ftula |> 
#     ggplot(aes(x = logFC)) +
#     geom_histogram(bins = 100) +
#     geom_density() +
#     theme_classic()
# da_results_Abau_vs_Ftula <- da_results_Abau_vs_Ftula |>
#     mutate(NhoodGroup = case_when(logFC > 0 & logFC < FCutoff & SpatialFDR < 0.05 ~ 'Abau_enriched_log2FC_0-5',
#                             logFC >= FCutoff  & SpatialFDR < 0.05 ~ 'Abau_enriched_log2FC_5',
#                             logFC < 0 & logFC > -FCutoff & SpatialFDR < 0.05 ~ 'Ftula_enriched_log2FC_0-5',
#                             logFC <= -FCutoff  & SpatialFDR < 0.05 ~ 'Ftula_enriched_log2FC_5',
#                             TRUE ~ 'Not_significant')) 
# # plotDAbeeswarm(da_results_Abau_vs_Ftula, group.by = 'NhoodGroup', alpha = 0.05)                            
# # plotNhoodGroups(trajectory_milo, da_results_Abau_vs_Ftula, layout = 'UMAP', alpha = 0.05)
# nhood_markers <- findNhoodGroupMarkers(trajectory_milo, da_results_Abau_vs_Ftula, aggregate.samples = F, sample_col = 'Samples')
# nhood_markers |> pivot_longer(cols = starts_with('adj'), names_to = 'neighborhood', values_to = 'adj_pval', names_prefix = 'adj.P.Val_')   |>
#     pivot_longer(cols = starts_with('log'), names_to = 'neighborhood2', values_to = 'logFC', names_prefix = 'logFC_')    |> filter(neighborhood == neighborhood2)
# nhood_markers <- nhood_markers |>
#     as_tibble() |>
#     pivot_longer(cols = starts_with('adj'), names_to = 'neighborhood', values_to = 'adj_pval', names_prefix = 'adj.P.Val_')  |>
#     pivot_longer(cols = starts_with('log'), names_to = 'neighborhood2', values_to = 'logFC', names_prefix = 'logFC_')   |>
#     filter(neighborhood == neighborhood2)     |>
#     select(-neighborhood2) |>
#     filter(adj_pval < 0.05 & logFC > 0.2) |> 
#     group_by(neighborhood) |>
#     slice_max(order_by = logFC, n = 25) |>
#     ungroup() |>
 #     arrange(neighborhood, desc(logFC))
# #Topn markers
# nhood_markers |> 
#     group_by(neighborhood)  |> 
#     arrange(desc(logFC)) |>
#     slice_head(n = 10) -> topn

trajectory_milo <- buildNhoodGraph(trajectory_milo)

da_results_Abau_vs_Ftula <- groupNhoods(trajectory_milo, da_results_Abau_vs_Ftula, max.lfc.delta = NULL, overlap = 2, da.fdr = 0.05, compute.new = T, original.behaviour = F )
plotDAbeeswarm(da_results_Abau_vs_Ftula, group.by = 'NhoodGroup', alpha = 0.05)
plotNhoodGroups(trajectory_milo, da_results_Abau_vs_Ftula, layout = 'UMAP', alpha = 0.05) 

# saveRDS(da_results_Abau_vs_Ftula, file = here(results_path, 'MiloR_da_results_Abau_vs_Ftula_neighborhood_groups.rds'))
# saveRDS(trajectory_milo, file = here(results_path, 'MiloR_trajectory_milo_neighborhood_groups.rds'))

da_results_Abau_vs_Ftula <- readRDS(file = here('data', 'MiloR_da_results_Abau_vs_Ftula_neighborhood_groups.rds'))
trajectory_milo <- readRDS(file = here('data', 'MiloR_trajectory_milo_neighborhood_groups.rds'))

nhood_markers <- findNhoodGroupMarkers(trajectory_milo, da_results_Abau_vs_Ftula, aggregate.samples = F)
nhood_markers |> pivot_longer(cols = starts_with('adj'), names_to = 'neighborhood', values_to = 'adj_pval', names_prefix = 'adj.P.Val_')   |>
    pivot_longer(cols = starts_with('log'), names_to = 'neighborhood2', values_to = 'logFC', names_prefix = 'logFC_')    |> filter(neighborhood == neighborhood2)
nhood_markers <- nhood_markers |>
    as_tibble() |>
    pivot_longer(cols = starts_with('adj'), names_to = 'neighborhood', values_to = 'adj_pval', names_prefix = 'adj.P.Val_')  |>
    pivot_longer(cols = starts_with('log'), names_to = 'neighborhood2', values_to = 'logFC', names_prefix = 'logFC_')   |>
    filter(neighborhood == neighborhood2)     |>
    select(-neighborhood2) |>
    filter(adj_pval < 0.05 & logFC > 0.2) |> 
    group_by(neighborhood) |>
    slice_max(order_by = logFC, n = 100) |> ## How many genes to include is setup here!!!
    ungroup() |>
    arrange(neighborhood, desc(logFC)) |>
    mutate(neighborhood = fct_inseq(neighborhood))
#Topn markers
nhood_markers |> 
    group_by(neighborhood)  |> 
    arrange(desc(logFC)) |>
    slice_head(n = 4) -> topn


```

```{r}

# pathway_enrichment <- c('Metascape', 'ClusterProfiler')
# Run pathway enrichment analysis
local_path_pathway_enrichment <- here(results_path, paste0('MiloR', '_neighborhood_identification_functional_analysis'))
unlink(local_path_pathway_enrichment, recursive = T)
dir.create(local_path_pathway_enrichment)
    if (!is.null(pathway_enrichment)) {
        if ('Metascape' %in% pathway_enrichment) {
            plot_list <- list()
            for (neighborhood in levels(nhood_markers$neighborhood)) {
                local_path_2 <- here(local_path_pathway_enrichment, neighborhood)
                dir.create(local_path_2, recursive = TRUE, showWarnings = FALSE)
                name <- paste0('Neighborhood_', neighborhood)
                significant_results <- nhood_markers |> 
                    filter(neighborhood == {{neighborhood}}) |>
                    pull(GeneID) |>
                    unique()
                    if (length(significant_results) < 3) {
                    next
                        }
                plot_list[[neighborhood]] <- Metascape_overrepresentation_analysis(significant_results,
                            local_path = local_path_2,
                            group = neighborhood,
                            filename = name,
                            grouping_var = 'Neighborhood',
                            nterms_to_plot_metascape = 5)
                }
            plots <- wrap_plots(plot_list) & 
            theme(plot.title = element_text(size = 9, hjust = 0.5),
                axis.text.y = element_text(size = 8),
                axis.text.x = element_text(size = 8),
                axis.title.x  = element_text(size = 8),
            axis.title.y  = element_text(size = 7))
        print(plots)
            }
            colnames(nhood_markers)
if ('ClusterProfiler' %in% pathway_enrichment) {
            all_genes <- Features(seurat[['RNA']]) |>unique()
            gene_lists <- nhood_markers %>%
                group_by(neighborhood) %>%
                arrange(desc(logFC)) |>
                slice_head(n = 100)  |>
                ungroup() |>
                select(neighborhood, GeneID)  |>
                group_by(neighborhood) |>
                mutate(id = row_number()) |>
                ungroup() |>
                pivot_wider(names_from = neighborhood, values_from = GeneID)  |>
                select(-id) |>
                as.list() |>
                map(~ .x[!is.na(.x)])
            overrepresentation_results <- GO_overrepresentation_analysis_multiple_lists(gene_lists, all_genes, local_path_pathway_enrichment, ontology = 'ALL', minGSSize = 5, maxGSSize = 400, filename = '',  drop_levels = F, levels_to_drop = c(1), simplify_function = min, simplify_by = 'p.adjust', simplify_terms = F, run_network = F, network_n_terms = 100, nterms_to_plot = 3, font_size = 10, grouping_var = 'Neighborhood')
  }  }
full_object_regressed$Milo_figures <- list()
full_object_regressed$Milo_figures$pathways <- overrepresentation_results$plot

```

```{r}
da_results_Abau_vs_Ftula <- da_results_Abau_vs_Ftula |>
    mutate(NhoodGroup = fct_inseq(NhoodGroup))
    # mutate(NhoodGroup = as.character(NhoodGroup))
da_results_Abau_vs_Ftula |> colnames()
full_object_regressed$Milo_figures$pathways_genes <-plotNhoodExpressionGroups(trajectory_milo, da_results_Abau_vs_Ftula,
                          features = topn$GeneID |> unique() |> rev(),
                          scale = T,
                        grid.space = 'fixed'
                          )
da_results_Abau_vs_Ftula <- da_results_Abau_vs_Ftula |>
    # mutate(NhoodGroup = fct_inseq(NhoodGroup))
    mutate(NhoodGroup = as.character(NhoodGroup))
full_object_regressed$Milo_figures$pathways_genes
    
```

### Milo figures

```{r}
#Visualize Neighborhoods
# full_object_regressed$Milo_figures <- list()
full_object_regressed$Milo_figures[['neighborhood_graph']] <- plotNhoodGraphDA(trajectory_milo, da_results_Abau_vs_Ftula, layout = 'UMAP', alpha = 0.1) + 
     scale_fill_continuous(palette = diverging_palette_2) +
     theme_classic() + 
     labs(title = 'MiloR: Abau vs Ftula', fill = 'log2FC\n(Abau/Ftula)', size = 'Neighborhood size', x = 'UMAP 1', y = 'UMAP 2')+ 
    guides(edge_width = 'none', size = 'none')
ggsave(plot = full_object_regressed$Milo_figures[['neighborhood_graph']], filename = 'MiloR_Neighborhoods_DA_Abau_vs_Ftula.pdf', path = figures_path, width = 5, height = 5)

full_object_regressed$Milo_figures[['beeswarm_graph']] <- plotDAbeeswarm(da_results_Abau_vs_Ftula, group.by = 'seurat_clusters', alpha = 0.05)
da.res <- full_object_regressed$Milo_figures[['beeswarm_graph']]@data
n_groups <- length(levels(da.res$group_by))
full_object_regressed$Milo_figures[['beeswarm_graph']] <- da.res |> 
    ggplot(aes(pos_x, pos_y, color=logFC_color)) +
    scale_color_gradient2() +
    guides(color="none") +
    xlab('seurat_clusters') + ylab("Log Fold Change") +
    scale_x_continuous(
      breaks = seq(1,n_groups),
      labels = setNames(levels(da.res$group_by), seq(1,n_groups))
      ) +
    geom_point(size = .5) +
    coord_flip() +
    theme_bw(base_size=22) +
    theme(strip.text.y =  element_text(angle=0)) +
    scale_color_continuous(palette = diverging_palette_2) +
    labs(x = 'Leiden clusters', y = 'log2FC(Abau/Ftula)', color = 'log2FC\n(Abau/Ftula)')
ggsave(plot = full_object_regressed$Milo_figures[['beeswarm_graph']], filename = 'MiloR_Beeswarm_DA_Abau_vs_Ftula.pdf', path = figures_path, width = 7, height = 5)

#Visualize Neighborhoods
full_object_regressed$Milo_figures[['neighborhood_cluster_graph']] <- plotNhoodGroups(trajectory_milo, da_results_Abau_vs_Ftula, layout = 'UMAP', alpha = 0.1) + 
    #  scale_fill_continuous(palette = diverging_palette_2) +
     theme_classic() + 
     labs(title = 'MiloR: Abau vs Ftula', size = 'Neighborhood size', x = 'UMAP 1', y = 'UMAP 2')+ 
    guides(edge_width = 'none', size = 'none')
ggsave(plot = full_object_regressed$Milo_figures[['neighborhood_cluster_graph']], filename = 'MiloR_Neighborhoods_Clusters_DA_Abau_vs_Ftula.pdf', path = figures_path, width = 5, height = 5)

da_results_Abau_vs_Ftula <- da_results_Abau_vs_Ftula |>
    mutate(NhoodGroup = fct_inseq(NhoodGroup))
full_object_regressed$Milo_figures[['beeswarm_graph_neighborhood_cluster']] <- plotDAbeeswarm(da_results_Abau_vs_Ftula, group.by = 'NhoodGroup',  alpha = 0.05)
da.res <- full_object_regressed$Milo_figures[['beeswarm_graph_neighborhood_cluster']]@data
n_groups <- length(levels(da.res$group_by))
full_object_regressed$Milo_figures[['beeswarm_graph_neighborhood_cluster']] <- da.res |> 
    ggplot(aes(pos_x, pos_y, color=logFC_color)) +
    scale_color_gradient2() +
    guides(color="none") +
    xlab('seurat_clusters') + ylab("Log Fold Change") +
    scale_x_continuous(
      breaks = seq(1,n_groups),
      labels = setNames(levels(da.res$group_by), seq(1,n_groups))
      ) +
    geom_point(size = .5) +
    coord_flip() +
    theme_bw(base_size=22) +
    theme(strip.text.y =  element_text(angle=0)) +
    scale_color_continuous(palette = diverging_palette_2) +
    labs(x = 'Neighborhood clusters', y = 'log2FC(Abau/Ftula)', color = 'log2FC\n(Abau/Ftula)')
ggsave(plot = full_object_regressed$Milo_figures[['beeswarm_graph_neighborhood_cluster']], filename = 'MiloR_Beeswarm_NhoodClusters_DA_Abau_vs_Ftula.pdf', path = figures_path, width = 7, height = 5)


# gene_list_plot <- topn |> pull(GeneID) |> unique() |> rev()
# Idents(seurat) <- 'seurat_clusters'
# DotPlot_scCustom(seurat,
#                 features = gene_list_plot,
#                 colors_use=sequential_palette_dotplot,
#                 flip_axes = T,
#                 dot.scale = 8,
#                 dot.min = 0,
#                 scale.min = 0,
#                 scale.max = 80,
#                 x_lab_rotate = T,
#                 y_lab_rotate = F) +
#     theme(axis.text.x = element_text(size = 14),
#         axis.text.y = element_text(size = 14),
#         legend.title = element_text(size = 18))

# local_path <- here(results_path, pcaste0('MiloR', '_neighborhood_identification_functional_analysis'))
# unlink(local_path, recursive = T)
# dir.create(local_path)
# for (neighborhood in unique(nhood_markers$neighborhood)) {
#      name <- paste0('Neighborhood_', neighborhood)
#      significant_results <- nhood_markers |> 
#           filter(neighborhood == {{neighborhood}}) |>
#           pull(GeneID) |>
#           unique()
#         if (length(significant_results) < 3) {
#           next
#               }
#   Metascape_overrepresentation_analysis(significant_genes_FC_ordered = significant_results, path = local_path, group = '', filename =  name,  grouping_var =  neighborhood )
#   }

full_object_regressed$Milo_figures |> walk(print)


```


## Differential gene expression analysis

```{r}
#| fig-width: 7
#| #| fig-height: 7
#| layout-ncol: 2
#| results: hide
seurat <- seurat_original
source(here('scripts/metascape_functions.r'))
# source(here('scripts/gProfiler2_functions.r'))
source(here('scripts/function_template.r'))
source(here('scripts/ClusterProfiler_functions.r'))

path <- here(results_path, 'DEG_pseudobulk_analysis_Abau_vs_Ftula')
unlink(path, recursive = T)
dir.create(path)
Idents(seurat) <- 'seurat_clusters'

seurat@meta.data <- seurat@meta.data |>
    mutate(Samples = str_remove(Samples, '_Mouse'))
# levels(seurat$Samples)
     
genes_to_label <- c('Sdc1', 'Zbtb7c', 'Acvr2a', 'Mmp25', 'Il7r', 'Cd164', 'Nr1d1', 'Ccr6', 'Hlf', 'Rorc', 'Ramp3', 'Mgat', 'Il1r1', 'Itgae', 'Tbx21', 'Gzma', 'Klrg1', 'Ifng', 'Gzmk', 'Lag3', 'Il1rl1', 'Plac8', 'Ifitm3', 'Batf3', 'Irf8', 'Bcat1') |> unique()
# genes_to_label <- list('Genes_of_interest' = genes_to_label)
# scRNAseq, comparison, group1, group2, cluster = 'all_clusters', 
#                        path = './', FC_threshold = 0.3, p_value_threshold = 0.05, genes_to_plot = NULL, 
#                        expression_threshold_for_gene_list = 20, minimum_cell_number = 10, 
#                        run_pathway_enrichment = NULL, genes_to_exclude = c(), ...

output <- pseudobulk(seurat, comparison='Samples', group1='Ftula', group2='Abau', cluster= 'all_MAITs' , path=path, label_threshold = 1000, distance_from_diagonal_threshold=0.5, max_overlaps = 15, my_colors =c('#d55e10','#0072b2', 'grey'), genes_to_plot = NULL, FC_threshold = 1, p_value_threshold = 0.05, run_pathway_enrichment = pathway_enrichment, minGSSize = 4, maxGSSize = , drop_levels = F, levels_to_drop = c(1), ontology = 'BP', simplify_terms=T, run_network=T, network_n_terms=70)


source(here('scripts/function_template.r'))
full_object_original$volcano_plot <- volcano_plot(output$results, comparison='Samples', group1='Ftula', group2='Abau', cluster= 'all_MAITs' , local_figures_path=path, label_size = 3, label_threshold = 1000, distance_from_diagonal_threshold=0.5, max_overlaps = 4, my_colors =c('#d55e10','#0072b2', 'grey'), genes_to_plot = genes_to_label, FC_threshold = 1, p_value_threshold = 0.05, run_pathway_enrichment = NULL, minGSSize = 4, maxGSSize = , drop_levels = F, levels_to_drop = c(1), ontology = 'BP', simplify_terms=T, run_network=T, network_n_terms=70, test_type = 'Pseudobulk', pt_size = 1) 
full_object_original$volcano_plot

full_object_original$scatter_plot <- scatterplot(output$results, comparison='Samples', group1='Ftula', group2='Abau', cluster= 'all_MAITs' , local_figures_path=path, label_size = 3, label_threshold = 1000, distance_from_diagonal_threshold=0.5, max_overlaps = 4, my_colors =c('#d55e10','#0072b2', 'grey'), genes_to_plot = genes_to_label, FC_threshold = 1, p_value_threshold = 0.05, run_pathway_enrichment = NULL, minGSSize = 4, maxGSSize = , drop_levels = F, levels_to_drop = c(1), ontology = 'BP', simplify_terms=T, run_network=T, network_n_terms=70, test_type = 'Pseudobulk', pt_size = 1)
full_object_original$scatter_plot

# results <- output$results
# GO_functional_analysis(results,  grouping_var = 'all_MAITs', path= path, FC_threshold = 1, p_value_threshold = 0.05, group1 = 'Ftula', group2 = 'Abau', minGSSize = 4, maxGSSize = 400, drop_levels = F, levels_to_drop = c(), ontology = 'BP', simplify_terms=T, run_network=T, network_n_terms=200)

# UP in Abau all MAITs
GO_results <- read_csv(here('results/Manuscript_figures/DEG_pseudobulk_analysis_Abau_vs_Ftula/GO_functional_analysis_UP_in_Abau_all_MAITs/GO_OverRepresentation_analysis_results_filtered_BP.csv')) |> select(-...1) |> mutate(GeneRatio = as.numeric(str_split_fixed(GeneRatio, '/', 2)[,1]) / as.numeric(str_split_fixed(GeneRatio, '/', 2)[,2]))
pathways_to_plot <- GO_results |>
    arrange(desc(GeneRatio)) |>
    slice_head(n = 10) |> 
    pull(Description)
pathways_to_replace <- c('regulation of hemopoiesis' = 'alpha-beta T cell differentiation')
pathways_to_plot <- pathways_to_plot |> recode(!!!pathways_to_replace)
pathways_to_plot
full_object_original$pathways_Abau <-  GO_results |>
    filter(Description %in% pathways_to_plot) |>
    ggplot(aes(fct_reorder(Description, -log10(p.adjust)), -log10(p.adjust) , fill = -log10(p.adjust))) +
    geom_col(width = 0.5) +
    # scale_fill_continuous(palette = sequential_palette ) +
    labs(title = 'Representative top enriched \nGO terms in Abau MAITs', x = NULL, y = '-log10(Adjusted p-value)') +
    guides(fill = 'none') +
    theme_classic() +
    scale_y_continuous(expand = expansion(mult = c(0.01, 0.1)))+ 
    # scale_x_discrete(labels = label_wrap(22))+
    coord_flip() 
full_object_original$pathways_Abau

t_cell_activation_terms <- c('alpha-beta T cell differentiation', 'positive regulation of cell adhesion')
wound_repair_terms <- c('response to wounding', 'morphogenesis of an epithelium')

# t_cell_activation_terms
genes_to_plot <- GO_results |>
    filter(Description %in% t_cell_activation_terms) |>
    pull(geneID) |>
    str_split(pattern = '/') |>
    unlist() |>
    unique() |>
    head(25) 
full_object_original$pathways_Abau_Tcelldifferentiation <- plot_pathways_heatmap(genes_to_plot, seurat |> subset(subset = Groups %in% c('CT'), invert = T), 'T cell activation terms', color_palette = diverging_palette_2)
full_object_original$pathways_Abau_Tcelldifferentiation

# wound_repair_terms
genes_to_plot <- GO_results |>
    filter(Description %in% wound_repair_terms) |>
    pull(geneID) |>
    str_split(pattern = '/') |>
    unlist() |>
    unique() |>
    head(25)  |>
    c('Areg')
full_object_original$pathways_Abau_woundrepair <- plot_pathways_heatmap(genes_to_plot, seurat |> subset(subset = Groups %in% c('CT'), invert = T), 'Wound repair terms', color_palette = diverging_palette_2)
full_object_original$pathways_Abau_woundrepair


# UP in Ftula all MAITs
GO_results <- read_csv(here('results/Manuscript_figures/DEG_pseudobulk_analysis_Abau_vs_Ftula/GO_functional_analysis_UP_in_Ftula_all_MAITs/GO_OverRepresentation_analysis_results_filtered_BP.csv')) |> select(-...1) |> mutate(GeneRatio = as.numeric(str_split_fixed(GeneRatio, '/', 2)[,1]) / as.numeric(str_split_fixed(GeneRatio, '/', 2)[,2]))

pathways_to_plot <- GO_results |>
    arrange(desc(GeneRatio)) |>
    slice_head(n = 20) |> 
    pull(Description)
pathways_to_remove <- c('nuclear chromosome segregation', 'nuclear division', 'DNA recombination', 'meiotic cell cycle process', 'mononuclear cell proliferation', 'defense response to virus', 'positive regulation of response to biotic stimulus', 'regulation of chromosome organization', 'cell cycle G1/S phase transition')
pathways_to_plot <- setdiff(pathways_to_plot, pathways_to_remove) |> c('cell killing')
full_object_original$pathways_Ftula <-  GO_results |>
    filter(Description %in% pathways_to_plot) |>
    ggplot(aes(fct_reorder(Description,  -log10(p.adjust) ),  -log10(p.adjust) , fill = -log10(p.adjust))) +
    geom_col(width = 0.5) +
    # scale_fill_continuous(palette = sequential_palette ) +
    labs(title = 'Representative top enriched\nGO terms in Ftula MAITs', x = NULL, y = '-log10(Adjusted p-value)') +
    guides(fill = 'none') +
    theme_classic() +
    scale_y_continuous(expand = expansion(mult = c(0.01, 0.1)))+ 
    # scale_x_discrete(labels = label_wrap(22))+
    coord_flip() 
full_object_original$pathways_Ftula

cytotoxicity_terms <- c('cell killing', 'lymphocyte mediated immunity', 'response to virus')

#cytotoxicity_terms
genes_to_plot <- GO_results |>
    filter(Description %in% cytotoxicity_terms) |>
    pull(geneID) |>
    str_split(pattern = '/') |>
    unlist() |>
    unique() |>
    head(30) 
full_object_original$pathways_Ftula_cytotoxicity <- plot_pathways_heatmap(genes_to_plot,  seurat |> subset(subset = Groups %in% c('CT'), invert = T), 'Cytotoxicity terms', color_palette = diverging_palette_2) 
full_object_original$pathways_Ftula_cytotoxicity

```

# Generate plots for Ftula object

```{r}
seurat <- seurat_Ftula
seurat@meta.data <- seurat@meta.data |>
    mutate(Samples = str_remove(Samples, '_Mouse')) 
Idents(seurat) <- 'seurat_clusters'

# Clusters UMAP plot
Ftula_object$DimPlot <- DimPlot(seurat, group.by = 'seurat_clusters', pt.size = UMAP_dot_size, label = F) +
     labs(title = 'MAIT cell Leiden clusters', 
        #   subtitle = 'Lung single-cell RNA-seq', 
        #   caption = 'n = 4 Abau, n = 5 Ftula, n = 4 CT',
     x = 'UMAP 1', y = 'UMAP 2')  
Ftula_object$DimPlot <- Ftula_object$DimPlot  |> LabelClusters(id = 'seurat_clusters', size = 3, color = 'black', repel = F, fontface = 'bold')

seurat <- PrepSCTFindMarkers(seurat)
top_genes_per_cluster_results <- top_genes_per_cluster(seurat, n_genes_to_plot = 3, object_annotations = '', tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = 'ClusterProfiler', nterms_to_plot = 3 ) 
Ftula_object$cluster_dotplot <- top_genes_per_cluster_results$plot 
ggsave(plot = Ftula_object$cluster_dotplot, filename = here(figures_path,  paste0('DotPlot_Top3_per_cluster', '.pdf')), width = 8, height = 12)
top_genes_per_cluster_results$ClusterProfiler_results

Ftula_object$cluster_heatmap <- top_genes_per_cluster_results$topn |> pull(gene) |> unique() |> plot_pathways_heatmap(seurat, pathway_name = 'top_3_per_cluster', color_palette = diverging_palette_2, grouping_var = 'seurat_clusters')
Ftula_object$cluster_signatures <- top_genes_per_cluster_results$ClusterProfiler_results

genes_to_plot <- c('Rorc', 'Il17a', 'Sdc1', 'Il7r', 'Il1r1', 'Ccr6', 'Tbx21', 'Ifng', 'Il12rb2', 'Cxcr3', 'Gzmb', 'Gzmk', 'Gzma', 'Gzmm', 'Klrg1', 'Klrd1', 'Ly6c2', 'Ramp3', 'Slamf7', 'Areg')
Ftula_object$feature_plots <- genes_to_plot |> 
    set_names() |>
    map(feature_plot )
```

# Generate plots for Abau object

```{r}
seurat <- seurat_Abau
seurat@meta.data <- seurat@meta.data |>
    mutate(Samples = str_remove(Samples, '_Mouse')) 
Idents(seurat) <- 'seurat_clusters'

# Clusters UMAP plot
Abau_object$DimPlot <- DimPlot(seurat, group.by = 'seurat_clusters', pt.size = UMAP_dot_size, label = F) +
     labs(title = 'MAIT cell Leiden clusters', 
        #   subtitle = 'Lung single-cell RNA-seq', 
        #   caption = 'n = 4 Abau, n = 5 Ftula, n = 4 CT',
     x = 'UMAP 1', y = 'UMAP 2')  
Abau_object$DimPlot <- Abau_object$DimPlot  |> LabelClusters(id = 'seurat_clusters', size = 3, color = 'black', repel = F, fontface = 'bold')

seurat <- PrepSCTFindMarkers(seurat)
top_genes_per_cluster_results <- top_genes_per_cluster(seurat, n_genes_to_plot = 3, object_annotations = '', tables_path = tables_path, figures_path = figures_path, results_path = results_path, run_pathway_enrichment = 'ClusterProfiler', nterms_to_plot = 3 ) 
Abau_object$cluster_dotplot <- top_genes_per_cluster_results$plot 
ggsave(plot = Abau_object$cluster_dotplot, filename = here(figures_path,  paste0('DotPlot_Top3_per_cluster', '.pdf')), width = 8, height = 12)
top_genes_per_cluster_results$ClusterProfiler_results

Abau_object$cluster_heatmap <- top_genes_per_cluster_results$topn |> pull(gene) |> unique() |> plot_pathways_heatmap(seurat, pathway_name = 'top_3_per_cluster', color_palette = diverging_palette_2, grouping_var = 'seurat_clusters')
Abau_object$cluster_signatures <- top_genes_per_cluster_results$ClusterProfiler_results

genes_to_plot <- c('Rorc', 'Il17a', 'Sdc1', 'Il7r', 'Il1r1', 'Ccr6', 'Tbx21', 'Ifng', 'Il12rb2', 'Cxcr3', 'Gzmb', 'Gzmk', 'Gzma', 'Gzmm', 'Klrg1', 'Klrd1', 'Ly6c2', 'Ramp3', 'Slamf7', 'Areg')
Abau_object$feature_plots <- genes_to_plot |> 
    set_names() |>
    map(feature_plot )
```

# cNMF results

```{r}
seurat <- seurat_cell_cycle
seurat@meta.data <- seurat@meta.data |>
    mutate(Samples = str_remove(Samples, '_Mouse'))
# colnames(seurat@meta.data)
```

```{r}
#| fig-width: 10
#| fig-height: 10
library(Matrix)
source(here('scripts/function_template.r'))
local_path <- here('results', 'cNMF')
# unlink(local_path, recursive = T)
dir.create(local_path)
local_path_input <- here('results', 'cNMF','input')
# unlink(local_path_input, recursive = T)
dir.create(local_path_input)
local_path_input

filtered_dir  <- local_path_input
data_dir <- local_path
runname = "cNMF_run"
cnmf_results <- run_cnmf_results(
    seurat,                    # Seurat object
    data_dir,                  # path to directory containing cNMF run folder (string)
    runname,                   # cNMF run name (string)
    k_used = 15,
    local_density_threshold = 0.1,
    do_cnmf = TRUE,            # whether to run system cnmf command
    sequential_palette = sequential_palette, # palette for FeaturePlot_scCustom
    sequential_palette_dotplot = sequential_palette_dotplot, # palette for DotPlot_scCustom
    local_path = local_path,
    object_annotations = '',
    top_n = 100,                # top genes per program to extract
    topn_plot = 3,             # top genes used in dotplot
    run_pathway_enrichment = pathway_enrichment       # run Metascape_overrepresentation_analysis
)


```

```{r}
k_used <- 15
dt_str <- '0_1'
runname = "cNMF_run"
data_dir <- here('results', 'cNMF')

seurat <- cnmf_results$seurat
GEP_to_plot <- str_c(c('cNMF_program_'), seq(1,12))
GEP_annotations <- c(
    'GEP 1:\nActin filament organization' = 'cNMF_program_1',
    'GEP 2:\nTCR signaling' = 'cNMF_program_2',
    'GEP 3:\nOxidative phosphorylation' = 'cNMF_program_3',
    'GEP 4:\nT cell differentiation' = 'cNMF_program_4',
    'GEP 5:\nStress response' = 'cNMF_program_5',
    'GEP 6:\nTranslation' = 'cNMF_program_6',
    'GEP 7:\nTranslation 2' = 'cNMF_program_7',
    'GEP 8:\nCell migration' = 'cNMF_program_8',
    'GEP 9:\nTranslation 3' = 'cNMF_program_9',
    'GEP 10:\nCytotoxicity' = 'cNMF_program_10',
    'GEP 11:\nCell division' = 'cNMF_program_11',
    'GEP 12:\nInterferon signaling' = 'cNMF_program_12'
)

GEP_to_plot <- c('GEP 1:\nActin filament organization', 
                 'GEP 2:\nTCR signaling', 
                 'GEP 3:\nOxidative phosphorylation', 
                 'GEP 4:\nT cell differentiation', 
                 'GEP 5:\nStress response', 
                 'GEP 6:\nTranslation',
                 'GEP 7:\nTranslation 2',
                 'GEP 8:\nCell migration',
                 'GEP 9:\nTranslation 3',
                 'GEP 10:\nCytotoxicity',
                 'GEP 11:\nCell division',
                 'GEP 12:\nInterferon signaling'                
                 )


seurat@meta.data <- seurat@meta.data |>
    rename(GEP_annotations)

full_object_regressed$cNMF_GEP_violin_plots <- map(GEP_to_plot, signature_violin_plot)
names(full_object_regressed$cNMF_GEP_violin_plots) <- GEP_to_plot

full_object_regressed$cNMF_GEP_feature_plots <- map(GEP_to_plot, feature_plot)
names(full_object_regressed$cNMF_GEP_feature_plots) <- GEP_to_plot

full_object_regressed$cNMF_GEP_feature_plots |> walk(print)


all_genes <- Features(seurat[['RNA']]) |>unique()
gene_list <- cnmf_results$top_colnames |> select(1:12) |> as.list() |>  map(~ .x[!is.na(.x)])
full_object_regressed$cNMF_GEP_pathways <- GO_overrepresentation_analysis_multiple_lists(gene_list, all_genes, local_path_pathway_enrichment, ontology = 'ALL', minGSSize = 5, maxGSSize = 400, filename = '',  drop_levels = F, levels_to_drop = c(), simplify_function = min, simplify_by = 'p.adjust', simplify_terms = F, run_network = F, network_n_terms = 100, nterms_to_plot = 4, font_size = 8)

```

# Assembling figures

```{r}
figures_path <- here('Manuscript_figures')
unlink(figures_path, recursive = T)
dir.create(figures_path, recursive = T, showWarnings = F)

axis_thickness <- 0.4
```

```{r}
#| fig-width: 8
#| #| fig-height: 10

full_object_original |> names()
full_object_regressed |> names()
full_object_regressed$Milo_figures |> names()

p1 <- textGrob('Figure 2', gp = gpar(fontsize = 20, fontface = 'bold'))

full_object_regressed$cluster_heatmap2 <- full_object_regressed$cluster_heatmap +
    scale_y_discrete(labels = c('g1' = '1', 'g2' = '2', 'g3' = '3', 'g4' = '4', 'g5' = '5', 'g6' = '6', 'g7' = '7', 'g8' = '8', 'g9' = '9')) +    
     labs(x = 'Genes', y = 'Leiden clusters', title = 'Top 3 genes per cluster') 

full_object_regressed$Milo_figures[['pathways']] <- full_object_regressed$Milo_figures[['pathways']] +
    scale_size(range = c(1,5)) +
    labs(title = 'MiloR: Neighborhood\nclusters - Pathway enrichment') +
    scale_y_discrete(labels = label_wrap(45)) 

full_object_regressed$Milo_figures$neighborhood_cluster_graph <- full_object_regressed$Milo_figures$neighborhood_cluster_graph +
    labs(x = 'UMAP 1', y = 'UMAP 2', title = 'MiloR: Neighborhood\nclusters') + theme(legend.title = element_text(size = 8))  +
    guides(fill = guide_legend(title = 'Clusters'))


plot_layout <- "
ABBB
"
row_one <- wrap_plots(list(full_object_regressed[['DimPlot']] ,
                            full_object_regressed[['DimPlot_groups']] 
                            ) ,
                            guides = 'collect', 
                            design = plot_layout) & 
                             guides(color = guide_legend(override.aes = list(size=2))) 
                            
 
row_milo <- wrap_plots(list(full_object_regressed$Milo_figures[['neighborhood_graph']] |> free(), 
                full_object_regressed$Milo_figures[['beeswarm_graph']] |> free(),
                plot_spacer() |> free()
                ), 
                # guides = 'collect',
                nrow = 1) +
                patchwork::plot_layout(widths = c(2, 1.5, 0.5))


row_feature_plots <- wrap_plots(list(
                full_object_regressed$feature_plots$Rorc |> free(),
                full_object_regressed$feature_plots$Tbx21  |> free(),
                full_object_regressed$feature_plots$Il17a  |> free(),
                full_object_regressed$feature_plots$Ifng |> free()
                ), nrow= 2)

plot_layout <- "
ABBE
ACDE
"
row_milo_neighborhoods <- wrap_plots(list(
                row_feature_plots |> free(),
                full_object_regressed$Milo_figures[['neighborhood_cluster_graph']] |> free(),
                full_object_regressed$Milo_figures[['beeswarm_graph_neighborhood_cluster']] |> free(),
                plot_spacer() |> free(),
                full_object_regressed$Milo_figures[['pathways']] |> free()
                ), 
                design= plot_layout) +
                patchwork::plot_layout(widths = c(2, 1, 0.2, 2.5))
            

row_feature_plots_2 <- wrap_plots(list(
                full_object_regressed$feature_plots$Gzmb,
                full_object_regressed$feature_plots$Gzmk,
                full_object_regressed$feature_plots$Ramp3,
                full_object_regressed$feature_plots$Sdc1,
                full_object_regressed$feature_plots$Slamf7,
                full_object_regressed$feature_plots$Klrg1
                ), nrow= 2) 

row_signature_violins <- wrap_plots(list(
                full_object_regressed$signature_violin_plots$MAIT1 |> free(),
                full_object_regressed$signature_violin_plots$MAIT17 |> free()
                ), nrow= 1)

row_signature_featureplots <- wrap_plots(list(
                full_object_regressed$signature_feature_plots$`Tissue repair` |> free(),
                full_object_regressed$signature_feature_plots$`Tissue residency` |> free(),
                full_object_regressed$signature_feature_plots$Circulating |> free()
                ), nrow= 1)

plot_layout <- "
    AAAABB
    CCCDDD
    EEEEEE
    EEEEEE
    FFGGGG
    #####Z
    "
figure_2 <- wrap_plots(list(
                A = row_one |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)), 
                B = full_object_regressed[['cluster_frequency']] |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                C = full_object_regressed$cluster_heatmap2 |> free(), 
                D = row_milo |>  free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                E = row_milo_neighborhoods |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                F =  row_signature_violins |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                G = row_signature_featureplots |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                # H = 
                Z = wrap_elements(p1) |> free()
                ),# |> map(\(x) free(x, c('label'))),
                design = plot_layout) +
    plot_annotation(tag_levels = list(c('A', 'B', 'C', 'D', 'E', 'F', 'G', '', '', '', 'H', 'I', 'J', 'K', '', 'L', '', '', ''))) &
    theme(
        base_family = 'Arial',
        text = element_text(size = 8, family = 'Arial'),
#         plot.title = element_text(size = 8, hjust = 0.5), 
        plot.title = element_text(size = 8, margin = margin(3, 3, 3, 3), hjust = 0.5),
        # plot.subtitle = element_text(size = 8),
        axis.text = element_text(size = 7),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title = element_text(size = 8),
        # axis.title.y = element_text(margin = margin(r = 0)),
        # axis.title.x = element_text(margin = margin(t = 0)),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        legend.key.size = unit(0.3, 'cm'),
        plot.margin = margin(4, 4, 4, 4),
        strip.text.x = element_text(size = 8, margin = margin(0, 0, 0, 0)),
        legend.box.spacing = unit(0.05, 'cm'),
        # axis.line = element_line(size = axis_thickness),
        # axis.ticks = element_line(size = axis_thickness),
        plot.tag = element_text(size = 12, face = 'bold')) 
ggsave(plot = figure_2, filename = 'Figure_2.pdf', path = figures_path, width = 12, height = 14, units = 'in')
figure_2

```

## Figure S2

```{r}
#| fig-width: 8
#| fig-height: 10

p1 <- textGrob('Figure S2', gp = gpar(fontsize = 20, fontface = 'bold'))

plot_layout <- "
ABBB
"
row_one <- wrap_plots(list(full_object_original[['DimPlot']],
                            full_object_original[['DimPlot_groups']]),
                            guides = 'collect', 
                            design = plot_layout) & 
                             guides(color = guide_legend(override.aes = list(size=2))) 
 
row_signature <- wrap_plots(list(
                (full_object_original$signature_feature_plots$Cycling_G2M+scale_color_continuous(palette = diverging_palette_3, breaks = c(-1, 0, 1)))  |> free(),
                full_object_original$signature_feature_plots$Cycling_S |> free(),
                full_object_original$signature_violin_plots$Cycling_G2M  |> free(),
                full_object_original$signature_violin_plots$Cycling_S  |> free()
                ), nrow= 2)

row_pseudobulk <- wrap_plots(list(
                C =  full_object_original$volcano_plot |> free(), 
                D = row_signature |> free() 
                ), nrow = 1
                ) #|> map(\(x) free(x, 'space', side = 'rl')),
                            

plot_layout <- "
    AAAABB
    CCCCCC
    CCCCCC
    FFGGHH
    DDEE#Z
    "
figure_S2 <- wrap_plots(list(
                A = row_one |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)), 
                B = full_object_original[['cluster_frequency']] |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                C = row_pseudobulk |>  free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                F = full_object_original$pathways_Abau |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                G = full_object_original$pathways_Abau_Tcelldifferentiation |> free(),
                H = full_object_original$pathways_Abau_woundrepair |> free(),
                D = full_object_original$pathways_Ftula |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                E = full_object_original$pathways_Ftula_cytotoxicity |> free(),
                Z = wrap_elements(p1) |> free()
                ),# |> map(\(x) free(x, c('label'))),
                design = plot_layout) +
    plot_annotation(tag_levels = list(c('A', 'B', 'C', 'D', 'E', '', 'F', '', 'G', 'H', 'I', 'J', 'K', ''))) &
    theme(
        # base_family = 'Arial',
        text = element_text(size = 8),
#         plot.title = element_text(size = 8, hjust = 0.5), 
        plot.title = element_text(size = 8, margin = margin(3, 1, 2, 1), hjust = 0.5),
        # plot.subtitle = element_text(size = 8),
        axis.text = element_text(size = 7),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title = element_text(size = 8),
        # axis.title.y = element_text(margin = margin(r = 0)),
        # axis.title.x = element_text(margin = margin(t = 0)),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        legend.key.size = unit(0.3, 'cm'),
        plot.margin = margin(4, 4, 4, 4),
        strip.text.x = element_text(size = 8, margin = margin(0, 0, 0, 0)),
        legend.box.spacing = unit(0.05, 'cm'),
        # axis.line = element_line(size = axis_thickness),
        # axis.ticks = element_line(size = axis_thickness),
        plot.tag = element_text(size = 12, face = 'bold')) 
ggsave(plot = figure_S2, filename = 'Figure_S2.pdf', path = figures_path, width = 13, height = 13, units = 'in')
figure_S2

```


## Figure S3

```{r}
#| fig-width: 8
#| #| fig-height: 10

full_object_original |> names()
full_object_regressed |> names()
full_object_regressed$Milo_figures |> names()

p1 <- textGrob('Figure S3', gp = gpar(fontsize = 20, fontface = 'bold'))

# full_object_regressed |> walk(print)

# full_object_regressed$signature_feature_plots$tissue_repair

full_object_regressed$cluster_signatures <- full_object_regressed$cluster_signatures + scale_y_discrete(label = label_wrap(40))

row_feature_plots_2 <- wrap_plots(list(
                full_object_regressed$feature_plots$Gzmb |> free(),
                full_object_regressed$feature_plots$Gzmk |> free(),
                full_object_regressed$feature_plots$Slamf7 |> free(),
                full_object_regressed$feature_plots$Klrg1 |> free(),
                full_object_regressed$feature_plots$Ramp3 |> free(),
                full_object_regressed$feature_plots$Sdc1 |> free()
                ), nrow= 1) 

row_signature_heatmaps <- wrap_plots(list(
                full_object_regressed$signature_heatmaps2$MAIT1 |>  free(),
                full_object_regressed$signature_heatmaps2$MAIT17 |>  free(),
                full_object_regressed$signature_heatmaps2$Cycling_G2M |>  free(),
                full_object_regressed$signature_heatmaps2$Cycling_S |>  free()
                ), nrow= 1)

plot_layout <- "
    AAABBB
    AAABBB
    AAABBB
    CCCCCC
    DDDEEE
    FFFGGG
    "
figure_S3 <- wrap_plots(list(
                A = full_object_regressed$cluster_signatures |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)), 
                B = full_object_regressed$Milo_figures$pathways_genes |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                C = row_feature_plots_2 |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                D = full_object_regressed$signature_heatmaps$MAIT1 |>  free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)), 
                E = full_object_regressed$signature_heatmaps$MAIT17 |>  free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                F = full_object_regressed$signature_heatmaps$Cycling_G2M |>  free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                G = full_object_regressed$signature_heatmaps$Cycling_S |>  free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                # D = full_object_regressed$cluster_heatmap2 |> free(),
                # # E = row_feature_plots |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                # F =  row_signature_violins |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)), 
                # G = row_milo_neighborhoods |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                # H = 
                Z = wrap_elements(p1) |> free()
                ),# |> map(\(x) free(x, c('label'))),
                design = plot_layout) +
    plot_annotation(tag_levels = list(c('A', 'B', 'C', '', '', 'D', '', '', 'E', 'F', 'G', 'H'))) &
    theme(
        base_family = 'Arial',
        text = element_text(size = 8, family = 'Arial'),
#         plot.title = element_text(size = 8, hjust = 0.5), 
        plot.title = element_text(size = 8, margin = margin(3, 3, 3, 3), hjust = 0.5),
        # plot.subtitle = element_text(size = 8),
        axis.text = element_text(size = 7),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title = element_text(size = 8),
        # axis.title.y = element_text(margin = margin(r = 0)),
        # axis.title.x = element_text(margin = margin(t = 0)),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        legend.key.size = unit(0.3, 'cm'),
        plot.margin = margin(4, 4, 4, 4),
        strip.text.x = element_text(size = 8, margin = margin(0, 0, 0, 0)),
        legend.box.spacing = unit(0.05, 'cm'),
        # axis.line = element_line(size = axis_thickness),
        # axis.ticks = element_line(size = axis_thickness),
        plot.tag = element_text(size = 12, face = 'bold')) 
ggsave(plot = figure_S3, filename = 'Figure_S3.pdf', path = figures_path, width = 12, height = 14, units = 'in')
figure_S3

```

## Figure S4

```{r}
p1 <- textGrob('Figure S4', gp = gpar(fontsize = 20, fontface = 'bold'))

Ftula_object$DimPlot <- Ftula_object$DimPlot + labs(title = 'Ftula MAIT Leiden clusters')
Abau_object$DimPlot <- Abau_object$DimPlot + labs(title = 'Abau MAIT Leiden clusters')

Ftula_object$cluster_signatures <- Ftula_object$cluster_signatures + scale_y_discrete(label = label_wrap(30))
Ftula_object$cluster_heatmap <- Ftula_object$cluster_heatmap + 
    scale_y_discrete(labels = c('g1' = '1', 'g2' = '2', 'g3' = '3', 'g4' = '4', 'g5' = '5', 'g6' = '6', 'g7' = '7')) +    
     labs(x = 'Genes', y = 'Leiden clusters', title = 'Top 3 genes per cluster')

Abau_object$cluster_signatures <- Abau_object$cluster_signatures + scale_y_discrete(label = label_wrap(30))
Abau_object$cluster_heatmap <- Abau_object$cluster_heatmap + 
    scale_y_discrete(labels = c('g1' = '1', 'g2' = '2', 'g3' = '3', 'g4' = '4', 'g5' = '5', 'g6' = '6', 'g7' = '7', 'g8' = '8', 'g9' = '9', 'g10' = '10')) +    
     labs(x = 'Genes', y = 'Leiden clusters', title = 'Top 3 genes per cluster')

row_one <- wrap_plots(list(Ftula_object[['DimPlot']]  |> free(),
                        Ftula_object$cluster_heatmap  |> free()
                            ))+
                        plot_layout(widths = c(1, 2.5))

row_feature_plots <- wrap_plots(list(
                Ftula_object$feature_plots$Rorc |> free(),
                Ftula_object$feature_plots$Tbx21  |> free(),
                Ftula_object$feature_plots$Il17a  |> free(),
                Ftula_object$feature_plots$Ifng |> free(),
                Ftula_object$feature_plots$Gzmb |> free(),
                Ftula_object$feature_plots$Gzmk |> free(),
                Ftula_object$feature_plots$Slamf7 |> free(),
                Ftula_object$feature_plots$Klrg1 |> free()                
                ), nrow= 2)

row_two <- wrap_plots(list(Abau_object[['DimPlot']]  |> free(),
                        Abau_object$cluster_heatmap  |> free()
                            ))+
                        plot_layout(widths = c(1, 2.5))

row_feature_plots2 <- wrap_plots(list(
                Abau_object$feature_plots$Rorc |> free(),
                Abau_object$feature_plots$Tbx21  |> free(),
                Abau_object$feature_plots$Il17a  |> free(),
                Abau_object$feature_plots$Ifng |> free(),
                Abau_object$feature_plots$Ly6c2 |> free(),
                Abau_object$feature_plots$Ramp3 |> free(),
                Abau_object$feature_plots$Sdc1 |> free(),
                Abau_object$feature_plots$Areg |> free()                
                ), nrow= 2)


plot_layout <- "
    AAAABB
    CCCCBB
    CCCCBB
    DDDDEE
    FFFFEE
    FFFFEE
    "
figure_S4 <- wrap_plots(list(
                A = row_one |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)), 
                B = Ftula_object$cluster_signatures |>  free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                C = row_feature_plots |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                D = row_two |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                E = Abau_object$cluster_signatures |>  free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                F = row_feature_plots2 |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                Z = wrap_elements(p1) |> free()
                ),# |> map(\(x) free(x, c('label'))),
                design = plot_layout) +
    plot_annotation(tag_levels = list(c('A', 'B', 'C', 'D', '', '', '', '', '', '', '', 'E', 'F', 'G', 'H', '', '', '', '', '', '', ''))) &
    theme(
        base_family = 'Arial',
        text = element_text(size = 8, family = 'Arial'),
#         plot.title = element_text(size = 8, hjust = 0.5), 
        plot.title = element_text(size = 8, margin = margin(3, 1, 2, 1), hjust = 0.5),
        # plot.subtitle = element_text(size = 8),
        axis.text = element_text(size = 7),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title = element_text(size = 8),
        # axis.title.y = element_text(margin = margin(r = 0)),
        # axis.title.x = element_text(margin = margin(t = 0)),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        legend.key.size = unit(0.3, 'cm'),
        plot.margin = margin(4, 4, 4, 4),
        strip.text.x = element_text(size = 8, margin = margin(0, 0, 0, 0)),
        legend.box.spacing = unit(0.05, 'cm'),
        # axis.line = element_line(size = axis_thickness),
        # axis.ticks = element_line(size = axis_thickness),
        plot.tag = element_text(size = 12, face = 'bold')) 
ggsave(plot = figure_S4, filename = 'Figure_S4.pdf', path = figures_path, width = 13, height = 13, units = 'in')
figure_S4
```

## Figure S5

```{r}
p1 <- textGrob('Figure S5', gp = gpar(fontsize = 20, fontface = 'bold'))

full_object_regressed$cNMF_GEP_pathways2 <- full_object_regressed$cNMF_GEP_pathways$plot +
    # scale_size(range = c(1,5)) +
    labs(title = 'cNMF: Gene expression program enriched GO terms', x = 'Gene expression programs (GEPs)') +
    scale_y_discrete(labels = label_wrap(45))

row_feature_plots <- wrap_plots(full_object_regressed$cNMF_GEP_feature_plots |> map(free), nrow= 4)
row_signatures <- wrap_plots(full_object_regressed$cNMF_GEP_violin_plots |> map(free), nrow= 4)

plot_layout <- "
    AAABBB
    AAABBB
    AAABBB
    CCC###
    CCC###
    CCC##Z
    "
figure_S5 <- wrap_plots(list(
                A = full_object_regressed$cNMF_GEP_pathways2 |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),  
                B = row_feature_plots |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                C = row_signatures |> free() & theme(axis.line = element_line(size = axis_thickness), axis.ticks = element_line(size = axis_thickness)),
                Z = wrap_elements(p1) |> free()
                ),# |> map(\(x) free(x, c('label'))),
                design = plot_layout) +
    plot_annotation(tag_levels = list(c('A', 'B', '', '', '', '', '', '', '', '', '', '', '', 'C', '', '', '', '', '', '', '', '', '', '', '', ''))) &
    theme(
        base_family = 'Arial',
        text = element_text(size = 8, family = 'Arial'),
#         plot.title = element_text(size = 8, hjust = 0.5), 
        plot.title = element_text(size = 8, margin = margin(3, 1, 2, 1), hjust = 0.5),
        # plot.subtitle = element_text(size = 8),
        axis.text = element_text(size = 7),
        axis.text.x = element_text(size = 7),
        axis.text.y = element_text(size = 7),
        axis.title = element_text(size = 8),
        # axis.title.y = element_text(margin = margin(r = 0)),
        # axis.title.x = element_text(margin = margin(t = 0)),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        legend.key.size = unit(0.3, 'cm'),
        plot.margin = margin(4, 4, 4, 4),
        strip.text.x = element_text(size = 8, margin = margin(0, 0, 0, 0)),
        legend.box.spacing = unit(0.05, 'cm'),
        # axis.line = element_line(size = axis_thickness),
        # axis.ticks = element_line(size = axis_thickness),
        plot.tag = element_text(size = 12, face = 'bold')) 
ggsave(plot = figure_S5, filename = 'Figure_S5.pdf', path = figures_path, width = 13, height = 13, units = 'in')
figure_S5
```